### 2019-03-04 manuscript.tex
move text into manuscript.tex file

\documentclass[11pt,letterpaper]{article}

\usepackage{threeparttable}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{pdfpages}
\usepackage{url}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[outercaption]{sidecap}
\doublespacing

\raggedright
\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0.0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.1 in
\parindent = 0.2 in

\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

%commands that were used to generate a PDF that was copied to make the .docx version for GSAB
%\pagestyle{empty}
%\usepackage[nomarkers,figuresonly]{endfloat}

\begin{document}

\begin{flushleft}
{\Large \textbf{The Precambrian paleogeography of Laurentia}}
\\

Nicholas L. Swanson-Hysell\textsuperscript{1}

\bigskip

\textsuperscript{1} Department of Earth and Planetary Science, University of California, Berkeley, CA 94720 USA

\end{flushleft}

\noindent\textit{This chapter is in preparation for the book Ancient Supercontinents and the Paleogeography of the Earth}

%\linenumbers

\section*{INTRODUCTION}

Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have been a constituent of both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography. There is a rich record of Precambrian paleomagnetic poles from Laurentia that is key to evaluating and developing paleogeographic models.

Laurentia refers to the craton that forms the Precambrian core of North America. Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly (Hoffman, 1988). Collision between the Superior Craton and the composite Slave+Rae+Hearne+Nain cratons that resulted in the the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Corrigan et al., 2009). Terminal collison recorded in the Trans-Hudson orogeny is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims (Skipton et al., 2016; Weller and St-Onge, 2017). A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collison of the Trans-Hudson orogeny. This story of rapid Paleoproterozoic amalgamation of  Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper *United Plates of America* (Hoffman, 1988) and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogeny associated with the collision between the Slave craton and the Rae craton ca. 2.0 to 1.9 Ga (Hoffman, 1988); the Snowbird orogeny associated with ca. 1.89 Ga collision between the Rae and Hearne cratons and associated terranes (Berman et al., 2007); the Nagssugtoqidian orogeny associated with ca. 1.86 Ga collision between the Rae and Nain cratons (St-Onge et al., 2009). As for the Wyoming craton, many models posit that it was conjoined with Hearne and associated cratons as the time of the Trans-Hudson orogeny (e.g. St-Onge et al., 2009; Pehrsson et al., 2015) although a contrasting view has been been proposed that it arrived ca. 1.72 Ga after the accretionary Yavapai and Penokean orogenies (Kilian et al. 2016).

In the paleogeographic model framework of Pehrsson et al. (2015), the cratonic and terrane collisions leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g. Skipton et al., 2016; Weller and St-Onge, 2017). This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/Sao Francisco, West Africa, and Amazonia (Pehrsson et al., 2015). Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (Whitmeyer and Karlstrom, 2007; \ref{fig:Laurentia_map}).

%The perphiral Penokean orogeny on the southern margin of the Superior craton coincided with the Trans-Hudson orogeny with the last evidence of the orogeny being undeformed plutons such as the ca. 1.78 Ga East Central Minnesota Batholith (Holm et al., 2005). The arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny

\begin{figure}
\centering
\includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}. The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}
\label{fig:Laurentia_map}
\end{figure}

\section{Paleomagnetic pole compilation}

This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them with grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are A and B. An A rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A B rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.

Prior to the termination of the Trans-Hudson orogeny orogeny (before 1.8 Ga), we consider paleomagnetic poles with respect to the individual Archean cratons. Following the termination of this orogeny (after 1.8 Ga), we consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.

For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone (Bates and Halls, 1991; Evans and Halls, 2010). Evans and Halls (2010) propose an Euler rotation of (51º N, 85º W, −14º CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga. I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame.

\end{{document}

### 2019-03-05 manuscript.tex
small updates to manuscript and notebook

@@
 Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
-This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (Whitmeyer and Karlstrom, 2007; \ref{fig:Laurentia_map}).
+This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
 
@@
 \includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}.
+\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}.
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}

### 2019-03-16 manuscript.tex
progress on notebook and manuscript

@@
 
-Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have been a constituent of both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
+Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have been a central constituent of both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography.
@@
 A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collison of the Trans-Hudson orogeny.
-This story of rapid Paleoproterozoic amalgamation of  Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper *United Plates of America* (Hoffman, 1988) and has been refined in the time since -- particularly with additional geochronological constraints.
+This story of rapid Paleoproterozoic amalgamation of  Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} (Hoffman, 1988) and has been refined in the time since -- particularly with additional geochronological constraints.
 Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogeny associated with the collision between the Slave craton and the Rae craton ca.
 2.0 to 1.9 Ga (Hoffman, 1988); the Snowbird orogeny associated with ca.
-1.89 Ga collision between the Rae and Hearne cratons and associated terranes (Berman et al., 2007); the Nagssugtoqidian orogeny associated with ca.
+1.89 Ga collision between the Rae and Hearne cratons and associated terranes (Berman et al., 2007); the XXXXXX orogeny associated with ca.
 1.86 Ga collision between the Rae and Nain cratons (St-Onge et al., 2009).
@@
 This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
+Major growth of Laurentia following the amalgmation of these Archean cratons occured associated with the arc-continent collision of the ca.
+1.71 to 1.68 Ga Yavapai orogeny whose accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
+These intrusions are hypothesized to have stablized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}.
+Subsequent accretionary orogenesis of the ca.
+1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic.
+Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
+An interval of major plutonism occured ca.
+1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces.
+In the latest Mesoproterozoic ca.
+1.11-1.08 Ga, a major intracontinental rift co-located with the a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces.
+This Midcontinent Rfit lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
+Major collisional orogenesis of the Grenvillian orogeny occured following Midcontinent Rift development
 

### 2019-03-16 manuscript.tex
updates to notebook and text related to Archean craton pole splits

@@
 \includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}.
+\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}
@@
 While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
-The categorizations used by the expert panel are A and B.
-An A rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions.
-A B rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
+The categorizations used by the expert panel are `A' and `B' with the last panel meeting occuring in Fall 2017 in Leirubakki, Iceland.
+An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions.
+A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
+Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are included and referred to as not-rated (`NR').
 
-Prior to the termination of the Trans-Hudson orogeny orogeny (before 1.8 Ga), we consider paleomagnetic poles with respect to the individual Archean cratons.
+Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), we consider paleomagnetic poles with respect to the individual Archean cratons.
+There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain cratons.
 Following the termination of this orogeny (after 1.8 Ga), we consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.
 
-For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone (Bates and Halls, 1991; Evans and Halls, 2010).
-Evans and Halls (2010) propose an Euler rotation of (51º N, 85º W, −14º CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.
+For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
+\cite{Evans2010a} propose an Euler rotation of (51º N, 85º W, −14º CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.
 I follow this interpretation and group the poles into Superior (West) and Superior (East).
@@
 
+Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton.
+It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
+\cite{Hall201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault.
+However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of thick plateau developed during continent-continent collision (e.g.
+\cite{}).
+Poles are also included in the composite that come from terranes that were once part of Laurentia, but have subsequently rifted away.
+These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction.
+\begin{itemize}
+\item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
+\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi2004
+\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin (Maloof et al., 2006 rotation)
+\end{itemize}
+
 \end{{document}

### 2019-07-04 manuscript.tex
Update manuscript.tex

@@
 Laurentia refers to the craton that forms the Precambrian core of North America.
-Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly (Hoffman, 1988).
-Collision between the Superior Craton and the composite Slave+Rae+Hearne+Nain cratons that resulted in the the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Corrigan et al., 2009).
+Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1988a, Whitmeyer2007a}.
+Collision between the Superior Craton and the composite Slave+Rae+Hearne+Nain cratons that resulted in the the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
 Terminal collison recorded in the Trans-Hudson orogeny is estimated to have been ca.
-1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims (Skipton et al., 2016; Weller and St-Onge, 2017).
+1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
 A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collison of the Trans-Hudson orogeny.
-This story of rapid Paleoproterozoic amalgamation of  Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} (Hoffman, 1988) and has been refined in the time since -- particularly with additional geochronological constraints.
+Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior cratron with the last evidence of that orogeny being ca.
+1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
+This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} \citep{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
 Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogeny associated with the collision between the Slave craton and the Rae craton ca.
@@
 
-In the paleogeographic model framework of Pehrsson et al.
-(2015), the cratonic and terrane collisions leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
+In the paleogeographic model framework of \cite{Pehrsson2015a}, the cratonic and terrane collisions leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g.
-Skipton et al., 2016; Weller and St-Onge, 2017).
-This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/Sao Francisco, West Africa, and Amazonia (Pehrsson et al., 2015).
+\citealp{Skipton2016a, Weller2017a}.
+This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/Sao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
 Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
@@
 This Midcontinent Rfit lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
-Major collisional orogenesis of the Grenvillian orogeny occured following Midcontinent Rift development
-
-%The perphiral Penokean orogeny on the southern margin of the Superior craton coincided with the Trans-Hudson orogeny with the last evidence of the orogeny being undeformed plutons such as the ca.
-1.78 Ga East Central Minnesota Batholith (Holm et al., 2005).
-The arc-continent collision of the ca.
-1.71 to 1.68 Ga Yavapai orogeny
+Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.
+The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
+1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a} and is typically to interpretted to have been a major collisional event associated with the amalgamation of continents into the supercontinent Rodinia \citep{Hoffman1991a}.
 
@@
 \caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}).
-The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}
+The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
+The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
 \label{fig:Laurentia_map}
@@
 This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
-The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them with grades meant to convey the confidence that the community has in these results.
+The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results.
 While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
@@
 A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
-Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are included and referred to as not-rated (`NR').
+Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
 

### 2019-07-05 manuscript.tex
Update manuscript.tex

@@
 
+\section*{Broad tectonic history overview}
+
 Laurentia refers to the craton that forms the Precambrian core of North America.
@@
 
-This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
+In this chapter, I will focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
 The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results.
@@
 Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
+These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}
 
-Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), we consider paleomagnetic poles with respect to the individual Archean cratons.
+Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons.
 There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain cratons.
-Following the termination of this orogeny (after 1.8 Ga), we consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.
+Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.
 
@@
 \cite{}).
-Poles are also included in the composite that come from terranes that were once part of Laurentia, but have subsequently rifted away.
+Poles with a B-rating are also included in the composite that come from terranes that were once part of Laurentia, but have subsequently rifted away.
 These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction.
+for which we used the following rotations
+
 \begin{itemize}

### 2019-07-05 manuscript.tex
progress on manuscript including first compile with references

@@
 
-Laurentia refers to the craton that forms the Precambrian core of North America.
+Laurentia refers to the craton that forms the Precambrian core of North America (Fig.
+\ref{fig:Laurentia_map}).
 Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1988a, Whitmeyer2007a}.
@@
 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
-A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collison of the Trans-Hudson orogeny.
-Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior cratron with the last evidence of that orogeny being ca.
+A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
+Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior craton with the last evidence of that orogeny being ca.
 1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
@@
 1.86 Ga collision between the Rae and Nain cratons (St-Onge et al., 2009).
-As for the Wyoming craton, many models posit that it was conjoined with Hearne and associated cratons as the time of the Trans-Hudson orogeny (e.g.
-St-Onge et al., 2009; Pehrsson et al., 2015) although a contrasting view has been been proposed that it arrived ca.
-1.72 Ga after the accretionary Yavapai and Penokean orogenies (Kilian et al.
-2016).
+Crustal growth also progressed at this time within the Wopmay orogen through the ca.
+1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
+1.88 to 1.84 Ga \citep{Hildebrand2009a}.
+As for the Wyoming craton, many models posit that it was conjoined with Hearne and associated cratons as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} although a contrasting view has been been proposed that it arrived ca.
+1.72 Ga after the accretionary Yavapai and Penokean orogenies \citep{Kilian2016a}.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{Figures/Fig1_map.pdf}
+\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}).
+The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
+The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
+\label{fig:Laurentia_map}
+\end{figure}
 
@@
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g.
-\citealp{Skipton2016a, Weller2017a}.
-This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/Sao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
+\citealp{Skipton2016a, Weller2017a}).
+This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
+
 Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
 This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
-Major growth of Laurentia following the amalgmation of these Archean cratons occured associated with the arc-continent collision of the ca.
+Major growth of Laurentia following the amalgamation of these Archean cratons occurred associated with the arc-continent collision of the ca.
 1.71 to 1.68 Ga Yavapai orogeny whose accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
-An interval of major plutonism occured ca.
+An interval of major plutonism occurred ca.
 1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces.
+This plutonism is likely due to crustal melting within a back-arc region of ca.
+1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}.
+Younger magmatic activity ca.
+1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
+
+\begin{figure}
+\centering
+\includegraphics[width=3in]{Figures/Tectonic_history.pdf}
+\caption{\small{\textbf{Summary of Laurentia's tectonic history over the past $\sim$1.8 billion years. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}}
+\label{fig:Laurentia_map}
+\end{figure}
+
  In the latest Mesoproterozoic ca.
@@
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a} and is typically to interpretted to have been a major collisional event associated with the amalgamation of continents into the supercontinent Rodinia \citep{Hoffman1991a}.
+1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a} and is typically to interpreted to have been a major collisional event associated with the amalgamation of continents into the supercontinent Rodinia \citep{Hoffman1991a}.
 
-\begin{figure}
-\centering
-\includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}).
-The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
-The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
-\label{fig:Laurentia_map}
-\end{figure}
+This tectonic history provides significant constraints on paleogeographic reconstructions.
+In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from recorded from the initiation of the Yavapai orogeny (ca.
+1.71 Ga) to the end of the Shawinigan orogeny (ca.
+1.06 Ga) requires an open active margin
+
+There was significantly less crustal growth on the western margin of Laurentia and the tectonic history is not as well elucidated as on the southern to eastern margin.
+A thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
+1.47 to 1.40 rifting \citep{}.
+This region is interpreted to have been deformed during a ca.
+1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+
+% Lu–Hf garnet ages have been interpreted to indicate that there were two metamorphic garnet growth events at 1330 Ma and 1080 Ma within metamorphic rocks of the Belt-Purcell Supergroup \citep{Nesheim2012a}.
+Ca.
+1.3 Ga orogenesis within the Belt Supergroup has been referred to as the East Kootenay Orogeny
 
@@
 For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
-\cite{Evans2010a} propose an Euler rotation of (51º N, 85º W, −14º CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.
+\cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.
 I follow this interpretation and group the poles into Superior (West) and Superior (East).
@@
 These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction.
-for which we used the following rotations
+for which I used the following rotations
 
@@
 
-\end{{document}
+\bibliographystyle{gsabull}
+\small\bibliography{../references/allrefs}
+
+\end{document}

### 2019-07-10 manuscript.tex
progress on tectonic history section

@@
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography.
-There is a rich record of Precambrian paleomagnetic poles from Laurentia that is key to evaluating and developing paleogeographic models.
+There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as a rich geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
 
@@
 \ref{fig:Laurentia_map}).
-Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1988a, Whitmeyer2007a}.
-Collision between the Superior Craton and the composite Slave+Rae+Hearne+Nain cratons that resulted in the the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
-Terminal collison recorded in the Trans-Hudson orogeny is estimated to have been ca.
+Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
+Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
+Terminal collison recorded in the Trans-Hudson orogen is estimated to have been ca.
 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
 A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
-Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior craton with the last evidence of that orogeny being ca.
+Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior province with the last evidence of that orogeny being ca.
 1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
-This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} \citep{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
-Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogeny associated with the collision between the Slave craton and the Rae craton ca.
-2.0 to 1.9 Ga (Hoffman, 1988); the Snowbird orogeny associated with ca.
-1.89 Ga collision between the Rae and Hearne cratons and associated terranes (Berman et al., 2007); the XXXXXX orogeny associated with ca.
-1.86 Ga collision between the Rae and Nain cratons (St-Onge et al., 2009).
-Crustal growth also progressed at this time within the Wopmay orogen through the ca.
+This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} \citep{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
+Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogen associated with the collision between the Slave province and the Rae province ca.
+2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca.
+1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; and the Nagssugtoqidian orogen due to the ca.
+1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}.
+Crustal growth also progressed at this time within the Wopmay orogen through ca.
 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
 1.88 to 1.84 Ga \citep{Hildebrand2009a}.
-As for the Wyoming craton, many models posit that it was conjoined with Hearne and associated cratons as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} although a contrasting view has been been proposed that it arrived ca.
-1.72 Ga after the accretionary Yavapai and Penokean orogenies \citep{Kilian2016a}.
+As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal and undergoing continued translation up to ca.
+1.80 Ga \citep{Whitmeyer2007a}.
+A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks arrived ca.
+1.72 Ga \citep{Kilian2016a} which is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
+1.75 to 1.72 Ga \citep{Condit2015a} and ca.
+1.72 tectonomagmatic activity in the Black Hills region \citep{}.
+However, the evidence for earlier orogenesis ca.
+1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca.
+1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca.
+1.72 Ga activity is a minor overprint on ca.
+1.75 terminal suturing between Wyoming and Superior.
+Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca.
+1.82 Ga amalgmation of Archean provinces with the Trans-Hudson orogen further to the northeast.
 
@@
 
-In the paleogeographic model framework of \cite{Pehrsson2015a}, the cratonic and terrane collisions leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
+In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g.
 \citealp{Skipton2016a, Weller2017a}).
-This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
+This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
 
 Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
-This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
-Major growth of Laurentia following the amalgamation of these Archean cratons occurred associated with the arc-continent collision of the ca.
+This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occured dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
+Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
 1.71 to 1.68 Ga Yavapai orogeny whose accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 Subsequent accretionary orogenesis of the ca.
-1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic.
+1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Whitmeyer2007a}.
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
@@
 
+Accretionary orogenesis continued along the south to southeast margin of Laurentia with Elzevirian
+
+The ca. 1190 to 1160 Ma Shawinigan orogeny is interpreted to be due to the accretion of a
+terrane comprised of amalgamated arc volcanics and associated metasediments \citep{McLelland2010a}
+
 In the latest Mesoproterozoic ca.
@@
 1.71 Ga) to the end of the Shawinigan orogeny (ca.
-1.06 Ga) requires an open active margin
+1.06 Ga) requires an long-lived open active margin up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
+This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructure into positions (e.g.
+Amazonia in \citealp{Elming2009a}) that are seemingly incompatible with this record of accretionary orogenesis.
+The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca.
+1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
 
@@
 1.47 to 1.40 rifting \citep{}.
-This region is interpreted to have been deformed during a ca.
+This region is interpreted to have been subsequently deformed during a ca.
 1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

### 2019-07-11 manuscript.tex
progress on Laurentia tectonic history section

@@
 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
-A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
-Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior province with the last evidence of that orogeny being ca.
-1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
-This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} \citep{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
-Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogen associated with the collision between the Slave province and the Rae province ca.
+A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
+This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
+Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca.
 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca.
-1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; and the Nagssugtoqidian orogen due to the ca.
-1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}.
-Crustal growth also progressed at this time within the Wopmay orogen through ca.
-1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
-1.88 to 1.84 Ga \citep{Hildebrand2009a}.
-As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal and undergoing continued translation up to ca.
+1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca.
+1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca.
+1.87 to
+1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}.
+As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior with still goingd undergoing continued translation up to ca.
 1.80 Ga \citep{Whitmeyer2007a}.
@@
 1.82 Ga amalgmation of Archean provinces with the Trans-Hudson orogen further to the northeast.
+Overall, the collision of these Archean microcontinents between ca.
+1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.
 
@@
 \end{figure}
+
+Crustal growth also progressed at this time through accretionary orogenesis.
+This accretion occured within the Wopmay orogen through ca.
+1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
+1.88 to 1.84 Ga \citep{Hildebrand2009a}.
+Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior province with the last evidence of that orogeny being ca.
+1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
 
@@
 
-Accretionary orogenesis continued along the south to southeast margin of Laurentia with Elzevirian
-
-The ca. 1190 to 1160 Ma Shawinigan orogeny is interpreted to be due to the accretion of a
-terrane comprised of amalgamated arc volcanics and associated metasediments \citep{McLelland2010a}
-
+Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca.
+1.25–1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
+The subsequent ca.
+1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a
+terrane comprised of amalgamated arc volcanics and associated metasediments \citep{McLelland2010a}.
 In the latest Mesoproterozoic ca.
-1.11-1.08 Ga, a major intracontinental rift co-located with the a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces.
-This Midcontinent Rfit lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
+1.11-1.08 Ga, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces.
+This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
 Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a} and is typically to interpreted to have been a major collisional event associated with the amalgamation of continents into the supercontinent Rodinia \citep{Hoffman1991a}.
+1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.
 
-This tectonic history provides significant constraints on paleogeographic reconstructions.
-In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from recorded from the initiation of the Yavapai orogeny (ca.
+There was significantly less crustal growth on the western margin of Laurentia and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
+A thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
+1.47 to 1.40 rifting which is typically interpretted to be intracratonic\citep{}.
+This region is interpreted to have been subsequently deformed during a ca.
+1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+
+This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions.
+In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca.
 1.71 Ga) to the end of the Shawinigan orogeny (ca.
-1.06 Ga) requires an long-lived open active margin up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
-This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructure into positions (e.g.
-Amazonia in \citealp{Elming2009a}) that are seemingly incompatible with this record of accretionary orogenesis.
+1.06 Ga) requires an long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
+This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructured into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
+Amazonia in \citealp{Elming2009a}).
 The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca.
 1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
+That the Laurentia margin experienced large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny that is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
 
-There was significantly less crustal growth on the western margin of Laurentia and the tectonic history is not as well elucidated as on the southern to eastern margin.
-A thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
-1.47 to 1.40 rifting \citep{}.
-This region is interpreted to have been subsequently deformed during a ca.
-1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting.
+Along the western margin of Laurentia small-scale rifting occured ca.
+780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountins of NW Laurentia \citep{Rooney2017a}.
+However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until near the ca.
+539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a}.
+The emplacement of the Gunbarrel large igneous province along this margin and the extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin initiating ca.
+780 Ma.
+If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the late Ediacaran.
+The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occured later than the timing of rifting typically implemented in models of Rodinia break-up.
+It is conceivable that the rifting that drove latest Ediacaran to Cambrian thermal subsidence was associated with a microcontinent fragment rather than a major conjugate continent, but it is an important constraint for paleogeographic models that the time of major lithospheric thinning on the western Laurentia margin is ca.
+539 Ma.
+The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin.
+Available geochronological constraints and thermal subsidence modeling indicate ca.
+820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a,Halverson2018a}.
+These data suggest that conjugate continental lithosphere rifted away from northeast Greenland at the time.
+Extensive rifting that was followed by thermal subsidence occured along the southeast to east Laurentia margin at the Neoproterozoic-Phanerozoic boundary.
+A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as long term Cambrian to Ordovician subsidence along the margin.
+The Argentine pre-Cordillera Cuyania terrane is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting.
+As with other rifts, it is difficult to distinguish the seperation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar.
+One interpretation is that there was break-up along the eastern margin that initationed ca.
+XXX associated with the Rome trough and prior to the ca.
+539 Oklahoma aulacogen rifting that liberated Cuyania.
 
@@
 1.3 Ga orogenesis within the Belt Supergroup has been referred to as the East Kootenay Orogeny
+
+Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic provides very strong evidence for mobile plate tectonics driving Laurentia evolution throughout the past 2 billion years.
+This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic rcord.
 
@@
 Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons.
-There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain cratons.
+There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces.
+These data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
 Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.
 
-For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
+For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
+This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny.
 \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.
@@
 These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame.
+Uncertainty remains with respect to whether the ca.
+1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference grame.
 
@@
 \cite{}).
-Poles with a B-rating are also included in the composite that come from terranes that were once part of Laurentia, but have subsequently rifted away.
-These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction.
-for which I used the following rotations
+Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland.
+These terranes were once part of contigous Laurentia, but have subsequently rifted away.
+These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:
 
@@
 
+To make an argument that differential plate tectonic motion began in the Neoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
+
 \bibliographystyle{gsabull}

### 2019-07-12 manuscript.tex
progress on tectonic history text

@@
 \ref{fig:Laurentia_map}).
-Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
+Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
 Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
-Terminal collison recorded in the Trans-Hudson orogen is estimated to have been ca.
+Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca.
 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
@@
 1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca.
-1.87 to
-1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}.
+1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}.
 As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior with still goingd undergoing continued translation up to ca.
@@
 Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca.
-1.82 Ga amalgmation of Archean provinces with the Trans-Hudson orogen further to the northeast.
+1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.
 Overall, the collision of these Archean microcontinents between ca.
@@
 \includegraphics[width=\textwidth]{Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}).
+\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 Crustal growth also progressed at this time through accretionary orogenesis.
-This accretion occured within the Wopmay orogen through ca.
+This accretion occurred within the Wopmay orogen through ca.
 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
@@
 \citealp{Skipton2016a, Weller2017a}).
-This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
+This model posits that this period terminal collision is not only associated with the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
 
@@
 This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occured dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
+Determining the extent of these belts is complicated by poor exposure of them in the Midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
 Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
-1.71 to 1.68 Ga Yavapai orogeny whose accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
-These intrusions are hypothesized to have stablized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}.
+1.71 to 1.68 Ga Yavapai orogeny which is interpreted to have resulted from the accretion of a series of arc terrances that collided with each and Laurentia \citep{Whitmeyer2007a}.
+Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
+These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}.
 Subsequent accretionary orogenesis of the ca.
@@
 \centering
-\includegraphics[width=3in]{Figures/Tectonic_history.pdf}
+\includegraphics[width=4in]{Figures/Tectonic_history.pdf}
 \caption{\small{\textbf{Summary of Laurentia's tectonic history over the past $\sim$1.8 billion years. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}}
@@
 The subsequent ca.
-1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a
-terrane comprised of amalgamated arc volcanics and associated metasediments \citep{McLelland2010a}.
+1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
 In the latest Mesoproterozoic ca.
@@
 There was significantly less crustal growth on the western margin of Laurentia and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
-A thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
-1.47 to 1.40 rifting which is typically interpretted to be intracratonic\citep{}.
+The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
+1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated.
+\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent while others envision it as being associated with continental rifting along the margin \citep[e.g.]{Jones2015a}.
 This region is interpreted to have been subsequently deformed during a ca.
@@
 1.71 Ga) to the end of the Shawinigan orogeny (ca.
-1.06 Ga) requires an long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
+1.06 Ga) requires a long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
 This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructured into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
@@
 The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting.
-Along the western margin of Laurentia small-scale rifting occured ca.
+Along the western margin of Laurentia small-scale rifting occurred ca.
 780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountins of NW Laurentia \citep{Rooney2017a}.
@@
 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a}.
-The emplacement of the Gunbarrel large igneous province along this margin and the extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin initiating ca.
-780 Ma.
-If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the late Ediacaran.
-The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occured later than the timing of rifting typically implemented in models of Rodinia break-up.
-It is conceivable that the rifting that drove latest Ediacaran to Cambrian thermal subsidence was associated with a microcontinent fragment rather than a major conjugate continent, but it is an important constraint for paleogeographic models that the time of major lithospheric thinning on the western Laurentia margin is ca.
-539 Ma.
+The emplacement of the ca.
+780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin.
+If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
+The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up.
+One possibility, along the lines of that proposed in \citet{Ross1981a}, is that ca.
+780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard.
+In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
 The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin.
 Available geochronological constraints and thermal subsidence modeling indicate ca.
-820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a,Halverson2018a}.
+820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}.
 These data suggest that conjugate continental lithosphere rifted away from northeast Greenland at the time.
-Extensive rifting that was followed by thermal subsidence occured along the southeast to east Laurentia margin at the Neoproterozoic-Phanerozoic boundary.
+
+Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
 A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as long term Cambrian to Ordovician subsidence along the margin.
-The Argentine pre-Cordillera Cuyania terrane is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting.
-As with other rifts, it is difficult to distinguish the seperation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar.
-One interpretation is that there was break-up along the eastern margin that initationed ca.
-XXX associated with the Rome trough and prior to the ca.
-539 Oklahoma aulacogen rifting that liberated Cuyania.
+The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this  history such as the rift development model of \citet{Burton2010a}.
+In this model, spatially-restricted rifting occurs ca.
+760 to 680 Ma in the region of modern-day North Carolina and Virginia.
+Ca.
+620-580 Ma rifting initiates in the region from modern-day New York to Newfoundland and by ca.
+580 to 550 Ma, rifting extends along the length of Laurentia's eastern margin.
+The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane.
+Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequent became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
+As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar.
+One interpretation is that there was successful break-up along the eastern margin during the ca.
+580 to 550 Ma interval of rifting prior to the ca.
+539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
+The Maz–Arequipa–Rio Apa (MARA) block MARA block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
+Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
+530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca.
+460 Famatinian orogeny.
 
-% Lu–Hf garnet ages have been interpreted to indicate that there were two metamorphic garnet growth events at 1330 Ma and 1080 Ma within metamorphic rocks of the Belt-Purcell Supergroup \citep{Nesheim2012a}.
-Ca.
-1.3 Ga orogenesis within the Belt Supergroup has been referred to as the East Kootenay Orogeny
+The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis.
+There are parallels between the Grenville orogenic interval and the Appalachian one in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian period) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
+These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision.
+In the case of the Grenville interval this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated to with the assembly of the supercontinent Pangea.
 
-Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic provides very strong evidence for mobile plate tectonics driving Laurentia evolution throughout the past 2 billion years.
-This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic rcord.
+Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic provides very strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
+This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.
 
-\section{Paleomagnetic pole compilation}
+\section*{Paleomagnetic pole compilation}
 
@@
 Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons.
-There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces.
+There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation.
 These data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
 Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.
+A lingering question raise in \citet{Hoffman198a} that still lingers is whether the Archean provicnes each had independent drit histories with significant speraation ro whether they have shared historieds before experiencing fragmantation and remamalgamation.
 

### 2019-07-12 manuscript.tex
edits to tectonic history section

@@
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography.
-There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as a rich geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
+There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
 
@@
 This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
-Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca.
+Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca.
 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca.
@@
 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}.
-As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior with still goingd undergoing continued translation up to ca.
+As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
 1.80 Ga \citep{Whitmeyer2007a}.
-A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks arrived ca.
-1.72 Ga \citep{Kilian2016a} which is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
+A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca.
+1.72 Ga \citep{Kilian2016a}.
+This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
 1.75 to 1.72 Ga \citep{Condit2015a} and ca.
-1.72 tectonomagmatic activity in the Black Hills region \citep{}.
+1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}.
 However, the evidence for earlier orogenesis ca.
@@
 \includegraphics[width=\textwidth]{Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}).
+\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 
-Crustal growth also progressed at this time through accretionary orogenesis.
+Crustal growth also progressed at this time in the Paleoproterozoic through accretionary orogenesis.
 This accretion occurred within the Wopmay orogen through ca.
@@
 In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
-The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g.
+The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+Nain provinces (often referred to as the Churchill domain or plate; e.g.
 \citealp{Skipton2016a, Weller2017a}).
-This model posits that this period terminal collision is not only associated with the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}.
+This model posits that this period terminal collision is not only associated with the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
 
 Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
-This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occured dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}).
+This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs.
+\ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
 Determining the extent of these belts is complicated by poor exposure of them in the Midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
 Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
-1.71 to 1.68 Ga Yavapai orogeny which is interpreted to have resulted from the accretion of a series of arc terrances that collided with each and Laurentia \citep{Whitmeyer2007a}.
+1.71 to 1.68 Ga Yavapai orogeny which is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Whitmeyer2007a}.
 Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 \includegraphics[width=4in]{Figures/Tectonic_history.pdf}
-\caption{\small{\textbf{Summary of Laurentia's tectonic history over the past $\sim$1.8 billion years. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}}
-\label{fig:Laurentia_map}
+\caption{\small{\textbf{Summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
+Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}}
+\label{fig:tectonic_history}
 \end{figure}
@@
 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
-In the latest Mesoproterozoic ca.
-1.11-1.08 Ga, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces.
+In the latest Mesoproterozoic (ca.
+1.11-1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces.
 This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
@@
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.
+1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen.
+The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.
 
-There was significantly less crustal growth on the western margin of Laurentia and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
+There was significantly less crustal growth on the western margin of Laurentia (Fig.
+\ref{fig:Laurentia_map}) and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
 The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
 1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated.
-\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent while others envision it as being associated with continental rifting along the margin \citep[e.g.]{Jones2015a}.
+\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep[e.g.]{Jones2015a}.
 This region is interpreted to have been subsequently deformed during a ca.
@@
 1.06 Ga) requires a long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
-This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but is sometimes other continental blocks are reconstructured into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
+This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructured into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
 Amazonia in \citealp{Elming2009a}).
@@
 The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting.
-Along the western margin of Laurentia small-scale rifting occurred ca.
-780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountins of NW Laurentia \citep{Rooney2017a}.
-However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until near the ca.
-539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a}.
+Along the western margin of Laurentia, small-scale rifting occurred ca.
+780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Rooney2017a}.
+However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran (closer to the ca.
+539 Ma Neoproterozoic-Phanerozoic boundary; \citealp{Bond1984a, Levy1991a}).
 The emplacement of the ca.
@@
 The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up.
-One possibility, along the lines of that proposed in \citet{Ross1981a}, is that ca.
+One possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca.
 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard.
@@
 Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
-A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as long term Cambrian to Ordovician subsidence along the margin.
-The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this  history such as the rift development model of \citet{Burton2010a}.
+A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a,Whitmeyer2007a}.
+The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this  history such as in the rift development model of \citet{Burton2010a}.
 In this model, spatially-restricted rifting occurs ca.
@@
 620-580 Ma rifting initiates in the region from modern-day New York to Newfoundland and by ca.
-580 to 550 Ma, rifting extends along the length of Laurentia's eastern margin.
+580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
 The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane.
-Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequent became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
+Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
 As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar.
@@
 The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis.
-There are parallels between the Grenville orogenic interval and the Appalachian one in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian period) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
+As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian one in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian period) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
 These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision.
@@
 These data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
-Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.
+Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.
 A lingering question raise in \citet{Hoffman198a} that still lingers is whether the Archean provicnes each had independent drit histories with significant speraation ro whether they have shared historieds before experiencing fragmantation and remamalgamation.
 
-For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
+For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
 This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny.
@@
 Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland.
-These terranes were once part of contigous Laurentia, but have subsequently rifted away.
+These terranes were once part of contiguous Laurentia, but have subsequently rifted away.
 These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:

### 2019-09-24 manuscript.tex
updates to project from summer

@@
 While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
-The categorizations used by the expert panel are `A' and `B' with the last panel meeting occuring in Fall 2017 in Leirubakki, Iceland.
+The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland.
 An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions.
@@
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
-\cite{Hall201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault.
-However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of thick plateau developed during continent-continent collision (e.g.
-\cite{}).
+\cite{Halls201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault.
+However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g.
+\citealp{Brown2012a}}).
 Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland.
@@
 
+The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
+
 To make an argument that differential plate tectonic motion began in the Neoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.

### 2019-09-24 manuscript.tex
edits to the Broad tectonic history overview section

@@
 
-Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have been a central constituent of both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
+Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography.
 There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
+This chapter seeks to provide a concise review of these records.
 
@@
 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}.
-As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
+As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
 1.80 Ga \citep{Whitmeyer2007a}.
@@
 \citealp{Skipton2016a, Weller2017a}).
-This model posits that this period terminal collision is not only associated with the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
+The \cite{Pehrsson2015a} model posits that this period terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
 
@@
 \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
-Determining the extent of these belts is complicated by poor exposure of them in the Midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
+Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
 Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
-1.71 to 1.68 Ga Yavapai orogeny which is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Whitmeyer2007a}.
+1.71 to 1.68 Ga Yavapai orogeny.
+Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Whitmeyer2007a}.
 Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 An interval of major plutonism occurred ca.
-1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces.
+1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
 This plutonism is likely due to crustal melting within a back-arc region of ca.
@@
 1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
+While an active margin interpretation with magmatism in back‐arc setting has gained traction within the literature with evidence, the tectonic setting is considered enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca.
-1.25–1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
+1.25-1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
 The subsequent ca.
@@
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen.
+1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}.
 The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.
@@
 1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated.
-\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep[e.g.]{Jones2015a}.
+\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep[e.g.]{Jones2015a}.
 This region is interpreted to have been subsequently deformed during a ca.
@@
 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}.
-These data suggest that conjugate continental lithosphere rifted away from northeast Greenland at the time.
+These data suggest that conjugate continental lithosphere rifted away from northeast Greenland ca.
+820 Ma.
 
@@
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
-The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane.
+The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
 Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
@@
 The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis.
-As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian one in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian period) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
+As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
 These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision.
-In the case of the Grenville interval this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated to with the assembly of the supercontinent Pangea.
+In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated to with the assembly of the supercontinent Pangea.
 

### 2019-10-03 manuscript.tex
develop the latex pole table

@@
 
-\section*{INTRODUCTION}
+\section*{Introduction}
 
@@
 \includegraphics[width=4in]{Figures/Tectonic_history.pdf}
-\caption{\small{\textbf{Summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
-Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}}
+\caption{\small{\textbf{Simplified summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
+Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}
 \label{fig:tectonic_history}
@@
 
-In this chapter, I will focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
+In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
 The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results.
@@
 Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
-These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}
+These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}.
+Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeography reconstructions.
+For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
+However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than for eruptive units or shallow-level intrusions.
+As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating.
+However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderence of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthnous origin discussed below).
 
-Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons.
+Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean provinces.
 There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation.
-These data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
+A lingering question raised in \citet{Hoffman198a} that still lingers is whether the Archean provinces each had independent drift histories with significant seperation or whether they have shared historieds before experiencing fragmantation and remamalgamation.
+Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
 Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.
-A lingering question raise in \citet{Hoffman198a} that still lingers is whether the Archean provicnes each had independent drit histories with significant speraation ro whether they have shared historieds before experiencing fragmantation and remamalgamation.
 
@@
 Uncertainty remains with respect to whether the ca.
-1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference grame.
+1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference frame.
 
@@
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
-\cite{Halls201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault.
+\cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault.
 However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g.
-\citealp{Brown2012a}}).
+\citealp{Brown2012a}).
 Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland.
@@
 \item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
-\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi2004
+\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi and Piper (2004)
 \item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin (Maloof et al., 2006 rotation)
 \end{itemize}
+
+{\scriptsize
+\begin{landscape}
+\input{pole_table_editted.tex}
+\end{landscape}
+}
+
+These paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia alone.
+Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:recons_time}.
+These reconstructions use the tectonic elements as
+
+As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
+While they provide constraints in this regard, they do not provide constraints in and of themselves
 

### 2019-10-04 manuscript.tex
Update manuscript.tex

@@
 There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation.
-A lingering question raised in \citet{Hoffman198a} that still lingers is whether the Archean provinces each had independent drift histories with significant seperation or whether they have shared historieds before experiencing fragmantation and remamalgamation.
+A lingering question raised in \citet{Hoffman198a} that still lingers is whether the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation.
 Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
@@
 For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
-This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny.
-\cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.
+This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that the Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny.
+\cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occurred in the time interval of 2.07 to 1.87 Ga.
 I follow this interpretation and group the poles into Superior (West) and Superior (East).
@@
 
-Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton.
+Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
-\cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault.
+\cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault.
 However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g.
@@
 
+\begin{table}[hbt]
+  \begin{tabular}{|l|l|l|l|l|l|l|}
+    & & \\
+    \hline
+Laurentia-Greenland &  1202 & 1800.0 & 67.5 & -118.5  -13.8 & 1000 & Rotation of Greenland to Laurentia \\
+    & &
+  \end{tabular}
+  \caption{}
+\end{table}
+
 \begin{itemize}
 \item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
-\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi and Piper (2004)
-\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin (Maloof et al., 2006 rotation)
+\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to \cite{Darabi2004a}
+\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}
 \end{itemize}
-
-{\scriptsize
-\begin{landscape}
-\input{pole_table_editted.tex}
-\end{landscape}
-}
 
@@
 Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:recons_time}.
-These reconstructions use the tectonic elements as
+These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth.
+
+Through the Proterozoic there are intervals when there
 
 As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
-While they provide constraints in this regard, they do not provide constraints in and of themselves
+While they provide constraints in this regard, they do not provide constraints in and of themselves of the longitudinal position of the block.
+Other approaches to obtain paleolongitude utilize geophysical hypotheses that are actualized through different methods such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed.
 
@@
 
+{\scriptsize
+\begin{landscape}
+\input{pole_table_editted.tex}
+\end{landscape}
+}
+
 \end{document}

### 2019-10-09 Manuscript/manuscript.tex
repository organization

--- 2019-10-04 manuscript.tex
@@
 
-%\linenumbers
+\linenumbers
 
@@
 1.06 Ga) requires a long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
-This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructured into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
+This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
 Amazonia in \citealp{Elming2009a}).
@@
 The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results.
-While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
+While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
 The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland.
@@
 As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating.
-However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderence of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthnous origin discussed below).
+However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderence of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{Figures/Laurentia_poles_combined.pdf}
+\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard).
+Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
+\label{fig:Laurentia_poles}
+\end{figure}
+
 Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
@@
 \begin{table}[hbt]
-  \begin{tabular}{|l|l|l|l|l|l|l|}
-    & & \\
+\begin{tabular}{|l|l|l|l|p{2 in}|}
   \hline
-Laurentia-Greenland &  1202 & 1800.0 & 67.5 & -118.5  -13.8 & 1000 & Rotation of Greenland to Laurentia \\
-    & &
+& Euler pole & Euler pole & rotation & note and \\
+Block & longitude & latitude & angle & citation \\
+\hline
+Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989} \\
+\hline
+Scotland & 161.9 & 78.6 & -31.0 & Reconstructing Atlantic opening following \cite{Torsvik2017a} \\
+\hline
+Svalbard & 125.0 & -81.0 & 68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
+\hline
 \end{tabular}
@@
 
-\begin{itemize}
-\item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
-\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to \cite{Darabi2004a}
-\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}
-\end{itemize}
+Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
+The implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+There are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca.
+1480 and 1380 from Laurentia (including Greenland poles).
+The best-constrained portion of the record is associated with 1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
+The late Tonian Period also has a number of poles the Gunbarrel LIP (ca.
+780 Ma) and Franklin LIP (ca.
+720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins.
+Overall the record of paleomagnetic poles has a lot of consistency with progressive paths and agreement between poles.
 
-These paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia alone.
+#A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{Figures/Laurentia_paleolatitude.pdf}
+\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW).
+The paleomagnetic poles are compiled in Table \ref{tab:Laurentia_poles}.
+Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
+\label{fig:Laurentia_paleolatitude}
+\end{figure}
+
+The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia.
+The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012}.
+Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.
+
+\section{Paleogeographic reconstructions}
+
 Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:recons_time}.
 These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth.
-
-Through the Proterozoic there are intervals when there
-
 As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
-While they provide constraints in this regard, they do not provide constraints in and of themselves of the longitudinal position of the block.
-Other approaches to obtain paleolongitude utilize geophysical hypotheses that are actualized through different methods such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed.
+While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block.
+Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
 

### 2019-10-09 Manuscript/pole_table.tex
repository organization

\begin{longtable}{p{1 in}p{1 in}rrrrrrr}
\toprule
                       terrane &                                          unit name & rating &  site lon &  site lat &  plon &  plat &  A\$\_\{95\}\$ &                    age \\
\midrule
\endhead
\midrule
\multicolumn{9}{r}{{Continued on next page}} \\
\midrule
\endfoot

\bottomrule
\endlastfoot
             Laurentia-Wyoming &                            Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
      Laurentia-Superior(East) &                       Otto Stock Dykes and Aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
               Laurentia-Slave &                                       Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
      Laurentia-Superior(East) &                                     PTARMIGAN MEAN &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
      Laurentia-Superior(East) &                                       MATACHEWAN R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ \\
      Laurentia-Superior(East) &                                       MATACHEWAN N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
               Laurentia-Slave &                                       Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
      Laurentia-Superior(East) &                                         SENNETERRE &      A &     283.0 &      49.0 & 284.3 & -15.3 &       6.0 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
      Laurentia-Superior(East) &                                       NIPISSING N1 &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
               Laurentia-Slave &                                       Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
      Laurentia-Superior(East) &                                        BISCOTASING &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ \\
               Laurentia-Slave &                                        Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ \\
      Laurentia-Superior(West) &                                         MARATHON N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
      Laurentia-Superior(West) &                                         MARATHON R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
      Laurentia-Superior(West) &                                       CAUCHON LAKE &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
      Laurentia-Superior(West) &                                       FORT FRANCES &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
      Laurentia-Superior(East) &                                         LAC ESPRIT &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
      Laurentia-Greenland-Nain &                                    Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ \\
               Laurentia-Slave &                                  Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
      Laurentia-Superior(East) &                                              MINTO &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
               Laurentia-Slave &                    Rifle (Western River) Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
                 Laurentia-Rae &                             Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ \\
             Laurentia-Wyoming &                         Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
               Laurentia-Slave &                                   Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ \\
               Laurentia-Slave &                           MEAN Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
               Laurentia-Slave &                     MEAN Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
      Laurentia-Superior(West) &                                        MOLSON B+C2 &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ \\
               Laurentia-Slave &                                 Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ \\
            Laurentia-Superior &                          MEAN Haig/Flaherty/Sutton &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
               Laurentia-Slave &  MEAN Pearson A/Peninsular sill/Kilohigok basin... &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
 Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &      B &     258.1 &      54.7 & 279.4 &  62.4 &       7.9 &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
                 Laurentia-Rae &                                      Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
                 Laurentia-Rae &                                   Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
                     Laurentia &                                      Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ \\
 Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
 Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
                     Laurentia &                                      Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
           Laurentia-Greenland &                        Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
                     Laurentia &                            Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &                 St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ \\
                     Laurentia &                      Michikamau Intrusion Combined &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
                     Laurentia &                                  Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ \\
                     Laurentia &                                 Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ \\
                     Laurentia &                    Tobacco Root Dykes - A combined &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ \\
                     Laurentia &                                       Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ \\
                     Laurentia &                     MEAN Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ \\
                     Laurentia &                                   Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ \\
                     Laurentia &                                 McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
                     Laurentia &         Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ \\
           Laurentia-Greenland &                                Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
           Laurentia-Greenland &                              Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
           Laurentia-Greenland &                      Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                                   Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ \\
           Laurentia-Greenland &                                  North Qoroq Intr. &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
           Laurentia-Greenland &                                  Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                         Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
           Laurentia-Greenland &                         West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ \\
           Laurentia-Greenland &                      West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ \\
                     Laurentia &                             Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
            Laurentia-Scotland &                                   MEAN Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ \\
           Laurentia-Greenland &                                     Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
           Laurentia-Greenland &                                 Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
           Laurentia-Greenland &                                  South Qoroq Intr. &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
           Laurentia-Greenland &                                 Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
           Laurentia-Greenland &                          NE-SW Trending Dyke Swarm &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
                     Laurentia &                                      Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                       MEAN Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1111\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
                     Laurentia &                           Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &                          Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
                     Laurentia &                           Upper Osler volcanics -R &      A &     272.4 &      48.7 & 201.6 &  42.5 &       3.7 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                 Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
                     Laurentia &              Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1098\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &                               North Shore lavas -N &      A &     268.7 &      46.3 & 181.3 &  34.5 &       2.8 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &                              Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                             Portage Lake Volcanics &      A &     271.2 &      47.0 & 178.0 &  26.7 &       4.7 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &                    Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
                     Laurentia &                           Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ \\
                     Laurentia &                        Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ \\
                     Laurentia &                                   Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1087\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                      Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
                     Laurentia &                                    Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ \\
                     Laurentia &                                     Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ \\
                     Laurentia &                              Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ \\
            Laurentia-Scotland &                                MEAN Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ \\
            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ \\
            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ \\
                     Laurentia &                     MEAN Wyoming "Gunbarrel" dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                           Tsezotene Sills Combined &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                               Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ \\
            Laurentia-Svalbard &                          Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ \\
                     Laurentia &                          Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
                     Laurentia &                                   Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
                     Laurentia &                         Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
                     Laurentia &                                   Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
                     Laurentia &                        Sept-Iles Layered Intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
\end{longtable}

### 2019-10-11 Manuscript/manuscript.tex
progress throughout project including the development of paleogeo recons using WK07 polygons

@@
 \centering
-\includegraphics[width=\textwidth]{Figures/Fig1_map.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
 \caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}).
@@
 \centering
-\includegraphics[width=4in]{Figures/Tectonic_history.pdf}
+\includegraphics[width=4in]{../Figures/Tectonic_history.pdf}
 \caption{\small{\textbf{Simplified summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
+\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard).
+Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
+\label{fig:Laurentia_poles}
+\end{figure}
+
 In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
@@
 There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation.
-A lingering question raised in \citet{Hoffman198a} that still lingers is whether the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation.
 Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
+A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation.
+An analysis of the record strengthened with a new paleomagnetic pole from the Slave province in \cite{Mitchell2014a}, supports significant differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga.
 Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{Figures/Laurentia_poles_combined.pdf}
-\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard).
-Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
-\label{fig:Laurentia_poles}
-\end{figure}
-
 Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
@@
 These terranes were once part of contiguous Laurentia, but have subsequently rifted away.
-These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:
+These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the  rotations shown in Table \ref{tab:terrane_rotations}.
+
+Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
+To visualize the temoporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+There are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca.
+1480 and 1380 from Laurentia (including Greenland poles).
+The best-constrained portion of the record is associated with 1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
+The late Tonian Period also has a number of poles the Gunbarrel LIP (ca.
+780 Ma) and Franklin LIP (ca.
+720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins.
+Overall the record of paleomagnetic poles has a lot of consistency with progressive paths and agreement between poles.
+
+%A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
+\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW).
+The paleomagnetic poles are compiled in Table \ref{tab:Laurentia_poles}.
+Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
+\label{fig:Laurentia_paleolatitude}
+\end{figure}
+
+The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia.
+The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012}.
+Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.
+
+\section*{Paleogeographic reconstructions}
+
+Seeking to developing comprehensive continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types.
+Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles.
+Paleomagnetic poles compilations themselves are evolving with better data and improved geochronology.
+Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
+\caption{\small{\textbf{Paleogeogrpahic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.
+These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
+Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.}}}
+\label{fig:Laurentia_reconstructions}
+\end{figure}
+
+ There is an overall lack of published continuous models in the literature for the Proterozoic that can be compared to the compilation of paleomagnetic poles presented herein.
+The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters.
+With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g.
+\citealp{Torsvik2012a}).
+
+ An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017a} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}.
+The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
+As noted above, the development of such models is challenging and the researchers need to balance varying constraints.
+The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia.
+The model does not honor the Grenville loop (e.g.
+go to moderately high southerly latitudes ca.
+1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models.
+Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole despite its `A' Nordic rating.
+The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.
+
+There are more published models that show snapshots at given time intervals \cite[e.g.]{Li2008a}.  The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
+
+In this work, I develop a continuous paleogeogrpahic reconstruciton of Laurentia constrained the paleomagnetic poles compilation.
+This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (900-800 Ma for example).
+One could use interpretations of paleogeographic connections with other cratons (e.g.
+Baltica in the Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself.
+The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:Laurentia_reconstructions.pdf}.
+These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth.
+As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
+While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block.
+Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
+In Figure \ref{fig:Laurentia_reconstructions.pdf}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
+
+To make an argument that differential plate tectonic motion began in the Neoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
+
+\section*{Acknowledgements}
+
+Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
+Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists as well as additional efforts of Lauri Pesonen in maintining the Paleomagia database \citep{Veikkolainen2014a}.
+GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}.
+Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive python environment \citep{Perez2007a}.
+This work was supported by NSF CAREER Grant 1847277 awarded to N.L.S.-H.
+
+\bibliographystyle{gsabull}
+\small\bibliography{../../references/allrefs}
 
@@
 
-Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
-The implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}.
-There are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca.
-1480 and 1380 from Laurentia (including Greenland poles).
-The best-constrained portion of the record is associated with 1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
-The late Tonian Period also has a number of poles the Gunbarrel LIP (ca.
-780 Ma) and Franklin LIP (ca.
-720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins.
-Overall the record of paleomagnetic poles has a lot of consistency with progressive paths and agreement between poles.
-
-#A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{Figures/Laurentia_paleolatitude.pdf}
-\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW).
-The paleomagnetic poles are compiled in Table \ref{tab:Laurentia_poles}.
-Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
-\label{fig:Laurentia_paleolatitude}
-\end{figure}
-
-The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia.
-The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012}.
-Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.
-
-\section{Paleogeographic reconstructions}
-
-Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:recons_time}.
-These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth.
-As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
-While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block.
-Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
-
-The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
-
-To make an argument that differential plate tectonic motion began in the Neoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
-
-\bibliographystyle{gsabull}
-\small\bibliography{../references/allrefs}
-
 {\scriptsize

### 2019-10-11 Manuscript/manuscript.tex
add Penokean details to text and add to tectonic summary

@@
 1.88 to 1.84 Ga \citep{Hildebrand2009a}.
-Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior province with the last evidence of that orogeny being ca.
-1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
+Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
+1.86 to 1.82 \cite{Schulz2007a}.
+Firm evidence of the end of the orogeny comes from the ca.
+1.78 undeformed plutons of the post-Penokean East Central Minnesota Batholith \citep{Holm2005a}.
 
@@
 \centering
-\includegraphics[width=4in]{../Figures/Tectonic_history.pdf}
-\caption{\small{\textbf{Simplified summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
-Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}
+\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
+\caption{\small{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
+A timeline of large igneous provinces  (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown.
+The interpreted age of paleomagnetic poles for Laurentia (not included separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.}}
 \label{fig:tectonic_history}

### 2019-10-11 Manuscript/manuscript.tex
add more details about Superior Slave pole comparison that now cites Buchan2009a

@@
 A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation.
-An analysis of the record strengthened with a new paleomagnetic pole from the Slave province in \cite{Mitchell2014a}, supports significant differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga.
+The strongest comparison in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a}.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either ca.
+2.17 Ga \citep{Mitchell2014a} or ca.
+2.02 Ga \citep{Buchan2009a}.
+These data provide paleomagnetic support for the hypothesis that the Trans-Hudson orogeny is the result of closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces as the result of differential plate tectonic motion.
 Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.

### 2019-10-11 Manuscript/manuscript.tex
add reference to Buchan2016a while discussing relative Superior and Slave motion

--- 2019-10-11 Manuscript/manuscript.tex
@@
 A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation.
-An analysis of the record strengthened with a new paleomagnetic pole from the Slave province in \cite{Mitchell2014a}, supports significant differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga.
+The strongest comparison in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga \citep{Buchan2016a}.
+These data provide paleomagnetic support for the hypothesis that the Trans-Hudson orogeny is the result of closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces as the result of differential plate tectonic motion.
 Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.

### 2019-10-17 Manuscript/manuscript.tex
Update manuscript.tex

@@
 A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca.
-1.72 Ga \citep{Kilian2016a}.
+1.72 Ga \citep{Kilian2016b}}.
 This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
@@
 In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
-The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results.
+The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume).
 While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
@@
 As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating.
-However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderence of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).
+However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).
 
@@
 Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
-A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and remamalgamation.
+A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and reamalgamation.
 The strongest comparison in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
@@
 I follow this interpretation and group the poles into Superior (West) and Superior (East).
-These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame.
 Uncertainty remains with respect to whether the ca.
@@
 
+%These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame. NEED TO FIGURE OUT HOW TO SHOW THIS
+
 Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
-\cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault.
+\cite{Halls2015b}} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault.
 However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g.
@@
 These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the  rotations shown in Table \ref{tab:terrane_rotations}.
+The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used).
+The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation of \citealp{Roest1989a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland.
+The reconstruction of Svalbard is more challenging given a multi-state tectonic history.
+The preferred Euler of \cite{Maloof2006a} is used here that designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a}.
 
@@
 To visualize the temoporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}.
-There are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca.
+The ages of poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}.
+Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both.
+In addition, intraplate magmatism resulting from plume-related large-igneous provinces can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quisence (e.g.
+the ca.
+1267 Ma Mackenzie LIP; Fig.
+\ref{fig:tectonic_history}.
+In particular, there are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12  poles between ca.
 1480 and 1380 from Laurentia (including Greenland poles).
-The best-constrained portion of the record is associated with 1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
-The late Tonian Period also has a number of poles the Gunbarrel LIP (ca.
+The best-constrained portion of the record is associated with ca.
+1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
+The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca.
 780 Ma) and Franklin LIP (ca.
 720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins.
-Overall the record of paleomagnetic poles has a lot of consistency with progressive paths and agreement between poles.
+Overall the record of paleomagnetic poles has a lot of internal consistency with progressive paths and agreement between poles such as for the Keweenawan Track \citep{Swanson-Hysell2019a} and the late Tonian path \citep{Eyster2019a}.
+
+Data from other terranes add resolution to the record.
+In particular, data from Greenland adds 12 poles between 1385 and 1160 Ma when there are only 4 poles from mainland Laurentia.
+Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:, these poles can be rotated into Laurentia coordinates and used for reconstruction of the entire continent.
 
@@
 \end{figure} 
+%ADD MORE RECONSTRUCTIONS. 1 per 100 million years perhaps with indication of which ones are well-constrained by models and
 
@@
 
-In this work, I develop a continuous paleogeogrpahic reconstruciton of Laurentia constrained the paleomagnetic poles compilation.
+In this work, I develop a continuous paleogeographic reconstruciton of Laurentia constrained the paleomagnetic poles compilation.
 This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (900-800 Ma for example).
@@
 Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
-In Figure \ref{fig:Laurentia_reconstructions.pdf}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
+In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
+
+\section*{Conclusion}
+
+The paleogeographic record is rich in constraints through the Precambrian -- certainly the
 
@@
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
-Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists as well as additional efforts of Lauri Pesonen in maintining the Paleomagia database \citep{Veikkolainen2014a}.
+Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists as well as additional efforts of Lauri Pesonen in maintaining the Paleomagia database \citep{Veikkolainen2014a}.
 GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}.
@@
 \hline
-Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989} \\
+Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
 \hline
@@
 \caption{}
+\label{tab:rotations}
 \end{table}

### 2019-10-19 Code/pole_table.tex
progress on reconstructions, now plotting with poles

\begin{longtable}{p{1 in}p{1 in}rrrrrrr}
\toprule
                       terrane &                                          unit name & rating &  site lon &  site lat &  plon &  plat &  A\$\_\{95\}\$ &                    age &                                     pole reference \\
\midrule
\endhead
\midrule
\multicolumn{10}{r}{{Continued on next page}} \\
\midrule
\endfoot

\bottomrule
\endlastfoot
             Laurentia-Wyoming &                            Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Selkin2008a\} \\
      Laurentia-Superior(East) &                       Otto Stock Dykes and Aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Pullaiah1975b\} \\
               Laurentia-Slave &                                       Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
      Laurentia-Superior(East) &                                     PTARMIGAN MEAN &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
      Laurentia-Superior(East) &                                       MATACHEWAN R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
      Laurentia-Superior(East) &                                       MATACHEWAN N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
               Laurentia-Slave &                                       Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2012a\} \\
      Laurentia-Superior(East) &                                         SENNETERRE &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
      Laurentia-Superior(East) &                                       NIPISSING N1 &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
               Laurentia-Slave &                                       Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
      Laurentia-Superior(East) &                                        BISCOTASING &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Kilian2015a\} \\
               Laurentia-Slave &                                        Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
      Laurentia-Superior(West) &                                         MARATHON N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
      Laurentia-Superior(West) &                                         MARATHON R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
      Laurentia-Superior(West) &                                       CAUCHON LAKE &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
      Laurentia-Superior(West) &                                       FORT FRANCES &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
      Laurentia-Superior(East) &                                         LAC ESPRIT &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
      Laurentia-Greenland-Nain &                                    Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                                NaN \\
               Laurentia-Slave &                                  Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Buchan2009a\} \\
      Laurentia-Superior(East) &                                              MINTO &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
               Laurentia-Slave &                    Rifle (Western River) Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans1981a\} \\
                 Laurentia-Rae &                             Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Halls1999a\} \\
             Laurentia-Wyoming &                         Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Kilian2016a\} \\
               Laurentia-Slave &                                   Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
               Laurentia-Slave &                           MEAN Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
               Laurentia-Slave &                     MEAN Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
      Laurentia-Superior(West) &                                        MOLSON B+C2 &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                \textbackslash citep\{Irving1979a\} \\
               Laurentia-Slave &                                 Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                \textbackslash citep\{Irving1979a\} \\
            Laurentia-Superior &                          MEAN Haig/Flaherty/Sutton &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
               Laurentia-Slave &  MEAN Pearson A/Peninsular sill/Kilohigok basin... &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
 Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                 \textbackslash cite\{Symons1999a\} \\
                 Laurentia-Rae &                                      Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{McGlynn1974a\} \\
                 Laurentia-Rae &                                   Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                  \textbackslash cite\{Evans1973a\} \\
                     Laurentia &                                      Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                    \textbackslash cite\{Park1973\} \\
 Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons2000a\} \\
 Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                   \textbackslash cite\{Gala1995a\} \\
                     Laurentia &                                      Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Irving2004\} \\
           Laurentia-Greenland &                        Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
                     Laurentia &                            Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Irving1972\} \\
                     Laurentia &                 St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &                                  \textbackslash cite\{Meert2002b\} \\
                     Laurentia &                      Michikamau Intrusion Combined &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Emslie1976a\} \\
                     Laurentia &                                  Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                  \textbackslash cite\{Elston2002\} \\
                     Laurentia &                                 Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                  \textbackslash cite\{Elston2002\} \\
                     Laurentia &                    Tobacco Root Dykes - A combined &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &                                 \textbackslash cite\{Harlan2008a\} \\
                     Laurentia &                                       Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Elston2002\} \\
                     Laurentia &                     MEAN Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &  mean pole calculated by Nordic workshop partic... \\
                     Laurentia &                                   Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                 \textbackslash cite\{Fahrig1976a\} \\
                     Laurentia &                                 McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Elston2002\} \\
                     Laurentia &         Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Elston2002\} \\
           Laurentia-Greenland &                                Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
           Laurentia-Greenland &                              Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
           Laurentia-Greenland &                      Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                                   Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash citep\{Murthy1978a\} \\
           Laurentia-Greenland &                                  North Qoroq Intr. &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
           Laurentia-Greenland &                                  Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                         Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
           Laurentia-Greenland &                         West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                                NaN \\
           Laurentia-Greenland &                      West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
                     Laurentia &                             Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
            Laurentia-Scotland &                                   MEAN Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                                                NaN \\
           Laurentia-Greenland &                                     Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
           Laurentia-Greenland &                                 Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
           Laurentia-Greenland &                                  South Qoroq Intr. &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
           Laurentia-Greenland &                                 Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
           Laurentia-Greenland &                          NE-SW Trending Dyke Swarm &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
                     Laurentia &                                      Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                       MEAN Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1111\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
                     Laurentia &                           Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
                     Laurentia &                          Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
                     Laurentia &                           Upper Osler volcanics -R &      A &     272.4 &      48.7 & 201.6 &  42.5 &       3.7 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                 Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
                     Laurentia &              Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1098\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
                     Laurentia &                               North Shore lavas -N &      A &     268.7 &      46.3 & 181.3 &  34.5 &       2.8 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
                     Laurentia &                              Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                             Portage Lake Volcanics &      A &     271.2 &      47.0 & 178.0 &  26.7 &       4.7 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
                     Laurentia &                    Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
                     Laurentia &                           Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &                                                NaN \\
                     Laurentia &                        Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
                     Laurentia &                                   Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1087\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                      Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
                     Laurentia &                                    Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                                NaN \\
                     Laurentia &                                     Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                                NaN \\
                     Laurentia &                              Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                                NaN \\
            Laurentia-Scotland &                                MEAN Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                                                NaN \\
            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                                NaN \\
            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
                     Laurentia &                     MEAN Wyoming "Gunbarrel" dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                           Tsezotene Sills Combined &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                               Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                                NaN \\
            Laurentia-Svalbard &                          Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                                NaN \\
                     Laurentia &                          Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
                     Laurentia &                                   Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                     Laurentia &                         Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
                     Laurentia &                                   Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
                     Laurentia &                        Sept-Iles Layered Intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
\end{longtable}

### 2019-10-19 Manuscript/manuscript.tex
add Neoproterozoic rifts to Fig 1

@@
 Along the western margin of Laurentia, small-scale rifting occurred ca.
-780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Rooney2017a}.
+780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Rooney2017a}.
 However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran (closer to the ca.
@@
 Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
-A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a,Whitmeyer2007a}.
+A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
+\ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a,Whitmeyer2007a}.
 The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}.
@@
 760 to 680 Ma in the region of modern-day North Carolina and Virginia.
-Ca.
-620-580 Ma rifting initiates in the region from modern-day New York to Newfoundland and by ca.
+Rifting ca.
+620-580 Ma initiates in the region from modern-day New York to Newfoundland and by ca.
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
-The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
+The last phases of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
 Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
@@
 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca.
-460 Famatinian orogeny.
+460 Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.
 
@@
 These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision.
-In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated to with the assembly of the supercontinent Pangea.
+In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated with the assembly of the supercontinent Pangea.
 
@@
 
-Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean provinces.
+Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), paleomagnetic poles need to be considered with respect to the individual Archean provinces.
 There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation.
@@
 A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and reamalgamation.
-The strongest comparison in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
+The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
 High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga \citep{Buchan2016a}.
@@
 
-Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
+Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
 To visualize the temoporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}.

### 2019-10-19 Manuscript/manuscript.tex
touch of text about paleogeo models

--- 2019-10-17 Manuscript/manuscript.tex
@@
 
-There are more published models that show snapshots at given time intervals \cite[e.g.]{Li2008a}.  The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
+There are more published models that show snapshots and publish rotation parameters associated with given time intervals (e.g.
+the rotation model of \citealp{Li2008a}).
+The position for Laurentia implied by the snapshot of the \cite{Li2008a} model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+
+ The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
 

### 2019-10-20 Manuscript/manuscript.tex
minor edits to text

--- 2019-10-19 Manuscript/manuscript.tex
@@
 A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca.
-1.72 Ga \citep{Kilian2016b}}.
+1.72 Ga \citep{Kilian2016b}.
 This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
@@
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
-\cite{Halls2015b}} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault.
+\cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault.
 However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g.
@@
 In particular, data from Greenland adds 12 poles between 1385 and 1160 Ma when there are only 4 poles from mainland Laurentia.
-Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:, these poles can be rotated into Laurentia coordinates and used for reconstruction of the entire continent.
+Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:rotations}) these poles can be rotated into Laurentia coordinates and used for reconstruction of the entire continent.
 
@@
 
-There are more published models that show snapshots at given time intervals \cite[e.g.]{Li2008a}.  The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
+There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, but did not publish parameters for a continuous model.
+The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude}.
+
+ The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
 

### 2019-10-20 Code/pole_table.tex
split Greenland into pieces and make Superior/Slave reconstructions

@@
                Laurentia-Slave &                                       Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2012a\} \\
-      Laurentia-Superior(East) &                                         SENNETERRE &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
-      Laurentia-Superior(East) &                                       NIPISSING N1 &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
+      Laurentia-Superior(East) &                                         Senneterre &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
+      Laurentia-Superior(East) &                                       Nipissing N1 &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
                Laurentia-Slave &                                       Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
-      Laurentia-Superior(East) &                                        BISCOTASING &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(East) &                                        Biscotasing &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
              Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Kilian2015a\} \\
@@
                      Laurentia &                                      Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Irving2004\} \\
-           Laurentia-Greenland &                        Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
+           Laurentia-Greenland &                        Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Halls2011a\} \\
                      Laurentia &                            Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Irving1972\} \\
@@
            Laurentia-Greenland &                                  Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                         Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
+                     Laurentia &                         Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
            Laurentia-Greenland &                         West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                                NaN \\
            Laurentia-Greenland &                      West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
-                     Laurentia &                             Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
-            Laurentia-Scotland &                                   MEAN Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                                                NaN \\
+                     Laurentia &                             Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Palmer1977a\} \\
+            Laurentia-Scotland &                                   MEAN Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                  Lulea Nordic workshop calculation \\
            Laurentia-Greenland &                                     Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
@@
            Laurentia-Greenland &                          NE-SW Trending Dyke Swarm &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
-                     Laurentia &                                      Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                       MEAN Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1111\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
-                     Laurentia &                           Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
-                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
-                     Laurentia &                          Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
-                     Laurentia &                           Upper Osler volcanics -R &      A &     272.4 &      48.7 & 201.6 &  42.5 &       3.7 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                 Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
-                     Laurentia &              Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1098\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
-                     Laurentia &                               North Shore lavas -N &      A &     268.7 &      46.3 & 181.3 &  34.5 &       2.8 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
-                     Laurentia &                              Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                             Portage Lake Volcanics &      A &     271.2 &      47.0 & 178.0 &  26.7 &       4.7 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
-                     Laurentia &                    Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
-                     Laurentia &                           Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &                                                NaN \\
-                     Laurentia &                        Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
-                     Laurentia &                                   Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1087\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                      Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
-                     Laurentia &                                    Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                                NaN \\
-                     Laurentia &                                     Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                                NaN \\
-                     Laurentia &                              Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                                NaN \\
-            Laurentia-Scotland &                                MEAN Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                                                NaN \\
-            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                                NaN \\
-            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
-                     Laurentia &                     MEAN Wyoming "Gunbarrel" dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                           Tsezotene Sills Combined &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                               Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                                NaN \\
-            Laurentia-Svalbard &                          Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                                NaN \\
-                     Laurentia &                          Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
-                     Laurentia &                                   Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
-                     Laurentia &                         Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
-                     Laurentia &                                   Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
-                     Laurentia &                        Sept-Iles Layered Intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
+                     Laurentia &                                      Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Ernst1993a\} \\
+                     Laurentia &                       MEAN Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                  Lulea Nordic workshop calculation \\
+                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1109\$\textasciicircum \{+2\}\$\$\_\{-3\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                           Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
+                     Laurentia &                          Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
+                     Laurentia &                           Upper Osler volcanics -R &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &     1105\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                         \textbackslash cite\{Swanson-Hysell2019a\} \\
+                     Laurentia &                 Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+3\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                               North Shore lavas -N &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &              \textbackslash cite\{Tauxe2009a,Swanson-Hysell2019a\} \\
+                     Laurentia &                              Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash citep\{Kean1997a\} \\
+                     Laurentia &                             Portage Lake Volcanics &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &  \textbackslash cite\{Books1972a, Hnat2006a\} as calculated in ...
+\\
+                     Laurentia &              Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094\$\textasciicircum \{+6\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                    Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                   \textbackslash cite\{Weil2003a\} \\
+                     Laurentia &                           Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
+                     Laurentia &                        Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                               \textbackslash cite\{Donadini2011a\} \\
+                     Laurentia &                                   Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
+                     Laurentia &                      Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
+                     Laurentia &                                     Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &    1080\$\textasciicircum \{+4\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
+                     Laurentia &                                    Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1070\$\textasciicircum \{+14\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
+                     Laurentia &                              Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash cite\{Warnock2000a\} \\
+            Laurentia-Scotland &                                MEAN Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                  Lulea Nordic workshop calculation \\
+            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                     MEAN Wyoming "Gunbarrel" dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Lulea Nordic workshop calculation based on dat...
+\\
+                     Laurentia &                           Tsezotene Sills Combined &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
+                     Laurentia &                               Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006a\} \\
+            Laurentia-Svalbard &                          Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                          Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                               \textbackslash cite\{Denyszyn2009a\} \\
+                     Laurentia &                                   Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\
+                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                         Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons1991a\} \\
+                     Laurentia &                                   Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Meert1994a\} \\
+                     Laurentia &                        Sept-Iles Layered Intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Tancyk1987a\} \\
 \end{longtable}

### 2019-10-20 Manuscript/manuscript.tex
small reference fixes to text

@@
 \includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a}).
+\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
-These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}.
+These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}.
 Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeography reconstructions.
@@
 The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia.
-The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012}.
+The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}.
 Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.
@@
 \includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
-\caption{\small{\textbf{Paleogeogrpahic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.
+\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.
 These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
@@
 
- An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017a} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}.
+ An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}.
 The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
@@
 \bibliographystyle{gsabull}
-\small\bibliography{../../references/allrefs}
+\footnotesize\bibliography{../../references/allrefs}
 

### 2019-10-20 Manuscript/manuscript.tex
add Superior_Slave_reconstructions and related text

@@
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga \citep{Buchan2016a}.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+\ref{fig:Superior_Slave_recons};\citealp{Buchan2016a}.
 These data provide paleomagnetic support for the hypothesis that the Trans-Hudson orogeny is the result of closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces as the result of differential plate tectonic motion.
+Reconstructions made of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.
+
+%The poles on this figure should be labeled
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Superior_Slave_reconstructions.pdf}
+\caption{\small{\textbf{Paleogeographic reconstructions using poles from the Superior, Slave and Rae provinces.
+Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
+Poles with ages that are within 25 million years of the given time slice are shown.
+The relatively well-resolved pole paths from the Superior and Slave provinces (Fig.
+\ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}}}
+\label{fig:Superior_Slave_recons}
+\end{figure}
+
  Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.

### 2019-10-21 Manuscript/pole_table.tex
clean up pole table

@@
-\begin{longtable}{p{1 in}p{1 in}rrrrrrr}
+\begin{longtable}{p{1 in}p{1 in}rrrrrrrr}
 \toprule
-                       terrane &                                          unit name & rating &  site lon &  site lat &  plon &  plat &  A\$\_\{95\}\$ &                    age \\
+                       terrane &                                        unit name & rating &  site lon &  site lat &  plon &  plat &  A\$\_\{95\}\$ &                    age &                                     pole reference \\
 \midrule
@@
 \midrule
-\multicolumn{9}{r}{{Continued on next page}} \\
+\multicolumn{10}{r}{{Continued on next page}} \\
 \midrule
@@
 \endlastfoot
-             Laurentia-Wyoming &                            Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-      Laurentia-Superior(East) &                       Otto Stock Dykes and Aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-               Laurentia-Slave &                                       Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-      Laurentia-Superior(East) &                                     PTARMIGAN MEAN &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-      Laurentia-Superior(East) &                                       MATACHEWAN R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ \\
-      Laurentia-Superior(East) &                                       MATACHEWAN N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-               Laurentia-Slave &                                       Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-      Laurentia-Superior(East) &                                         SENNETERRE &      A &     283.0 &      49.0 & 284.3 & -15.3 &       6.0 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
-      Laurentia-Superior(East) &                                       NIPISSING N1 &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-               Laurentia-Slave &                                       Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-      Laurentia-Superior(East) &                                        BISCOTASING &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ \\
-               Laurentia-Slave &                                        Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ \\
-      Laurentia-Superior(West) &                                         MARATHON N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-      Laurentia-Superior(West) &                                         MARATHON R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-      Laurentia-Superior(West) &                                       CAUCHON LAKE &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-      Laurentia-Superior(West) &                                       FORT FRANCES &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-      Laurentia-Superior(East) &                                         LAC ESPRIT &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-      Laurentia-Greenland-Nain &                                    Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ \\
-               Laurentia-Slave &                                  Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-      Laurentia-Superior(East) &                                              MINTO &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-               Laurentia-Slave &                    Rifle (Western River) Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
-                 Laurentia-Rae &                             Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ \\
-             Laurentia-Wyoming &                         Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-               Laurentia-Slave &                                   Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ \\
-               Laurentia-Slave &                           MEAN Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-               Laurentia-Slave &                     MEAN Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-      Laurentia-Superior(West) &                                        MOLSON B+C2 &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
-               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ \\
-               Laurentia-Slave &                                 Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ \\
-            Laurentia-Superior &                          MEAN Haig/Flaherty/Sutton &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-               Laurentia-Slave &  MEAN Pearson A/Peninsular sill/Kilohigok basin... &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
- Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &      B &     258.1 &      54.7 & 279.4 &  62.4 &       7.9 &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-                 Laurentia-Rae &                                      Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-                 Laurentia-Rae &                                   Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-                     Laurentia &                                      Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ \\
- Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
- Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-                     Laurentia &                                      Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-           Laurentia-Greenland &                        Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-                     Laurentia &                            Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &                 St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ \\
-                     Laurentia &                      Michikamau Intrusion Combined &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-                     Laurentia &                                  Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ \\
-                     Laurentia &                                 Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ \\
-                     Laurentia &                    Tobacco Root Dykes - A combined &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ \\
-                     Laurentia &                                       Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ \\
-                     Laurentia &                     MEAN Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ \\
-                     Laurentia &                                   Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ \\
-                     Laurentia &                                 McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ \\
-                     Laurentia &         Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ \\
-           Laurentia-Greenland &                                Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-           Laurentia-Greenland &                              Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-           Laurentia-Greenland &                      Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                                   Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ \\
-           Laurentia-Greenland &                                  North Qoroq Intr. &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-           Laurentia-Greenland &                                  Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                         Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-           Laurentia-Greenland &                         West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ \\
-           Laurentia-Greenland &                      West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ \\
-                     Laurentia &                             Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-            Laurentia-Scotland &                                   MEAN Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ \\
-           Laurentia-Greenland &                                     Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-           Laurentia-Greenland &                                 Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-           Laurentia-Greenland &                                  South Qoroq Intr. &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-           Laurentia-Greenland &                                 Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-           Laurentia-Greenland &                          NE-SW Trending Dyke Swarm &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-                     Laurentia &                                      Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                       MEAN Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1111\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-                     Laurentia &                           Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &                          Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
-                     Laurentia &                           Upper Osler volcanics -R &      A &     272.4 &      48.7 & 201.6 &  42.5 &       3.7 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                 Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-                     Laurentia &              Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1098\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &                               North Shore lavas -N &      A &     268.7 &      46.3 & 181.3 &  34.5 &       2.8 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &                              Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                             Portage Lake Volcanics &      A &     271.2 &      47.0 & 178.0 &  26.7 &       4.7 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &                    Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-                     Laurentia &                           Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ \\
-                     Laurentia &                        Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ \\
-                     Laurentia &                                   Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1087\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                      Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ \\
-                     Laurentia &                                    Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ \\
-                     Laurentia &                                     Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &   1050\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ \\
-                     Laurentia &                              Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ \\
-            Laurentia-Scotland &                                MEAN Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ \\
-            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ \\
-            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ \\
-                     Laurentia &                     MEAN Wyoming "Gunbarrel" dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                           Tsezotene Sills Combined &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                               Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ \\
-            Laurentia-Svalbard &                          Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ \\
-                     Laurentia &                          Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ \\
-                     Laurentia &                                   Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ \\
-                     Laurentia &                         Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-                     Laurentia &                                   Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ \\
-                     Laurentia &                        Sept-Iles Layered Intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ \\
+             Laurentia-Wyoming &                          Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Selkin2008a\} \\
+      Laurentia-Superior(East) &                     Otto Stock dykes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Pullaiah1975b\} \\
+               Laurentia-Slave &                                     Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
+      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dykes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(East) &                               Matachewan dykes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(East) &                               Matachewan dykes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &                                     Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2012a\} \\
+      Laurentia-Superior(East) &                                 Senneterre dykes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
+      Laurentia-Superior(East) &                               Nipissing N1 sills &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
+               Laurentia-Slave &                                     Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
+      Laurentia-Superior(East) &                                Biscotasing dykes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Kilian2015a\} \\
+               Laurentia-Slave &                                      Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
+      Laurentia-Superior(West) &                                 Marathon dykes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
+      Laurentia-Superior(West) &                                 Marathon dykes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
+      Laurentia-Superior(West) &                               Cauchon Lake dykes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(West) &                               Fort Frances dykes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(East) &                                 Lac Esprit dykes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Greenland-Nain &                                  Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                 \textbackslash cite\{Fahrig1976b\} \\
+               Laurentia-Slave &                                Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Buchan2009a\} \\
+      Laurentia-Superior(East) &                                      Minto dykes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &                                  Rifle Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans1981a\} \\
+                 Laurentia-Rae &                           Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Halls1999a\} \\
+             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Kilian2016a\} \\
+               Laurentia-Slave &                                 Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
+               Laurentia-Slave &                         Mean Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
+               Laurentia-Slave &                   Mean Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
+      Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                \textbackslash citep\{Irving1979a\} \\
+               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                \textbackslash citep\{Irving1979a\} \\
+            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+\\
+               Laurentia-Slave &             Pearson A/Peninsular/Kilohigok sills &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
+ Laurentia-Trans-Hudson orogen &                              Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                 \textbackslash cite\{Symons1999a\} \\
+                 Laurentia-Rae &                                    Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{McGlynn1974a\} \\
+                 Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                  \textbackslash cite\{Evans1973a\} \\
+                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                    \textbackslash cite\{Park1973\} \\
+ Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons2000a\} \\
+ Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                   \textbackslash cite\{Gala1995a\} \\
+                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Irving2004\} \\
+           Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Halls2011a\} \\
+                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Irving1972\} \\
+                     Laurentia &               St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &                                  \textbackslash cite\{Meert2002b\} \\
+                     Laurentia &                             Michikamau Intrusion &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Emslie1976a\} \\
+                     Laurentia &                                Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &                                 \textbackslash cite\{Harlan2008a\} \\
+                     Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &                        Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+\\
+                     Laurentia &                                 Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                 \textbackslash cite\{Fahrig1976a\} \\
+                     Laurentia &                               McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+           Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
+           Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
+           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
+                     Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash citep\{Murthy1978a\} \\
+           Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
+           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
+                     Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
+           Laurentia-Greenland &                       West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
+           Laurentia-Greenland &                    West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
+                     Laurentia &                           Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Palmer1977a\} \\
+            Laurentia-Scotland &                                      Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                        Nordic workshop calculation \\
+           Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                                South Qoroq Intr.
+&      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
+           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
+                     Laurentia &                        Ontario lamprophyre dykes &      A &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\
+                     Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Ernst1993a\} \\
+                     Laurentia &                          Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+\\
+                     Laurentia &           Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1109\$\textasciicircum \{+2\}\$\$\_\{-3\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                         Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
+                     Laurentia &                        Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
+                     Laurentia &                         Upper Osler volcanics -R &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &     1105\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  \textbackslash cite\{Halls1974a, Swanson-Hysell2014b, Swanson...
+\\
+                     Laurentia &               Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+3\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &   Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                             North Shore lavas -N &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &              \textbackslash cite\{Tauxe2009a,Swanson-Hysell2019a\} \\
+                     Laurentia &                           Portage Lake Volcanics &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &  \textbackslash cite\{Books1972a, Hnat2006a\} as calculated in ...
+\\
+                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash citep\{Kean1997a\} \\
+                     Laurentia &            Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094\$\textasciicircum \{+6\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                  Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                   \textbackslash cite\{Weil2003a\} \\
+                     Laurentia &                         Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
+                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                               \textbackslash cite\{Donadini2011a\} \\
+                     Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
+                     Laurentia &                    Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
+                     Laurentia &                                   Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &    1080\$\textasciicircum \{+4\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
+                     Laurentia &                                  Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1070\$\textasciicircum \{+14\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
+                     Laurentia &                            Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash cite\{Warnock2000a\} \\
+            Laurentia-Scotland &                                   Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                        Nordic workshop calculation \\
+            Laurentia-Svalbard &                    Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+            Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+\\
+                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
+                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006a\} \\
+            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                               \textbackslash cite\{Denyszyn2009a\} \\
+                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\
+                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons1991a\} \\
+                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Meert1994a\} \\
+                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Tancyk1987a\} \\
 \end{longtable}

### 2019-10-21 Manuscript/pole_table_edited.tex
clean up pole table

\begin{longtable}{p{1 in}p{1 in}rrrrrrrp{1.5 in}}
\toprule
                       terrane &                                        unit name & rating &  site lon &  site lat &  plon &  plat &  A$_{95}$ &                    age &                                     pole reference \\ \hline
\midrule
\endhead
\midrule
\multicolumn{10}{r}{{Continued on next page}} \\ \hline
\midrule
\endfoot

\bottomrule
\endlastfoot
             Laurentia-Wyoming &                          Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705$^{+4}_{-4}$ &                                 \cite{Selkin2008a} \\ \hline
      Laurentia-Superior(East) &                     Otto Stock dykes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676$^{+5}_{-5}$ &                               \cite{Pullaiah1975b} \\ \hline
               Laurentia-Slave &                                     Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625$^{+5}_{-5}$ &                               \cite{Mitchell2014a} \\ \hline
      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dykes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
      Laurentia-Superior(East) &                               Matachewan dykes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466$^{+23}_{-23}$ &                                  \cite{Evans2010a} \\ \hline
      Laurentia-Superior(East) &                               Matachewan dykes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
               Laurentia-Slave &                                     Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231$^{+2}_{-2}$ &                                 \cite{Buchan2012a} \\ \hline
      Laurentia-Superior(East) &                                 Senneterre dykes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218$^{+6}_{-6}$ &                                 \cite{Buchan1993a} \\ \hline
      Laurentia-Superior(East) &                               Nipissing N1 sills &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217$^{+4}_{-4}$ &                                 \cite{Buchan2000a} \\ \hline
               Laurentia-Slave &                                     Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193$^{+2}_{-2}$ &                               \cite{Mitchell2014a} \\ \hline
      Laurentia-Superior(East) &                                Biscotasing dykes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160$^{+11}_{-8}$ &                                 \cite{Kilian2015a} \\ \hline
               Laurentia-Slave &                                      Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126$^{+3}_{-18}$ &                                 \cite{Buchan2016a} \\ \hline
      Laurentia-Superior(West) &                                 Marathon dykes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
      Laurentia-Superior(West) &                                 Marathon dykes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
      Laurentia-Superior(West) &                               Cauchon Lake dykes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
      Laurentia-Superior(West) &                               Fort Frances dykes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077$^{+5}_{-5}$ &                                  \cite{Evans2010a} \\ \hline
      Laurentia-Superior(East) &                                 Lac Esprit dykes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069$^{+1}_{-1}$ &                                  \cite{Evans2010a} \\ \hline
      Laurentia-Greenland-Nain &                                  Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042$^{+12}_{-12}$ &                                 \cite{Fahrig1976b} \\ \hline
               Laurentia-Slave &                                Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026$^{+5}_{-5}$ &                                 \cite{Buchan2009a} \\ \hline
      Laurentia-Superior(East) &                                      Minto dykes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
               Laurentia-Slave &                                  Rifle Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963$^{+6}_{-6}$ &                                  \cite{Evans1981a} \\ \hline
                 Laurentia-Rae &                           Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917$^{+7}_{-7}$ &                                  \cite{Halls1999a} \\ \hline
             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899$^{+5}_{-5}$ &                                 \cite{Kilian2016a} \\ \hline
               Laurentia-Slave &                                 Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887$^{+5}_{-9}$ &                                 \cite{Buchan2016a} \\ \hline
               Laurentia-Slave &                         Mean Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885$^{+5}_{-5}$ &                               \cite{Mitchell2010c} \\ \hline
               Laurentia-Slave &                   Mean Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882$^{+4}_{-4}$ &                               \cite{Mitchell2010c} \\ \hline
      Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879$^{+6}_{-6}$ &                                  \cite{Evans2010a} \\ \hline
               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876$^{+10}_{-10}$ &                                \cite{Irving1979a} \\ \hline
               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876$^{+10}_{-10}$ &                                \cite{Irving1979a} \\ \hline
            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870$^{+1}_{-1}$ &  Nordic workshop calculation based on data of \cite{Schmidt1980a, Schwarz1982a} \\ \hline
               Laurentia-Slave &             Pearson A/Peninsular/Kilohigok sills &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870$^{+4}_{-4}$ &                               \cite{Mitchell2010c} \\ \hline
 Laurentia-Trans-Hudson orogen &                              Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838$^{+1}_{-1}$ &                                 \cite{Symons1999a} \\ \hline
                 Laurentia-Rae &                                    Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827$^{+4}_{-4}$ &                                \cite{McGlynn1974a} \\ \hline
                 Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818$^{+4}_{-4}$ &                                  \cite{Evans1973a} \\ \hline
                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785$^{+35}_{-35}$ &                                    \cite{Park1973} \\ \hline
 Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766$^{+5}_{-5}$ &                                 \cite{Symons2000a} \\ \hline
 Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758$^{+1}_{-1}$ &                                   \cite{Gala1995a} \\ \hline
                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741$^{+5}_{-5}$ &                                  \cite{Irving2004} \\ \hline
           Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633$^{+5}_{-5}$ &                                  \cite{Halls2011a} \\ \hline
                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590$^{+3}_{-3}$ &                                  \cite{Irving1972} \\ \hline
                     Laurentia &               St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476$^{+16}_{-16}$ &                                  \cite{Meert2002b} \\ \hline
                     Laurentia &                             Michikamau Intrusion &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460$^{+5}_{-5}$ &                                 \cite{Emslie1976a} \\ \hline
                     Laurentia &                                Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458$^{+13}_{-13}$ &                                 \cite{Elston2002a} \\ \hline
                     Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450$^{+14}_{-14}$ &                                 \cite{Elston2002a} \\ \hline
                     Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448$^{+49}_{-49}$ &                                 \cite{Harlan2008a} \\ \hline
                     Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443$^{+7}_{-7}$ &                                 \cite{Elston2002a} \\ \hline
                     Laurentia &                        Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430$^{+15}_{-15}$ &  Nordic workshop calculation based on data of \cite{Harlan1994a,Harlan1998a} \\ \hline
                     Laurentia &                                 Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425$^{+25}_{-25}$ &                                 \cite{Fahrig1976a} \\ \hline
                     Laurentia &                               McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401$^{+6}_{-6}$ &                                 \cite{Elston2002a} \\ \hline
                     Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385$^{+23}_{-23}$ &                                 \cite{Elston2002a} \\ \hline
           Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382$^{+2}_{-2}$ &                              \cite{Marcussen1983a} \\ \hline
           Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382$^{+2}_{-2}$ &                              \cite{Marcussen1983a} \\ \hline
           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382$^{+2}_{-2}$ &                             \cite{Abrahamsen1987a} \\ \hline
                     Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305$^{+15}_{-15}$ &                                \cite{Murthy1978a} \\ \hline
           Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275$^{+1}_{-1}$ &                                  \cite{Piper1992a} \\ \hline
           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275$^{+2}_{-2}$ &                                  \cite{Piper1977b} \\ \hline
                     Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267$^{+2}_{-2}$ &                                 \cite{Buchan2000a} \\ \hline
           Laurentia-Greenland &                       West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244$^{+8}_{-8}$ &                                  \cite{Piper1977b} \\ \hline
           Laurentia-Greenland &                    West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238$^{+11}_{-11}$ &                                  \cite{Piper1977b} \\ \hline
                     Laurentia &                           Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237$^{+5}_{-5}$ &                                 \cite{Palmer1977a} \\ \hline
            Laurentia-Scotland &                                      Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199$^{+70}_{-70}$ &                        Nordic workshop calculation \\ \hline
           Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
           Laurentia-Greenland &                                South Qoroq Intr. &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1992a} \\ \hline
           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1977a} \\ \hline
           Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160$^{+5}_{-5}$ &                                  \cite{Piper1992a} \\ \hline
                     Laurentia &                        Ontario lamprophyre dykes &      A &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143$^{+10}_{-10}$ &                                 \cite{Piispa2018a} \\ \hline
                     Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141$^{+2}_{-2}$ &                                  \cite{Ernst1993a} \\ \hline
                     Laurentia &                          Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109$^{+2}_{-2}$ &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
                     Laurentia &           Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1109$^{+2}_{-3}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
                     Laurentia &                         Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108$^{+3}_{-3}$ &                         \cite{Swanson-Hysell2014b} \\ \hline
                     Laurentia &                        Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107$^{+4}_{-4}$ &                         \cite{Swanson-Hysell2014b} \\ \hline
                     Laurentia &                         Upper Osler volcanics -R &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &     1105$^{+1}_{-1}$ &  \cite{Halls1974a, Swanson-Hysell2014b, Swanson-Hysell2019a} \\ \hline
                     Laurentia &               Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105$^{+3}_{-4}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
                     Laurentia &   Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101$^{+1}_{-1}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
                     Laurentia &                             North Shore lavas -N &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &     1097$^{+3}_{-3}$ &              \cite{Tauxe2009a,Swanson-Hysell2019a} \\ \hline
                     Laurentia &                           Portage Lake Volcanics &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &     1095$^{+3}_{-3}$ &  \cite{Books1972a, Hnat2006a} as calculated in \cite{Swanson-Hysell2019a} \\ \hline
                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095$^{+2}_{-2}$ &                                  \cite{Kean1997a} \\ \hline
                     Laurentia &            Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094$^{+6}_{-4}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
                     Laurentia &                  Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091$^{+5}_{-5}$ &                                   \cite{Weil2003a} \\ \hline
                     Laurentia &                         Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090$^{+2}_{-7}$ &                              \cite{Fairchild2017a} \\ \hline
                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088$^{+11}_{-11}$ &                               \cite{Donadini2011a} \\ \hline
                     Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086$^{+1}_{-1}$ &                                \cite{Kulakov2013a} \\ \hline
                     Laurentia &                    Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084$^{+1}_{-1}$ &                              \cite{Fairchild2017a} \\ \hline
                     Laurentia &                                   Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &    1080$^{+4}_{-10}$ &                                  \cite{Henry1977a} \\ \hline
                     Laurentia &                                  Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1070$^{+14}_{-10}$ &                                  \cite{Henry1977a} \\ \hline
                     Laurentia &                            Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015$^{+15}_{-15}$ &                                \cite{Warnock2000a} \\ \hline
            Laurentia-Scotland &                                   Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925$^{+145}_{-145}$ &                        Nordic workshop calculation \\ \hline
            Laurentia-Svalbard &                    Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831$^{+20}_{-20}$ &                                 \cite{Maloof2006a} \\ \hline
            Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800$^{+11}_{-11}$ &                                 \cite{Maloof2006a} \\ \hline
                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778$^{+2}_{-2}$ &  Nordic workshop calculation based on data of \cite{Harlan1993a, Harlan1997a} \\ \hline
                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778$^{+2}_{-2}$ &                                   \cite{Park1989a} \\ \hline
                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775$^{+25}_{-25}$ &                                   \cite{Weil2006a} \\ \hline
            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760$^{+30}_{-30}$ &                                 \cite{Maloof2006a} \\ \hline
                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline
                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575$^{+5}_{-5}$ &                                 \cite{Symons1991a} \\ \hline
                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572$^{+5}_{-5}$ &                                  \cite{Meert1994a} \\ \hline
                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565$^{+4}_{-4}$ &                                 \cite{Tanczyk1987a} \\ \hline
\end{longtable}

### 2019-10-21 Manuscript/manuscript.tex
generate running mean for Laurentia, update .rot for Laurentia (a bit) and do a full run through of notesbooks saving output

@@
 
-In this work, I develop a continuous paleogeographic reconstruciton of Laurentia constrained the paleomagnetic poles compilation.
+In this work, I develop a continuous paleogeographic reconstruction of Laurentia constrained the paleomagnetic poles compilation.
 This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (900-800 Ma for example).
@@
 GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}.
-Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive python environment \citep{Perez2007a}.
-This work was supported by NSF CAREER Grant 1847277 awarded to N.L.S.-H.
+Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
+This work was supported by NSF CAREER Grant EAR-1847277 awarded to N.L.S.-H.
 
@@
 \begin{table}[hbt]
+\caption{Rotations of separated terranes}
+{\scriptsize
 \begin{tabular}{|l|l|l|l|p{2 in}|}
@@
 \end{tabular}
-\caption{}
+}
 \label{tab:rotations}
@@
 \begin{landscape}
-\input{pole_table_editted.tex}
+\input{pole_table_edited.tex}
 \end{landscape}

### 2019-10-21 Manuscript/pole_table.tex
generate running mean for Laurentia, update .rot for Laurentia (a bit) and do a full run through of notesbooks saving output

@@
       Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                \textbackslash citep\{Irving1979a\} \\
-               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                \textbackslash citep\{Irving1979a\} \\
+               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
+               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
             Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
@@
                      Laurentia &                             Michikamau Intrusion &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Emslie1976a\} \\
-                     Laurentia &                                Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                  \textbackslash cite\{Elston2002\} \\
-                     Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &                                Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
+                     Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
                      Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &                                 \textbackslash cite\{Harlan2008a\} \\
-                     Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
                      Laurentia &                        Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
@@
                      Laurentia &                                 Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                 \textbackslash cite\{Fahrig1976a\} \\
-                     Laurentia &                               McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Elston2002\} \\
-                     Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Elston2002\} \\
+                     Laurentia &                               McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
+                     Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
            Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
@@
            Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
-                     Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash citep\{Murthy1978a\} \\
+                     Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                 \textbackslash cite\{Murthy1978a\} \\
            Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
@@
 \\
-                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash citep\{Kean1997a\} \\
+                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Kean1997a\} \\
                      Laurentia &            Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094\$\textasciicircum \{+6\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
@@
                      Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                               \textbackslash cite\{Donadini2011a\} \\
-                     Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
+                     Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                \textbackslash cite\{Kulakov2013a\} \\
                      Laurentia &                    Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
@@
                      Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Meert1994a\} \\
-                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Tancyk1987a\} \\
+                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{Tanczyk1987a\} \\
 \end{longtable}

### 2019-10-21 Manuscript/manuscript.tex
small fixes to references

@@
  An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}.
-The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
+The extent to which the implied position of Laurentia in \cite{Merdith2017b}} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
 As noted above, the development of such models is challenging and the researchers need to balance varying constraints.

### 2019-10-22 Manuscript/pole_table.tex
add f factor calculations and strat adding updated Chuar poles

@@
                  Laurentia-Rae &                           Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Halls1999a\} \\
-             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Kilian2016a\} \\
+             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Kilian2016b\} \\
                Laurentia-Slave &                                 Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
@@
                  Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                  \textbackslash cite\{Evans1973a\} \\
-                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                    \textbackslash cite\{Park1973\} \\
+                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                   \textbackslash cite\{Park1973a\} \\
  Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons2000a\} \\
  Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                   \textbackslash cite\{Gala1995a\} \\
-                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Irving2004\} \\
+                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Irving2004a\} \\
            Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Halls2011a\} \\
-                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Irving1972\} \\
+                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                 \textbackslash cite\{Irving1972a\} \\
                      Laurentia &               St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &                                  \textbackslash cite\{Meert2002b\} \\
@@
            Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
+           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
            Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
-           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
                      Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                 \textbackslash cite\{Murthy1978a\} \\
+           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
            Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
                      Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
@@
             Laurentia-Scotland &                                      Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                        Nordic workshop calculation \\
+           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
-           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                                South Qoroq Intr.
 &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
@@
                      Laurentia &                         Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
-                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                               \textbackslash cite\{Donadini2011a\} \\
+                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                               \textbackslash cite\{Donadini2011b\} \\
                      Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                \textbackslash cite\{Kulakov2013a\} \\
@@
             Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
                      Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
 \\
-                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
-                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006a\} \\
+                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006b\} \\
             Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
@@
                      Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\
+                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
                      Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
                      Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons1991a\} \\

### 2019-10-22 Manuscript/manuscript.tex
edits up through pre-Laurentia amalgamation section (some some further along as well)

--- 2019-10-21 Manuscript/manuscript.tex
@@
 \documentclass[11pt,letterpaper]{article}
-
+\setcounter{section}{2}
 \usepackage{threeparttable}
@@
 \begin{flushleft}
-{\Large \textbf{The Precambrian paleogeography of Laurentia}}
-\\
+\section{The Precambrian paleogeography of Laurentia}
 
@@
 
-\section*{Introduction}
+\subsection{Introduction}
 
@@
 
-\section*{Broad tectonic history overview}
+\subsection{Broad tectonic history overview}
 
@@
 Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
-Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
+Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
 Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca.
@@
 This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
-Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca.
+Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca.
 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca.
-1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca.
-1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}.
+1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca.
+1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.
 As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
@@
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
-The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
+The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
 \label{fig:Laurentia_map}
@@
 
-Crustal growth also progressed at this time in the Paleoproterozoic through accretionary orogenesis.
+Crustal growth also progressed in the Paleoproterozoic through accretionary orogenesis.
 This accretion occurred within the Wopmay orogen through ca.
@@
 Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
-1.86 to 1.82 \cite{Schulz2007a}.
+1.86 to 1.82 \citep{Schulz2007a}.
 Firm evidence of the end of the orogeny comes from the ca.
@@
 In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
-The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+Nain provinces (often referred to as the Churchill domain or plate; e.g.
+The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
 \citealp{Skipton2016a, Weller2017a}).
@@
 1.71 to 1.68 Ga Yavapai orogeny.
-Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Whitmeyer2007a}.
+Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}.
 Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 Subsequent accretionary orogenesis of the ca.
-1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Whitmeyer2007a}.
+1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
@@
 1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
-While an active margin interpretation with magmatism in back‐arc setting has gained traction within the literature with evidence, the tectonic setting is considered enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+While an active margin interpretation with magmatism in back‐arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 A timeline of large igneous provinces  (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown.
-The interpreted age of paleomagnetic poles for Laurentia (not included separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.}}
+The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.}}
 \label{fig:tectonic_history}
@@
 
-There was significantly less crustal growth on the western margin of Laurentia (Fig.
+There is significantly less preserved crustal growth on the western margin of Laurentia (Fig.
 \ref{fig:Laurentia_map}) and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
-The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca.
+The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
 1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated.
-\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep[e.g.]{Jones2015a}.
+\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}.
 This region is interpreted to have been subsequently deformed during a ca.
@@
 1.06 Ga) requires a long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
-This constraint is incorporated into models such as that of \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
+This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
 Amazonia in \citealp{Elming2009a}).
@@
 1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
-That the Laurentia margin experienced large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny that is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
+Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea.
+This extensive and major collision orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
+The term Grenville orogeny or Grenville belt is used rather loosely throuhgout much of the literature to refer to any late Mesoproterozoic orogenic belt.
+The timeline of orogenesis on the Laurentia margin has more nuanced constraints that this which can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
+\ref{fig:tectonic_history}).
 
-The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting.
+The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig.
+\ref{fig:tectonic_history}).
 Along the western margin of Laurentia, small-scale rifting occurred ca.
@@
 In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
-The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin.
+The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
+\ref{fig:tectonic_history}).
 Available geochronological constraints and thermal subsidence modeling indicate ca.
@@
 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
-The Maz–Arequipa–Rio Apa (MARA) block MARA block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
+The Maz–Arequipa–Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
 Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
@@
 
-\section*{Paleomagnetic pole compilation}
+\subsection{Paleomagnetic pole compilation}
 
@@
 
-In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications.
+In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig.
+\ref{fig:Laurentia_poles} and Table 2).
 The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume).
@@
 These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}.
-Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeography reconstructions.
+Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions.
+However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age.
 For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
-However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than for eruptive units or shallow-level intrusions.
-As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating.
+However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions.
+As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca.
+1015 Ma Haliburton intrusions.
 However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).
 
+\subsection{Differential motion before Laurentia amalgamation}
+
 Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), paleomagnetic poles need to be considered with respect to the individual Archean provinces.
-There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation.
-Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
-A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or whether they have shared histories before experiencing fragmentation and reamalgamation.
-The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
-\ref{fig:Superior_Slave_recons};\citealp{Buchan2016a}.
-These data provide paleomagnetic support for the hypothesis that the Trans-Hudson orogeny is the result of closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces as the result of differential plate tectonic motion.
-Reconstructions made of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.
+For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
+This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that the Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny.
+\cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occurred in the time interval of 2.07 to 1.87 Ga.
+I follow this interpretation and group the poles into Superior (West) and Superior (East).
+Uncertainty remains with respect to whether the ca.
+1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time being should be considered solely in the western Superior province reference frame.
 
@@
 \includegraphics[width=\textwidth]{../Figures/Superior_Slave_reconstructions.pdf}
-\caption{\small{\textbf{Paleogeographic reconstructions using poles from the Superior, Slave and Rae provinces.
+\caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.
 Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
@@
 
- Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.
+There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation (Fig.
+\ref{fig:Laurentia_poles} and Table 2).
+Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
+A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation.
+The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+\ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}.
+These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion.
+They also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces.
+Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.
 
-For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}.
-This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that the Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny.
-\cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occurred in the time interval of 2.07 to 1.87 Ga.
-I follow this interpretation and group the poles into Superior (West) and Superior (East).
-Uncertainty remains with respect to whether the ca.
-1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference frame.
+\subsection{Paleogeography of an assembled Laurentia}
 
-%These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame. NEED TO FIGURE OUT HOW TO SHOW THIS
-
-Following Laurentia's amalgamation, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
+Following the amalgmation of the Archean provinces in Laurentia  ca.
+1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
@@
 The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used).
-The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation of \citealp{Roest1989a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland.
-The reconstruction of Svalbard is more challenging given a multi-state tectonic history.
+The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland.
+The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting.
 The preferred Euler of \cite{Maloof2006a} is used here that designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a}.
@@
 Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
-To visualize the temoporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+The poles are shown in Figure \ref{fig:Laurentia_poles} where they are colored by age.
+To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}.
 The ages of poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}.
 Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both.
-In addition, intraplate magmatism resulting from plume-related large-igneous provinces can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quisence (e.g.
+In addition, intraplate magmatism resulting from plume-related large-igneous provinces can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g.
 the ca.
@@
 780 Ma) and Franklin LIP (ca.
-720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins.
-Overall the record of paleomagnetic poles has a lot of internal consistency with progressive paths and agreement between poles such as for the Keweenawan Track \citep{Swanson-Hysell2019a} and the late Tonian path \citep{Eyster2019a}.
+720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
+Overall the record of paleomagnetic poles has a lot of internal consistency in intervals for which there is high-resolution coverage.
+These data result in the progressive paths agreement between poles such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian ( ca.
+path \citep{Eyster2019a}.
 
@@
 In particular, data from Greenland adds 12 poles between 1385 and 1160 Ma when there are only 4 poles from mainland Laurentia.
-Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:rotations}) these poles can be rotated into Laurentia coordinates and used for reconstruction of the entire continent.
+Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}) these poles can be rotated into Laurentia coordinates and used for reconstruction of the entire continent.
+There is good agreement between the ca.
+1633 Ma Melville Bugt diabase dykes pole from Greenland \citep{Halls2011a} and the ca.
+1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}.
+Similarly, there is the good agreement between the Mackenzie dykes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the North Qoroq intrusives \citep{Piper1992a} and the Kungnat Ring dyke \citep{Piper1977a}.
+Furthermore, the Greenland poles with ages between the ca.
+1237 Ma Sudbury dikes and ca.
+1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.
+
+The exception to this overall agreement between poles from Greenland and mainland Laurentia are at ca.
+1382 Ma.
+There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}.
+However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g.
+the McNamara Formation; \citealp{Elston2002a}.
+Additionally, the Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton.
+There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation.
+However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland.
+Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite.
+As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed.
+However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts.
+There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite) however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+The positions between the are in such conflict with one another that taking a mean between the two is not satisfactory.
+In general, the poles from the Belt Supergroup imply a lower latitude than
+
+ Poles from Greenland provide additional constraints particularly in the Mesoproterozoic and are included in the running mean.
+
+When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied by this model.
+Where there are no constraints a simple interpolation between constraints is made.
+While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain.
+These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca.
+810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
+
+In portions of the path where there are sparse constraints,
 
@@
 \caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW).
-The paleomagnetic poles are compiled in Table \ref{tab:Laurentia_poles}.
+The paleomagnetic poles are compiled in Table 2.
 Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
@@
 
-\section*{Paleogeographic reconstructions}
+\subsection{Paleogeographic reconstructions}
 
@@
  An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}.
-The extent to which the implied position of Laurentia in \cite{Merdith2017a} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
+The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
 As noted above, the development of such models is challenging and the researchers need to balance varying constraints.
@@
 
- The synthesis of \citet{Pesonen2012a} developed reconstructions at various Proterozoic time slices.
+In this work, I develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles.
+The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example).
+One could use interpretations of paleogeographic connections with other cratons (e.g.
+Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself.
+In portions of the record with a more dense record of poles such as ca.
+1450, a calculated running mean is used to integrate constraints from multiple poles.
+A similar approach to that often taken in the Phanerozoic (e.g.
+\citealp{Torsvik2012a} is utilized where a running mean with a 20 Myr window that is moved forward in 10 Myr steps is calculated.
+An approach that seeks to synthesize data in this way is particularly needed ca.
+1400 Ma where there are numerous constraints, but where not every individual pole can be satisfied without jerking Laurentia back and forth.
+In contrast, through the time interval of the Midcontinent Rift there is a continuous progression of paleomagnetic poles.
+As shown in \cite{Swanson-Hysell2019a}, a running mean approach can reduce motion along an apparent polar wander path by bring the mean in towards where there is a concentration of poles.
+For the portion of the reconstruction during the interval of time ca.
+1110 to 1070 Ma, I utilize an Euler pole inversion from \cite{Swanson-Hysell2019a}.
 
-In this work, I develop a continuous paleogeographic reconstruction of Laurentia constrained the paleomagnetic poles compilation.
-This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (900-800 Ma for example).
-One could use interpretations of paleogeographic connections with other cratons (e.g.
-Baltica in the Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself.
-The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
-Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:Laurentia_reconstructions.pdf}.
+ Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:Laurentia_reconstructions}.
 These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth.
@@
 
-\section*{Conclusion}
+\subsection{Discussion}
 
-The paleogeographic record is rich in constraints through the Precambrian -- certainly the
+The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles.
+As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of developing paleogeographic models through the Precambrian.
 
-To make an argument that differential plate tectonic motion began in the Neoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
+There is strong evidence both in the geological and paleomagnetic record of Laurentia for differential plate tectonic motion between 2.2 and 1.8 Ga.
+The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well.
+While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca.
+0.8 Ga \citep{Hamilton2011a, Stern2018a}.
+These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}.
+An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}.
+While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
+As an example, consider the record of Laurentia in the late Mesoproterozoic.
+As can be seen in Table 2, there is a fantastic quantity of high-quality paleomagnetic poles between 1110 and 1070 Ma.
+These poles
 
@@
 This work was supported by NSF CAREER Grant EAR-1847277 awarded to N.L.S.-H.
+The code, data, and reconstructions used in this paper are openly available in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
 
@@
 }
-\label{tab:rotations}
+\label{tab:terrane_rotations}
 \end{table}

### 2019-10-22 Manuscript/manuscript.tex
progress on paleogeography of Laurentia section

@@
 
-Following the amalgmation of the Archean provinces in Laurentia  ca.
+Following the amalgamation of the Archean provinces in Laurentia  ca.
 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
@@
 
-Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse.
-The poles are shown in Figure \ref{fig:Laurentia_poles} where they are colored by age.
+Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse (shown colored by age in Fig.
+\ref{fig:Laurentia_poles}).
 To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}.
@@
 1267 Ma Mackenzie LIP; Fig.
-\ref{fig:tectonic_history}.
-In particular, there are many high-quality paleomagnetic poles in the early Mesoproterozoic with 12 poles between ca.
-1480 and 1380 from Laurentia (including Greenland poles).
-The best-constrained portion of the record is associated with ca.
-1110 to 1083 Ma magmatism within the Midcontinent Rift and southwest Laurentia that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
+\ref{fig:tectonic_history}).
+Intracontinental rifts have led to the highest density of poles both in the case of the ca.
+1.4 Ga Belt Supergroup and the ca.
+1.1 Ga Midcontinent Rift (Fig.
+\ref{fig:tectonic_history}).
+The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formations that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
 The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca.
 780 Ma) and Franklin LIP (ca.
-720 Ma) as well as contemporaneous sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
-Overall the record of paleomagnetic poles has a lot of internal consistency in intervals for which there is high-resolution coverage.
-These data result in the progressive paths agreement between poles such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian ( ca.
-path \citep{Eyster2019a}.
+720 Ma), as well as similiarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
+Overall, the record of paleomagnetic poles has a lot of internal consistency in intervals for which there is high-resolution coverage.
+These data result in the progressive paths agreement between poles such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian (ca.
+775 to 720 Ma) path \citep{Eyster2019a}.
 
@@
 In particular, data from Greenland adds 12 poles between 1385 and 1160 Ma when there are only 4 poles from mainland Laurentia.
-Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}) these poles can be rotated into Laurentia coordinates and used for reconstruction of the entire continent.
-There is good agreement between the ca.
+Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent.
+The reliability of this approach gains credence through the good agreement between the ca.
 1633 Ma Melville Bugt diabase dykes pole from Greenland \citep{Halls2011a} and the ca.
 1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}.
-Similarly, there is the good agreement between the Mackenzie dykes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the North Qoroq intrusives \citep{Piper1992a} and the Kungnat Ring dyke \citep{Piper1977a}.
-Furthermore, the Greenland poles with ages between the ca.
+Similarly, there is good agreement between the ca.
+1267 Ma Mackenzie dykes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca.
+1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dyke \citep{Piper1977a}.
+Furthermore, the Greenland poles with ages that fall between the ca.
 1237 Ma Sudbury dikes and ca.
-1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.
+1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Fig.
+\ref{fig:Logan_Loop}).
 
-The exception to this overall agreement between poles from Greenland and mainland Laurentia are at ca.
+An exception to this overall agreement between poles from Greenland and mainland Laurentia occurs ca.
 1382 Ma.
@@
 However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g.
-the McNamara Formation; \citealp{Elston2002a}.
-Additionally, the Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton.
+the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}).
+Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton.
 There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation.
-However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland.
+However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton.
 Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite.
@@
 There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite) however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
-The positions between the are in such conflict with one another that taking a mean between the two is not satisfactory.
-In general, the poles from the Belt Supergroup imply a lower latitude than
+At present, it is unclear which poles are a better representation of Laurentia's position ca.
+1400 Ma.
 
- Poles from Greenland provide additional constraints particularly in the Mesoproterozoic and are included in the running mean.
-
-When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied by this model.
-Where there are no constraints a simple interpolation between constraints is made.
-While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain.
-These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca.
-810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
-
-In portions of the path where there are sparse constraints,
+Another challenging portion of the Laurentia record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs.
+\ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitudes}).
+As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma.
+One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}.
+Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Bono2019a}.
+Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig.
+\ref{fig:Laurentia_paleolatitude}).
 
@@
 
-The combined paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia.
-The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}.
-Such an approach is more challenging when there are significant gaps in the record as is the case through the early Neoproterozoic Era and the late Paleoproterozoic into early Mesoproterozoic.
+Synthesizing the compilation of paleomagnetic poles for Laurentia into a composite path over the past 1.8 billion years presents a challenge given the highly variable temporal coverage.
+The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting spherical splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}.
+Applying such an approach can reduce the effect of spurious poles in regions of high data density where seeking to satisfy every mean pole position would result in jerky motion.
 
-\subsection{Paleogeographic reconstructions}
+A synthesized pole path for Laurentia is developed here and used to develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles.
+The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example).
+One could use interpretations of paleogeographic connections with other cratons (e.g.
+Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself.
+In portions of the record with a more dense record of poles, such as ca.
+1450, a calculated running mean is used to integrate constraints from multiple poles.
+This method follows the approach taken in the Phanerozoic (e.g.
+\citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps.
+When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model.
+Where there are no constraints a simple interpolation between constraints is made.
+While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain.
+These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca.
+810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
 
-Seeking to developing comprehensive continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types.
+One downside of a running mean approach is that it pulls the mean to regions of high data density.
+As was shown in \cite{Swanson-Hysell2019a}, this behavior can reduce motion along an apparent polar wander path.
+As a result, for the portion of the reconstruction during the interval of time ca.
+1110 to 1070 Ma, I utilize an Euler pole inversion from \cite{Swanson-Hysell2019a}.
+
+ Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}.
+These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth.
+As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
+While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block.
+Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
+In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
+
+\subsection{Discussion}
+
+Seeking to developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types.
 Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles.
@@
 The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude}.
-
-In this work, I develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles.
-The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
-This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example).
-One could use interpretations of paleogeographic connections with other cratons (e.g.
-Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself.
-In portions of the record with a more dense record of poles such as ca.
-1450, a calculated running mean is used to integrate constraints from multiple poles.
-A similar approach to that often taken in the Phanerozoic (e.g.
-\citealp{Torsvik2012a} is utilized where a running mean with a 20 Myr window that is moved forward in 10 Myr steps is calculated.
-An approach that seeks to synthesize data in this way is particularly needed ca.
-1400 Ma where there are numerous constraints, but where not every individual pole can be satisfied without jerking Laurentia back and forth.
-In contrast, through the time interval of the Midcontinent Rift there is a continuous progression of paleomagnetic poles.
-As shown in \cite{Swanson-Hysell2019a}, a running mean approach can reduce motion along an apparent polar wander path by bring the mean in towards where there is a concentration of poles.
-For the portion of the reconstruction during the interval of time ca.
-1110 to 1070 Ma, I utilize an Euler pole inversion from \cite{Swanson-Hysell2019a}.
-
- Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:Laurentia_reconstructions}.
-These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's growth.
-As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
-While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block.
-Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
-In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
-
-\subsection{Discussion}
 

### 2019-10-22 Manuscript/manuscript.tex
regenerate figures and make new reconstruction figure prior to submission

@@
 However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).
+In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2).
+Additional not-rated poles included in the present compilation is the new pole for the ca.
+1144 Ma Ontario lamprophyre dykes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca.
+1140 Ma Abitibi dikes \citep{Ernst1993a}.
+This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
+Poles from the Chuar Group as presented in \cite{Eyster2019a} are also included.
 
@@
 
-\subsection{Discussion}
+\subsection{Comparing paleogeographic models to the paleomagnetic compilation}
 
@@
 These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
-Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.}}}
+Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.
+Paleomagnetic poles within 25 million years of each reconstruction time are plotted.}}}
 \label{fig:Laurentia_reconstructions}
 \end{figure}
-%ADD MORE RECONSTRUCTIONS. 1 per 100 million years perhaps with indication of which ones are well-constrained by models and
 
@@
 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models.
-Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole despite its `A' Nordic rating.
+Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating.
 The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.
@@
 
-The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles.
-As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of developing paleogeographic models through the Precambrian.
+\subsection{Conclusion}
 
@@
 As can be seen in Table 2, there is a fantastic quantity of high-quality paleomagnetic poles between 1110 and 1070 Ma.
-These poles
+These poles constrain rapid motion of Laurentia along
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
+\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}.
+The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
+\label{fig:Laurentia_reconstructions}
+\end{figure}
+
+The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles.
+As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of developing paleogeographic models through the Precambrian and will continue to be moving forward.
 

### 2019-10-22 Manuscript/pole_table.tex
regenerate figures and make new reconstruction figure prior to submission

@@
                Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
-            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
 \\
@@
                      Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
+           Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
            Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
            Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
-           Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
                      Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                 \textbackslash cite\{Murthy1978a\} \\
@@
            Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
-           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                                South Qoroq Intr.
 &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
+           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-                     Laurentia &                        Ontario lamprophyre dykes &      A &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\
+                     Laurentia &                        Ontario lamprophyre dykes &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\
                      Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Ernst1993a\} \\
@@
                      Laurentia &                             North Shore lavas -N &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &              \textbackslash cite\{Tauxe2009a,Swanson-Hysell2019a\} \\
+                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Kean1997a\} \\
                      Laurentia &                           Portage Lake Volcanics &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &  \textbackslash cite\{Books1972a, Hnat2006a\} as calculated in ...
 \\
-                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Kean1997a\} \\
                      Laurentia &            Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094\$\textasciicircum \{+6\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
@@
                      Laurentia &                            Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash cite\{Warnock2000a\} \\
+                     Laurentia &                      Adirondack fayalite granite &     NR &     285.5 &      44.0 & 132.7 & -28.4 &       6.9 &    990\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\
+                     Laurentia &             Adirondack metamorphic\textbackslash nanorthosites &     NR &     286.0 &      44.0 & 149.0 & -25.1 &      11.6 &    970\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\
+                     Laurentia &                     Adirondack Microcline gneiss &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &    960\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\
             Laurentia-Scotland &                                   Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                        Nordic workshop calculation \\
@@
             Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
-                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
-                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Calculation from \textbackslash cite\{Eyster2019a\} based on d...
 \\
-                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006b\} \\
-            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
-                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                               \textbackslash cite\{Denyszyn2009a\} \\
+            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    770\$\textasciicircum \{+19\}\$\$\_\{-40\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &     760\$\textasciicircum \{+6\}\$\$\_\{-10\}\$ &                                   \textbackslash cite\{Weil2006b\} \\
+                     Laurentia &                                    Carbon Canyon &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &      757\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &  \textbackslash cite\{Weil2004a\} as calculated in \textbackslash cite\{Eyster...
+\\
+                     Laurentia &                             Carbon Butte/Awatubi &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &      751\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Eyster2019a\} \\
+                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      718\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Denyszyn2009b\} \\
                      Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\

### 2019-10-22 Manuscript/pole_table_edited.tex
fix a few table references
small fixes to references
update table with Chuar poles

@@
 \toprule
-                       terrane &                                        unit name & rating &  site lon &  site lat &  plon &  plat &  A$_{95}$ &                    age &                                     pole reference \\ \hline
+                       terrane &                                        unit name & rating &  site lon (\textdegree) &  site lat (\textdegree) &  plon (\textdegree) &  plat (\textdegree) &  A$_{95}$ &                    age (Ma)&                                     pole reference \\ \hline
 \midrule
@@
                  Laurentia-Rae &                           Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917$^{+7}_{-7}$ &                                  \cite{Halls1999a} \\ \hline
-             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899$^{+5}_{-5}$ &                                 \cite{Kilian2016a} \\ \hline
+             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899$^{+5}_{-5}$ &                                 \cite{Kilian2016b} \\ \hline
                Laurentia-Slave &                                 Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887$^{+5}_{-9}$ &                                 \cite{Buchan2016a} \\ \hline
@@
                  Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818$^{+4}_{-4}$ &                                  \cite{Evans1973a} \\ \hline
-                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785$^{+35}_{-35}$ &                                    \cite{Park1973} \\ \hline
+                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785$^{+35}_{-35}$ &                                    \cite{Park1973a} \\ \hline
  Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766$^{+5}_{-5}$ &                                 \cite{Symons2000a} \\ \hline
  Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758$^{+1}_{-1}$ &                                   \cite{Gala1995a} \\ \hline
-                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741$^{+5}_{-5}$ &                                  \cite{Irving2004} \\ \hline
+                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741$^{+5}_{-5}$ &                                  \cite{Irving2004a} \\ \hline
            Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633$^{+5}_{-5}$ &                                  \cite{Halls2011a} \\ \hline
-                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590$^{+3}_{-3}$ &                                  \cite{Irving1972} \\ \hline
+                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590$^{+3}_{-3}$ &                                 \cite{Irving1972a} \\ \hline
                      Laurentia &               St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476$^{+16}_{-16}$ &                                  \cite{Meert2002b} \\ \hline
@@
            Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160$^{+5}_{-5}$ &                                  \cite{Piper1992a} \\ \hline
-                     Laurentia &                        Ontario lamprophyre dykes &      A &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143$^{+10}_{-10}$ &                                 \cite{Piispa2018a} \\ \hline
+                     Laurentia &                        Ontario lamprophyre dykes &      NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143$^{+10}_{-10}$ &                                 \cite{Piispa2018a} \\ \hline
                      Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141$^{+2}_{-2}$ &                                  \cite{Ernst1993a} \\ \hline
@@
                      Laurentia &                         Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090$^{+2}_{-7}$ &                              \cite{Fairchild2017a} \\ \hline
-                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088$^{+11}_{-11}$ &                               \cite{Donadini2011a} \\ \hline
+                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088$^{+11}_{-11}$ &                               \cite{Donadini2011b} \\ \hline
                      Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086$^{+1}_{-1}$ &                                \cite{Kulakov2013a} \\ \hline
@@
             Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800$^{+11}_{-11}$ &                                 \cite{Maloof2006a} \\ \hline
-                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778$^{+2}_{-2}$ &  Nordic workshop calculation based on data of \cite{Harlan1993a, Harlan1997a} \\ \hline
-                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778$^{+2}_{-2}$ &                                   \cite{Park1989a} \\ \hline
-                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775$^{+25}_{-25}$ &                                   \cite{Weil2006a} \\ \hline
-            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760$^{+30}_{-30}$ &                                 \cite{Maloof2006a} \\ \hline
-                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
+            Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778$^{+2}_{-2}$ &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a}  \\ \hline
+   Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    770$^{+19}_{-40}$ &                                 \cite{Maloof2006a} \\ \hline
+            Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &     760$^{+6}_{-10}$ &                                   \cite{Weil2006b} \\ \hline
+            Laurentia &                                    Carbon Canyon &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &      757$^{+7}_{-7}$ &  \cite{Weil2004a} as calculated in \cite{Eyster2019a} \\ \hline
+            Laurentia &                             Carbon Butte/Awatubi &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &      751$^{+8}_{-8}$ &                                 \cite{Eyster2019a} \\                   Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
                      Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline

### 2019-10-22 Manuscript/manuscript.tex
submission version of manuscript

@@
 As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
-While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block.
+While they provide constraints in this regard, they do not provide constraints in and of themselves of the longitudinal position of the block.
 Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
 In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
+The continuous paleolatitude implied by this model is shown as the simple Laurentia pole interpolation mode in Figure \ref{fig:Laurentia_paleolatitude}.
 
@@
 
-Seeking to developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types.
+Developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types.
 Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles.
 Paleomagnetic poles compilations themselves are evolving with better data and improved geochronology.
-Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress.
+Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress in Precambrian paleogeography.
 
@@
 Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.
-Paleomagnetic poles within 25 million years of each reconstruction time are plotted.}}}
+Paleomagnetic poles within 25 million years of each reconstruction time are plotted.
+All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}}}
 \label{fig:Laurentia_reconstructions}
@@
 Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating.
-The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.
+The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.
 
 There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, but did not publish parameters for a continuous model.
-The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude}.
+The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record.
 
@@
 
-There is strong evidence both in the geological and paleomagnetic record of Laurentia for differential plate tectonic motion between 2.2 and 1.8 Ga.
+There is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga.
 The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well.
@@
 As an example, consider the record of Laurentia in the late Mesoproterozoic.
-As can be seen in Table 2, there is a fantastic quantity of high-quality paleomagnetic poles between 1110 and 1070 Ma.
-These poles constrain rapid motion of Laurentia along
+As can be seen in Table 2, there are many high-quality paleomagnetic poles between 1110 and 1070 Ma.
+These poles constrain rapid motion of Laurentia leading up to collisional orogenesis as illustrated in Figure \ref{fig:Grenville_reconstructions} providing strong evidence for differential plate motion at the time.
 
@@
 The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
-\label{fig:Laurentia_reconstructions}
+\label{fig:Grenville_reconstructions}
 \end{figure}
@@
 
+\newpage
+
+\textbf{Table 2}: Compilation of paleomagnetic poles from Laurentia
 {\scriptsize

### 2019-10-22 Manuscript/pole_table_edited.tex
submission version of manuscript

@@
 \toprule
-                       terrane &                                        unit name & rating &  site lon (\textdegree) &  site lat (\textdegree) &  plon (\textdegree) &  plat (\textdegree) &  A$_{95}$ &                    age (Ma)&                                     pole reference \\ \hline
+                       terrane &                                        unit name & Nordic rating &  site lon (\textdegree) &  site lat (\textdegree) &  plon (\textdegree) &  plat (\textdegree) &  A$_{95}$ (\textdegree) &                    age (Ma)&                                     pole reference \\ \hline
 \midrule

### 2019-10-24 Manuscript/manuscript.tex
add text referring to THO ecologies

@@
 While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
-As an example, consider the record of Laurentia in the late Mesoproterozoic.
+From a paleomagnetic perspective there is strong paleomagnetic support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
+From a geological perspective, the the Trans-Hudson orogenic cycle, the Grenville orogenic cycle, and the Appalachian orogenic cycle are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig.
+\ref{fig:tectonic_history}).
+One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics.
+However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also ecologites preserved in the Trans-Hudson orogen that preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
+1.8 Ga \citep{Weller2017a}.
+Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}.
+Outside of Laurentia, there are examples of ecologites with geochemical affinity to oceanic crust such as that documented in the ca.
+1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.
+
+As an additional example, consider the record of Laurentia in the late Mesoproterozoic.
 As can be seen in Table 2, there are many high-quality paleomagnetic poles between 1110 and 1070 Ma.

### 2019-10-27 Manuscript/manuscript.tex
add intermittent stagnant lid text to disscussion and start abstract

@@
 
-\subsection{Introduction}
+\subsection{Abstract}
 
@@
 There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
-This chapter seeks to provide a concise review of these records.
+Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics
 
-\subsection{Broad tectonic history overview}
+\subsection{Introduction and broad tectonic history}
 
@@
 0.8 Ga \citep{Hamilton2011a, Stern2018a}.
-These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}.
+These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}.
 An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}.
 While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
-From a paleomagnetic perspective there is strong paleomagnetic support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
-From a geological perspective, the the Trans-Hudson orogenic cycle, the Grenville orogenic cycle, and the Appalachian orogenic cycle are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig.
+From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
+From a geological perspective, the Trans-Hudson orogenic cycle, the Grenville orogenic cycle, and the Appalachian orogenic cycle are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig.
 \ref{fig:tectonic_history}).
@@
 
-As an additional example, consider the record of Laurentia in the late Mesoproterozoic.
-As can be seen in Table 2, there are many high-quality paleomagnetic poles between 1110 and 1070 Ma.
-These poles constrain rapid motion of Laurentia leading up to collisional orogenesis as illustrated in Figure \ref{fig:Grenville_reconstructions} providing strong evidence for differential plate motion at the time.
+Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}.
+In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion.
+In particular, it was been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}.
+The accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
+\ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
+An additional constraint comes from the presence of reliable paleomagnetic poles themselves that are complete with baked contact tests.
+In a stagnant lid regime, there would not be sufficient heat flow to sustain a geodynamo and Earth should not have had a geomagnetic field that would have enable the preservation of paleomagnetic poles with positive field tests.
+The record of these poles also show that there was motion of Laurentia (Figs.
+\ref{fig:Laurentia_paleolatitude} and \ref{fig:Laurentia_reconstructions}).
+Using data from Laurentia alone, however, it is difficult to ascertain whether this motion is due to plate tectonic motion or rotation of the entire solid Earth through true polar wander.
+True polar wander can lead to changing position relative to the spin axis even with a stagnant lid.
+One interval when the Laurentian paleomagnetic record demands that some of the motion is through differential plate tectonics is in the latest Mesoproterozoic.
+At that time, the pole path is very well-resolved with many high-quality paleomagnetic poles between 1110 and 1070 Ma (Table 2; \ref{fig:Laurentia_poles}).
+The progression of the poles requires rotation about an Euler pole that is distinct from a great circle path which would result if the motion were solely due to true polar wander \citep{Swanson-Hysell2019a}.
+These poles constrain rapid motion of Laurentia leading up to collisional orogenesis associated with the Grenvillian orogeny as illustrated in Figure \ref{fig:Grenville_reconstructions}.
+These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid.
+Rather the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
 

### 2019-10-27 Manuscript/manuscript.tex
progress on abstract

@@
 There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
-Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics
+Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic (2.3 to 1.8 Ga) leading up to their amalgamation through the Trans-Hudson orogeny that led to the formation of the core of Laurentia.
+Subsequent crustal growth proceeded through accretionary orogenesis until the rapid plate motion and continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary at 1.0 Ga.
 

### 2019-10-28 Manuscript/manuscript.tex
Extend short introductory paragraph into a more comprehensive abstract

--- 2019-10-27 Manuscript/manuscript.tex
@@
 
-Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
+Laurentia was a major continent throughout the majority of the Proterozoic.
+It is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography.
-There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models.
-Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics
+There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism.
+These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models.
+Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny.
+The collisions of these and other Archean provinces led to the formation of the core of Laurentia.
+Subsequent crustal growth occurred through multiple intervals of accretionary orogenesis through the late Paleoproterozoic and Mesoproterozoic until the continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary (1.0 Ga).
+The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles.
+Following, a return to high-latitudes as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent.
+Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia–Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
+While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.
+In each case, accretionary collision of arc terranes was followed by continent-continent collision.
+The major difference is that the collisions of the Grenvillian and Appalachian orogenic cycles resulted in relatively minor crustal growth compared to the Trans-Hudson.
+Break-up following the Grenvillian and Appalachian orogenic cycles occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic cycle have remained sutured.
+As a result, Laurentia has been a formidable continent for the past 1.8 billion years whose position and history is key for reconstructing global paleogeography.
+The evidence from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.
 

### 2019-10-28 Manuscript/manuscript.tex
small edits through manuscript text

@@
 1.71 Ga) to the end of the Shawinigan orogeny (ca.
-1.06 Ga) requires a long-lived open margin without a major conjugate continent up until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
+1.06 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
 This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
@@
 This extensive and major collision orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
-The term Grenville orogeny or Grenville belt is used rather loosely throuhgout much of the literature to refer to any late Mesoproterozoic orogenic belt.
-The timeline of orogenesis on the Laurentia margin has more nuanced constraints that this which can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
+The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt.
+The timeline of orogenesis on the Laurentia margin has more nuanced constraints that this usage and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
 \ref{fig:tectonic_history}).
@@
 780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Rooney2017a}.
-However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran (closer to the ca.
+However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran Period (closer to the ca.
 539 Ma Neoproterozoic-Phanerozoic boundary; \citealp{Bond1984a, Levy1991a}).
@@
 A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
-\ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a,Whitmeyer2007a}.
+\ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}.
 The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}.
@@
 Rifting ca.
-620-580 Ma initiates in the region from modern-day New York to Newfoundland and by ca.
+620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca.
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
@@
 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca.
-460 Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.
+460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.
 
@@
 
-Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic provides very strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
+Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides very strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
 This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.
@@
 However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age.
-For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
+For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
 However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions.
@@
 1015 Ma Haliburton intrusions.
-However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).
+However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
 In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2).
@@
 This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
-Poles from the Chuar Group as presented in \cite{Eyster2019a} are also included.
+Poles from the Neoproterozoic Chuar Group as presented in \cite{Eyster2019a} are also included.
 
@@
 Uncertainty remains with respect to whether the ca.
-1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time being should be considered solely in the western Superior province reference frame.
+1.88 Ga Molson dikes pole pre-dates or post-dates this rotation \citep{Evans2010a} and thus for the time being should be considered solely in the western Superior province reference frame.
 
@@
 
-There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation (Fig.
+There are poles in the compilation for the Slave, Wyoming, Rae, Superior and North Atlantic provinces prior to Laurentia amalgamation (Fig.
 \ref{fig:Laurentia_poles} and Table 2).
@@
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
 \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}.
@@
 The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting.
-The preferred Euler of \cite{Maloof2006a} is used here that designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a}.
+The preferred Euler of \cite{Maloof2006a} is used here.
+This Euler is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
 
@@
 To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}.
-The ages of poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}.
+The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}.
 Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both.
@@
 \ref{fig:tectonic_history}).
-The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formations that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
+The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
 The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca.
@@
 720 Ma), as well as similiarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
-Overall, the record of paleomagnetic poles has a lot of internal consistency in intervals for which there is high-resolution coverage.
-These data result in the progressive paths agreement between poles such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian (ca.
+Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage.
+These data result in progressive paths such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian (ca.
 775 to 720 Ma) path \citep{Eyster2019a}.

### 2019-10-28 Manuscript/manuscript.tex
updates to manuscript including revised paleolatitude figure

@@
 
-There is an overall lack of published continuous models in the literature for the Proterozoic that can be compared to the compilation of paleomagnetic poles presented herein.
-The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters.
+There is an overall lack of models in the literature for the Proterozoic have published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein.
+The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters although some studies have published the Euler rotations associated with specified times.
 With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g.
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
+\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}.
+The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
+\label{fig:Grenville_reconstructions}
+\end{figure}
+
 Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}.
@@
 Rather the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
-\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}.
-The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
-\label{fig:Grenville_reconstructions}
-\end{figure}
 

### 2019-10-31 Manuscript/manuscript.tex
add note about polarity chosen in Superior-Slave figure
add details and Nimmo2000a citation regarding a dynamo implying plate tectonics

@@
 \caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.
+The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
 Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
@@
 
-\subsection{Conclusion}
+\subsection{The record implies plate tectonics throughout the Proterozoic}
 
@@
 An additional constraint comes from the presence of reliable paleomagnetic poles themselves that are complete with baked contact tests.
-In a stagnant lid regime, there would not be sufficient heat flow to sustain a geodynamo and Earth should not have had a geomagnetic field that would have enable the preservation of paleomagnetic poles with positive field tests.
-The record of these poles also show that there was motion of Laurentia (Figs.
+In a stagnant lid regime, there would not be sufficient heat flow to sustain a geodynamo \citep{Nimmo2000a}.
+Therefore, the interpretation of stagnant lid for an extended interval implies that Earth should not have had a geomagnetic field that would have enable the preservation of paleomagnetic poles with positive field tests.
+The record of poles with positive baked contact tests through the Paleoproterozoic and Mesoproterozoic necessitates a geomagnetic field which itself implies operational plate tectonics.
+This interpretation of a significant persistent geomagnetic field through much of the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g.
+\citealp{Macouin2006a}) and elsewhere.
+
+The record of these poles also show that there was progressive motion of Laurentia through the Proterozoic (Figs.
 \ref{fig:Laurentia_paleolatitude} and \ref{fig:Laurentia_reconstructions}).
@@
 Rather the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
+
+\section{Conclusion}
 

### 2019-12-06 Manuscript/manuscript.tex
small edits to manuscript text including adding mention of LEH2013 model

@@
 Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
-Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia–Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
 While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.
@@
 1.80 Ga \citep{Whitmeyer2007a}.
-A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca.
+A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca.
 1.72 Ga \citep{Kilian2016b}.
@@
 In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
-The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
+The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
 \citealp{Skipton2016a, Weller2017a}).
@@
 Subsequent accretionary orogenesis of the ca.
-1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
+1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
 An interval of major plutonism occurred ca.
-1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
+1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
 This plutonism is likely due to crustal melting within a back-arc region of ca.
@@
 Younger magmatic activity ca.
-1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
-While an active margin interpretation with magmatism in back‐arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
+While an active margin interpretation with magmatism in back-arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca.
-1.25-1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
+1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
 The subsequent ca.
@@
 In the latest Mesoproterozoic (ca.
-1.11-1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces.
-This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
+1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces \citep{Cannon1992b}.
+This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
 Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.
@@
 The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt.
-The timeline of orogenesis on the Laurentia margin has more nuanced constraints that this usage and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
+The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
 \ref{fig:tectonic_history}).
@@
 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
-The Maz–Arequipa–Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
+The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
 Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
@@
 
-Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides very strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
+Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
 This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.
@@
 These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}.
-Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions.
+Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions.
 However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age.
@@
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
 \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}.
 These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion.
-They also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces.
+The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces.
 Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.
@@
 
-Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is quite sparse (shown colored by age in Fig.
+Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig.
 \ref{fig:Laurentia_poles}).
@@
 1237 Ma Sudbury dikes and ca.
-1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Fig.
-\ref{fig:Logan_Loop}).
+1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.
 
@@
 Another challenging portion of the Laurentia record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs.
-\ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitudes}).
+\ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
 As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma.
@@
 \includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
-\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79ºN, lon=92.10ºW).
+\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W).
 The paleomagnetic poles are compiled in Table 2.
@@
 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
+The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.
 
@@
 As a result, for the portion of the reconstruction during the interval of time ca.
-1110 to 1070 Ma, I utilize an Euler pole inversion from \cite{Swanson-Hysell2019a}.
+1110 to 1070 Ma where there is high data density from the Midcontinent Rift, I utilize an Euler pole inversion from \cite{Swanson-Hysell2019a}.
 
@@
 As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block).
-While they provide constraints in this regard, they do not provide constraints in and of themselves of the longitudinal position of the block.
+While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block.
 Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
@@
 The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record.
+The compilation also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of (\citealp{Li2013a}; while the model parameters were not published with that study they have now been made available by the authors).
 
@@
 In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion.
-In particular, it was been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}.
+In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}.
 The accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
 \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
-An additional constraint comes from the presence of reliable paleomagnetic poles themselves that are complete with baked contact tests.
-In a stagnant lid regime, there would not be sufficient heat flow to sustain a geodynamo \citep{Nimmo2000a}.
-Therefore, the interpretation of stagnant lid for an extended interval implies that Earth should not have had a geomagnetic field that would have enable the preservation of paleomagnetic poles with positive field tests.
-The record of poles with positive baked contact tests through the Paleoproterozoic and Mesoproterozoic necessitates a geomagnetic field which itself implies operational plate tectonics.
+An additional constraint comes from the paleomagnetic record and paleomagnetic poles supported with baked contact tests through the Mesoproterozoic.
+In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to is necessary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.
+Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction.
+Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
+Therefore, the record supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic (Table 2) which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
 This interpretation of a significant persistent geomagnetic field through much of the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g.
@@
 The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles.
-As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of developing paleogeographic models through the Precambrian and will continue to be moving forward.
+As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.
 
@@
 
-\textbf{Table 2}: Compilation of paleomagnetic poles from Laurentia
 {\scriptsize
 \begin{landscape}
+\textbf{Table 2}: Compilation of paleomagnetic poles from Laurentia
 \input{pole_table_edited.tex}

### 2020-03-23 Manuscript/manuscript.tex
implement more edits associated with Evans review

@@
 \documentclass[11pt,letterpaper]{article}
-\setcounter{section}{2}
+\setcounter{section}{3}
 \usepackage{threeparttable}
@@
 
-Laurentia was a major continent throughout the majority of the Proterozoic.
-It is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
-The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography.
+Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic.
+The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
 There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism.
 These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models.
-Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny.
-The collisions of these and other Archean provinces led to the formation of the core of Laurentia.
-Subsequent crustal growth occurred through multiple intervals of accretionary orogenesis through the late Paleoproterozoic and Mesoproterozoic until the continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary (1.0 Ga).
-The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles.
-Following, a return to high-latitudes as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent.
-Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
-Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
-While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.
-In each case, accretionary collision of arc terranes was followed by continent-continent collision.
-The major difference is that the collisions of the Grenvillian and Appalachian orogenic cycles resulted in relatively minor crustal growth compared to the Trans-Hudson.
-Break-up following the Grenvillian and Appalachian orogenic cycles occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic cycle have remained sutured.
-As a result, Laurentia has been a formidable continent for the past 1.8 billion years whose position and history is key for reconstructing global paleogeography.
-The evidence from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.
+These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.
 
@@
 
-Laurentia refers to the craton that forms the Precambrian core of North America (Fig.
+Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig.
 \ref{fig:Laurentia_map}).
-Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
+Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
 Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
@@
 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
-A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
+A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
 This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
@@
 Overall, the collision of these Archean microcontinents between ca.
-1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.
+1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton.
 
@@
 \citealp{Skipton2016a, Weller2017a}).
-The \cite{Pehrsson2015a} model posits that this period terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
+The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
 
@@
 Subsequent accretionary orogenesis of the ca.
-1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
+1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
 An interval of major plutonism occurred ca.
-1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
+1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
 This plutonism is likely due to crustal melting within a back-arc region of ca.
 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}.
+Geologic data from northern New Mexico have been interpreted to indicate an interval of ca.
+1.49 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a}.
 Younger magmatic activity ca.
-1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
-While an active margin interpretation with magmatism in back-arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
+While an active margin interpretation with magmatism in back-arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 
-Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca.
+Accretionary orogenesis continued along the east margin of Laurentia with the arc-continent collision of the ca.
 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
 The subsequent ca.
-1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
+1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprising amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
 In the latest Mesoproterozoic (ca.
-1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces \citep{Cannon1992b}.
+1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}.
 This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
@@
 There is significantly less preserved crustal growth on the western margin of Laurentia (Fig.
-\ref{fig:Laurentia_map}) and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
+\ref{fig:Laurentia_map}), and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
 The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
-1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated.
-\citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}.
+1.47 to 1.40 rift \citep{Evans2000c}.
+While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated.
+\citet{Hoffman1989a} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}.
 This region is interpreted to have been subsequently deformed during a ca.
-1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
 
@@
 Amazonia in \citealp{Elming2009a}).
-The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca.
-1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
+The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca.
+1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
 Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea.
-This extensive and major collision orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
+This extensive and major collisional orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
 The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt.
-The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
+The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage, and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
 \ref{fig:tectonic_history}).
@@
 Along the western margin of Laurentia, small-scale rifting occurred ca.
-780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Rooney2017a}.
-However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran Period (closer to the ca.
+780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Rooney2017a}.
+However, this extensional basin development is relatively minor and predates the more significant rifting that led to passive margin thermal subsidence that did not occur until the Ediacaran Period (closer to the ca.
 539 Ma Neoproterozoic-Phanerozoic boundary; \citealp{Bond1984a, Levy1991a}).
@@
 
-Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
+Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
 A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
@@
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
-The last phases of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
-Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
-As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar.
+The last phase of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
+Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
+As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton, as the record that lingers on the craton is similar.
 One interpretation is that there was successful break-up along the eastern margin during the ca.
@@
 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
-The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
+The Maz-“Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
 Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
@@
 As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
-These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision.
-In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated with the assembly of the supercontinent Pangea.
+These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until continent-continent collision.
+In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia, and in the Appalachian interval it is interpreted to be associated with the assembly of the supercontinent Pangea.
 
@@
 The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland.
-An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions.
-A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
+The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions.
+The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
 Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
@@
 Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions.
-However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age.
+However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age.
 For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
@@
 1015 Ma Haliburton intrusions.
-However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
+However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
 In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2).
-Additional not-rated poles included in the present compilation is the new pole for the ca.
+An additional not-rated pole included in the present compilation is the new pole for the ca.
 1144 Ma Ontario lamprophyre dykes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca.
@@
 Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly.
-A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation.
+A lingering question raised in \citet{Hoffman1988a} is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation.
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
-\ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by five time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+\ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}).
 These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion.
@@
 The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles.
-As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.
+Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny.
+The collisions of these and other Archean provinces led to the formation of the core of Laurentia.
+Subsequent crustal growth occurred through multiple intervals of accretionary orogenesis through the late Paleoproterozoic and Mesoproterozoic until the continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary (1.0 Ga).
+The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles.
+Following, a return to high latitudes as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent.
+Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-“Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
+While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.
+In each case, accretionary collision of arc terranes was followed by continent-continent collision.
+The major difference is that the collisions of the Grenvillian and Appalachian orogenic cycles resulted in relatively minor crustal growth compared to the Trans-Hudson.
+Break-up following the Grenvillian and Appalachian orogenic cycles occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic cycle have remained sutured.
+As a result, Laurentia has been a formidable continent for the past 1.8 billion years.
+As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), the constraints from Laurentia are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.
 

### 2020-04-03 Manuscript/manuscript.tex
small edits to address review comments

@@
 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.
-As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
+
+As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
 1.80 Ga \citep{Whitmeyer2007a}.
@@
 
-In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna.
+The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}.
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
 \citealp{Skipton2016a, Weller2017a}).
-The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
+The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
 
@@
 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}.
-The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.
+The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca.
+1090 to 1030 Ma; \citealp{Rivers2008a}).
+Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca.
+980 Ma \citep{Hynes2010a}.
 
@@
 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
-The Maz-“Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
+The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
 Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
@@
 The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
+This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of narrower Manikewan with a pivot-like closure.
 Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
@@
 These terranes were once part of contiguous Laurentia, but have subsequently rifted away.
-These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the rotations shown in Table \ref{tab:terrane_rotations}.
+These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}.
 The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used).
@@
 The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting.
-The preferred Euler of \cite{Maloof2006a} is used here.
-This Euler is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
+The preferred Euler pole parameters of \cite{Maloof2006a} is used here.
+This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
 
@@
 \ref{fig:tectonic_history}).
-The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
+The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
 The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca.
@@
 Data from other terranes add resolution to the record.
-In particular, data from Greenland adds 12 poles between 1385 and 1160 Ma when there are only 4 poles from mainland Laurentia.
+In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia.
 Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent.
@@
 However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts.
-There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite) however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
 At present, it is unclear which poles are a better representation of Laurentia's position ca.
@@
 
-Another challenging portion of the Laurentia record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs.
+Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
@@
 One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}.
-Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Bono2019a}.
+Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}.
 Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig.
@@
 The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting spherical splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}.
-Applying such an approach can reduce the effect of spurious poles in regions of high data density where seeking to satisfy every mean pole position would result in jerky motion.
+Applying such an approach can reduce the influence of spurious poles.
+Such synthesis is particularly important in regions of high data density where seeking to satisfy every mean pole position would result in jerky motion.
 
@@
 In portions of the record with a more dense record of poles, such as ca.
-1450, a calculated running mean is used to integrate constraints from multiple poles.
+1450 Ma, a calculated running mean is used to integrate constraints from multiple poles.
 This method follows the approach taken in the Phanerozoic (e.g.
@@
 Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles.
-Paleomagnetic poles compilations themselves are evolving with better data and improved geochronology.
+Paleomagnetic pole compilations themselves are evolving with better data and improved geochronology.
 Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress in Precambrian paleogeography.
@@
 
-There is an overall lack of models in the literature for the Proterozoic have published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein.
-The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters although some studies have published the Euler rotations associated with specified times.
+There is an overall lack of models in the literature for the Proterozoic with published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein.
+The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters, although some studies have published the Euler rotations associated with specified times.
 With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g.
@@
 The model does not honor the Grenville loop (e.g.
-go to moderately high southerly latitudes ca.
-1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models.
+Laurentia going to moderately high southerly latitudes ca.
+1000 Ma), which is a striking departure from the paleomagnetic record and standard paleogeographic models.
 Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating.
-The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.
+The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence both for the supercontinent Rodinia being equatorial and for ice sheets associated with the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.
 
-There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, but did not publish parameters for a continuous model.
+There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model.
 The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record.
-The compilation also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of (\citealp{Li2013a}; while the model parameters were not published with that study they have now been made available by the authors).
+The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of (\citealp{Li2013a}; while the model parameters were not published with that study they have now been made available by the authors).
 
@@
 One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics.
-However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also ecologites preserved in the Trans-Hudson orogen that preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
-1.8 Ga \citep{Weller2017a}.
+However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}.
+These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
+1.8 Ga.
 Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}.
@@
 An additional constraint comes from the paleomagnetic record and paleomagnetic poles supported with baked contact tests through the Mesoproterozoic.
-In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to is necessary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.
+In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.
 Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction.
@@
 The progression of the poles requires rotation about an Euler pole that is distinct from a great circle path which would result if the motion were solely due to true polar wander \citep{Swanson-Hysell2019a}.
-These poles constrain rapid motion of Laurentia leading up to collisional orogenesis associated with the Grenvillian orogeny as illustrated in Figure \ref{fig:Grenville_reconstructions}.
+These poles constrain rapid motion of Laurentia leading up to collisional orogenesis associated with the Grenvillian orogeny, as illustrated in Figure \ref{fig:Grenville_reconstructions}.
 These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid.
-Rather the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
+Rather, the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
 
@@
 Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
-Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-“Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
 While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.

### 2020-04-04 Manuscript/manuscript.tex
additional edits related to review comments

@@
 \ref{fig:tectonic_history}).
-Along the western margin of Laurentia, small-scale rifting occurred ca.
-780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Rooney2017a}.
-However, this extensional basin development is relatively minor and predates the more significant rifting that led to passive margin thermal subsidence that did not occur until the Ediacaran Period (closer to the ca.
-539 Ma Neoproterozoic-Phanerozoic boundary; \citealp{Bond1984a, Levy1991a}).
+Along the western margin of Laurentia, rifting occurred ca.
+780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Dehler2017a, Rooney2017a}.
+However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca.
+539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}.
 The emplacement of the ca.
-780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin.
+780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia).
 If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
-The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up.
-One possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca.
+The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
+The record of Neoproterozoic basin development lead \cite{Yonkee2014a} to propose that the early ca.
+780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models).
+Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca.
 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard.
 In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
+
 The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
@@
 The last phase of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
+
+The tectonic history of terrane transfer can provide valuable paleogeographic constraints.
 Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
@@
 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
-The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
+The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
 Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
@@
 An additional not-rated pole included in the present compilation is the new pole for the ca.
-1144 Ma Ontario lamprophyre dykes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca.
+1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca.
 1140 Ma Abitibi dikes \citep{Ernst1993a}.
@@
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by five time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
 \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}).
@@
 The reliability of this approach gains credence through the good agreement between the ca.
-1633 Ma Melville Bugt diabase dykes pole from Greenland \citep{Halls2011a} and the ca.
+1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca.
 1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}.
 Similarly, there is good agreement between the ca.
-1267 Ma Mackenzie dykes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca.
-1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dyke \citep{Piper1977a}.
+1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca.
+1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}.
 Furthermore, the Greenland poles with ages that fall between the ca.
 1237 Ma Sudbury dikes and ca.
-1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.
+1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.
 

### 2020-04-06 Manuscript/manuscript.tex
small edits in response to reviewer line-by-line suggestions

@@
 A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown.
-The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.}}
+The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.
+Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}}
 \label{fig:tectonic_history}
@@
 \citealp{Brown2012a}).
-Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland.
+Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland.
 These terranes were once part of contiguous Laurentia, but have subsequently rifted away.
@@
 The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting.
-The preferred Euler pole parameters of \cite{Maloof2006a} is used here.
+The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction.
 This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
@@
 However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts.
-There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
 At present, it is unclear which poles are a better representation of Laurentia's position ca.
@@
 As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma.
-One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}.
+One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}.
 Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}.
@@
 
-A synthesized pole path for Laurentia is developed here and used to develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles.
+A synthesized pole path for Laurentia is developed here and used to develop a paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles.
 The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
@@
 Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
-In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.
-The continuous paleolatitude implied by this model is shown as the simple Laurentia pole interpolation mode in Figure \ref{fig:Laurentia_paleolatitude}.
+In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig.
+\ref{fig:Laurentia_paleolatitude}).
 
@@
 In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}.
-The accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
+The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
 \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
-An additional constraint comes from the paleomagnetic record and paleomagnetic poles supported with baked contact tests through the Mesoproterozoic.
+
+An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig.
+\ref{fig:baked_contact}).
 In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.

### 2020-05-14 Manuscript/manuscript.tex
small edits and addition of NW Laurentia basin development paragraph

@@
 Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic.
-The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
+The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic-Mesoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
 There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism.
 These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models.
-These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.
+These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.
 
@@
 Overall, the collision of these Archean microcontinents between ca.
-1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton.
+1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig.
+\ref{fig:Laurentia_map}).
 
@@
 1.88 to 1.84 Ga \citep{Hildebrand2009a}.
-Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
+Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
 1.86 to 1.82 \citep{Schulz2007a}.
@@
 \citealp{Skipton2016a, Weller2017a}).
-The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
+The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia.
 
@@
 1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
-While an active margin interpretation with magmatism in back-arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+While an active margin interpretation with magmatism in a back-arc setting has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 
-Accretionary orogenesis continued along the east margin of Laurentia with the arc-continent collision of the ca.
+Accretionary orogenesis continued along the east margin of Laurentia with the amalgamation and accretion of arcs associated with the ca.
 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
 The subsequent ca.
-1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprising amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
+1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of terrane comprised of a previously rifted fragment of Laurentia and led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}.
+The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
 In the latest Mesoproterozoic (ca.
@@
 1.71 Ga) to the end of the Shawinigan orogeny (ca.
-1.06 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
+1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
 This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
@@
 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
-Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea.
+Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and up to the Labrador Sea \citep{Rivers2008a}.
 This extensive and major collisional orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
@@
 Along the western margin of Laurentia, rifting occurred ca.
-780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2012a, Dehler2017a, Rooney2017a}.
+780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}.
 However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca.
@@
 
-The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
+In northwest Laurentia, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
+Tectonic extension is recorded in units of the lower Fifteenmile Group with maximum depositional ages of ca.
+1050 Ma with ongoing basin development ca.
+812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}.
+Earlier basin development recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age -- older than the Fifteenmile Group and younger than the ca.
+1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).
+
+Another margin that did experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
 \ref{fig:tectonic_history}).
 Available geochronological constraints and thermal subsidence modeling indicate ca.
-820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}.
-These data suggest that conjugate continental lithosphere rifted away from northeast Greenland ca.
+850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}.
+These data suggest that conjugate continental lithosphere had rifted away from northeast Greenland by ca.
 820 Ma.
@@
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
-The last phase of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
+The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
 
@@
 
-The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis.
-As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
+The eastern margin of Laurentia then went through the multiple phases of Appalachian orogenesis.
+As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
 These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until continent-continent collision.

### 2020-05-14 Manuscript/manuscript.tex
add NW Laurentia rift details to tectonic history

@@
 
-In northwest Laurentia, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
+In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
 Tectonic extension is recorded in units of the lower Fifteenmile Group with maximum depositional ages of ca.
@@
 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}.
-Earlier basin development recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age -- older than the Fifteenmile Group and younger than the ca.
+Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age -- older than the Fifteenmile Group and younger than the ca.
 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).

### 2020-05-17 Manuscript/manuscript.tex
add text related to the Llano uplift and microcontinent collision interpretation

@@
 However, the evidence for earlier orogenesis ca.
-1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca.
+1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade metamorphism as early as ca.
 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca.
@@
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
-\citealp{Skipton2016a, Weller2017a}).
+\citealp{Skipton2016a, Weller2017a}; Fig.
+\ref{fig:Superior_Slave_recons}).
 The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia.
@@
 Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
-1.71 to 1.68 Ga Yavapai orogeny.
+1.71 to 1.68 Ga Yavapai orogeny (Fig.
+\ref{fig:tectonic_history}).
 Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}.
@@
 
-Accretionary orogenesis continued along the east margin of Laurentia with the amalgamation and accretion of arcs associated with the ca.
+Accretionary orogenesis continued along the eastern margin of Laurentia with the amalgamation and accretion of arcs associated with the ca.
 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
@@
 The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
-In the latest Mesoproterozoic (ca.
-1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}.
-This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
-Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.
+An exception to this quiescence during the interval between the Shawinigan and Grenvillian orogenies is ca.
+1.15 to 1.12 Ga orogenesis in the Llano uplift of the southern Laurentia margin \citep{Mosher1998a}.
+Llano orogenesis is interpreted to have resulted from collision of continental lithosphere along with an accreted arc \citep{Mosher1998a}.
+This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}.
+Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision similar to that associated with the Paleozoic Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval \citep{fig:tectonic_history}.
+
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}.
+Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, up through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and up to the Labrador Sea \citep{Rivers2008a}.
 The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca.
@@
 Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca.
-980 Ma \citep{Hynes2010a}.
+980 Ma \citep{Hynes2010a}
+
+In the latest Mesoproterozoic (ca.
+1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}.
+This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
+Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.
 
@@
 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
-Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and up to the Labrador Sea \citep{Rivers2008a}.
 This extensive and major collisional orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
+Collisional orogenic activity in the same time interval such as in the
+
 The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt.
@@
 The tectonic history of terrane transfer can provide valuable paleogeographic constraints.
-Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
+Cuyania is widely interpreted be a rifted fragment of southeast Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
 As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton, as the record that lingers on the craton is similar.
@@
 While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block.
-Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
+Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes with specified paleolongitudes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
 In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig.
@@
 
-\section{Conclusion}
+%\subsection{Evaluating Laurentia's paleogeographic neighbors}
+%
+%\subsubsection{Baltica}
+%
+%\subsubsection{North China}
+%
+%\subsubsection{Australia}
+%
+%\subsubsection{Kalahari}
+% U-Pb zircon ages from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica have been interpreted to indicate continental arc magmatism between ca. 1160 to
+%1070 Ma with a a peak of activity ca.
+1110–1090 Ma  \citep{Wang2020a}.
+These age constraints as well as additional geochemical data have been interpreted to indicate subduction under Kalahari in the late Mesoproterozoic.
+% This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca. 1080 and 1030 Ma \citep{Wang2020a}.
+
+\subsection{Conclusion}
 
@@
 Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
-Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
 While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.
@@
 
-\section*{Acknowledgements}
+\subsection*{Acknowledgements}
 

### 2020-05-21 Manuscript/manuscript.tex
text improvements in Introduction and broad tectonic history section

@@
 This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}.
-Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision similar to that associated with the Paleozoic Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval \citep{fig:tectonic_history}.
+Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision.
+If this interpretation is correct, it is similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (\ref{fig:tectonic_history}).
 
@@
 Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca.
-980 Ma \citep{Hynes2010a}
+980 Ma \citep{Hynes2010a}.
 
@@
 
-There is significantly less preserved crustal growth on the western margin of Laurentia (Fig.
-\ref{fig:Laurentia_map}), and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin.
+There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig.
+\ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin.
 The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
@@
 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
-This extensive and major collisional orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
-Collisional orogenic activity in the same time interval such as in the
-
-The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt.
-The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage, and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
+This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
+Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage.
+These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
 \ref{fig:tectonic_history}).
@@
 
-Another margin that did experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
+Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
 \ref{fig:tectonic_history}).
@@
 
-Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
+Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
 A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
@@
 
-The tectonic history of terrane transfer can provide valuable paleogeographic constraints.
+The tectonic history of terrane transfer, such as Cuyania, can provide valuable paleogeographic constraints.
 Cuyania is widely interpreted be a rifted fragment of southeast Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
@@
 The eastern margin of Laurentia then went through the multiple phases of Appalachian orogenesis.
-As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval).
+As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano orogenies in the Grenville interval; Acadian orogeny in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian orogeny in the Appalachian interval).
 These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until continent-continent collision.

### 2020-05-23 Manuscript/manuscript.tex
add initial North China neighbor text

@@
 1015 Ma Haliburton intrusions.
-However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
-In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2).
+However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
+In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2).
 An additional not-rated pole included in the present compilation is the new pole for the ca.
@@
 This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
-Poles from the Neoproterozoic Chuar Group as presented in \cite{Eyster2019a} are also included.
+Poles from the Neoproterozoic Chuar Group of southerwest Laurentia as presented in \cite{Eyster2019a} are also included.
 
@@
 The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used).
-The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland.
-The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting.
+The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland.
+The reconstruction of Svalbard is more challenging given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting.
 The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction.
@@
 %\subsection{Evaluating Laurentia's paleogeographic neighbors}
+%%
+%%\subsubsection{Baltica}
+%%
+%\subsubsection{North China}
 %
-%\subsubsection{Baltica}
-%
-%\subsubsection{North China}
+%The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume).
+While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca.
+1110 Ma into the early Neoproterozoic.
+North China would have been at polar latitudes ca.
+1110 Ma and moved rapidly with Laurentia as it transited towards the equator.
+\cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
+In particular, sediment transport from Laurentia could provide a source for ca.
+1.18 Ga zircons (from the Shawinigan orogen) and ca.
+1.08 Ga zircons (from the Grenville orogen).
+In the Laurentia basins, ca.
+1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g.
+\citealp{Wang2020a}).
+If North China was in this position, the timing of its arrival of North China is unclear.
+The ca.
+1220 Ma dikes pole of the North China craton is not coincident with the ca.
+1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al.
+(this volume) to suggest that North China arrived on the Laurentian margin between ca.
+1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
+In terms of departing from this position, one possibility is that
+
+%CHECK THE SUDBURY POINT FOR MYSELF
+
 %

### 2020-05-23 Manuscript/manuscript.tex
add reviewer acknowledgements and (CC BY 4.0) license info

@@
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
+The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Francis Macdonald.
 This work was supported by NSF CAREER Grant EAR-1847277 awarded to N.L.S.-H.
-The code, data, and reconstructions used in this paper are openly available in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
+The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse (CC BY 4.0) with attribution in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
 

### 2020-06-12 Manuscript/manuscript.tex
remove Cuyania text per Athena's suggestion; add Paleo-Iapetus hypothesis as explanation of Laurentia's eastern margin rift history; add Amazonia section to the in-development paleogeographic neighbors section

@@
 The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
-
-The tectonic history of terrane transfer, such as Cuyania, can provide valuable paleogeographic constraints.
-Cuyania is widely interpreted be a rifted fragment of southeast Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}.
-As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton, as the record that lingers on the craton is similar.
-One interpretation is that there was successful break-up along the eastern margin during the ca.
-580 to 550 Ma interval of rifting prior to the ca.
-539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent.
-The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting.
-Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca.
-530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca.
-460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.
+As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar.
+Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca.
+700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia).
+This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Thethys).
+Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca.
+600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.
 
@@
 
+\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
+\textit{Note that this section is currently under development and is incomplete}
+
+\subsubsection{Amazonia}
+
+In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}.
+On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed.
+In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers.
+This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a,Cawood2017a}.
+In this model, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca.
+1080 initiating the Ottawan phase of the Grenvillian orogeny (Fig.
+\ref{fig:tectonic_history}).
+Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart.
+Departure of Amazonia potentially occurred as early as ca.
+700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.
+
+%%\subsubsection{Baltica}
+%%
+\subsubsection{North China}
+
+The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume).
+While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca.
+1110 Ma into the early Neoproterozoic.
+North China would have been at polar latitudes ca.
+1110 Ma and moved rapidly with Laurentia as it transited towards the equator.
+\cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
+In particular, sediment transport from Laurentia could provide a source for ca.
+1.18 Ga zircons (from the Shawinigan orogen) and ca.
+1.08 Ga zircons (from the Grenville orogen).
+In the Laurentia basins, ca.
+1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g.
+\citealp{Wang2020a}).
+If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear.
+The ca.
+1220 Ma dikes pole of the North China craton is not coincident with the ca.
+1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al.
+(this volume) to suggest that North China arrived on the Laurentian margin between ca.
+1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
+In terms of departing from this position, one possibility is that its departure is associated with early Neoproterozoic extension in northwest Laurentia.
+
+%CHECK THE SUDBURY POINT FOR MYSELF
+
+%
+%\subsubsection{Australia}
+%
+%\subsubsection{Kalahari}
+% U-Pb zircon ages from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica have been interpreted to indicate continental arc magmatism between ca. 1160 to
+%1070 Ma with a a peak of activity ca.
+1110–1090 Ma  \citep{Wang2020a}.
+These age constraints as well as additional geochemical data have been interpreted to indicate subduction under Kalahari in the late Mesoproterozoic.
+% This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca. 1080 and 1030 Ma \citep{Wang2020a}.
+
 \subsection{The record implies plate tectonics throughout the Proterozoic}
@@
 Rather, the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
-
-%\subsection{Evaluating Laurentia's paleogeographic neighbors}
-%%
-%%\subsubsection{Baltica}
-%%
-%\subsubsection{North China}
-%
-%The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume).
-While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca.
-1110 Ma into the early Neoproterozoic.
-North China would have been at polar latitudes ca.
-1110 Ma and moved rapidly with Laurentia as it transited towards the equator.
-\cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
-In particular, sediment transport from Laurentia could provide a source for ca.
-1.18 Ga zircons (from the Shawinigan orogen) and ca.
-1.08 Ga zircons (from the Grenville orogen).
-In the Laurentia basins, ca.
-1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g.
-\citealp{Wang2020a}).
-If North China was in this position, the timing of its arrival of North China is unclear.
-The ca.
-1220 Ma dikes pole of the North China craton is not coincident with the ca.
-1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al.
-(this volume) to suggest that North China arrived on the Laurentian margin between ca.
-1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
-In terms of departing from this position, one possibility is that
-
-%CHECK THE SUDBURY POINT FOR MYSELF
-
-%
-%\subsubsection{Australia}
-%
-%\subsubsection{Kalahari}
-% U-Pb zircon ages from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica have been interpreted to indicate continental arc magmatism between ca. 1160 to
-%1070 Ma with a a peak of activity ca.
-1110–1090 Ma  \citep{Wang2020a}.
-These age constraints as well as additional geochemical data have been interpreted to indicate subduction under Kalahari in the late Mesoproterozoic.
-% This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca. 1080 and 1030 Ma \citep{Wang2020a}.
 

### 2020-09-02 Manuscript/manuscript.tex
flip the polarity of Ediacaran poles within the figures and table, add more detail to Ediacaran pole discussion

@@
 \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-
-%commands that were used to generate a PDF that was copied to make the .docx version for GSAB
-%\pagestyle{empty}
-%\usepackage[nomarkers,figuresonly]{endfloat}
 
@@
 
-Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs.
+Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there are inconsistencies between poles of similar age (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
-As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma.
+As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 583 and 565 Ma (Fig.
+\ref{fig:Laurentia_paleolatitude}).
+This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g.
+\citealp{Pisarevsky2001a,Li2008a}).
 One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}.
-Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}.
+Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization and that interpret steeper directions also recovered from the intrusives as the result of remagnetization.
+They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca.
+583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization.
+Notably the lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole.
+Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}.
+Data from the ca.
+585 Ma Grenville dyke swarm interpreted as primary reveal $\sim$90\textdegree$\;$differences in direction within dikes dated within to high-precision 4 million years of one another \citep{Halls2015a}.
+The rates of $>$20\textdegree /Myr ($>$220 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
 Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig.

### 2020-09-02 Manuscript/pole_table.tex
flip the polarity of Ediacaran poles within the figures and table, add more detail to Ediacaran pole discussion

@@
              Laurentia-Wyoming &                            Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Selkin2008a\} \\
-      Laurentia-Superior(East) &                     Otto Stock dykes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Pullaiah1975b\} \\
+  Baltica-Fennoscandia-Karelia &                                Koitere sanukitoids &      B &      30.6 &      63.3 & 192.5 & -67.5 &      19.5 &     2684\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
+      Laurentia-Superior(East) &                       Otto Stock dikes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Pullaiah1975b\} \\
                Laurentia-Slave &                                       Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
-      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dykes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(East) &                               Matachewan dykes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(East) &                               Matachewan dykes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-               Laurentia-Slave &                                     Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2012a\} \\
-      Laurentia-Superior(East) &                                 Senneterre dykes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
+          Baltica-Fennoscandia &                   Shalskiy thick gabbronorite dike &      B &      36.0 &      61.9 & 222.1 &  22.7 &      11.5 &     2511\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
+      Laurentia-Superior(East) &                         Ptarmigan-Mistassini dikes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+          Baltica-Fennoscandia &                     Generalskaya Layered Intrusion &      B &      31.0 &      69.4 & 292.7 & -42.5 &      10.4 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                              Monchegorsk Intrusion &      B &      33.0 &      68.0 & 265.3 &   1.3 &       9.9 &     2504\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &  Avdeev gabbronorite and thin Shalskiy diabase ...
+&      B &      36.0 &      61.9 & 243.5 & -12.3 &      14.0 &   2476\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                                NaN \\
+      Laurentia-Superior(East) &                                 Matachewan dikes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+          Baltica-Fennoscandia &                          Imandra Layered Intrusion &      B &      33.1 &      67.6 & 280.3 & -16.1 &       7.8 &   2446\$\textasciicircum \{+39\}\$\$\_\{-39\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                                     Karelian dikes &      B &      30.7 &      66.0 & 278.7 & -19.9 &       6.1 &     2446\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
+      Laurentia-Superior(East) &                                 Matachewan dikes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+          Baltica-Fennoscandia &                            Taivalkoski 2.3 Ga dike &     C+ &      28.1 &      65.5 & 257.3 &  20.4 &       6.3 &   2339\$\textasciicircum \{+18\}\$\$\_\{-18\}\$ &                                                NaN \\
+               Laurentia-Slave &                                       Malley dikes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2012a\} \\
+      Laurentia-Superior(East) &                                   Senneterre dikes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
       Laurentia-Superior(East) &                                 Nipissing N1 sills &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
-               Laurentia-Slave &                                     Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
-      Laurentia-Superior(East) &                                Biscotasing dykes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Kilian2015a\} \\
-               Laurentia-Slave &                                      Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
-      Laurentia-Superior(West) &                                 Marathon dykes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
-      Laurentia-Superior(West) &                                 Marathon dykes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
-      Laurentia-Superior(West) &                               Cauchon Lake dykes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(West) &                               Fort Frances dykes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(East) &                                 Lac Esprit dykes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Greenland-Nain &                                  Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                 \textbackslash cite\{Fahrig1976b\} \\
-               Laurentia-Slave &                                Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Buchan2009a\} \\
-      Laurentia-Superior(East) &                                      Minto dykes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &                                       Dogrib dikes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
+      Laurentia-Superior(East) &                                  Biscotasing dikes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path dikes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Kilian2015a\} \\
+               Laurentia-Slave &                                        Indin dikes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
+      Laurentia-Superior(West) &                                   Marathon dikes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
+      Laurentia-Superior(West) &                                   Marathon dikes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
+      Laurentia-Superior(West) &                                 Cauchon Lake dikes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(West) &                                 Fort Frances dikes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(East) &                                   Lac Esprit dikes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+              Baltica-Sarmatia &               Volyn-Dniestr-Bug intrusions A-group &      B &      29.8 &      49.6 & 182.9 &  15.7 &      13.7 &   2061\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                               Kuetsyarvi Formation &      B &      29.5 &      69.5 & 300.8 &  24.7 &      16.4 &     2060\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                                NaN \\
+      Laurentia-Greenland-Nain &                                    Kangamiut dikes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                 \textbackslash cite\{Fahrig1976b\} \\
+               Laurentia-Slave &                                  Lac de Gras dikes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Buchan2009a\} \\
+              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &      B &      27.4 &      50.9 & 140.4 &  64.4 &      12.7 &   2000\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
+      Laurentia-Superior(East) &                                        Minto dikes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+             Baltica - Karelia &                           Onego 1975 Ma Intrusions &      A &      34.0 &      62.1 & 101.5 &  44.4 &       6.3 &     1976\$\textasciicircum \{+9\}\$\$\_\{-9\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                                    Konchozero Sill &      B &      34.0 &      62.1 & 281.9 & -14.2 &      10.4 &   1974\$\textasciicircum \{+27\}\$\$\_\{-27\}\$ &                                                NaN \\
                Laurentia-Slave &                                    Rifle Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans1981a\} \\
+          Baltica-Fennoscandia &             Tsuomasavarri Gabbro-Diorite Intrusion &      B &      28.3 &      69.9 & 247.3 &  40.2 &       6.0 &     1931\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                                NaN \\
                  Laurentia-Rae &                             Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Halls1999a\} \\
@@
                Laurentia-Slave &                                   Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
+          Baltica-Fennoscandia &                         Kiuruvesi-Pielavesi gabbro &      B &      26.8 &      63.2 & 231.3 &  41.1 &       5.4 &     1886\$\textasciicircum \{+6\}\$\$\_\{-5\}\$ &                                                NaN \\
                Laurentia-Slave &                           Mean Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
                Laurentia-Slave &                     Mean Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
-      Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(West) &                                Molson (B+C2) dikes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &                                 Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
                Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
-               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
+               Laurentia-Slave &               Pearson A/Peninsular/Kilohigok sills &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
             Laurentia-Superior &                         Haig/Flaherty/Sutton Mean  &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
 \\
-               Laurentia-Slave &             Pearson A/Peninsular/Kilohigok sills &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
+        Baltica - Fennoscandia &                                      Keuruu dikes  &     B  &      24.7 &      62.3 & 230.9 &  45.7 &       5.5 &   1869\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
  Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                 \textbackslash cite\{Symons1999a\} \\
-                 Laurentia-Rae &                                    Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{McGlynn1974a\} \\
+                 Laurentia-Rae &                                      Sparrow dikes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{McGlynn1974a\} \\
                  Laurentia-Rae &                                   Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                  \textbackslash cite\{Evans1973a\} \\
+          Baltica-Fennoscandia &                                    Notträsk gabbro &      B &      21.8 &      65.8 & 216.2 &  43.5 &       6.1 &     1806\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                                      Kallax gabbro &      B &      22.0 &      65.5 & 209.0 &  49.0 &       nan &     1800\$\textasciicircum \{+5\}\$\$\_\{-6\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                 Lake Ladoga, Mean intr.
+\& dikes, A &      B &      30.0 &      61.3 & 229.1 &  50.9 &       nan &   1800\$\textasciicircum \{+19\}\$\$\_\{-56\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                                      Hoting gabbro &      A &      16.2 &      64.2 & 233.3 &  43.0 &      10.9 &   1786\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
                      Laurentia &                                      Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                   \textbackslash cite\{Park1973a\} \\
+          Baltica-Fennoscandia &                              Småland intrusives -C &      A &      15.7 &      57.1 & 182.7 &  45.7 &       8.0 &     1777\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                                 Shoksha Sandstones &      B &      33.8 &      61.3 & 221.1 &  39.7 &       4.0 &   1775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                                NaN \\
  Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons2000a\} \\
  Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                   \textbackslash cite\{Gala1995a\} \\
-                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Irving2004a\} \\
-           Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Halls2011a\} \\
+              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &      A &      28.6 &      50.6 & 169.1 &  26.5 &       3.9 &   1755\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                                NaN \\
+          Baltica-Fennoscandia &                                MEAN Ropruchey sill &      B &      35.3 &      61.4 & 217.0 &  39.1 &       8.6 &     1751\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
+                     Laurentia &                                      Cleaver dikes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Irving2004a\} \\
+              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &      B &      29.6 &      50.0 & 163.2 &  10.7 &      10.2 &   1722\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                                NaN \\
+        Baltica - Fennoscandia &                                   Häme dike swarm  &      B &      25.1 &      61.4 & 209.8 &  23.6 &      14.7 &   1647\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                                NaN \\
+                       Baltica &                        Sipoo Quartz Porphyry dikes &      B &      25.3 &      60.3 & 180.4 &  26.4 &       9.4 &   1633\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
+           Laurentia-Greenland &                        Melville Bugt diabase dikes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Halls2011a\} \\
+                       Baltica &                              Quartz porphyry dikes &      B &      26.8 &      61.3 & 175.4 &  30.2 &       9.4 &   1631\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
+                       Baltica &                                 Hoting basic dikes &      B &      16.2 &      64.2 & 146.7 &  21.9 &      13.7 &   1614\$\textasciicircum \{+24\}\$\$\_\{-24\}\$ &                                                NaN \\
                      Laurentia &                            Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                 \textbackslash cite\{Irving1972a\} \\
+                       Baltica &                                   Åland dike swarm &      A &      20.0 &      60.0 & 191.4 &  23.7 &       2.8 &   1578\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                                NaN \\
+                       Baltica &                            Satakunta dike swarm -C &      A &      21.5 &      62.0 & 188.1 &  29.3 &       6.6 &   1578\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                                NaN \\
+                       Baltica &                                  Ragunda Formation &      B &      16.1 &      63.3 & 166.6 &  51.6 &       7.1 &   1506\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                                NaN \\
                      Laurentia &                 St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &                                  \textbackslash cite\{Meert2002b\} \\
+                       Baltica &                           MEAN Bunkris/Glysjon/Oje &      B &      13.5 &      61.5 & 179.8 &  28.3 &      13.2 &     1469\$\textasciicircum \{+9\}\$\$\_\{-9\}\$ &                                                NaN \\
                      Laurentia &                               Michikamau Intrusion &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Emslie1976a\} \\
                      Laurentia &                                  Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
+                       Baltica &                    Lake Ladoga basalt, sill, dikes &      A &      31.2 &      61.5 & 177.1 &  15.2 &       5.5 &   1452\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                                NaN \\
                      Laurentia &                                 Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
-                     Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &                                 \textbackslash cite\{Harlan2008a\} \\
+                     Laurentia &                                 Tobacco Root dikes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &                                 \textbackslash cite\{Harlan2008a\} \\
                      Laurentia &                                       Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
@@
                      Laurentia &         Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
+           Laurentia-Greenland &                                Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
+           Laurentia-Greenland &                      Victoria Fjord dolerite dikes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
            Laurentia-Greenland &                              Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
-           Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
-           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
+                       Baltica &                                             Mashak &      B &      57.0 &      54.0 & 193.0 &   1.8 &      14.8 &   1376\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
                      Laurentia &                                   Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                 \textbackslash cite\{Murthy1978a\} \\
-           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
            Laurentia-Greenland &                             North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-                     Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
-           Laurentia-Greenland &                       West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
-           Laurentia-Greenland &                    West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
-                     Laurentia &                           Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Palmer1977a\} \\
+           Laurentia-Greenland &                                  Kungnat Ring dike &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
+                     Laurentia &                         Mackenzie dikes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
+                       Baltica &                       MEAN post-Jotnian intrusions &      A &      18.1 &      62.4 & 159.1 &  -1.8 &       3.4 &   1258\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                                NaN \\
+           Laurentia-Greenland &                         West Gardar Dolerite dikes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
+           Laurentia-Greenland &                      West Gardar Lamprophyre dikes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
+                     Laurentia &                             Sudbury dikes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Palmer1977a\} \\
             Laurentia-Scotland &                                        Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                        Nordic workshop calculation \\
-           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                                 Hviddal Giant dike &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                                     Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
@@
 &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
-           Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-                     Laurentia &                        Ontario lamprophyre dykes &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\
-                     Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Ernst1993a\} \\
+           Laurentia-Greenland &                                 Giant Gabbro dikes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                               NE-SW Trending dikes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
+                     Laurentia &                          Ontario lamprophyre dikes &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\
+                     Laurentia &                                      Abitibi dikes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Ernst1993a\} \\
+                       Baltica &                        Salla dike (width 60-100 m) &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &     1122\$\textasciicircum \{+5\}\$\$\_\{-3\}\$ &                                                NaN \\
                      Laurentia &                            Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
@@
                      Laurentia &                                    Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1070\$\textasciicircum \{+14\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
+                       Baltica &                             MEAN Bamble Intrusions &     C+ &       9.5 &      58.9 & 222.5 & -15.1 &      19.3 &   1048\$\textasciicircum \{+73\}\$\$\_\{-73\}\$ &                                                NaN \\
+                       Baltica &                           Laanila-Ristijarvi dikes &      B &      28.1 &      68.7 & 212.2 &  -2.1 &      13.8 &   1044\$\textasciicircum \{+52\}\$\$\_\{-52\}\$ &                                                NaN \\
                      Laurentia &                              Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash cite\{Warnock2000a\} \\
@@
                      Laurentia &                       Adirondack Microcline gneiss &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &    960\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\
+                       Baltica &                            MEAN Blekinga dolerites &      B &      15.0 &      56.1 & 247.0 &  13.0 &      16.0 &      950\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
+                       Baltica &               dikes of central and southern Sweden &      A &      14.2 &      58.4 & 240.7 &  -0.9 &       6.7 &    945\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
             Laurentia-Scotland &                                     Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                        Nordic workshop calculation \\
+        Baltica - Fennoscandia &                Bjerkreim-Sokndal layered intrusion &      B &       6.1 &      58.5 & 217.9 & -35.9 &       6.3 &    921\$\textasciicircum \{+17\}\$\$\_\{-17\}\$ &                                                NaN \\
+                       Baltica &                    Bratton and Algon Igneous Rocks &      B &      11.7 &      57.9 & 249.0 &   5.0 &       3.9 &    916\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                                NaN \\
+                       Baltica &                           MEAN Rogaland igneous cx &      B &       6.9 &      58.3 & 207.9 & -43.2 &      10.1 &    903\$\textasciicircum \{+33\}\$\$\_\{-33\}\$ &                                                NaN \\
+                       Baltica &                                   Hunnedalen dikes &      B &       7.0 &      58.9 & 222.0 & -41.0 &      15.6 &    848\$\textasciicircum \{+27\}\$\$\_\{-27\}\$ &                                                NaN \\
             Laurentia-Svalbard &                      Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
             Laurentia-Svalbard &                      Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
-                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Calculation from \textbackslash cite\{Eyster2019a\} based on d...
+                       Baltica &                                    Katav Formation &      B &      57.4 &      54.8 & 169.9 &  35.7 &      11.4 &  800\$\textasciicircum \{+100\}\$\$\_\{-100\}\$ &                                                NaN \\
+                     Laurentia &                                    Gunbarrel dikes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Calculation from \textbackslash cite\{Eyster2019a\} based on d...
 \\
@@
                      Laurentia &                          Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      718\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Denyszyn2009b\} \\
-                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
-                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons1991a\} \\
-                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Meert1994a\} \\
-                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{Tanczyk1987a\} \\
+                       Baltica &                                     Egersund dikes &      A &       5.9 &      58.8 &  44.1 &  31.4 &      15.6 &      616\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                                NaN \\
+                     Laurentia &                                   Long Range dikes &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\
+                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 152.7 & -42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                           Baie des Moutons complex &      B &     301.0 &      50.8 & 141.5 &  34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                         Callander Alkaline Complex &      B &     280.6 &      46.2 & 121.4 & -46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons1991a\} \\
+                     Laurentia &                                   Catoctin Basalts &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Meert1994a\} \\
+                       Baltica &                                       Volhyn lavas &      B &      26.0 &      51.2 &   4.4 &  19.8 &      28.3 &    571\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
+                     Laurentia &                        Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{Tanczyk1987a\} \\
+                       Baltica &                               Kurgashlya formation &      A &      57.5 &      53.3 & 314.5 &  50.9 &       5.3 &      565\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
+                       Baltica &                                  Bakeevo formation &      A &      58.2 &      54.9 & 299.1 &  42.3 &       5.3 &      565\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
+                       Baltica &                        Basu-Kukkarauk formation -C &      B &      57.0 &      54.0 & 187.3 &   1.1 &       5.8 &      562\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                                NaN \\
+                 Baltica-Urals &                          Basu formation  (precise) &      B &      56.9 &      53.9 & 186.1 &   1.7 &       3.8 &    560\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                                NaN \\
+                       Baltica &                 Chernokamenskay group sediments -C &      B &      58.4 &      58.5 & 306.7 &  17.3 &       6.0 &    557\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                                NaN \\
+                       Baltica &                             MEAN Vendian sediments &      B &      40.0 &      65.5 & 293.0 &  31.3 &       9.9 &      556\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                                NaN \\
+                 Baltica-Urals &                                Zigan HTC thin beds &      B &      56.7 &      54.4 & 138.3 & -18.0 &      10.7 &   548\$\textasciicircum \{+27\}\$\$\_\{--12\}\$ &                                                NaN \\
+                       Baltica &                   Zigan formation clastic rocks -C &      B &      56.7 &      53.7 & 138.4 & -16.2 &       4.1 &      548\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                                NaN \\
 \end{longtable}

### 2020-09-02 Manuscript/pole_table_edited.tex
flip the polarity of Ediacaran poles within the figures and table, add more detail to Ediacaran pole discussion

@@
             Laurentia &                                    Carbon Canyon &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &      757$^{+7}_{-7}$ &  \cite{Weil2004a} as calculated in \cite{Eyster2019a} \\ \hline
-            Laurentia &                             Carbon Butte/Awatubi &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &      751$^{+8}_{-8}$ &                                 \cite{Eyster2019a} \\                   Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
-                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
-                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575$^{+5}_{-5}$ &                                 \cite{Symons1991a} \\ \hline
-                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572$^{+5}_{-5}$ &                                  \cite{Meert1994a} \\ \hline
-                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565$^{+4}_{-4}$ &                                 \cite{Tanczyk1987a} \\ \hline
+            Laurentia &                             Carbon Butte/Awatubi &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &      751$^{+8}_{-8}$ &                                 \cite{Eyster2019a} \\ \hline
+            Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
+                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline
+                     Laurentia &                         Baie des Moutons complex (A) &      B &     301.0 &      50.8 & 152.7 & -42.6 &      12.0 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
+                     Laurentia &                         Baie des Moutons complex (B) &      B &     301.0 &      50.8 & 141.5 &  34.2 &      15.4 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
+                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 121.4 & -46.3 &       6.0 &      575$^{+5}_{-5}$ &                                 \cite{Symons1991a} \\ \hline
+                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &      572$^{+5}_{-5}$ &                                  \cite{Meert1994a} \\ \hline
+                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &      565$^{+4}_{-4}$ &                                 \cite{Tanczyk1987a} \\ \hline
 \end{longtable}

### 2020-09-10 Manuscript/manuscript.tex
add Proterozoic evaporite text

@@
 700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia).
-This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Thethys).
+This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys).
 Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca.
@@
 
-%A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
 \begin{figure}
@@
 
+\subsection{Paleoenvironmental constraints on paleolatitude}
+
+Sedimentary rocks whose deposition is associated with certain climatic conditions have the potential to provide insight into paleolatitude. 
+Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation.
+Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}.
+Evaporite deposits are particularly compelling as a paleolatitude constraint given that their deposition is associated arid regions resulting from large scale Hadley cell downwelling \citep{Evans2006a}.
+While moisture in the subtropics can change along with Earth's climate the overall pattern of $\sim$15-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}.
+Using compiled data that paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga that large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones.
+This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale Hadley cell circulation.
+
+Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
+\begin{itemize}
+%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}.
+The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited
+\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict preserved gypsum and anhydrite within secondary silica \citep{White1984a}.
+The correlative Prichard Formation is
+intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}.
+Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}.
+Further up in the Belt Supergroup , within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
+These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and likely similar in age to the 1454 $\pm$ 9 Ma tuff within Helena Formation \citep{Evans2000c}.
+\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated) and shale with halite casts \citep{Kah2001a}.
+These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}  While gypsum beds can reach thicknesses of 2.5 meters \citep{Kah2001a}, evaporites in the Iqqittuq Formation are interpreted to be volumetrically minor compared to units such as the Tonian Ten Stone Formation \citep{Turner2009a}.
+\item the Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} .
+These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca.
+1005 Ma detrital zircons \citep{Turner2016a}.
+The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca.
+820 Ma \citep{Turner2016a}.
+These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
+\item the Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
+The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum bearing siltstone\citep{Jefferson1989a, Jones2010a}.
+The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
+\end{itemize}
+
 \subsection{Comparing paleogeographic models to the paleomagnetic compilation}
@@
 \subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
+Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models.
+This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
 \textit{Note that this section is currently under development and is incomplete}
@@
 In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers.
-This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a,Cawood2017a}.
-In this model, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca.
+This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a, Cawood2017a}.
+While the lack of ca.
+1100 and ca.
+1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca.
+1200 Nova Floresta pole and ca.
+1150 Fortuna Formation pole; D'Agrella-Filho et al., this volume) as shown in \cite{Evans2013a}.
+In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca.
 1080 initiating the Ottawan phase of the Grenvillian orogeny (Fig.
@@
 
-%%\subsubsection{Baltica}
-%%
+\subsubsection{Baltica}
+Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration.
+This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland (\citealp{Gower1990a}, Salminen et al., this volume).
+In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin \citep{Karlstrom2001a}.
+This reconstruction is brings paleomagnetic poles
+
+compatible with existing paleomagnetic constraints from ca.
+1750 to 1270 Ma \cite{Evans2008a} and these conjoined cratons feature as a major component of the hypothesized Nuna supercontinent (\citealp{Evans2011a, Zhang2012a}; .
+
 \subsubsection{North China}

### 2020-09-10 Manuscript/manuscript.tex
progress on adding evaporite paleolatitudes to analysis and manuscript

@@
 
-Sedimentary rocks whose deposition is associated with certain climatic conditions have the potential to provide insight into paleolatitude. 
+Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude.
 Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation.
 Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}.
-Evaporite deposits are particularly compelling as a paleolatitude constraint given that their deposition is associated arid regions resulting from large scale Hadley cell downwelling \citep{Evans2006a}.
-While moisture in the subtropics can change along with Earth's climate the overall pattern of $\sim$15-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}.
-Using compiled data that paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga that large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones.
-This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale Hadley cell circulation.
+Evaporite deposits are particularly compelling as paleolatitude constraints given that their deposition is interpreted to be associated with arid regions resulting from large-scale Hadley cell downwelling \citep{Evans2006a}.
+While moisture in the subtropics can change along with Earth's climate, the overall pattern of $\sim$10-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}.
+Using a compilation of paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones.
+This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale convection circulation cells.
+
+There is high evaporation in the subtropics and tropics.
+Within the tropical rain belt (0 to $\sim$10\textdegree$\;$latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree$\;$ of the equator (within $\sim$8\textdegree$\;$ of the equator over land) with evaporation exceeding precipitation from those latitudes towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree$\;$ \citep{Park2020b}.
+However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to the precipitation of evaporites.
+For example, Lake Magadi in Kenya at a latitude of 1.9\textdegree$\;$S is a saline lake where thick bedded evaporites have accumulated \citep{Eugster1980a}.
+Caution is therefore needed when interpreting paleolatitude from evaporites in terrestrial and intracratonic settings given that they could occur both in tropical and subtropical latitudes.
 
@@
 The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited
-\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict preserved gypsum and anhydrite within secondary silica \citep{White1984a}.
-The correlative Prichard Formation is
-intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}.
+\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preservedwithin secondary silica \citep{White1984a}.
+The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}.
 Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}.
-Further up in the Belt Supergroup , within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
-These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and likely similar in age to the 1454 $\pm$ 9 Ma tuff within Helena Formation \citep{Evans2000c}.
-\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated) and shale with halite casts \citep{Kah2001a}.
-These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}  While gypsum beds can reach thicknesses of 2.5 meters \citep{Kah2001a}, evaporites in the Iqqittuq Formation are interpreted to be volumetrically minor compared to units such as the Tonian Ten Stone Formation \citep{Turner2009a}.
-\item the Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} .
+Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
+There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.
+These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 within the Helena Formation \citep{Evans2000c}.
+\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.
+These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
+\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} .
 These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca.
@@
 These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
-\item the Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
-The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum bearing siltstone\citep{Jefferson1989a, Jones2010a}.
+\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
+The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}.
 The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
 \end{itemize}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Laurentia_evaporite_figure.pdf}
+\caption{\small{\textbf{Left panel: The paleolatitude of evaporite deposits following the amalgamation of Laurentia as reconstructed by the simple Laurentia model shown in Fig.
+\ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic.
+Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}.
+The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.
+Right panel: The present-day location of evaporites reconstructed to the left.
+1: Altyn Formation (Belt Supergroup), 2: Wallace/Helena Formations (Belt Supergroup), 3: Iqqittuq Formation (Bylot Supergroup), 4: Ten Stone Formation (Mackenzie Mountains Supergroup), 5: Minto Inlet Formation (Shaler Supergroup), 6: Redstone River Formation (Mackenzie Mountains Supergroup), 7: Kilian Formation (Shaler Supergroup), 8: Carboniferous Canadian Maritime, 9: Carboniferous Sverdrup, 10: Permian Midcontinental USA.}}}
+\label{fig:Laurentia_evaporites}
+\end{figure}
+
+In Figure \ref{fig:Laurentia_evaporites}, the paleolatitude of these evaporite deposits are reconstructed using the
 

### 2020-09-15 Manuscript/manuscript.tex
Pesonen review edits and starting glossary

@@
 \usepackage[outercaption]{sidecap}
+\usepackage{glossaries}
 \doublespacing
@@
 \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+
+\makeglossaries
+
+\newglossaryentry{latex}
+{
+    name=latex,
+    description={Is a mark up language specially suited 
+    for scientific documents}
+}
+
+\newglossaryentry{Laurentia}
+{
+    name=Laurentia,
+    description={craton that forms the Precambrian core of North America}
+}
+
+\newglossaryentry{Nuna}
+{
+    name=Nuna,
+    description={hypothesized Paleoproterozoic to Mesoproterozoic supercontinent}
+}
+
+\newglossaryentry{Archean}
+{
+    name=Archean,
+    description={Geologic eon spanning from 4,000 to 2,500 million years ago (4 to 2.5 Ga).}
+}
 
@@
 
-Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic.
+Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its ca.
+1.8 Ga amalgamation.
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic-Mesoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
@@
 These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models.
-These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.
+These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.
 
@@
 \ref{fig:Laurentia_map}).
-Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
+Laurentia comprises multiple \gls{Archean} provinces that had unique histories prior to their amalgamation in the Paleoproterozoic ca.
+1.8 Ga, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
 Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
@@
 \ref{fig:Superior_Slave_recons}).
-The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia.
+The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
+In this volume, \cite{Elming2021a} put forward an alternate scenario wherein Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}).
+In the \cite{Elming2021a} model, this core subsequently grows to be a semi-supercontinent with India and Australia, however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
 
@@
 While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
-The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland.
+The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}.
 The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions.
@@
 This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
-Poles from the Neoproterozoic Chuar Group of southerwest Laurentia as presented in \cite{Eyster2019a} are also included.
+Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca.
+760 Ma)  as presented in \cite{Eyster2019a} are also included.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Superior_Slave_reconstructions.pdf}
+\includegraphics[width=6 in]{../Figures/Superior_Slave_reconstructions.pdf}
 \caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.
@@
 These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion.
-The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces.
+The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of the Manikewan Ocean between the Superior province and the Hearne+Rae+Slave provinces.
 Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.
@@
 Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both.
-In addition, intraplate magmatism resulting from plume-related large-igneous provinces can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g.
+In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g.
 the ca.
@@
 780 Ma) and Franklin LIP (ca.
-720 Ma), as well as similiarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
+720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
 Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage.
@@
 However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts.
-There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
 At present, it is unclear which poles are a better representation of Laurentia's position ca.
@@
 One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}.
-Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization and that interpret steeper directions also recovered from the intrusives as the result of remagnetization.
+Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca.
+565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization and that interpret steeper directions also recovered from the intrusives as the result of remagnetization.
 They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca.
@@
 The rates of $>$20\textdegree /Myr ($>$220 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
+Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak which could support a deviation from stable axial dipolar geomagnetic field behavior at the time \citep{Thallner2020a}.
 Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig.
@@
 In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca.
-1080 initiating the Ottawan phase of the Grenvillian orogeny (Fig.
+1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig.
 \ref{fig:tectonic_history}).
@@
 
+\clearpage
+\printglossaries
+
 \begin{table}[hbt]

### 2020-09-15 Manuscript/manuscript.tex
progress on glossary

@@
 \usepackage[outercaption]{sidecap}
-\usepackage{glossaries}
 \doublespacing
@@
 \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-
-\makeglossaries
-
-\newglossaryentry{latex}
-{
-    name=latex,
-    description={Is a mark up language specially suited 
-    for scientific documents}
-}
-
-\newglossaryentry{Laurentia}
-{
-    name=Laurentia,
-    description={craton that forms the Precambrian core of North America}
-}
-
-\newglossaryentry{Nuna}
-{
-    name=Nuna,
-    description={hypothesized Paleoproterozoic to Mesoproterozoic supercontinent}
-}
-
-\newglossaryentry{Archean}
-{
-    name=Archean,
-    description={Geologic eon spanning from 4,000 to 2,500 million years ago (4 to 2.5 Ga).}
-}
 
@@
 \ref{fig:Laurentia_map}).
-Laurentia comprises multiple \gls{Archean} provinces that had unique histories prior to their amalgamation in the Paleoproterozoic ca.
+Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic ca.
 1.8 Ga, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
@@
 1.80 Ga \citep{Whitmeyer2007a}.
-A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca.
+A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca.
 1.72 Ga \citep{Kilian2016b}.
@@
 However, the evidence for earlier orogenesis ca.
-1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade metamorphism as early as ca.
+1.78 to 1.75 in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca.
 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca.
@@
 
-\clearpage
-\printglossaries
+\section*{Glossary}
+\noindent\textbf{accretionary orogeny } Lithospheric deformation associated with the subduction of oceanic lithosphere and the addition of material from the downgoing plate such as island arcs.
+
+\noindent\textbf{Archean } A geologic eon spanning from 4,000 to 2,500 million years ago (4 to 2.5 Ga)
+
+\noindent\textbf{Archean province } A contiguous area of Archean continental lithosphere typically surrounded by Proterozoic orogens inferred to be suture zones (e.g. Superior province). \\
+
+\noindent\textbf{collisional orogeny } Lithospheric deformation resulting from the collision of two significant provinces of continental lithosphere.
+
+\noindent\textbf{Ga } Giga-annum, one billion (10$^9$) years. This term is used as an abbreviation for ``billions of years before present.''
+
+\noindent\textbf{Laurentia } The cratonic core of the North America continent that formed through the amalgamation of Archean provinces in the Paleoproterozoic.
+
+\noindent\textbf{lithosphere } The rigid outermost layer of the Earth that is broken into tectonic plates and responds to the emplacement of a load by flexural bending.
+
+\noindent\textbf{Ma } Mega-annum, one million (10$^6$) years. This term is used as an abbreviation for ``millions of years before present.''
+
+\noindent\textbf{Manikewan Ocean } An ocean basin interpreted to have existed between the  Slave+Rae+Hearne+North Atlantic provinces and Superior province that closed leading up to the Trans-Hudson orogeny.
+
+\noindent\textbf{Mesoproterozoic } A geologic era spanning from 1,600 to 1,000 million years ago.
+
+\noindent\textbf{monazite } A phosphate mineral (Ce,La,Nd,Th)(PO\textsubscript{4},SiO\textsubscript{4}) found as an accessory phase in metamorphic rocks that can be targeted by U-Pb geochronology to date metamorphic events.
+
+\noindent\textbf{Neoproterozoic } A geologic era spanning from 1,000 to 541 million years ago.
+
+\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and broke apart in the middle of the Mesoproterozoic.
+
+\noindent\textbf{orogen } A region of lithosphere that has undergone deformation during a mountain-building event (an orogeny).
+
+\noindent\textbf{Paleoproterozoic } A geologic era spanning from 2,500 to 1,600 million years ago.
+
+\noindent\textbf{Precambrian } A commonly used informal term to refer to geologic time prior to the Cambrian Period that started 541 million years ago.
+
+\noindent\textbf{province } A spatial entity with a shared geologic history.
+The term is used in this chapter to refer to Archean provinces that moved as independent cratonic blocks prior to Laurentia's amalgmation (e.g.
+Superior province).
+It is also used to refer to zones of crustal growth associated with orogens and the products of contemporaneous magmatic activity (large igneous provinces).
+
+\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era and broke apart in the middle of the Neoproterozoic.
+
+\noindent\textbf{Superior province } The largest Archean province of Laurentia framed by the Trans-Hudson orogen to the west, the Grenvillian orogen to the east and the Penokean orogen to the south.
+It is also referred to as the Superior craton.
+
+\noindent\textbf{Thelon orogen } An orogen resulting from a Paleoproterozoic collision orogeny between the Slave and Rae provinces.
+
+\noindent\textbf{Wopmay orogen } An orogen resulting from collision between the Hottah terrane, a continental magmatic arc, and the west margin of the Slave province.
+
+\noindent\textbf{zircon } A nesosilicate mineral with the chemical name of zirconium silicate and a chemical formula of ZrSiO\textsubscript{4}.
 

### 2020-09-15 Manuscript/manuscript.tex
smalls edits to tectonic into and additions to glossary

@@
 1.8 Ga amalgamation.
-The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic-Mesoproterozoic Nuna and Neoproterozoic Rodinia supercontinents.
+The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia.
 There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism.
@@
 \ref{fig:Laurentia_map}).
-Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic ca.
-1.8 Ga, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
+Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca.
+1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
 Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
 Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca.
-1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}.
+1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}.
 A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
@@
 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca.
-1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca.
+1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca.
 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca.
-1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.
+1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.
 
-As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca.
+As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca.
 1.80 Ga \citep{Whitmeyer2007a}.
@@
 1.72 Ga activity is a minor overprint on ca.
-1.75 terminal suturing between Wyoming and Superior.
+1.75 terminal suturing between the Wyoming and Superior provinces.
 Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca.
@@
 \caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}).
+Proterozoic orogens are labeled with \textit{italicized text} (or.
+?
+orogen).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
-In this volume, \cite{Elming2021a} put forward an alternate scenario wherein Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}).
-In the \cite{Elming2021a} model, this core subsequently grows to be a semi-supercontinent with India and Australia, however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
+In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography.
+In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}).
+This core then subsequently grows to be a semi-supercontinent with India and Australia, however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
 
@@
 
+\singlespacing
 \bibliographystyle{gsabull}
@@
 
-\noindent\textbf{Archean province } A contiguous area of Archean continental lithosphere typically surrounded by Proterozoic orogens inferred to be suture zones (e.g. Superior province). \\
+\noindent\textbf{Archean province } A contiguous area of Archean continental lithosphere typically surrounded by Proterozoic orogens inferred to be suture zones (e.g. Superior province).
+
+\noindent\textbf{Canadian shield } Large area of exposed Precambrian rock of Laurentia, or rock covered by thin soil, that is well-exposed due to Pleistocene glaciation erosion
 
@@
 
+\noindent\textbf{craton } The stable and relatively immobile continental lithosphere in the interior of continents.
+In this chapter, craton is predominantly used in reference to Laurentia which formed through the collision of Archean provinces and grew further through subsequent accretionary and collisional orogenesis.
+
 \noindent\textbf{Ga } Giga-annum, one billion (10$^9$) years. This term is used as an abbreviation for ``billions of years before present.''
 
-\noindent\textbf{Laurentia } The cratonic core of the North America continent that formed through the amalgamation of Archean provinces in the Paleoproterozoic.
+\noindent\textbf{Grenvillian orogen } An orogen resulting from collision orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.
+
+\noindent\textbf{Hearne province } An Archean province of Laurentia extending from southern Alberta, Canada to Hudson Bay.
+It is framed by the Rae province to the northwest and the Trans-Hudson orogen to the southeast.
+It is also referred to as the Hearne craton.
+
+\noindent\textbf{Laurentia } The Precambrian cratonic core of the North America continent and Greenland that formed through the amalgamation of Archean provinces in the Paleoproterozoic.
 
@@
 
+\noindent\textbf{Medicine Hat province } An Archean province of Laurentia extending from northern Montana, USA into southern Alberta and Saskatchewan, Canada.
+It is framed by a suture with the Hearne province to the north, the Trans-Hudson orogen to the east, and the Great Falls tectonic zone to the south.
+It is also referred to as the Medicine Hat block
+
 \noindent\textbf{Mesoproterozoic } A geologic era spanning from 1,600 to 1,000 million years ago.
+
+\noindent\textbf{Meta Incognita province } A province of Archean basement rocks that comprises most of southern Baffin Island. It is also referred to as the Meta Incognita microcontinent.
 
@@
 
+\noindent\textbf{Nagssugtoqidian orogen } An orogen resulting from the Paleoproterozoic collision between the Rae and North Atlantic provinces.
+
 \noindent\textbf{Neoproterozoic } A geologic era spanning from 1,000 to 541 million years ago.
+
+\noindent\textbf{North Atlantic province } An Archean province of Laurentia in southernmost Greenland and northeastern Labrador, Canada. It is also referred to as the North Atlantic craton.
 
@@
 
+\noindent\textbf{Penokean orogen } An orogen resulting from Paleoproterozoic accretion of an oceanic arc and the Marshfield terrane continental block along the southern margin of the Superior province.
+
 \noindent\textbf{Precambrian } A commonly used informal term to refer to geologic time prior to the Cambrian Period that started 541 million years ago.
+
+\noindent\textbf{Proterozoic } A geologic eon spanning from 2,500 to 541 million years ago.
 
@@
 
+\noindent\textbf{Rae province } An Archean province of Laurentia extending the region of Lake Athabasca northeast to northern Baffin Island in arctic Canada.
+It is framed by the Thelon orogen to the west, the Taltson orogen to the southwest and the Hearne province to the east.
+It is also referred to as the Rae craton.
+
 \noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era and broke apart in the middle of the Neoproterozoic.
+
+\noindent\textbf{Slave province } An Archean province of Laurentia extending to the north from the region of Great Slave Lake in northern Canada.
+It is framed by the Thelon orogen to the east and the Great Bear Arc to the west.
+It is also referred to as the Slave craton.
+
+\noindent\textbf{Snowbird orogen } An orogen resulting from Paleoproterozoic collision between the Rae and Hearne provinces prior to the Trans-Hudson orogeny.
+The Snowbird tectonic zone is part of the orogen.
 
@@
 
-\noindent\textbf{Thelon orogen } An orogen resulting from a Paleoproterozoic collision orogeny between the Slave and Rae provinces.
+\noindent\textbf{Thelon orogen } An orogen resulting from the Paleoproterozoic collision orogeny between the Slave and Rae provinces.
 
-\noindent\textbf{Wopmay orogen } An orogen resulting from collision between the Hottah terrane, a continental magmatic arc, and the west margin of the Slave province.
+\noindent\textbf{Torngat orogen } An orogen resulting from the Paleoproterozoic collision orogeny between the Meta Incognita and North Atlantic provinces.
+
+\noindent\textbf{Trans-Hudson orogen } An orogen resulting from the Paleoproterozoic collision between the composite Slave+Rae+Hearne provinces and the Superior province.
+
+\noindent\textbf{Wopmay orogen } An orogen resulting from Paleoproterozoic collision between the Hottah terrane, a continental magmatic arc, and the west margin of the Slave province.
+
+\noindent\textbf{Wyoming province } An Archean province of Laurentia underlying much of Wyoming, USA and southeast Montana, USA.
+It is framed by the Trans-Hudson orogen to the east (sometimes referred to the Black Hills orogen within the USA) and the Great Falls tectonic zone to the north.
+It is also referred to as the Wyoming craton.
 

### 2020-09-15 Manuscript/manuscript.tex
add Yavapai and Mazatzal to glossary

@@
 
+\noindent\textbf{Mazatzal orogen } An orogen resulting from latest Paleoproterozoic accretion of volcanic arc and back-arc terranes with southern Laurentia.
+
 \noindent\textbf{Medicine Hat province } An Archean province of Laurentia extending from northern Montana, USA into southern Alberta and Saskatchewan, Canada.
@@
 
+\noindent\textbf{Yavapai orogen } An orogen resulting from Paleoproterozoic collision and accretion of oceanic arc terranes with southern Laurentia.
+
 \noindent\textbf{zircon } A nesosilicate mineral with the chemical name of zirconium silicate and a chemical formula of ZrSiO\textsubscript{4}.

### 2020-09-16 Manuscript/manuscript.tex
add Kalahari paleogeographic neighbor subsection

@@
 
-\linenumbers
+%\linenumbers
 
@@
 Proterozoic orogens are labeled with \textit{italicized text} (or.
-?
-orogen).
+-- orogen).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
-This plutonism is likely due to crustal melting within a back-arc region of ca.
-1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}.
-Geologic data from northern New Mexico have been interpreted to indicate an interval of ca.
-1.49 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a}.
+This plutonism is interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region Laurentia's long-live active margin \citep{Bickford2015a}.
+Metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca.
+1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}.
 Younger magmatic activity ca.
-1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.
-While an active margin interpretation with magmatism in a back-arc setting has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}.
+While an active margin interpretation for the Granite-Rhyolite Province with arc and back-arc magmatism in a back-arc setting has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 
+\subsubsection{Australia}
+
+Could discuss Gibson2019b model related to hypothesized Paleoproterozoic tectonic history
+
 \subsubsection{Baltica}
@@
 1750 to 1270 Ma \cite{Evans2008a} and these conjoined cratons feature as a major component of the hypothesized Nuna supercontinent (\citealp{Evans2011a, Zhang2012a}; .
+
+\subsubsection{Kalahari}
+
+High-quality paleomagnetic poles constrain the coeval Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree$\;$ of latitude at the time they were emplaced with the craton margins separated by more than 30\textdegree$\;$ of latitude \citep{Swanson-Hysell2015a}.
+These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone.
+However, with the subsequent rapid motion of Laurentia to low latitudes it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny.
+This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g.
+\citealp{Li2008a}).
+Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift.
+The preferred interpretation of \cite{Swanson-Hysell2015a} and \citep{Kasbohm2015a} was that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
+\citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia.
+The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them become conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}.
+The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca.
+1090 followed by peak granulite metamorphism ca.
+1065 to 1045 Ma \citep{Diener2013, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean.
+If they indeed became conjoined in Rodinia, the separation of Kalahari from Laurentia likely initiated ca.
+795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a}.
 
@@
 
+\noindent\textbf{large igneous province (LIP) } Voluminous and rapidly emplaced volcanics and intrusions that are typically of mafic composition.
+These provinces are often interpreted to be due to decompression melting of an upwelling mantle plume.
+
 \noindent\textbf{lithosphere } The rigid outermost layer of the Earth that is broken into tectonic plates and responds to the emplacement of a load by flexural bending.

### 2020-09-16 Manuscript/manuscript.tex
small edits as well as additions to glossary

@@
 Proterozoic orogens are labeled with \textit{italicized text} (or.
--- orogen).
+-- orogen; THO  -- Trans-Hudson orogen; prov.
+-- province).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 1.71 Ga) to the end of the Shawinigan orogeny (ca.
-1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}.
+1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}.
 This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
-Amazonia in \citealp{Elming2009a}).
+Amazonia in \citealp{Elming2009a, Elming2021a}).
 The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca.
@@
 The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
-The record of Neoproterozoic basin development lead \cite{Yonkee2014a} to propose that the early ca.
+The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models).
@@
 This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys).
-Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca.
+Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia margin ca.
 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.
@@
 \ref{fig:Laurentia_poles} and Table 2).
-The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume).
+The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021b}.
 While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
@@
 For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
-However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions.
+However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen.
+Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions.
 As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca.
-1015 Ma Haliburton intrusions.
+1015 Ma Haliburton intrusions (Table 2).
 However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
@@
 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca.
-1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}.
+1590 Ma Western Channel diabase pole of mainland Laurentia (\citealp{Irving1972a}; Figs.
+\ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
 Similarly, there is good agreement between the ca.
@@
 1237 Ma Sudbury dikes and ca.
-1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.
+1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs.
+\ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
 
@@
 
-Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there are inconsistencies between poles of similar age (Figs.
+Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic era where there are inconsistencies between poles of similar age (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
@@
 Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca.
-565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization and that interpret steeper directions also recovered from the intrusives as the result of remagnetization.
+565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.
+\cite{Bono2015a} interpret steeper directions also recovered from the intrusives as the result of remagnetization.
 They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca.
@@
 Data from the ca.
-585 Ma Grenville dyke swarm interpreted as primary reveal $\sim$90\textdegree$\;$differences in direction within dikes dated within to high-precision 4 million years of one another \citep{Halls2015a}.
-The rates of $>$20\textdegree /Myr ($>$220 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
+585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree$\;$differences in direction within dikes dated within to 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
+The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
 Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak which could support a deviation from stable axial dipolar geomagnetic field behavior at the time \citep{Thallner2020a}.
-Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig.
+Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig.
 \ref{fig:Laurentia_paleolatitude}).
@@
 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
-The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.
+The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China (albeit one that needs to be revisited given updates to the paleomagnetic and geochronologic record from South China; \citealp{Zhang2021a}).
 
@@
 Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes with specified paleolongitudes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
-In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig.
+In Figure \ref{fig:Laurentia_reconstructions}, the map projections are centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig.
 \ref{fig:Laurentia_paleolatitude}).
@@
 \subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
-Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models.
+Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia.
 This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
@@
 
-Could discuss Gibson2019b model related to hypothesized Paleoproterozoic tectonic history
+%Could discuss Gibson2019b model related to hypothesized Paleoproterozoic tectonic history
 
@@
 
+\noindent\textbf{allochthonous } An adjective denoting that a rock or terrane originated in a position at significant distance from the lithospheric block that it is currently a part of.
+
 \noindent\textbf{Archean } A geologic eon spanning from 4,000 to 2,500 million years ago (4 to 2.5 Ga)
@@
 
+\noindent\textbf{conjugate } Adjective referring to continents or continental margins that were previously conjoined.
+
 \noindent\textbf{craton } The stable and relatively immobile continental lithosphere in the interior of continents.
@@
 
+\noindent\textbf{Elzevirian orogen } The orogen resulting from the Mesoproterozoic Elzevirian orogeny when there was accretion of arc terranes to eastern Laurentia.
+
+\noindent\textbf{evaporite } Minerals that crystallize from water that becomes supersaturated in salts through evaporation.
+
 \noindent\textbf{Ga } Giga-annum, one billion (10$^9$) years. This term is used as an abbreviation for ``billions of years before present.''
 
+\noindent\textbf{geocentric axial dipole hypothesis } The hypothesis that when it is time-averaged, Earth's magnetic field is dominantly a dipole aligned with the spin axis.
+
+\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from southerneastern Ontario, Canada to west Texas, USA.
+
 \noindent\textbf{Grenvillian orogen } An orogen resulting from collision orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.
+
+\noindent\textbf{Hadley cell } Large-scale atmospheric circulation where air rises near the equator, flows poleward, and descends in the subtropics.
+This circulation drives convective tropical precipitation and the dry downwelling air leads to aridity in the subtropics.
 
@@
 It is also referred to as the Hearne craton.
+
+\noindent\textbf{hematite } An iron oxide mineral with a formula of Fe\textsubscript{2}O\textsubscript{3} that commonly holds magnetization in geologic materials, particularly oxidized sedimentary rocks.
+
+\noindent\textbf{juvenile } Rocks that have formed through melt recently extracted from the mantle.
 
@@
 
+\noindent\textbf{Midcontinent Rift } A major Mesoproterozoic intracratonic rift where there was co-location of large igneous province magmatism and extension in Laurentia's interior centered on the Lake Superior region.
+
 \noindent\textbf{monazite } A phosphate mineral (Ce,La,Nd,Th)(PO\textsubscript{4},SiO\textsubscript{4}) found as an accessory phase in metamorphic rocks that can be targeted by U-Pb geochronology to date metamorphic events.
@@
 
+\noindent\textbf{paleolatitude } The past latitude of a given point on Earth's surface at a given time.
+
+\noindent\textbf{paleomagnetic pole } A calculated position from paleomagnetic data that is interpreted to correspond to the ancient position of Earth's spin axis (the north pole) through application of the geocentric axial dipole hypothesis.
+The uncertainty on the pole position is given as a circle with a radius of a given angle ($A_{95}$).
+
 \noindent\textbf{Paleoproterozoic } A geologic era spanning from 2,500 to 1,600 million years ago.
@@
 \noindent\textbf{Penokean orogen } An orogen resulting from Paleoproterozoic accretion of an oceanic arc and the Marshfield terrane continental block along the southern margin of the Superior province.
+
+\noindent\textbf{Phanerozoic } A geologic eon spanning from 541 million years ago to the present day.
+
+\noindent\textbf{Picuris orogen } An orogen resulting from a Mesoproterozoic orogeny interpreted from metamorphic rocks with Mesoproterozoic-aged protoliths in northern New Mexico, USA.
 
@@
 
+\noindent\textbf{Shawinigan orogen } The orogen resulting from the Mesoproterozoic Shawinigan orogeny when there was accretion of terranes to eastern Laurentia.
+
 \noindent\textbf{Snowbird orogen } An orogen resulting from Paleoproterozoic collision between the Rae and Hearne provinces prior to the Trans-Hudson orogeny.
@@
 \noindent\textbf{Thelon orogen } An orogen resulting from the Paleoproterozoic collision orogeny between the Slave and Rae provinces.
+
+\noindent\textbf{thermal remanent magnetization } Magnetization acquired by magnetic minerals rocks as they cool following crystallization from magma.
 

### 2020-09-17 Manuscript/pole_table_edited.tex
uniform spelling of dike

@@
+%make dike throughout for consistency
+
 \begin{longtable}{p{1 in}p{1 in}rrrrrrrp{1.5 in}}
@@
              Laurentia-Wyoming &                          Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705$^{+4}_{-4}$ &                                 \cite{Selkin2008a} \\ \hline
-      Laurentia-Superior(East) &                     Otto Stock dykes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676$^{+5}_{-5}$ &                               \cite{Pullaiah1975b} \\ \hline
+      Laurentia-Superior(East) &                     Otto Stock dikes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676$^{+5}_{-5}$ &                               \cite{Pullaiah1975b} \\ \hline
                Laurentia-Slave &                                     Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625$^{+5}_{-5}$ &                               \cite{Mitchell2014a} \\ \hline
-      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dykes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(East) &                               Matachewan dykes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466$^{+23}_{-23}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(East) &                               Matachewan dykes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
-               Laurentia-Slave &                                     Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231$^{+2}_{-2}$ &                                 \cite{Buchan2012a} \\ \hline
-      Laurentia-Superior(East) &                                 Senneterre dykes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218$^{+6}_{-6}$ &                                 \cite{Buchan1993a} \\ \hline
+      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dikes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                               Matachewan dikes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466$^{+23}_{-23}$ &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                               Matachewan dikes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                     Malley dikes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231$^{+2}_{-2}$ &                                 \cite{Buchan2012a} \\ \hline
+      Laurentia-Superior(East) &                                 Senneterre dikes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218$^{+6}_{-6}$ &                                 \cite{Buchan1993a} \\ \hline
       Laurentia-Superior(East) &                               Nipissing N1 sills &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217$^{+4}_{-4}$ &                                 \cite{Buchan2000a} \\ \hline
-               Laurentia-Slave &                                     Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193$^{+2}_{-2}$ &                               \cite{Mitchell2014a} \\ \hline
-      Laurentia-Superior(East) &                                Biscotasing dykes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
-             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160$^{+11}_{-8}$ &                                 \cite{Kilian2015a} \\ \hline
-               Laurentia-Slave &                                      Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126$^{+3}_{-18}$ &                                 \cite{Buchan2016a} \\ \hline
-      Laurentia-Superior(West) &                                 Marathon dykes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
-      Laurentia-Superior(West) &                                 Marathon dykes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
-      Laurentia-Superior(West) &                               Cauchon Lake dykes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(West) &                               Fort Frances dykes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077$^{+5}_{-5}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(East) &                                 Lac Esprit dykes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069$^{+1}_{-1}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Greenland-Nain &                                  Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042$^{+12}_{-12}$ &                                 \cite{Fahrig1976b} \\ \hline
-               Laurentia-Slave &                                Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026$^{+5}_{-5}$ &                                 \cite{Buchan2009a} \\ \hline
-      Laurentia-Superior(East) &                                      Minto dykes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                     Dogrib dikes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193$^{+2}_{-2}$ &                               \cite{Mitchell2014a} \\ \hline
+      Laurentia-Superior(East) &                                Biscotasing dikes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
+             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path dikes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160$^{+11}_{-8}$ &                                 \cite{Kilian2015a} \\ \hline
+               Laurentia-Slave &                                      Indin dikes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126$^{+3}_{-18}$ &                                 \cite{Buchan2016a} \\ \hline
+      Laurentia-Superior(West) &                                 Marathon dikes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
+      Laurentia-Superior(West) &                                 Marathon dikes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
+      Laurentia-Superior(West) &                               Cauchon Lake dikes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(West) &                               Fort Frances dikes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077$^{+5}_{-5}$ &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                 Lac Esprit dikes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069$^{+1}_{-1}$ &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Greenland-Nain &                                  Kangamiut dikes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042$^{+12}_{-12}$ &                                 \cite{Fahrig1976b} \\ \hline
+               Laurentia-Slave &                                Lac de Gras dikes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026$^{+5}_{-5}$ &                                 \cite{Buchan2009a} \\ \hline
+      Laurentia-Superior(East) &                                      Minto dikes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
                Laurentia-Slave &                                  Rifle Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963$^{+6}_{-6}$ &                                  \cite{Evans1981a} \\ \hline
@@
                Laurentia-Slave &                   Mean Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882$^{+4}_{-4}$ &                               \cite{Mitchell2010c} \\ \hline
-      Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879$^{+6}_{-6}$ &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(West) &                              Molson (B+C2) dikes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879$^{+6}_{-6}$ &                                  \cite{Evans2010a} \\ \hline
                Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876$^{+10}_{-10}$ &                                \cite{Irving1979a} \\ \hline
@@
  Laurentia-Trans-Hudson orogen &                              Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838$^{+1}_{-1}$ &                                 \cite{Symons1999a} \\ \hline
-                 Laurentia-Rae &                                    Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827$^{+4}_{-4}$ &                                \cite{McGlynn1974a} \\ \hline
+                 Laurentia-Rae &                                    Sparrow dikes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827$^{+4}_{-4}$ &                                \cite{McGlynn1974a} \\ \hline
                  Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818$^{+4}_{-4}$ &                                  \cite{Evans1973a} \\ \hline
@@
  Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758$^{+1}_{-1}$ &                                   \cite{Gala1995a} \\ \hline
-                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741$^{+5}_{-5}$ &                                  \cite{Irving2004a} \\ \hline
-           Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633$^{+5}_{-5}$ &                                  \cite{Halls2011a} \\ \hline
+                     Laurentia &                                    Cleaver dikes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741$^{+5}_{-5}$ &                                  \cite{Irving2004a} \\ \hline
+           Laurentia-Greenland &                      Melville Bugt diabase dikes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633$^{+5}_{-5}$ &                                  \cite{Halls2011a} \\ \hline
                      Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590$^{+3}_{-3}$ &                                 \cite{Irving1972a} \\ \hline
@@
                      Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450$^{+14}_{-14}$ &                                 \cite{Elston2002a} \\ \hline
-                     Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448$^{+49}_{-49}$ &                                 \cite{Harlan2008a} \\ \hline
+                     Laurentia &                               Tobacco Root dikes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448$^{+49}_{-49}$ &                                 \cite{Harlan2008a} \\ \hline
                      Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443$^{+7}_{-7}$ &                                 \cite{Elston2002a} \\ \hline
@@
            Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382$^{+2}_{-2}$ &                              \cite{Marcussen1983a} \\ \hline
-           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382$^{+2}_{-2}$ &                             \cite{Abrahamsen1987a} \\ \hline
+           Laurentia-Greenland &                    Victoria Fjord dolerite dikes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382$^{+2}_{-2}$ &                             \cite{Abrahamsen1987a} \\ \hline
                      Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305$^{+15}_{-15}$ &                                \cite{Murthy1978a} \\ \hline
            Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275$^{+1}_{-1}$ &                                  \cite{Piper1992a} \\ \hline
-           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275$^{+2}_{-2}$ &                                  \cite{Piper1977b} \\ \hline
-                     Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267$^{+2}_{-2}$ &                                 \cite{Buchan2000a} \\ \hline
-           Laurentia-Greenland &                       West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244$^{+8}_{-8}$ &                                  \cite{Piper1977b} \\ \hline
-           Laurentia-Greenland &                    West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238$^{+11}_{-11}$ &                                  \cite{Piper1977b} \\ \hline
-                     Laurentia &                           Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237$^{+5}_{-5}$ &                                 \cite{Palmer1977a} \\ \hline
+           Laurentia-Greenland &                                Kungnat Ring dike &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275$^{+2}_{-2}$ &                                  \cite{Piper1977b} \\ \hline
+                     Laurentia &                       Mackenzie dikes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267$^{+2}_{-2}$ &                                 \cite{Buchan2000a} \\ \hline
+           Laurentia-Greenland &                       West Gardar Dolerite dikes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244$^{+8}_{-8}$ &                                  \cite{Piper1977b} \\ \hline
+           Laurentia-Greenland &                    West Gardar Lamprophyre dikes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238$^{+11}_{-11}$ &                                  \cite{Piper1977b} \\ \hline
+                     Laurentia &                           Sudbury dikes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237$^{+5}_{-5}$ &                                 \cite{Palmer1977a} \\ \hline
             Laurentia-Scotland &                                      Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199$^{+70}_{-70}$ &                        Nordic workshop calculation \\ \hline
            Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
-           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                               Hviddal Giant dike &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
            Laurentia-Greenland &                                South Qoroq Intr.
 &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1992a} \\ \hline
-           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1977a} \\ \hline
-           Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160$^{+5}_{-5}$ &                                  \cite{Piper1992a} \\ \hline
-                     Laurentia &                        Ontario lamprophyre dykes &      NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143$^{+10}_{-10}$ &                                 \cite{Piispa2018a} \\ \hline
-                     Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141$^{+2}_{-2}$ &                                  \cite{Ernst1993a} \\ \hline
+           Laurentia-Greenland &                               Giant Gabbro dikes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                             NE-SW Trending dikes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160$^{+5}_{-5}$ &                                  \cite{Piper1992a} \\ \hline
+                     Laurentia &                        Ontario lamprophyre dikes &      NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143$^{+10}_{-10}$ &                                 \cite{Piispa2018a} \\ \hline
+                     Laurentia &                                    Abitibi dikes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141$^{+2}_{-2}$ &                                  \cite{Ernst1993a} \\ \hline
                      Laurentia &                          Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109$^{+2}_{-2}$ &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
@@
             Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800$^{+11}_{-11}$ &                                 \cite{Maloof2006a} \\ \hline
-            Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778$^{+2}_{-2}$ &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a}  \\ \hline
+            Laurentia &                                  Gunbarrel dikes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778$^{+2}_{-2}$ &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a}  \\ \hline
    Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    770$^{+19}_{-40}$ &                                 \cite{Maloof2006a} \\ \hline
@@
             Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
-                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline
+                     Laurentia &                                 Long Range dikes &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline
                      Laurentia &                         Baie des Moutons complex (A) &      B &     301.0 &      50.8 & 152.7 & -42.6 &      12.0 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline

### 2020-09-17 Manuscript/pole_table.tex
updated pole table with Duluth paleolatitude column

--- 2019-10-22 Manuscript/pole_table.tex
@@
 \toprule
-                       terrane &                                        unit name & rating &  site lon &  site lat &  plon &  plat &  A\$\_\{95\}\$ &                    age &                                     pole reference \\
+                       terrane &                                          unit name &                    age & rating &  site lon &  site lat &  plon &  plat &  A\$\_\{95\}\$ &  Duluth lat &                                     pole reference \\
 \midrule
@@
 \midrule
-\multicolumn{10}{r}{{Continued on next page}} \\
+\multicolumn{11}{r}{{Continued on next page}} \\
 \midrule
@@
 \endlastfoot
-             Laurentia-Wyoming &                          Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Selkin2008a\} \\
-      Laurentia-Superior(East) &                     Otto Stock dykes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Pullaiah1975b\} \\
-               Laurentia-Slave &                                     Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
-      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dykes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(East) &                               Matachewan dykes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(East) &                               Matachewan dykes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-               Laurentia-Slave &                                     Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2012a\} \\
-      Laurentia-Superior(East) &                                 Senneterre dykes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Buchan1993a\} \\
-      Laurentia-Superior(East) &                               Nipissing N1 sills &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
-               Laurentia-Slave &                                     Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Mitchell2014a\} \\
-      Laurentia-Superior(East) &                                Biscotasing dykes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Kilian2015a\} \\
-               Laurentia-Slave &                                      Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
-      Laurentia-Superior(West) &                                 Marathon dykes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
-      Laurentia-Superior(West) &                                 Marathon dykes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                  \textbackslash cite\{Halls2008a\} \\
-      Laurentia-Superior(West) &                               Cauchon Lake dykes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(West) &                               Fort Frances dykes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Superior(East) &                                 Lac Esprit dykes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-      Laurentia-Greenland-Nain &                                  Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &                                 \textbackslash cite\{Fahrig1976b\} \\
-               Laurentia-Slave &                                Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Buchan2009a\} \\
-      Laurentia-Superior(East) &                                      Minto dykes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-               Laurentia-Slave &                                  Rifle Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans1981a\} \\
-                 Laurentia-Rae &                           Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                  \textbackslash cite\{Halls1999a\} \\
-             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Kilian2016b\} \\
-               Laurentia-Slave &                                 Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &                                 \textbackslash cite\{Buchan2016a\} \\
-               Laurentia-Slave &                         Mean Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
-               Laurentia-Slave &                   Mean Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
-      Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                  \textbackslash cite\{Evans2010a\} \\
-               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
-               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Irving1979a\} \\
-            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+             Laurentia-Wyoming &                            Stillwater Complex - C2 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &         nan &                                 \textbackslash cite\{Selkin2008a\} \\
+  Baltica-Fennoscandia-Karelia &                                Koitere sanukitoids &     2684\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      30.6 &      63.3 & 192.5 & -67.5 &      19.5 &         nan &                                                NaN \\
+      Laurentia-Superior(East) &                       Otto Stock dikes and aureole &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &         nan &                               \textbackslash cite\{Pullaiah1975b\} \\
+               Laurentia-Slave &                                       Defeat Suite &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &         nan &                               \textbackslash cite\{Mitchell2014a\} \\
+          Baltica-Fennoscandia &                   Shalskiy thick gabbronorite dike &     2511\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      36.0 &      61.9 & 222.1 &  22.7 &      11.5 &         nan &                                                NaN \\
+      Laurentia-Superior(East) &                         Ptarmigan-Mistassini dikes &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
+          Baltica-Fennoscandia &                     Generalskaya Layered Intrusion &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      31.0 &      69.4 & 292.7 & -42.5 &      10.4 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                              Monchegorsk Intrusion &     2504\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      33.0 &      68.0 & 265.3 &   1.3 &       9.9 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &  Avdeev gabbronorite and thin Shalskiy diabase ...
+&   2476\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &      B &      36.0 &      61.9 & 243.5 & -12.3 &      14.0 &         nan &                                                NaN \\
+      Laurentia-Superior(East) &                                 Matachewan dikes R &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
+          Baltica-Fennoscandia &                          Imandra Layered Intrusion &   2446\$\textasciicircum \{+39\}\$\$\_\{-39\}\$ &      B &      33.1 &      67.6 & 280.3 & -16.1 &       7.8 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                                     Karelian dikes &     2446\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &      30.7 &      66.0 & 278.7 & -19.9 &       6.1 &         nan &                                                NaN \\
+      Laurentia-Superior(East) &                                 Matachewan dikes N &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
+          Baltica-Fennoscandia &                            Taivalkoski 2.3 Ga dike &   2339\$\textasciicircum \{+18\}\$\$\_\{-18\}\$ &     C+ &      28.1 &      65.5 & 257.3 &  20.4 &       6.3 &         nan &                                                NaN \\
+               Laurentia-Slave &                                       Malley dikes &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &         nan &                                 \textbackslash cite\{Buchan2012a\} \\
+      Laurentia-Superior(East) &                                   Senneterre dikes &     2218\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &         nan &                                 \textbackslash cite\{Buchan1993a\} \\
+      Laurentia-Superior(East) &                                 Nipissing N1 sills &     2217\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &         nan &                                 \textbackslash cite\{Buchan2000a\} \\
+               Laurentia-Slave &                                       Dogrib dikes &     2193\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &         nan &                               \textbackslash cite\{Mitchell2014a\} \\
+      Laurentia-Superior(East) &                                  Biscotasing dikes &     2170\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
+             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path dikes &    2160\$\textasciicircum \{+11\}\$\$\_\{-8\}\$ &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &         nan &                                 \textbackslash cite\{Kilian2015a\} \\
+               Laurentia-Slave &                                        Indin dikes &    2126\$\textasciicircum \{+3\}\$\$\_\{-18\}\$ &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &         nan &                                 \textbackslash cite\{Buchan2016a\} \\
+      Laurentia-Superior(West) &                                   Marathon dikes N &     2124\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &        43.3 &                                  \textbackslash cite\{Halls2008a\} \\
+      Laurentia-Superior(West) &                                   Marathon dikes R &     2104\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &        38.8 &                                  \textbackslash cite\{Halls2008a\} \\
+      Laurentia-Superior(West) &                                 Cauchon Lake dikes &     2091\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &        37.5 &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(West) &                                 Fort Frances dikes &     2077\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &        33.6 &                                  \textbackslash cite\{Evans2010a\} \\
+      Laurentia-Superior(East) &                                   Lac Esprit dikes &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
+              Baltica-Sarmatia &               Volyn-Dniestr-Bug intrusions A-group &   2061\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &      B &      29.8 &      49.6 & 182.9 &  15.7 &      13.7 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                               Kuetsyarvi Formation &     2060\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &      B &      29.5 &      69.5 & 300.8 &  24.7 &      16.4 &         nan &                                                NaN \\
+      Laurentia-Greenland-Nain &                                    Kangamiut dikes &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &        57.6 &                                 \textbackslash cite\{Fahrig1976b\} \\
+               Laurentia-Slave &                                  Lac de Gras dikes &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &         nan &                                 \textbackslash cite\{Buchan2009a\} \\
+              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &   2000\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      27.4 &      50.9 & 140.4 &  64.4 &      12.7 &         nan &                                                NaN \\
+      Laurentia-Superior(East) &                                        Minto dikes &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
+             Baltica - Karelia &                           Onego 1975 Ma Intrusions &     1976\$\textasciicircum \{+9\}\$\$\_\{-9\}\$ &      A &      34.0 &      62.1 & 101.5 &  44.4 &       6.3 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                                    Konchozero Sill &   1974\$\textasciicircum \{+27\}\$\$\_\{-27\}\$ &      B &      34.0 &      62.1 & 281.9 & -14.2 &      10.4 &         nan &                                                NaN \\
+               Laurentia-Slave &                                    Rifle Formation &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &         nan &                                  \textbackslash cite\{Evans1981a\} \\
+          Baltica-Fennoscandia &             Tsuomasavarri Gabbro-Diorite Intrusion &     1931\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      28.3 &      69.9 & 247.3 &  40.2 &       6.0 &         nan &                                                NaN \\
+                 Laurentia-Rae &                             Clearwater Anorthosite &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &         nan &                                  \textbackslash cite\{Halls1999a\} \\
+             Laurentia-Wyoming &                         Sourdough mafic dike swarm &     1899\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &         nan &                                 \textbackslash cite\{Kilian2016b\} \\
+               Laurentia-Slave &                                   Ghost Dike Swarm &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &         nan &                                 \textbackslash cite\{Buchan2016a\} \\
+          Baltica-Fennoscandia &                         Kiuruvesi-Pielavesi gabbro &     1886\$\textasciicircum \{+6\}\$\$\_\{-5\}\$ &      B &      26.8 &      63.2 & 231.3 &  41.1 &       5.4 &         nan &                                                NaN \\
+               Laurentia-Slave &                           Mean Seton/Akaitcho/Mara &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &         nan &                               \textbackslash cite\{Mitchell2010c\} \\
+               Laurentia-Slave &                     Mean Kahochella, Peacock Hills &     1882\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &         nan &                               \textbackslash cite\{Mitchell2010c\} \\
+      Laurentia-Superior(West) &                                Molson (B+C2) dikes &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &        47.6 &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &                                 Takiyuak Formation &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &         nan &                                 \textbackslash cite\{Irving1979a\} \\
+               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &         nan &                                 \textbackslash cite\{Irving1979a\} \\
+               Laurentia-Slave &               Pearson A/Peninsular/Kilohigok sills &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &         nan &                               \textbackslash cite\{Mitchell2010c\} \\
+            Laurentia-Superior &                         Haig/Flaherty/Sutton Mean  &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &         nan &  Nordic workshop calculation based on data of \textbackslash ...
 \\
-               Laurentia-Slave &             Pearson A/Peninsular/Kilohigok sills &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                               \textbackslash cite\{Mitchell2010c\} \\
- Laurentia-Trans-Hudson orogen &                              Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                 \textbackslash cite\{Symons1999a\} \\
-                 Laurentia-Rae &                                    Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{McGlynn1974a\} \\
-                 Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                  \textbackslash cite\{Evans1973a\} \\
-                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &                                   \textbackslash cite\{Park1973a\} \\
- Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons2000a\} \\
- Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                   \textbackslash cite\{Gala1995a\} \\
-                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Irving2004a\} \\
-           Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Halls2011a\} \\
-                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                                 \textbackslash cite\{Irving1972a\} \\
-                     Laurentia &               St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &                                  \textbackslash cite\{Meert2002b\} \\
-                     Laurentia &                             Michikamau Intrusion &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Emslie1976a\} \\
-                     Laurentia &                                Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
-                     Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
-                     Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &                                 \textbackslash cite\{Harlan2008a\} \\
-                     Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
-                     Laurentia &                        Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+        Baltica - Fennoscandia &                                      Keuruu dikes  &   1869\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &     B  &      24.7 &      62.3 & 230.9 &  45.7 &       5.5 &         nan &                                                NaN \\
+ Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &        73.8 &                                 \textbackslash cite\{Symons1999a\} \\
+                 Laurentia-Rae &                                      Sparrow dikes &     1827\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &         nan &                                \textbackslash cite\{McGlynn1974a\} \\
+                 Laurentia-Rae &                                   Martin Formation &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &         nan &                                  \textbackslash cite\{Evans1973a\} \\
+          Baltica-Fennoscandia &                                    Notträsk gabbro &     1806\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      B &      21.8 &      65.8 & 216.2 &  43.5 &       6.1 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                                      Kallax gabbro &     1800\$\textasciicircum \{+5\}\$\$\_\{-6\}\$ &      B &      22.0 &      65.5 & 209.0 &  49.0 &       nan &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                 Lake Ladoga, Mean intr.
+\& dikes, A &   1800\$\textasciicircum \{+19\}\$\$\_\{-56\}\$ &      B &      30.0 &      61.3 & 229.1 &  50.9 &       nan &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                                      Hoting gabbro &   1786\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      A &      16.2 &      64.2 & 233.3 &  43.0 &      10.9 &         nan &                                                NaN \\
+                     Laurentia &                                      Dubawnt Group &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &        49.4 &                                   \textbackslash cite\{Park1973a\} \\
+          Baltica-Fennoscandia &                              Småland intrusives -C &     1777\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &      A &      15.7 &      57.1 & 182.7 &  45.7 &       8.0 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                                 Shoksha Sandstones &   1775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &      B &      33.8 &      61.3 & 221.1 &  39.7 &       4.0 &         nan &                                                NaN \\
+ Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &        68.9 &                                 \textbackslash cite\{Symons2000a\} \\
+ Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &        67.3 &                                   \textbackslash cite\{Gala1995a\} \\
+              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &   1755\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      A &      28.6 &      50.6 & 169.1 &  26.5 &       3.9 &         nan &                                                NaN \\
+          Baltica-Fennoscandia &                                MEAN Ropruchey sill &     1751\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      B &      35.3 &      61.4 & 217.0 &  39.1 &       8.6 &         nan &                                                NaN \\
+                     Laurentia &                                      Cleaver dikes &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &        61.7 &                                 \textbackslash cite\{Irving2004a\} \\
+              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &   1722\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      B &      29.6 &      50.0 & 163.2 &  10.7 &      10.2 &         nan &                                                NaN \\
+        Baltica - Fennoscandia &                                   Häme dike swarm  &   1647\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &      B &      25.1 &      61.4 & 209.8 &  23.6 &      14.7 &         nan &                                                NaN \\
+                       Baltica &                        Sipoo Quartz Porphyry dikes &   1633\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      25.3 &      60.3 & 180.4 &  26.4 &       9.4 &         nan &                                                NaN \\
+           Laurentia-Greenland &                        Melville Bugt diabase dikes &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &        45.5 &                                  \textbackslash cite\{Halls2011a\} \\
+                       Baltica &                              Quartz porphyry dikes &   1631\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      26.8 &      61.3 & 175.4 &  30.2 &       9.4 &         nan &                                                NaN \\
+                       Baltica &                                 Hoting basic dikes &   1614\$\textasciicircum \{+24\}\$\$\_\{-24\}\$ &      B &      16.2 &      64.2 & 146.7 &  21.9 &      13.7 &         nan &                                                NaN \\
+                     Laurentia &                            Western Channel Diabase &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &        47.5 &                                 \textbackslash cite\{Irving1972a\} \\
+                       Baltica &                                   Åland dike swarm &   1578\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      20.0 &      60.0 & 191.4 &  23.7 &       2.8 &         nan &                                                NaN \\
+                       Baltica &                            Satakunta dike swarm -C &   1578\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      21.5 &      62.0 & 188.1 &  29.3 &       6.6 &         nan &                                                NaN \\
+                       Baltica &                                  Ragunda Formation &   1506\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &      B &      16.1 &      63.3 & 166.6 &  51.6 &       7.1 &         nan &                                                NaN \\
+                     Laurentia &                 St.Francois Mountains Acidic Rocks &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &        15.8 &                                  \textbackslash cite\{Meert2002b\} \\
+                       Baltica &                           MEAN Bunkris/Glysjon/Oje &     1469\$\textasciicircum \{+9\}\$\$\_\{-9\}\$ &      B &      13.5 &      61.5 & 179.8 &  28.3 &      13.2 &         nan &                                                NaN \\
+                     Laurentia &                               Michikamau Intrusion &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &        24.7 &                                 \textbackslash cite\{Emslie1976a\} \\
+                     Laurentia &                                  Spokane Formation &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &         4.2 &                                 \textbackslash cite\{Elston2002a\} \\
+                       Baltica &                    Lake Ladoga basalt, sill, dikes &   1452\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      31.2 &      61.5 & 177.1 &  15.2 &       5.5 &         nan &                                                NaN \\
+                     Laurentia &                                 Snowslip Formation &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &         1.4 &                                 \textbackslash cite\{Elston2002a\} \\
+                     Laurentia &                                 Tobacco Root dikes &   1448\$\textasciicircum \{+49\}\$\$\_\{-49\}\$ &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &        31.9 &                                 \textbackslash cite\{Harlan2008a\} \\
+                     Laurentia &                                       Purcell Lava &     1443\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &         5.3 &                                 \textbackslash cite\{Elston2002a\} \\
+                     Laurentia &                          Rocky Mountain intrusions &   1430\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &        16.0 &  Nordic workshop calculation based on data of \textbackslash ...
 \\
-                     Laurentia &                                 Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                 \textbackslash cite\{Fahrig1976a\} \\
-                     Laurentia &                               McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
-                     Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &                                 \textbackslash cite\{Elston2002a\} \\
-           Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
-           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{Abrahamsen1987a\} \\
-           Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                              \textbackslash cite\{Marcussen1983a\} \\
-                     Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                 \textbackslash cite\{Murthy1978a\} \\
-           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
-           Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-                     Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Buchan2000a\} \\
-           Laurentia-Greenland &                       West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
-           Laurentia-Greenland &                    West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                  \textbackslash cite\{Piper1977b\} \\
-                     Laurentia &                           Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Palmer1977a\} \\
-            Laurentia-Scotland &                                      Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &                        Nordic workshop calculation \\
-           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
-           Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
-           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1977a\} \\
+                     Laurentia &                                   Mistastin Pluton &   1425\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &        15.1 &                                 \textbackslash cite\{Fahrig1976a\} \\
+                     Laurentia &                                 McNamara Formation &     1401\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &         9.6 &                                 \textbackslash cite\{Elston2002a\} \\
+                     Laurentia &         Pilcher, Garnet Range and Libby Formations &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &         8.8 &                                 \textbackslash cite\{Elston2002a\} \\
+           Laurentia-Greenland &                                Zig-Zag Dal Basalts &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &        43.8 &                              \textbackslash cite\{Marcussen1983a\} \\
+           Laurentia-Greenland &                      Victoria Fjord dolerite dikes &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &        36.6 &                             \textbackslash cite\{Abrahamsen1987a\} \\
+           Laurentia-Greenland &                              Midsommersoe Dolerite &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &        39.0 &                              \textbackslash cite\{Marcussen1983a\} \\
+                       Baltica &                                             Mashak &   1376\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      57.0 &      54.0 & 193.0 &   1.8 &      14.8 &         nan &                                                NaN \\
+                     Laurentia &                                   Nain Anorthosite &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &        28.1 &                                 \textbackslash cite\{Murthy1978a\} \\
+           Laurentia-Greenland &                             North Qoroq intrusives &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &        21.0 &                                  \textbackslash cite\{Piper1992a\} \\
+           Laurentia-Greenland &                                  Kungnat Ring dike &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &        11.1 &                                  \textbackslash cite\{Piper1977b\} \\
+                     Laurentia &                         Mackenzie dikes grand mean &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &        11.2 &                                 \textbackslash cite\{Buchan2000a\} \\
+                       Baltica &                       MEAN post-Jotnian intrusions &   1258\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      18.1 &      62.4 & 159.1 &  -1.8 &       3.4 &         nan &                                                NaN \\
+           Laurentia-Greenland &                         West Gardar Dolerite dikes &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &        17.1 &                                  \textbackslash cite\{Piper1977b\} \\
+           Laurentia-Greenland &                      West Gardar Lamprophyre dikes &   1238\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &        15.9 &                                  \textbackslash cite\{Piper1977b\} \\
+                     Laurentia &                             Sudbury dikes Combined &     1237\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &         8.3 &                                 \textbackslash cite\{Palmer1977a\} \\
+            Laurentia-Scotland &                                        Stoer Group &   1199\$\textasciicircum \{+70\}\$\$\_\{-70\}\$ &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &        43.9 &                        Nordic workshop calculation \\
+           Laurentia-Greenland &                                 Hviddal Giant dike &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &        43.3 &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                                     Narssaq Gabbro &     1184\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &        48.8 &                                  \textbackslash cite\{Piper1977a\} \\
            Laurentia-Greenland &                                  South Qoroq Intr.
-&      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-           Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Piper1992a\} \\
-                     Laurentia &                        Ontario lamprophyre dykes &      A &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\
-                     Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                  \textbackslash cite\{Ernst1993a\} \\
-                     Laurentia &                          Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+&     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &        48.7 &                                  \textbackslash cite\{Piper1992a\} \\
+           Laurentia-Greenland &                                 Giant Gabbro dikes &     1163\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &        55.5 &                                  \textbackslash cite\{Piper1977a\} \\
+           Laurentia-Greenland &                               NE-SW Trending dikes &     1160\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &        53.5 &                                  \textbackslash cite\{Piper1992a\} \\
+                     Laurentia &                          Ontario lamprophyre dikes &   1143\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &        61.2 &                                 \textbackslash cite\{Piispa2018a\} \\
+                     Laurentia &                                      Abitibi dikes &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \textbackslash cite\{Ernst1993a\} \\
+                       Baltica &                        Salla dike (width 60-100 m) &     1122\$\textasciicircum \{+5\}\$\$\_\{-3\}\$ &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &         nan &                                                NaN \\
+                     Laurentia &                            Nipigon sills and lavas &     1109\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \textbackslash ...
 \\
-                     Laurentia &           Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1109\$\textasciicircum \{+2\}\$\$\_\{-3\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
-                     Laurentia &                         Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
-                     Laurentia &                        Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
-                     Laurentia &                         Upper Osler volcanics -R &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &     1105\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  \textbackslash cite\{Halls1974a, Swanson-Hysell2014b, Swanson...
+                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &     1109\$\textasciicircum \{+2\}\$\$\_\{-3\}\$ &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &        62.9 &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                           Lower Osler volcanics -R &     1108\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &        54.6 &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
+                     Laurentia &                          Middle Osler volcanics -R &     1107\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &        50.5 &                         \textbackslash cite\{Swanson-Hysell2014b\} \\
+                     Laurentia &                           Upper Osler volcanics -R &     1105\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &        45.1 &  \textbackslash cite\{Halls1974a, Swanson-Hysell2014b, Swanson...
 \\
-                     Laurentia &               Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105\$\textasciicircum \{+3\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
-                     Laurentia &   Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
-                     Laurentia &                             North Shore lavas -N &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &              \textbackslash cite\{Tauxe2009a,Swanson-Hysell2019a\} \\
-                     Laurentia &                           Portage Lake Volcanics &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &  \textbackslash cite\{Books1972a, Hnat2006a\} as calculated in ...
+                     Laurentia &                 Lower Mamainse Point volcanics -R2 &     1105\$\textasciicircum \{+3\}\$\$\_\{-4\}\$ &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &        43.9 &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &     1101\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &        32.9 &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                               North Shore lavas -N &     1097\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &        24.5 &              \textbackslash cite\{Tauxe2009a,Swanson-Hysell2019a\} \\
+                     Laurentia &                              Chengwatana Volcanics &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &        27.3 &                                   \textbackslash cite\{Kean1997a\} \\
+                     Laurentia &                             Portage Lake Volcanics &     1095\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &        22.7 &  \textbackslash cite\{Books1972a, Hnat2006a\} as calculated in ...
 \\
-                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Kean1997a\} \\
-                     Laurentia &            Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094\$\textasciicircum \{+6\}\$\$\_\{-4\}\$ &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
-                     Laurentia &                  Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                   \textbackslash cite\{Weil2003a\} \\
-                     Laurentia &                         Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
-                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                               \textbackslash cite\{Donadini2011b\} \\
-                     Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                                \textbackslash cite\{Kulakov2013a\} \\
-                     Laurentia &                    Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &                              \textbackslash cite\{Fairchild2017a\} \\
-                     Laurentia &                                   Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &    1080\$\textasciicircum \{+4\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
-                     Laurentia &                                  Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1070\$\textasciicircum \{+14\}\$\$\_\{-10\}\$ &                                  \textbackslash cite\{Henry1977a\} \\
-                     Laurentia &                            Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &                                \textbackslash cite\{Warnock2000a\} \\
-            Laurentia-Scotland &                                   Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &                        Nordic workshop calculation \\
-            Laurentia-Svalbard &                    Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
-            Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
-                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
-                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ...
+                     Laurentia &              Uppermost Mamainse Point volcanics -N &     1094\$\textasciicircum \{+6\}\$\$\_\{-4\}\$ &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &        25.6 &                         \textbackslash cite\{Swanson-Hysell2014a\} \\
+                     Laurentia &                    Cardenas Basalts and Intrusions &     1091\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &        27.3 &                                   \textbackslash cite\{Weil2003a\} \\
+                     Laurentia &                           Schroeder Lutsen Basalts &     1090\$\textasciicircum \{+2\}\$\$\_\{-7\}\$ &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &        25.9 &                              \textbackslash cite\{Fairchild2017a\} \\
+                     Laurentia &                        Central Arizona diabases -N &   1088\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &         9.6 &                               \textbackslash cite\{Donadini2011b\} \\
+                     Laurentia &                                   Lake Shore Traps &     1086\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &        22.3 &                                \textbackslash cite\{Kulakov2013a\} \\
+                     Laurentia &                      Michipicoten Island Formation &     1084\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &        10.2 &                              \textbackslash cite\{Fairchild2017a\} \\
+                     Laurentia &                                     Nonesuch Shale &    1080\$\textasciicircum \{+4\}\$\$\_\{-10\}\$ &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &         5.7 &                                  \textbackslash cite\{Henry1977a\} \\
+                     Laurentia &                                    Freda Sandstone &   1070\$\textasciicircum \{+14\}\$\$\_\{-10\}\$ &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &         2.4 &                                  \textbackslash cite\{Henry1977a\} \\
+                       Baltica &                             MEAN Bamble Intrusions &   1048\$\textasciicircum \{+73\}\$\$\_\{-73\}\$ &     C+ &       9.5 &      58.9 & 222.5 & -15.1 &      19.3 &         nan &                                                NaN \\
+                       Baltica &                           Laanila-Ristijarvi dikes &   1044\$\textasciicircum \{+52\}\$\$\_\{-52\}\$ &      B &      28.1 &      68.7 & 212.2 &  -2.1 &      13.8 &         nan &                                                NaN \\
+                     Laurentia &                              Haliburton Intrusions &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &       -47.0 &                                \textbackslash cite\{Warnock2000a\} \\
+                     Laurentia &                        Adirondack fayalite granite &    990\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &     NR &     285.5 &      44.0 & 132.7 & -28.4 &       6.9 &       -50.7 &                                  \textbackslash cite\{Brown2012a\} \\
+                     Laurentia &               Adirondack metamorphic\textbackslash nanorthosites &    970\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &     NR &     286.0 &      44.0 & 149.0 & -25.1 &      11.6 &       -37.5 &                                  \textbackslash cite\{Brown2012a\} \\
+                     Laurentia &                       Adirondack Microcline gneiss &    960\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &       -31.5 &                                  \textbackslash cite\{Brown2012a\} \\
+                       Baltica &                            MEAN Blekinga dolerites &      950\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &      15.0 &      56.1 & 247.0 &  13.0 &      16.0 &         nan &                                                NaN \\
+                       Baltica &               dikes of central and southern Sweden &    945\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      A &      14.2 &      58.4 & 240.7 &  -0.9 &       6.7 &         nan &                                                NaN \\
+            Laurentia-Scotland &                                     Torridon Group &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &        -8.6 &                        Nordic workshop calculation \\
+        Baltica - Fennoscandia &                Bjerkreim-Sokndal layered intrusion &    921\$\textasciicircum \{+17\}\$\$\_\{-17\}\$ &      B &       6.1 &      58.5 & 217.9 & -35.9 &       6.3 &         nan &                                                NaN \\
+                       Baltica &                    Bratton and Algon Igneous Rocks &    916\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &      B &      11.7 &      57.9 & 249.0 &   5.0 &       3.9 &         nan &                                                NaN \\
+                       Baltica &                           MEAN Rogaland igneous cx &    903\$\textasciicircum \{+33\}\$\$\_\{-33\}\$ &      B &       6.9 &      58.3 & 207.9 & -43.2 &      10.1 &         nan &                                                NaN \\
+                       Baltica &                                   Hunnedalen dikes &    848\$\textasciicircum \{+27\}\$\$\_\{-27\}\$ &      B &       7.0 &      58.9 & 222.0 & -41.0 &      15.6 &         nan &                                                NaN \\
+            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &        -5.3 &                                 \textbackslash cite\{Maloof2006a\} \\
+            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &        11.5 &                                 \textbackslash cite\{Maloof2006a\} \\
+                       Baltica &                                    Katav Formation &  800\$\textasciicircum \{+100\}\$\$\_\{-100\}\$ &      B &      57.4 &      54.8 & 169.9 &  35.7 &      11.4 &         nan &                                                NaN \\
+                     Laurentia &                                    Gunbarrel dikes &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &       -18.4 &  Calculation from \textbackslash cite\{Eyster2019a\} based on d...
 \\
-                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006b\} \\
-            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
-                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                               \textbackslash cite\{Denyszyn2009a\} \\
-                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                 \textbackslash cite\{Murthy1992a\} \\
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                             \textbackslash cite\{McCausland2011a\} \\
-                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                 \textbackslash cite\{Symons1991a\} \\
-                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &                                  \textbackslash cite\{Meert1994a\} \\
-                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &                                \textbackslash cite\{Tanczyk1987a\} \\
+            Laurentia-Svalbard &                          Svanbergfjellet Formation &    770\$\textasciicircum \{+19\}\$\$\_\{-40\}\$ &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &        12.8 &                                 \textbackslash cite\{Maloof2006a\} \\
+                     Laurentia &                               Uinta Mountain Group &     760\$\textasciicircum \{+6\}\$\$\_\{-10\}\$ &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &       -10.7 &                                   \textbackslash cite\{Weil2006b\} \\
+                     Laurentia &                                      Carbon Canyon &      757\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &        -8.5 &  \textbackslash cite\{Weil2004a\} as calculated in \textbackslash cite\{Eyster...
+\\
+                     Laurentia &                               Carbon Butte/Awatubi &      751\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &         1.0 &                                 \textbackslash cite\{Eyster2019a\} \\
+                     Laurentia &                          Franklin event grand mean &      718\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &        -5.7 &                               \textbackslash cite\{Denyszyn2009b\} \\
+                       Baltica &                                     Egersund dikes &      616\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &       5.9 &      58.8 &  44.1 &  31.4 &      15.6 &         nan &                                                NaN \\
+                     Laurentia &                                   Long Range dikes &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &       -15.5 &                                 \textbackslash cite\{Murthy1992a\} \\
+                     Laurentia &                           Baie des Moutons complex &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     301.0 &      50.8 & 152.7 & -42.6 &      12.0 &       -45.1 &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                           Baie des Moutons complex &      583\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     301.0 &      50.8 & 141.5 &  34.2 &      15.4 &         4.2 &                             \textbackslash cite\{McCausland2011a\} \\
+                     Laurentia &                         Callander Alkaline Complex &      575\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     280.6 &      46.2 & 121.4 & -46.3 &       6.0 &       -67.1 &                                 \textbackslash cite\{Symons1991a\} \\
+                     Laurentia &                                   Catoctin Basalts &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &       -69.0 &                                  \textbackslash cite\{Meert1994a\} \\
+                       Baltica &                                       Volhyn lavas &    571\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      26.0 &      51.2 &   4.4 &  19.8 &      28.3 &         nan &                                                NaN \\
+                     Laurentia &                        Sept-Iles layered intrusion &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &        -7.9 &                                \textbackslash cite\{Tanczyk1987a\} \\
+                       Baltica &                               Kurgashlya formation &      565\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &      57.5 &      53.3 & 314.5 &  50.9 &       5.3 &         nan &                                                NaN \\
+                       Baltica &                                  Bakeevo formation &      565\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &      58.2 &      54.9 & 299.1 &  42.3 &       5.3 &         nan &                                                NaN \\
+                       Baltica &                        Basu-Kukkarauk formation -C &      562\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &      57.0 &      54.0 & 187.3 &   1.1 &       5.8 &         nan &                                                NaN \\
+                 Baltica-Urals &                          Basu formation  (precise) &    560\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      56.9 &      53.9 & 186.1 &   1.7 &       3.8 &         nan &                                                NaN \\
+                       Baltica &                 Chernokamenskay group sediments -C &    557\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &      B &      58.4 &      58.5 & 306.7 &  17.3 &       6.0 &         nan &                                                NaN \\
+                       Baltica &                             MEAN Vendian sediments &      556\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &      40.0 &      65.5 & 293.0 &  31.3 &       9.9 &         nan &                                                NaN \\
+                 Baltica-Urals &                                Zigan HTC thin beds &   548\$\textasciicircum \{+27\}\$\$\_\{--12\}\$ &      B &      56.7 &      54.4 & 138.3 & -18.0 &      10.7 &         nan &                                                NaN \\
+                       Baltica &                   Zigan formation clastic rocks -C &      548\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &      56.7 &      53.7 & 138.4 & -16.2 &       4.1 &         nan &                                                NaN \\
 \end{longtable}

### 2020-09-17 Manuscript/pole_table_edited.tex
updated pole table with Duluth paleolatitude column

--- 2019-10-22 Manuscript/pole_table_edited.tex
@@
-\begin{longtable}{p{1 in}p{1 in}rrrrrrrp{1.5 in}}
+\begin{ThreePartTable}
+\begin{TableNotes}
+\footnotesize
+site lon -- longitude of paleomagnetic locality;
+site lat -- latitude of paleomagnetic locality;
+plon -- longitude of the paleomagnetic pole position;
+plat -- latitude of the paleomagnetic pole position;
+A$_{95}$ -- angle of 95$\%$ confidence on the pole position;
+Duluth lat -- latitude of Duluth, MN implied by the paleomagnetic pole
+\end{TableNotes}
+
+\begin{longtable}{p{1.4 in}p{1.2 in}rrrrrrrrp{1.2 in}}
 \toprule
-                       terrane &                                        unit name & Nordic rating &  site lon (\textdegree) &  site lat (\textdegree) &  plon (\textdegree) &  plat (\textdegree) &  A$_{95}$ (\textdegree) &                    age (Ma)&                                     pole reference \\ \hline
+terrane & unit name & age (Ma) & rating & site lon & site lat & plon & plat & A$_{95}$ & Duluth lat & pole reference \\ \hline
 \midrule
@@
 \midrule
-\multicolumn{10}{r}{{Continued on next page}} \\ \hline
+\multicolumn{11}{r}{{Continued on next page}} \\ \hline
 \midrule
@@
 \bottomrule
+\insertTableNotes
 \endlastfoot
-             Laurentia-Wyoming &                          Stillwater Complex - C2 &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &     2705$^{+4}_{-4}$ &                                 \cite{Selkin2008a} \\ \hline
-      Laurentia-Superior(East) &                     Otto Stock dykes and aureole &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &     2676$^{+5}_{-5}$ &                               \cite{Pullaiah1975b} \\ \hline
-               Laurentia-Slave &                                     Defeat Suite &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &     2625$^{+5}_{-5}$ &                               \cite{Mitchell2014a} \\ \hline
-      Laurentia-Superior(East) &                       Ptarmigan-Mistassini dykes &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &     2505$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(East) &                               Matachewan dykes R &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &   2466$^{+23}_{-23}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(East) &                               Matachewan dykes N &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &     2446$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
-               Laurentia-Slave &                                     Malley dykes &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &     2231$^{+2}_{-2}$ &                                 \cite{Buchan2012a} \\ \hline
-      Laurentia-Superior(East) &                                 Senneterre dykes &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &     2218$^{+6}_{-6}$ &                                 \cite{Buchan1993a} \\ \hline
-      Laurentia-Superior(East) &                               Nipissing N1 sills &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &     2217$^{+4}_{-4}$ &                                 \cite{Buchan2000a} \\ \hline
-               Laurentia-Slave &                                     Dogrib dykes &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &     2193$^{+2}_{-2}$ &                               \cite{Mitchell2014a} \\ \hline
-      Laurentia-Superior(East) &                                Biscotasing dykes &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &     2170$^{+3}_{-3}$ &                                  \cite{Evans2010a} \\ \hline
-             Laurentia-Wyoming &  Rabbit Creek, Powder River and South Path Dykes &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &    2160$^{+11}_{-8}$ &                                 \cite{Kilian2015a} \\ \hline
-               Laurentia-Slave &                                      Indin dykes &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &    2126$^{+3}_{-18}$ &                                 \cite{Buchan2016a} \\ \hline
-      Laurentia-Superior(West) &                                 Marathon dykes N &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &     2124$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
-      Laurentia-Superior(West) &                                 Marathon dykes R &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &     2104$^{+3}_{-3}$ &                                  \cite{Halls2008a} \\ \hline
-      Laurentia-Superior(West) &                               Cauchon Lake dykes &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &     2091$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(West) &                               Fort Frances dykes &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &     2077$^{+5}_{-5}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Superior(East) &                                 Lac Esprit dykes &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &     2069$^{+1}_{-1}$ &                                  \cite{Evans2010a} \\ \hline
-      Laurentia-Greenland-Nain &                                  Kangamiut Dykes &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &   2042$^{+12}_{-12}$ &                                 \cite{Fahrig1976b} \\ \hline
-               Laurentia-Slave &                                Lac de Gras dykes &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &     2026$^{+5}_{-5}$ &                                 \cite{Buchan2009a} \\ \hline
-      Laurentia-Superior(East) &                                      Minto dykes &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &     1998$^{+2}_{-2}$ &                                  \cite{Evans2010a} \\ \hline
-               Laurentia-Slave &                                  Rifle Formation &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &     1963$^{+6}_{-6}$ &                                  \cite{Evans1981a} \\ \hline
-                 Laurentia-Rae &                           Clearwater Anorthosite &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &     1917$^{+7}_{-7}$ &                                  \cite{Halls1999a} \\ \hline
-             Laurentia-Wyoming &                       Sourdough mafic dike swarm &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &     1899$^{+5}_{-5}$ &                                 \cite{Kilian2016b} \\ \hline
-               Laurentia-Slave &                                 Ghost Dike Swarm &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &     1887$^{+5}_{-9}$ &                                 \cite{Buchan2016a} \\ \hline
-               Laurentia-Slave &                         Mean Seton/Akaitcho/Mara &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &     1885$^{+5}_{-5}$ &                               \cite{Mitchell2010c} \\ \hline
-               Laurentia-Slave &                   Mean Kahochella, Peacock Hills &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &     1882$^{+4}_{-4}$ &                               \cite{Mitchell2010c} \\ \hline
-      Laurentia-Superior(West) &                              Molson (B+C2) dykes &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &     1879$^{+6}_{-6}$ &                                  \cite{Evans2010a} \\ \hline
-               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &   1876$^{+10}_{-10}$ &                                \cite{Irving1979a} \\ \hline
-               Laurentia-Slave &                               Takiyuak Formation &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &   1876$^{+10}_{-10}$ &                                \cite{Irving1979a} \\ \hline
-            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870$^{+1}_{-1}$ &  Nordic workshop calculation based on data of \cite{Schmidt1980a, Schwarz1982a} \\ \hline
-               Laurentia-Slave &             Pearson A/Peninsular/Kilohigok sills &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &     1870$^{+4}_{-4}$ &                               \cite{Mitchell2010c} \\ \hline
- Laurentia-Trans-Hudson orogen &                              Boot-Phantom Pluton &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &     1838$^{+1}_{-1}$ &                                 \cite{Symons1999a} \\ \hline
-                 Laurentia-Rae &                                    Sparrow Dykes &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &     1827$^{+4}_{-4}$ &                                \cite{McGlynn1974a} \\ \hline
-                 Laurentia-Rae &                                 Martin Formation &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &     1818$^{+4}_{-4}$ &                                  \cite{Evans1973a} \\ \hline
-                     Laurentia &                                    Dubawnt Group &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &   1785$^{+35}_{-35}$ &                                    \cite{Park1973a} \\ \hline
- Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &     1766$^{+5}_{-5}$ &                                 \cite{Symons2000a} \\ \hline
- Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &     1758$^{+1}_{-1}$ &                                   \cite{Gala1995a} \\ \hline
-                     Laurentia &                                    Cleaver Dykes &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &     1741$^{+5}_{-5}$ &                                  \cite{Irving2004a} \\ \hline
-           Laurentia-Greenland &                      Melville Bugt diabase dykes &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &     1633$^{+5}_{-5}$ &                                  \cite{Halls2011a} \\ \hline
-                     Laurentia &                          Western Channel Diabase &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &     1590$^{+3}_{-3}$ &                                 \cite{Irving1972a} \\ \hline
-                     Laurentia &               St.Francois Mountains Acidic Rocks &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &   1476$^{+16}_{-16}$ &                                  \cite{Meert2002b} \\ \hline
-                     Laurentia &                             Michikamau Intrusion &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &     1460$^{+5}_{-5}$ &                                 \cite{Emslie1976a} \\ \hline
-                     Laurentia &                                Spokane Formation &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &   1458$^{+13}_{-13}$ &                                 \cite{Elston2002a} \\ \hline
-                     Laurentia &                               Snowslip Formation &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &   1450$^{+14}_{-14}$ &                                 \cite{Elston2002a} \\ \hline
-                     Laurentia &                               Tobacco Root dykes &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &   1448$^{+49}_{-49}$ &                                 \cite{Harlan2008a} \\ \hline
-                     Laurentia &                                     Purcell Lava &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &     1443$^{+7}_{-7}$ &                                 \cite{Elston2002a} \\ \hline
-                     Laurentia &                        Rocky Mountain intrusions &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &   1430$^{+15}_{-15}$ &  Nordic workshop calculation based on data of \cite{Harlan1994a,Harlan1998a} \\ \hline
-                     Laurentia &                                 Mistastin Pluton &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &   1425$^{+25}_{-25}$ &                                 \cite{Fahrig1976a} \\ \hline
-                     Laurentia &                               McNamara Formation &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &     1401$^{+6}_{-6}$ &                                 \cite{Elston2002a} \\ \hline
-                     Laurentia &       Pilcher, Garnet Range and Libby Formations &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &   1385$^{+23}_{-23}$ &                                 \cite{Elston2002a} \\ \hline
-           Laurentia-Greenland &                              Zig-Zag Dal Basalts &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &     1382$^{+2}_{-2}$ &                              \cite{Marcussen1983a} \\ \hline
-           Laurentia-Greenland &                            Midsommersoe Dolerite &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &     1382$^{+2}_{-2}$ &                              \cite{Marcussen1983a} \\ \hline
-           Laurentia-Greenland &                    Victoria Fjord dolerite dykes &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &     1382$^{+2}_{-2}$ &                             \cite{Abrahamsen1987a} \\ \hline
-                     Laurentia &                                 Nain Anorthosite &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &   1305$^{+15}_{-15}$ &                                \cite{Murthy1978a} \\ \hline
-           Laurentia-Greenland &                           North Qoroq intrusives &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &     1275$^{+1}_{-1}$ &                                  \cite{Piper1992a} \\ \hline
-           Laurentia-Greenland &                                Kungnat Ring Dyke &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &     1275$^{+2}_{-2}$ &                                  \cite{Piper1977b} \\ \hline
-                     Laurentia &                       Mackenzie dykes grand mean &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &     1267$^{+2}_{-2}$ &                                 \cite{Buchan2000a} \\ \hline
-           Laurentia-Greenland &                       West Gardar Dolerite Dykes &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &     1244$^{+8}_{-8}$ &                                  \cite{Piper1977b} \\ \hline
-           Laurentia-Greenland &                    West Gardar Lamprophyre Dykes &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &   1238$^{+11}_{-11}$ &                                  \cite{Piper1977b} \\ \hline
-                     Laurentia &                           Sudbury Dykes Combined &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &     1237$^{+5}_{-5}$ &                                 \cite{Palmer1977a} \\ \hline
-            Laurentia-Scotland &                                      Stoer Group &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &   1199$^{+70}_{-70}$ &                        Nordic workshop calculation \\ \hline
-           Laurentia-Greenland &                                   Narssaq Gabbro &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
-           Laurentia-Greenland &                               Hviddal Giant Dyke &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &     1184$^{+5}_{-5}$ &                                  \cite{Piper1977a} \\ \hline
+             Laurentia-Wyoming &                            Stillwater Complex - C2 &     2705$^{+4}_{-4}$ &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &          &                                 \cite{Selkin2008a} \\ \hline
+      Laurentia-Superior(East) &                       Otto Stock dikes and aureole &     2676$^{+5}_{-5}$ &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &          &                               \cite{Pullaiah1975b} \\ \hline
+               Laurentia-Slave &                                       Defeat Suite &     2625$^{+5}_{-5}$ &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &          &                               \cite{Mitchell2014a} \\ \hline
+      Laurentia-Superior(East) &                         Ptarmigan-Mistassini dikes &     2505$^{+2}_{-2}$ &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &          &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                 Matachewan dikes R &   2466$^{+23}_{-23}$ &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &          &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                 Matachewan dikes N &     2446$^{+3}_{-3}$ &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &          &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                       Malley dikes &     2231$^{+2}_{-2}$ &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &          &                                 \cite{Buchan2012a} \\ \hline
+      Laurentia-Superior(East) &                                   Senneterre dikes &     2218$^{+6}_{-6}$ &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &          &                                 \cite{Buchan1993a} \\ \hline
+      Laurentia-Superior(East) &                                 Nipissing N1 sills &     2217$^{+4}_{-4}$ &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &          &                                 \cite{Buchan2000a} \\ \hline
+               Laurentia-Slave &                                       Dogrib dikes &     2193$^{+2}_{-2}$ &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &          &                               \cite{Mitchell2014a} \\ \hline
+      Laurentia-Superior(East) &                                  Biscotasing dikes &     2170$^{+3}_{-3}$ &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &          &                                  \cite{Evans2010a} \\ \hline
+             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path dikes &    2160$^{+11}_{-8}$ &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &          &                                 \cite{Kilian2015a} \\ \hline
+               Laurentia-Slave &                                        Indin dikes &    2126$^{+3}_{-18}$ &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &          &                                 \cite{Buchan2016a} \\ \hline
+      Laurentia-Superior(West) &                                   Marathon dikes N &     2124$^{+3}_{-3}$ &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &        43.3 &                                  \cite{Halls2008a} \\ \hline
+      Laurentia-Superior(West) &                                   Marathon dikes R &     2104$^{+3}_{-3}$ &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &        38.8 &                                  \cite{Halls2008a} \\ \hline
+      Laurentia-Superior(West) &                                 Cauchon Lake dikes &     2091$^{+2}_{-2}$ &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &        37.5 &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(West) &                                 Fort Frances dikes &     2077$^{+5}_{-5}$ &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &        33.6 &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                   Lac Esprit dikes &     2069$^{+1}_{-1}$ &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &          &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Greenland-Nain &                                    Kangamiut dikes &   2042$^{+12}_{-12}$ &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &         &                                 \cite{Fahrig1976b} \\ \hline
+               Laurentia-Slave &                                  Lac de Gras dikes &     2026$^{+5}_{-5}$ &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &          &                                 \cite{Buchan2009a} \\ \hline
+      Laurentia-Superior(East) &                                        Minto dikes &     1998$^{+2}_{-2}$ &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &          &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                    Rifle Formation &     1963$^{+6}_{-6}$ &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &          &                                  \cite{Evans1981a} \\ \hline
+                 Laurentia-Rae &                             Clearwater Anorthosite &     1917$^{+7}_{-7}$ &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &          &                                  \cite{Halls1999a} \\ \hline
+             Laurentia-Wyoming &                         Sourdough mafic dike swarm &     1899$^{+5}_{-5}$ &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &          &                                 \cite{Kilian2016b} \\ \hline
+               Laurentia-Slave &                                   Ghost Dike Swarm &     1887$^{+5}_{-9}$ &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &          &                                 \cite{Buchan2016a} \\ \hline
+               Laurentia-Slave &                           Mean Seton/Akaitcho/Mara &     1885$^{+5}_{-5}$ &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &          &                               \cite{Mitchell2010c} \\ \hline
+               Laurentia-Slave &                     Mean Kahochella, Peacock Hills &     1882$^{+4}_{-4}$ &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &          &                               \cite{Mitchell2010c} \\ \hline
+      Laurentia-Superior(West) &                                Molson (B+C2) dikes &     1879$^{+6}_{-6}$ &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &        47.6 &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                 Takiyuak Formation &   1876$^{+10}_{-10}$ &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &          &                                 \cite{Irving1979a} \\ \hline
+               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &   1876$^{+10}_{-10}$ &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &          &                                 \cite{Irving1979a} \\ \hline
+               Laurentia-Slave &               Pearson A/Peninsular/Kilohigok sills &     1870$^{+4}_{-4}$ &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &          &                               \cite{Mitchell2010c} \\ \hline
+            Laurentia-Superior &                         Haig/Flaherty/Sutton Mean  &     1870$^{+1}_{-1}$ &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &          &  Nordic workshop calculation based on data of \cite{Schmidt1980a, Schwarz1982a} \\ \hline
+ Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &     1838$^{+1}_{-1}$ &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &        73.8 &                                 \cite{Symons1999a} \\ \hline
+                 Laurentia-Rae &                                      Sparrow dikes &     1827$^{+4}_{-4}$ &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &          &                                \cite{McGlynn1974a} \\ \hline
+                 Laurentia-Rae &                                   Martin Formation &     1818$^{+4}_{-4}$ &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &          &                                  \cite{Evans1973a} \\ \hline
+                     Laurentia &                                      Dubawnt Group &   1785$^{+35}_{-35}$ &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &        49.4 &                                   \cite{Park1973a} \\ \hline
+ Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &     1766$^{+5}_{-5}$ &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &        68.9 &                                 \cite{Symons2000a} \\ \hline
+ Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &     1758$^{+1}_{-1}$ &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &        67.3 &                                   \cite{Gala1995a} \\ \hline
+                     Laurentia &                                      Cleaver dikes &     1741$^{+5}_{-5}$ &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &        61.7 &                                 \cite{Irving2004a} \\ \hline
+           Laurentia-Greenland &                        Melville Bugt diabase dikes &     1633$^{+5}_{-5}$ &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &        45.5 &                                  \cite{Halls2011a} \\ \hline
+                     Laurentia &                            Western Channel Diabase &     1590$^{+3}_{-3}$ &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &        47.5 &                                 \cite{Irving1972a} \\ \hline
+                     Laurentia &                 St.Francois Mountains Acidic Rocks &   1476$^{+16}_{-16}$ &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &        15.8 &                                  \cite{Meert2002b} \\ \hline
+                     Laurentia &                               Michikamau Intrusion &     1460$^{+5}_{-5}$ &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &        24.7 &                                 \cite{Emslie1976a} \\ \hline
+                     Laurentia &                                  Spokane Formation &   1458$^{+13}_{-13}$ &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &         4.2 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &                                 Snowslip Formation &   1450$^{+14}_{-14}$ &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &         1.4 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &                                 Tobacco Root dikes &   1448$^{+49}_{-49}$ &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &        31.9 &                                 \cite{Harlan2008a} \\ \hline
+                     Laurentia &                                       Purcell Lava &     1443$^{+7}_{-7}$ &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &         5.3 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &                          Rocky Mountain intrusions &   1430$^{+15}_{-15}$ &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &        16.0 &  Nordic workshop calculation based on data of \cite{Harlan1994a,Harlan1998a} \\ \hline
+                     Laurentia &                                   Mistastin Pluton &   1425$^{+25}_{-25}$ &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &        15.1 &                                 \cite{Fahrig1976a} \\ \hline
+                     Laurentia &                                 McNamara Formation &     1401$^{+6}_{-6}$ &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &         9.6 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &         Pilcher, Garnet Range and Libby Formations &   1385$^{+23}_{-23}$ &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &         8.8 &                                 \cite{Elston2002a} \\ \hline
+           Laurentia-Greenland &                                Zig-Zag Dal Basalts &     1382$^{+2}_{-2}$ &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &        43.8 &                              \cite{Marcussen1983a} \\ \hline
+           Laurentia-Greenland &                      Victoria Fjord dolerite dikes &     1382$^{+2}_{-2}$ &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &        36.6 &                             \cite{Abrahamsen1987a} \\ \hline
+           Laurentia-Greenland &                              Midsommersoe Dolerite &     1382$^{+2}_{-2}$ &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &        39.0 &                              \cite{Marcussen1983a} \\ \hline
+                     Laurentia &                                   Nain Anorthosite &   1305$^{+15}_{-15}$ &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &        28.1 &                                 \cite{Murthy1978a} \\ \hline
+           Laurentia-Greenland &                             North Qoroq intrusives &     1275$^{+1}_{-1}$ &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &        21.0 &                                  \cite{Piper1992a} \\ \hline
+           Laurentia-Greenland &                                  Kungnat Ring dike &     1275$^{+2}_{-2}$ &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &        11.1 &                                  \cite{Piper1977b} \\ \hline
+                     Laurentia &                         Mackenzie dikes grand mean &     1267$^{+2}_{-2}$ &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &        11.2 &                                 \cite{Buchan2000a} \\ \hline
+           Laurentia-Greenland &                         West Gardar Dolerite dikes &     1244$^{+8}_{-8}$ &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &        17.1 &                                  \cite{Piper1977b} \\ \hline
+           Laurentia-Greenland &                      West Gardar Lamprophyre dikes &   1238$^{+11}_{-11}$ &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &        15.9 &                                  \cite{Piper1977b} \\ \hline
+                     Laurentia &                             Sudbury dikes Combined &     1237$^{+5}_{-5}$ &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &         8.3 &                                 \cite{Palmer1977a} \\ \hline
+            Laurentia-Scotland &                                        Stoer Group &   1199$^{+70}_{-70}$ &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &        43.9 &                        Nordic workshop calculation \\ \hline
+           Laurentia-Greenland &                                 Hviddal Giant dike &     1184$^{+5}_{-5}$ &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &        43.3 &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                                     Narssaq Gabbro &     1184$^{+5}_{-5}$ &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &        48.8 &                                  \cite{Piper1977a} \\ \hline
            Laurentia-Greenland &                                  South Qoroq Intr.
-&      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1992a} \\ \hline
-           Laurentia-Greenland &                               Giant Gabbro Dykes &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &     1163$^{+2}_{-2}$ &                                  \cite{Piper1977a} \\ \hline
-           Laurentia-Greenland &                             NE-SW Trending dykes &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &     1160$^{+5}_{-5}$ &                                  \cite{Piper1992a} \\ \hline
-                     Laurentia &                        Ontario lamprophyre dykes &      NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143$^{+10}_{-10}$ &                                 \cite{Piispa2018a} \\ \hline
-                     Laurentia &                                    Abitibi Dykes &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &     1141$^{+2}_{-2}$ &                                  \cite{Ernst1993a} \\ \hline
-                     Laurentia &                          Nipigon sills and lavas &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &     1109$^{+2}_{-2}$ &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
-                     Laurentia &           Lowermost Mamainse Point volcanics -R1 &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &     1109$^{+2}_{-3}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
-                     Laurentia &                         Lower Osler volcanics -R &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &     1108$^{+3}_{-3}$ &                         \cite{Swanson-Hysell2014b} \\ \hline
-                     Laurentia &                        Middle Osler volcanics -R &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &     1107$^{+4}_{-4}$ &                         \cite{Swanson-Hysell2014b} \\ \hline
-                     Laurentia &                         Upper Osler volcanics -R &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &     1105$^{+1}_{-1}$ &  \cite{Halls1974a, Swanson-Hysell2014b, Swanson-Hysell2019a} \\ \hline
-                     Laurentia &               Lower Mamainse Point volcanics -R2 &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &     1105$^{+3}_{-4}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
-                     Laurentia &   Mamainse Point volcanics -C (lower N, upper R) &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &     1101$^{+1}_{-1}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
-                     Laurentia &                             North Shore lavas -N &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &     1097$^{+3}_{-3}$ &              \cite{Tauxe2009a,Swanson-Hysell2019a} \\ \hline
-                     Laurentia &                           Portage Lake Volcanics &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &     1095$^{+3}_{-3}$ &  \cite{Books1972a, Hnat2006a} as calculated in \cite{Swanson-Hysell2019a} \\ \hline
-                     Laurentia &                            Chengwatana Volcanics &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &     1095$^{+2}_{-2}$ &                                  \cite{Kean1997a} \\ \hline
-                     Laurentia &            Uppermost Mamainse Point volcanics -N &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &     1094$^{+6}_{-4}$ &                         \cite{Swanson-Hysell2014a} \\ \hline
-                     Laurentia &                  Cardenas Basalts and Intrusions &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &     1091$^{+5}_{-5}$ &                                   \cite{Weil2003a} \\ \hline
-                     Laurentia &                         Schroeder Lutsen Basalts &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &     1090$^{+2}_{-7}$ &                              \cite{Fairchild2017a} \\ \hline
-                     Laurentia &                      Central Arizona diabases -N &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &   1088$^{+11}_{-11}$ &                               \cite{Donadini2011b} \\ \hline
-                     Laurentia &                                 Lake Shore Traps &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &     1086$^{+1}_{-1}$ &                                \cite{Kulakov2013a} \\ \hline
-                     Laurentia &                    Michipicoten Island Formation &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &     1084$^{+1}_{-1}$ &                              \cite{Fairchild2017a} \\ \hline
-                     Laurentia &                                   Nonesuch Shale &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &    1080$^{+4}_{-10}$ &                                  \cite{Henry1977a} \\ \hline
-                     Laurentia &                                  Freda Sandstone &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &   1070$^{+14}_{-10}$ &                                  \cite{Henry1977a} \\ \hline
-                     Laurentia &                            Haliburton Intrusions &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &   1015$^{+15}_{-15}$ &                                \cite{Warnock2000a} \\ \hline
-            Laurentia-Scotland &                                   Torridon Group &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &  925$^{+145}_{-145}$ &                        Nordic workshop calculation \\ \hline
-            Laurentia-Svalbard &                    Lower Grusdievbreen Formation &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &    831$^{+20}_{-20}$ &                                 \cite{Maloof2006a} \\ \hline
-            Laurentia-Svalbard &                    Upper Grusdievbreen Formation &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &    800$^{+11}_{-11}$ &                                 \cite{Maloof2006a} \\ \hline
-            Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778$^{+2}_{-2}$ &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a}  \\ \hline
-   Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    770$^{+19}_{-40}$ &                                 \cite{Maloof2006a} \\ \hline
-            Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &     760$^{+6}_{-10}$ &                                   \cite{Weil2006b} \\ \hline
-            Laurentia &                                    Carbon Canyon &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &      757$^{+7}_{-7}$ &  \cite{Weil2004a} as calculated in \cite{Eyster2019a} \\ \hline
-            Laurentia &                             Carbon Butte/Awatubi &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &      751$^{+8}_{-8}$ &                                 \cite{Eyster2019a} \\                   Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724$^{+3}_{-3}$ &                               \cite{Denyszyn2009a} \\ \hline
-                     Laurentia &                                 Long Range Dykes &      B &     303.3 &      53.7 & 355.3 &  19.0 &      17.4 &      615$^{+2}_{-2}$ &                                 \cite{Murthy1992a} \\ \hline
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 321.5 & -34.2 &      15.4 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
-                     Laurentia &                         Baie des Moutons complex &      B &     301.0 &      50.8 & 332.7 &  42.6 &      12.0 &      583$^{+2}_{-2}$ &                             \cite{McCausland2011a} \\ \hline
-                     Laurentia &                       Callander Alkaline Complex &      B &     280.6 &      46.2 & 301.4 &  46.3 &       6.0 &      575$^{+5}_{-5}$ &                                 \cite{Symons1991a} \\ \hline
-                     Laurentia &                                 Catoctin Basalts &      B &     281.8 &      38.5 & 296.7 &  42.0 &      17.5 &      572$^{+5}_{-5}$ &                                  \cite{Meert1994a} \\ \hline
-                     Laurentia &                      Sept-Iles layered intrusion &      B &     293.5 &      50.2 & 321.0 & -20.0 &       6.7 &      565$^{+4}_{-4}$ &                                 \cite{Tanczyk1987a} \\ \hline
+&     1163$^{+2}_{-2}$ &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &        48.7 &                                  \cite{Piper1992a} \\ \hline
+           Laurentia-Greenland &                                 Giant Gabbro dikes &     1163$^{+2}_{-2}$ &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &        55.5 &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                               NE-SW Trending dikes &     1160$^{+5}_{-5}$ &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &        53.5 &                                  \cite{Piper1992a} \\ \hline
+                     Laurentia &                          Ontario lamprophyre dikes &   1143$^{+12}_{-12}$ &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &        61.2 &                                 \cite{Piispa2018a} \\ \hline
+                     Laurentia &                                      Abitibi dikes &     1141$^{+2}_{-2}$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \cite{Ernst1993a} \\ \hline                       Baltica &                        Salla dike (width 60-100 m) &     1122$^{+5}_{-3}$ &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &          &                                                 \\ \hline
+                     Laurentia &                            Nipigon sills and lavas &     1109$^{+2}_{-2}$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
+                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &     1109$^{+2}_{-3}$ &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &        62.9 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &                           Lower Osler volcanics -R &     1108$^{+3}_{-3}$ &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &        54.6 &                         \cite{Swanson-Hysell2014b} \\ \hline
+                     Laurentia &                          Middle Osler volcanics -R &     1107$^{+4}_{-4}$ &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &        50.5 &                         \cite{Swanson-Hysell2014b} \\ \hline
+                     Laurentia &                           Upper Osler volcanics -R &     1105$^{+1}_{-1}$ &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &        45.1 &  \cite{Halls1974a, Swanson-Hysell2014b, Swanson-Hysell2019a} \\ \hline
+                     Laurentia &                 Lower Mamainse Point volcanics -R2 &     1105$^{+3}_{-4}$ &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &        43.9 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &     1101$^{+1}_{-1}$ &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &        32.9 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &                               North Shore lavas -N &     1097$^{+3}_{-3}$ &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &        24.5 &              \cite{Tauxe2009a,Swanson-Hysell2019a} \\ \hline
+                     Laurentia &                              Chengwatana Volcanics &     1095$^{+2}_{-2}$ &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &        27.3 &                                   \cite{Kean1997a} \\ \hline
+                     Laurentia &                             Portage Lake Volcanics &     1095$^{+3}_{-3}$ &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &        22.7 &  \cite{Books1972a, Hnat2006a} as calculated in \cite{Swanson-Hysell2019a} \\ \hline
+                     Laurentia &              Uppermost Mamainse Point volcanics -N &     1094$^{+6}_{-4}$ &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &        25.6 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &                    Cardenas Basalts and Intrusions &     1091$^{+5}_{-5}$ &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &        27.3 &                                   \cite{Weil2003a} \\ \hline
+                     Laurentia &                           Schroeder Lutsen Basalts &     1090$^{+2}_{-7}$ &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &        25.9 &                              \cite{Fairchild2017a} \\ \hline
+                     Laurentia &                        Central Arizona diabases -N &   1088$^{+11}_{-11}$ &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &         9.6 &                               \cite{Donadini2011b} \\ \hline
+                     Laurentia &                                   Lake Shore Traps &     1086$^{+1}_{-1}$ &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &        22.3 &                                \cite{Kulakov2013a} \\ \hline
+                     Laurentia &                      Michipicoten Island Formation &     1084$^{+1}_{-1}$ &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &        10.2 &                              \cite{Fairchild2017a} \\ \hline
+                     Laurentia &                                     Nonesuch Shale &    1080$^{+4}_{-10}$ &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &         5.7 &                                  \cite{Henry1977a} \\ \hline
+                     Laurentia &                                    Freda Sandstone &   1070$^{+14}_{-10}$ &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &         2.4 &                                  \cite{Henry1977a} \\ \hline
+                     Laurentia &                              Haliburton Intrusions &   1015$^{+15}_{-15}$ &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &       -47.0 &                                \cite{Warnock2000a} \\ \hline
+                     Laurentia &                        Adirondack fayalite granite &    990$^{+20}_{-20}$ &     NR &     285.5 &      44.0 & 132.7 & -28.4 &       6.9 &       -50.7 &                                  \cite{Brown2012a} \\ \hline
+                     Laurentia &               Adirondack metamorphic\ orthosites &    970$^{+20}_{-20}$ &     NR &     286.0 &      44.0 & 149.0 & -25.1 &      11.6 &       -37.5 &                                  \cite{Brown2012a} \\ \hline
+                     Laurentia &                       Adirondack Microcline gneiss &    960$^{+20}_{-20}$ &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &       -31.5 &                                  \cite{Brown2012a} \\ \hline
+            Laurentia-Scotland &                                     Torridon Group &  925$^{+145}_{-145}$ &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &        -8.6 &                        Nordic workshop calculation \\ \hline
+            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &    831$^{+20}_{-20}$ &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &        -5.3 &                                 \cite{Maloof2006a} \\ \hline
+            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &    800$^{+11}_{-11}$ &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &        11.5 &                                 \cite{Maloof2006a} \\ \hline
+                     Laurentia &                                    Gunbarrel dikes &      778$^{+2}_{-2}$ &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &       -18.4 &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a} \\ \hline
+            Laurentia-Svalbard &                          Svanbergfjellet Formation &    770$^{+19}_{-40}$ &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &        12.8 &                                 \cite{Maloof2006a} \\ \hline
+                     Laurentia &                               Uinta Mountain Group &     760$^{+6}_{-10}$ &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &       -10.7 &                                   \cite{Weil2006b} \\ \hline
+                     Laurentia &                                      Carbon Canyon &      757$^{+7}_{-7}$ &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &        -8.5 &  \cite{Weil2004a} as calculated in \cite{Eyster2019a} \\ \hline
+                     Laurentia &                               Carbon Butte/Awatubi &      751$^{+8}_{-8}$ &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &         1.0 &                                 \cite{Eyster2019a} \\ \hline
+                     Laurentia &                          Franklin event grand mean &      718$^{+2}_{-2}$ &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &        -5.7 &                               \cite{Denyszyn2009b} \\ \hline
+                     Laurentia &                                   Long Range dikes &      615$^{+2}_{-2}$ &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &       -15.5 &                                 \cite{Murthy1992a} \\ \hline
+                     Laurentia &                           Baie des Moutons complex &      583$^{+2}_{-2}$ &      B &     301.0 &      50.8 & 152.7 & -42.6 &      12.0 &       -45.1 &                             \cite{McCausland2011a} \\ \hline
+                     Laurentia &                           Baie des Moutons complex &      583$^{+2}_{-2}$ &      B &     301.0 &      50.8 & 141.5 &  34.2 &      15.4 &         4.2 &                             \cite{McCausland2011a} \\ \hline
+                     Laurentia &                         Callander Alkaline Complex &      575$^{+5}_{-5}$ &      B &     280.6 &      46.2 & 121.4 & -46.3 &       6.0 &       -67.1 &                                 \cite{Symons1991a} \\ \hline
+                     Laurentia &                                   Catoctin Basalts &      572$^{+5}_{-5}$ &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &       -69.0 &                                  \cite{Meert1994a} \\ \hline
+                     Laurentia &                        Sept-Iles layered intrusion &      565$^{+4}_{-4}$ &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &        -7.9 &                                \cite{Tanczyk1987a} \\ \hline
 \end{longtable}
+\end{ThreePartTable}

### 2020-09-24 Manuscript/Laurentia_Paleogeo_Manuscript.tex
rough Baltica text added and some additional edits

--- 2020-09-16 Manuscript/manuscript.tex
@@
-\documentclass[11pt,letterpaper]{article}
-\setcounter{section}{3}
-\usepackage{threeparttable}
+%% arara directives
+% arara: xelatex
+% arara: bibtex
+% arara: xelatex
+% arara: xelatex
+
+%\documentclass{article} % One-column default
+\documentclass[twocolumn, switch]{article} % Method A for two-column formatting
+
+\usepackage{preprint}
+
+%% Math packages
+%\usepackage{amsmath, amsthm, amssymb, amsfonts}
+
+%% Bibliography options
+%\usepackage[numbers,square]{natbib}
+%\bibliographystyle{unsrtnat}
+\usepackage{natbib}
+%\bibliographystyle{Geology}
+
+%% General packages
+\usepackage[utf8]{inputenc}	% allow utf-8 input
+\usepackage[T1]{fontenc}	% use 8-bit T1 fonts
+\usepackage{xcolor}		% colors for hyperlinks
+\usepackage[colorlinks = true,
+            linkcolor = purple,
+            urlcolor  = blue,
+            citecolor = black,
+            anchorcolor = black]{hyperref}	% Color links to references, figures, etc.
+\usepackage{booktabs} 		% professional-quality tables
+\usepackage{nicefrac}		% compact symbols for 1/2, etc.
+%\usepackage{microtype}		% microtypography
+%\usepackage{lineno}		% Line numbers
+\usepackage{float}			% Allows for figures within multicol
 \usepackage{textcomp,marvosym}
-\usepackage{amsmath,amssymb}
-\usepackage[left]{lineno}
-\usepackage{changepage}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{setspace}
-\usepackage{fancyhdr}
-\usepackage{graphicx}
-\usepackage{booktabs}
-\usepackage{pdfpages}
-\usepackage{url}
+%\usepackage{multicol}		% Multiple columns (Method B)
+
+%\usepackage{lipsum}		%  Filler text
+\usepackage{lettrine}
+\usepackage{pdflscape}
 \usepackage{longtable}
-\usepackage{pdflscape}
-\usepackage[outercaption]{sidecap}
-\doublespacing
+\usepackage{threeparttablex}
 
-\raggedright
-\textwidth = 6.5 in
-\textheight = 9 in
-\oddsidemargin = 0.0 in
-\evensidemargin = 0.0 in
-\topmargin = 0.0 in
-\headheight = 0.0 in
-\headsep = 0.0 in
-\parskip = 0.1 in
-\parindent = 0.2 in
+ %% Special figure caption options
+\usepackage{newfloat}
+\DeclareFloatingEnvironment[name={Supplementary Figure}]{suppfigure}
+\usepackage{sidecap}
+\sidecaptionvpos{figure}{c}
 
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+% Section title spacing  options
+\usepackage{titlesec}
+\titlespacing\section{0pt}{12pt plus 3pt minus 3pt}{1pt plus 1pt minus 1pt}
+\titlespacing\subsection{0pt}{10pt plus 3pt minus 3pt}{1pt plus 1pt minus 1pt}
+\titlespacing\subsubsection{0pt}{8pt plus 3pt minus 3pt}{1pt plus 1pt minus 1pt}
 
+%%%%%%%%%%%%%%%%   Title   %%%%%%%%%%%%%%%%
+\title{The Precambrian paleogeography of Laurentia}
+
+% Add watermark with submission status
+%\usepackage{xwatermark}
+%% Left watermark
+%\newwatermark[firstpage,color=gray!60,angle=90,scale=0.32, xpos=-4.05in,ypos=0]{\href{https://doi.org/}{\color{gray}{Publication doi}}}
+%% Right watermark
+%\newwatermark[firstpage,color=gray!60,angle=90,scale=0.32, xpos=3.9in,ypos=0]{\href{https://doi.org/}{\color{gray}{Preprint doi}}}
+% Bottom watermark
+%\newwatermark[firstpage,color=gray!90,angle=0,scale=0.28, xpos=0in,ypos=-5in]{*correspondence: \texttt{swanson-hysell@berkeley.edu}}
+
+%%%%%%%%%%%%%%%  Author list  %%%%%%%%%%%%%%%
+\usepackage{authblk}
+\renewcommand*{\Authfont}{\bfseries}
+\author[]{Nicholas L. Swanson-Hysell}
+\affil[]{Department of Earth and Planetary Science, University of California, Berkeley, CA 94720 USA}
+
+\setcounter{section}{4}
+
+%%%%%%%%%%%%%%    Front matter    %%%%%%%%%%%%%%
 \begin{document}
 
-\begin{flushleft}
-\section{The Precambrian paleogeography of Laurentia}
+\twocolumn[ % Method A for two-column formatting
+  \begin{@twocolumnfalse} % Method A for two-column formatting
 
-Nicholas L. Swanson-Hysell\textsuperscript{1}
+\maketitle
 
-\bigskip
-
-\textsuperscript{1} Department of Earth and Planetary Science, University of California, Berkeley, CA 94720 USA
-
-\end{flushleft}
-
-\noindent\textit{This chapter is in preparation for the book Ancient Supercontinents and the Paleogeography of the Earth}
-
-%\linenumbers
-
-\subsection{Abstract}
-
-Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its ca.
-1.8 Ga amalgamation.
+\begin{abstract}
+Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its amalgamation 1.8 billion years ago.
 The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia.
@@
 These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.
+\end{abstract}
+%\keywords{First keyword \and Second keyword \and More} % (optional)
+\textit{This manuscript is a preprint of the chapter: \vspace{0.1 cm} \\
+Swanson-Hysell, N.
+L.
+(2021) The Precambrian paleogeography of Laurentia.
+In: Pesonen, L.J., Salminen, J., Evans, D.A.D., Elming, S.-Å., Veikkolainen, T.
+(eds.) Ancient Supercontinents and the Paleogeography of the Earth.}
+\vspace{0.3 cm}
+  \end{@twocolumnfalse} % Method A for two-column formatting
+] % Method A for two-column formatting
+
+%\begin{multicols}{2} % Method B for two-column formatting (doesn't play well with line numbers), comment out if using method A
+
+%%%%%%%%%%%%%%%  Main text   %%%%%%%%%%%%%%%
+% \linenumbers
 
 \subsection{Introduction and broad tectonic history}
-
-Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig.
+\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig.
 \ref{fig:Laurentia_map}).
 Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca.
-1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}.
+1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly (Fig.
+\ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}.
 Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
@@
 This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
-Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
-2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca.
+Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
+2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca.
 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca.
@@
 1.75 terminal suturing between the Wyoming and Superior provinces.
-Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca.
+Regardless, in both of these interpretations Wyoming is a later addition to Laurentia with final suturing post-dating ca.
 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
 \includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
-\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}).
+\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}.
 Proterozoic orogens are labeled with \textit{italicized text} (or.
@@
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
-The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
+The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a}.}
 \label{fig:Laurentia_map}
-\end{figure}
+\end{figure*}
 
@@
 1.86 to 1.82 \citep{Schulz2007a}.
-Firm evidence of the end of the orogeny comes from the ca.
-1.78 undeformed plutons of the post-Penokean East Central Minnesota Batholith \citep{Holm2005a}.
+Firm evidence of the end of the Penokean orogeny comes from the ca.
+1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
 
@@
 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
+Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where is called the Labradorian orogeny that is interpreted to have been active from ca.
+1.71 to 1.60 Ma \citep{Gower1992a, Gower2008b}.
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
 \includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
-\caption{\small{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
+\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
 A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown.
 The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.
-Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}}
+Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}
 \label{fig:tectonic_history}
-\end{figure}
+\end{figure*}
 
@@
 While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated.
-\citet{Hoffman1989a} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}.
+\citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}.
 This region is interpreted to have been subsequently deformed during a ca.
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
 \includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
-\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard).
-Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
+\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard).
+Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}
 \label{fig:Laurentia_poles}
-\end{figure}
+\end{figure*}
 
@@
 \ref{fig:Laurentia_poles} and Table 2).
-The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021b}.
+The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021a}.
 While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
@@
 %The poles on this figure should be labeled
-\begin{figure}
+\begin{figure*}
 \centering
 \includegraphics[width=6 in]{../Figures/Superior_Slave_reconstructions.pdf}
-\caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.
-The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
+\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
 This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of narrower Manikewan with a pivot-like closure.
@@
 The relatively well-resolved pole paths from the Superior and Slave provinces (Fig.
-\ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}}}
+\ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
 \label{fig:Superior_Slave_recons}
-\end{figure}
+\end{figure*}
 
@@
 This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
+
+\begin{table}[hbt]
+\caption{Rotations of separated terranes}
+{\scriptsize
+\begin{tabular}{|l|l|l|l|p{1.1 in}|}
+  \hline
+& Euler pole & Euler pole & rotation & note and \\
+Block & longitude & latitude & angle & citation \\
+\hline
+Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
+\hline
+Scotland & 161.9 & 78.6 & -31.0 & Reconstructing Atlantic opening following \cite{Torsvik2017a} \\
+\hline
+Svalbard & 125.0 & -81.0 & 68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
+\hline
+\end{tabular}
+}
+\label{tab:terrane_rotations}
+\end{table}
 
@@
 Data from the ca.
-585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree$\;$differences in direction within dikes dated within to 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
+585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree\ differences in direction within dikes dated within to 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
 The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
-\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W).
+\includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
+\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W).
 The paleomagnetic poles are compiled in Table 2.
-Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
+Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.
+TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}.
+Bottom panel: velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
 \label{fig:Laurentia_paleolatitude}
-\end{figure}
+\end{figure*}
 
@@
 
-\subsection{Paleoenvironmental constraints on paleolatitude}
-
-Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude.
-Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation.
-Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}.
-Evaporite deposits are particularly compelling as paleolatitude constraints given that their deposition is interpreted to be associated with arid regions resulting from large-scale Hadley cell downwelling \citep{Evans2006a}.
-While moisture in the subtropics can change along with Earth's climate, the overall pattern of $\sim$10-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}.
-Using a compilation of paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones.
-This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale convection circulation cells.
-
-There is high evaporation in the subtropics and tropics.
-Within the tropical rain belt (0 to $\sim$10\textdegree$\;$latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree$\;$ of the equator (within $\sim$8\textdegree$\;$ of the equator over land) with evaporation exceeding precipitation from those latitudes towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree$\;$ \citep{Park2020b}.
-However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to the precipitation of evaporites.
-For example, Lake Magadi in Kenya at a latitude of 1.9\textdegree$\;$S is a saline lake where thick bedded evaporites have accumulated \citep{Eugster1980a}.
-Caution is therefore needed when interpreting paleolatitude from evaporites in terrestrial and intracratonic settings given that they could occur both in tropical and subtropical latitudes.
-
-Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
-\begin{itemize}
-%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}.
-The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited
-\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preservedwithin secondary silica \citep{White1984a}.
-The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}.
-Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}.
-Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
-There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.
-These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 within the Helena Formation \citep{Evans2000c}.
-\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.
-These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
-\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} .
-These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca.
-1005 Ma detrital zircons \citep{Turner2016a}.
-The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca.
-820 Ma \citep{Turner2016a}.
-These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
-\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
-The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}.
-The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
-\end{itemize}
-
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=\textwidth]{../Figures/Laurentia_evaporite_figure.pdf}
-\caption{\small{\textbf{Left panel: The paleolatitude of evaporite deposits following the amalgamation of Laurentia as reconstructed by the simple Laurentia model shown in Fig.
-\ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic.
-Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}.
-The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.
-Right panel: The present-day location of evaporites reconstructed to the left.
-1: Altyn Formation (Belt Supergroup), 2: Wallace/Helena Formations (Belt Supergroup), 3: Iqqittuq Formation (Bylot Supergroup), 4: Ten Stone Formation (Mackenzie Mountains Supergroup), 5: Minto Inlet Formation (Shaler Supergroup), 6: Redstone River Formation (Mackenzie Mountains Supergroup), 7: Kilian Formation (Shaler Supergroup), 8: Carboniferous Canadian Maritime, 9: Carboniferous Sverdrup, 10: Permian Midcontinental USA.}}}
-\label{fig:Laurentia_evaporites}
-\end{figure}
-
-In Figure \ref{fig:Laurentia_evaporites}, the paleolatitude of these evaporite deposits are reconstructed using the
+\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
+\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
+Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.
+Paleomagnetic poles within 25 million years of each reconstruction time are plotted.
+All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}
+\label{fig:Laurentia_reconstructions}
+\end{figure*}
 
@@
 Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles.
-Paleomagnetic pole compilations themselves are evolving with better data and improved geochronology.
+Paleomagnetic pole compilations themselves are evolving with better data and improved geochronology \citep{Evans2021a}.
 Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress in Precambrian paleogeography.
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
-\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.
-These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
-Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.
-Paleomagnetic poles within 25 million years of each reconstruction time are plotted.
-All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}}}
-\label{fig:Laurentia_reconstructions}
-\end{figure}
 
@@
 With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g.
-\citealp{Torsvik2012a}).
+\citealp{Torsvik2017a}).
 
@@
 
+\subsection{Paleoenvironmental constraints on paleolatitude}
+
+Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude.
+Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation.
+Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003b}.
+Evaporite deposits are particularly compelling as paleolatitude constraints given that their deposition is interpreted to be associated with arid regions resulting from large-scale Hadley cell downwelling \citep{Evans2006a}.
+While moisture in the subtropics can change along with Earth's climate, the overall pattern of $\sim$10-35\textdegree\ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}.
+Using a compilation of paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 billion years large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to the latitudes of modern arid zones.
+This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale convection circulation cells.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Laurentia_evaporite_figure.pdf}
+\caption{\textbf{Paleolatitude of Laurentia evaporites.} Left panel: The paleolatitude of evaporite deposits following the amalgamation of Laurentia as reconstructed by the simple Laurentia model shown in Fig.
+\ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic.
+Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}.
+The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.
+Right panel: The present-day location of evaporites reconstructed to the left.
+1: Altyn Formation (Belt Supergroup), 2: Wallace/Helena Formations (Belt Supergroup), 3: Iqqittuq Formation (Bylot Supergroup), 4: Ten Stone Formation (Mackenzie Mountains Supergroup), 5: Minto Inlet Formation (Shaler Supergroup), 6: Redstone River Formation (Mackenzie Mountains Supergroup), 7: Kilian Formation (Shaler Supergroup), 8: Carboniferous Canadian Maritime, 9: Carboniferous Sverdrup, 10: Permian Midcontinental USA.}
+\label{fig:Laurentia_evaporites}
+\end{figure*}
+
+There is high evaporation in the subtropics and tropics.
+Within the tropical rain belt (0 to $\sim$10\textdegree\ latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree\ of the equator (within $\sim$8\textdegree$\;$ of the equator over land) with evaporation exceeding precipitation from those latitudes towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree\ \citep{Park2020b}.
+However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to the precipitation of evaporites.
+For example, Lake Magadi in Kenya at a latitude of 1.9\textdegree S is a saline lake where thick bedded evaporites have accumulated \citep{Eugster1980a}.
+Caution is therefore needed when interpreting paleolatitude from evaporites in terrestrial and intracratonic settings given that they could occur both in tropical and subtropical latitudes.
+
+Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
+\begin{itemize}
+%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}.
+The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited
+\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preserved within secondary silica \citep{White1984a}.
+The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}.
+Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}.
+Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
+There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.
+These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 Ma within the Helena Formation \citep{Evans2000c}.
+\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.
+These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
+\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} .
+These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca.
+1005 Ma detrital zircons \citep{Turner2016a}.
+The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca.
+820 Ma \citep{Turner2016a}.
+These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
+\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in an intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
+The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}.
+The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation \citep{Rooney2014a}.
+\end{itemize}
+
+In Figure \ref{fig:Laurentia_evaporites}, the paleolatitude of these evaporite deposits are reconstructed using the simple Laurentia model developed in this work as well as with the \cite{Li2013a} model for the late Mesoproterozoic to Neoproterozoic.
+The position of major Phanerozoic evaporite basins of North America are also shown with their paeleolatitude reconstructed with the paleogeographic mode of \cite{Torsvik2017a}.
+These paleogeographic models reconstruct evaporite deposition to have been within 30\textdegree\ of the equator in both the Phanerozoic and Proterozoic.
+In the Tonian period, evaporite deposition may have occurred equatorward of 10\textdegree\ (Fig.
+\ref{fig:Laurentia_evaporites} which may reflect increased aridity in the tropical interior of the Rodinia supercontinent.
+
 \subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
+
 Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia.
-This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
+This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models -- particularly those following the assembly of Laurentia.
+
 \textit{Note that this section is currently under development and is incomplete}
+
+\subsubsection{Paleogeographic connections prior to Laurentia assembly}
+
+Within this volume, \cite{Salminen2021a} describe proposed Neoarchean to Paleoproterozoic supercratons (terminology of \cite{Bleeker2003a}) prior to Laurentia assembly.
+In particular, they discuss the hypothesis of Superia (named after the Superior province) wherein the Superior province of Laurentia is central to a group of Archean lithospheric blocks \citep{Bleeker2006a}, including the Kola and Karelia provinces of Baltica and the Wyoming province of Laurentia, that broke up prior to 2.0 Ga.
+This hypothesis has largely been based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeast margin of the Superior Province in present-day coordinates \citep{Davey2020a}.
+While \cite{Salminen2021a}  argue that the originally proposed Superia of fit is not consistent with Baltica paleomagnetic poles, they propose a Superia (II) fit where there is a connection between the blocks between 2.68 and 2.05 Ga, but with some internal rotations.
+Another proposed connection, also evaluated by \cite{Salminen2021a}, based on the correlation of mafic dikes is one between the Slave province of Laurentia and the Dharwar province of India \citep{French2010a} as part of Sclavia (named after the Slave province; \citealp{Bleeker2003a}).
+This Slave-Dharwar connection is found to be consistent with ca.
+2.23 Ga and ca.
+2.19 Ga pairs of paleomagnetic poles if modified into a Sclavia (II) orientation and these blocks would have had a distinct drift history from Superia for most of the Paleoproterozoic \citep{Salminen2021a}.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
+\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic}.
+This reconstruction is modified from \cite{Swanson-Hysell2019a}.
+The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
+\label{fig:Grenville_reconstructions}
+\end{figure*}
 
@@
 In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}.
-On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed.
+On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed.
 In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers.
-This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a, Cawood2017a}.
+This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Hoffman1991a, Evans2013b, Cawood2017a}.
 While the lack of ca.
@@
 1200 Nova Floresta pole and ca.
-1150 Fortuna Formation pole; D'Agrella-Filho et al., this volume) as shown in \cite{Evans2013a}.
+1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}.
 In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca.
@@
 
-\subsubsection{Australia}
+\subsubsection{Australia and East Antarctica}
+
+It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia.
+Such correlations has led to a number of different proposed connections that place Australia + East Antarctica that
+
+Goodge2017a
+
+The
+
+\textit{Goodge2008a}
 
@@
 \subsubsection{Baltica}
+
 Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration.
-This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland (\citealp{Gower1990a}, Salminen et al., this volume).
-In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin \citep{Karlstrom2001a}.
-This reconstruction is brings paleomagnetic poles
+This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
+In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogeny of Baltica is a continuation of the the Mazatzal-Labradorian orogenies of Laurentia\citep{Karlstrom2001a}.
+Accretionary orogenesis within the Gothian orogeny was active ca.
+1.66 to 1.52 Ga \citep{Bergstrom2020a}.
+Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}.
+Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca.
+1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca.
+1.12 Ga if a virtual geomagnetic pole from
+the Salla dike is considered to be representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
+These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.
+
+Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic.
+This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b}.
+The timing of the main phase of the Sveconorwegian orogeny from ca.
+1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}.
+However, it is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}.
+\cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca.
+1.0 Ga.
+However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca.
+1.06 Ga.
+
+A clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g.
+\citealp{Evans2009a}).
+
+ orogen in Labrador         as advocated by \cite{Gower1990b}
+
+Tonian period (ca.
+975 to 915 Ma; \citealp{McClelland2019a}) intrusive rocks in Svalbard and East Greenland are interpreted to be related to subduction and accretionary orogenesis that has been termed the Valhalla orogeny \citep{Cawood2010a}.
+
+The NENA connection
+
+ -- a constraint that is honored in . This reconstruction is brings paleomagnetic poles
 
@@
 
-High-quality paleomagnetic poles constrain the coeval Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree$\;$ of latitude at the time they were emplaced with the craton margins separated by more than 30\textdegree$\;$ of latitude \citep{Swanson-Hysell2015a}.
+High-quality paleomagnetic poles constrain the coeval ca.
+1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced with the craton margins separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}.
 These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone.
-However, with the subsequent rapid motion of Laurentia to low latitudes it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny.
+However, with the subsequent rapid motion of Laurentia to low latitudes (Figs.
+\ref{Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny.
 This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g.
@@
 Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift.
-The preferred interpretation of \cite{Swanson-Hysell2015a} and \citep{Kasbohm2015a} was that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
+The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} was that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
 \citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia.
@@
 1090 followed by peak granulite metamorphism ca.
-1065 to 1045 Ma \citep{Diener2013, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean.
+1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean.
 If they indeed became conjoined in Rodinia, the separation of Kalahari from Laurentia likely initiated ca.
 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a}.
+
+%This interpretation is also consistent with a record of continental arc magmatism ca.
+1160 to 1070 Ma with a a peak of activity ca.
+1110 to 1090 Ma from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica  \citep{Wang2020a}.
+This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca.
+1080 and 1030 Ma consistent with orogenesis at the time of the Grenvillian orogeny \citep{Wang2020a}.
+They argue that this continued as an accretionary orogen with the TOAST coming in, but that Namaqua-Natal was on the lower plate with Maud belt on the upper plate.
 
@@
 
-%
-%\subsubsection{Australia}
-%
-%\subsubsection{Kalahari}
-% U-Pb zircon ages from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica have been interpreted to indicate continental arc magmatism between ca. 1160 to
-%1070 Ma with a a peak of activity ca.
-1110–1090 Ma  \citep{Wang2020a}.
-These age constraints as well as additional geochemical data have been interpreted to indicate subduction under Kalahari in the late Mesoproterozoic.
-% This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca. 1080 and 1030 Ma \citep{Wang2020a}.
+\subsubsection{Siberia}
 
 \subsection{The record implies plate tectonics throughout the Proterozoic}
+\label{sec:plate_tectonics}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
-\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}.
-The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
-\label{fig:Grenville_reconstructions}
-\end{figure}
-
 Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}.
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{../Figures/baked_contact_timeline_all.pdf}
+\caption{\textbf{Paleomagnetic poles with positive baked contact tests from Laurentia and other cratons}.
+This timeline shows the age of paleomagnetic poles with positive baked contact tests within the Nordic Paleogeography Workshop compilation \citep{Evans2021a}.
+.}
+\label{fig:baked_contact}
+\end{figure*}
+
 An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig.
@@
 Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
-Therefore, the record supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic (Table 2) which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
-This interpretation of a significant persistent geomagnetic field through much of the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g.
+Therefore, the record of abundant positive baked contact tests (Fig.
+\ref{fig:baked_contact}) and coherent paleomagnetic poles (Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
+This interpretation of a significant persistent geomagnetic field throughout the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g.
 \citealp{Macouin2006a}) and elsewhere.
@@
 Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
-While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities.
+While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic intervals bear similarities.
 In each case, accretionary collision of arc terranes was followed by continent-continent collision.
-The major difference is that the collisions of the Grenvillian and Appalachian orogenic cycles resulted in relatively minor crustal growth compared to the Trans-Hudson.
+The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson.
 Break-up following the Grenvillian and Appalachian orogenic cycles occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic cycle have remained sutured.
 As a result, Laurentia has been a formidable continent for the past 1.8 billion years.
-As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), the constraints from Laurentia are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.
+As can be seen in the Chapters on Archean paleogeography \citep{Salminen2021b}, Nuna \citep{Elming2021a}, and Rodinia \citep{Evans2021b}, the constraints from Laurentia are at the center of paleogeographic models through the Precambrian and will continue to be as the next generation of paleogeographic models are developed.
 
+%%%%%%%%%%%% Supplementary Methods %%%%%%%%%%%%
+%\footnotesize
+%\section*{Methods}
+
+%%%%%%%%%%%%% Acknowledgements %%%%%%%%%%%%%
+\footnotesize
 \subsection*{Acknowledgements}
-
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
@@
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
-The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Francis Macdonald.
-This work was supported by NSF CAREER Grant EAR-1847277 awarded to N.L.S.-H.
-The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse (CC BY 4.0) with attribution in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
+The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Yuem Park and Francis Macdonald.
+This work was supported by NSF CAREER Grant EAR-1847277.
+The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0)  in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
 
-\singlespacing
-\bibliographystyle{gsabull}
-\footnotesize\bibliography{../../references/allrefs}
-
-\section*{Glossary}
+\subsection*{Glossary}
 \noindent\textbf{accretionary orogeny } Lithospheric deformation associated with the subduction of oceanic lithosphere and the addition of material from the downgoing plate such as island arcs.
@@
 In this chapter, craton is predominantly used in reference to Laurentia which formed through the collision of Archean provinces and grew further through subsequent accretionary and collisional orogenesis.
+Note that in other usages the term can be focused on stable Archean lithosphere such as the individual Archean provinces of Laurentia.
 
@@
 \noindent\textbf{geocentric axial dipole hypothesis } The hypothesis that when it is time-averaged, Earth's magnetic field is dominantly a dipole aligned with the spin axis.
+
+\noindent\textbf{geodynamo } The mechanism whereby convective flow in Earth's fluid outer core generates Earth's magnetic field
 
@@
 
+\noindent\textbf{plate tectonics } A process where the lithosphere is in distinct pieces that move relative to one another.
+
 \noindent\textbf{Precambrian } A commonly used informal term to refer to geologic time prior to the Cambrian Period that started 541 million years ago.
@@
 
+\noindent\textbf{supercontinent } A large continent where most of Earth's continental lithosphere has been concentrated into a large landmass.
+The supercontinent Pangea that existed ca.
+200 Ma is the archtypal supercontinent.
+A threshold of 75$\%$ of extant continental crust has been proposed for a continent to be considered a supercontinent \citep{Meert2012a}.
+Gondwana (Australia + India + Africa + South America) constituted $\sim$60$\%$ and was a constituent of Pangea such that \cite{Evans2016a} proposed that it and similar landmasses should be called \textbf{semi-supercontinents}.
+
+\noindent\textbf{supercraton } A landmass that subsequently split into constituent crustal provinces. The term is typically applied to Archean provinces (cratons).
+
 \noindent\textbf{Superior province } The largest Archean province of Laurentia framed by the Trans-Hudson orogen to the west, the Grenvillian orogen to the east and the Penokean orogen to the south.
 It is also referred to as the Superior craton.
+
+\noindent\textbf{stagnant lid } A planetary state where there is a single lithospheric plate (`lid').
+The lithosphere is relatively stable and immobile in comparison to a planet with active plate tectonics where there is motion between multiple lithospheric plates.
 
@@
 \noindent\textbf{Torngat orogen } An orogen resulting from the Paleoproterozoic collision orogeny between the Meta Incognita and North Atlantic provinces.
+
+\noindent\textbf{true polar wander } The rotation of the solid Earth about the liquid outer core to maintain rotational equilibrium.
+This process results in Earth's lithosphere undergoing a single coherent rotation relative to the spin axis.
 
@@
 
-\begin{table}[hbt]
-\caption{Rotations of separated terranes}
-{\scriptsize
-\begin{tabular}{|l|l|l|l|p{2 in}|}
-  \hline
-& Euler pole & Euler pole & rotation & note and \\
-Block & longitude & latitude & angle & citation \\
-\hline
-Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
-\hline
-Scotland & 161.9 & 78.6 & -31.0 & Reconstructing Atlantic opening following \cite{Torsvik2017a} \\
-\hline
-Svalbard & 125.0 & -81.0 & 68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
-\hline
-\end{tabular}
-}
-\label{tab:terrane_rotations}
-\end{table}
+%%%%%%%%%%%%%%   Bibliography   %%%%%%%%%%%%%%
+\bibliographystyle{gsabull}
+\footnotesize\bibliography{../../references/allrefs}
 
@@
 
+%%%%%%%%%%%%  Supplementary Figures  %%%%%%%%%%%%
+%\clearpage
+
+%%%%%%%%%%%%%%%%   End   %%%%%%%%%%%%%%%%
+%\end{multicols}  % Method B for two-column formatting (doesn't play well with line numbers), comment out if using method A
 \end{document}

### 2020-09-24 Manuscript/pole_table.tex
tighten up Laurentia pole compilation notebook and regenerate output

@@
              Laurentia-Wyoming &                          Stillwater Complex - C2 &     2705\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &         nan &                                 \textbackslash cite\{Selkin2008a\} \\
-  Baltica-Fennoscandia-Karelia &                                Koitere sanukitoids &     2684\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      30.6 &      63.3 & 192.5 & -67.5 &      19.5 &         nan &                                                NaN \\
       Laurentia-Superior(East) &                     Otto Stock dikes and aureole &     2676\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &         nan &                               \textbackslash cite\{Pullaiah1975b\} \\
                Laurentia-Slave &                                     Defeat Suite &     2625\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &         nan &                               \textbackslash cite\{Mitchell2014a\} \\
-          Baltica-Fennoscandia &                   Shalskiy thick gabbronorite dike &     2511\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      36.0 &      61.9 & 222.1 &  22.7 &      11.5 &         nan &                                                NaN \\
       Laurentia-Superior(East) &                       Ptarmigan-Mistassini dikes &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
-          Baltica-Fennoscandia &                     Generalskaya Layered Intrusion &     2505\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      31.0 &      69.4 & 292.7 & -42.5 &      10.4 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                              Monchegorsk Intrusion &     2504\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      33.0 &      68.0 & 265.3 &   1.3 &       9.9 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &  Avdeev gabbronorite and thin Shalskiy diabase ...
-&   2476\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &      B &      36.0 &      61.9 & 243.5 & -12.3 &      14.0 &         nan &                                                NaN \\
       Laurentia-Superior(East) &                               Matachewan dikes R &   2466\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
-          Baltica-Fennoscandia &                          Imandra Layered Intrusion &   2446\$\textasciicircum \{+39\}\$\$\_\{-39\}\$ &      B &      33.1 &      67.6 & 280.3 & -16.1 &       7.8 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                                     Karelian dikes &     2446\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &      30.7 &      66.0 & 278.7 & -19.9 &       6.1 &         nan &                                                NaN \\
       Laurentia-Superior(East) &                               Matachewan dikes N &     2446\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
-          Baltica-Fennoscandia &                            Taivalkoski 2.3 Ga dike &   2339\$\textasciicircum \{+18\}\$\$\_\{-18\}\$ &     C+ &      28.1 &      65.5 & 257.3 &  20.4 &       6.3 &         nan &                                                NaN \\
                Laurentia-Slave &                                     Malley dikes &     2231\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &         nan &                                 \textbackslash cite\{Buchan2012a\} \\
@@
       Laurentia-Superior(East) &                                 Lac Esprit dikes &     2069\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
-              Baltica-Sarmatia &               Volyn-Dniestr-Bug intrusions A-group &   2061\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &      B &      29.8 &      49.6 & 182.9 &  15.7 &      13.7 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                               Kuetsyarvi Formation &     2060\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &      B &      29.5 &      69.5 & 300.8 &  24.7 &      16.4 &         nan &                                                NaN \\
       Laurentia-Greenland-Nain &                                  Kangamiut dikes &   2042\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &        57.6 &                                 \textbackslash cite\{Fahrig1976b\} \\
                Laurentia-Slave &                                Lac de Gras dikes &     2026\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &         nan &                                 \textbackslash cite\{Buchan2009a\} \\
-              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &   2000\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      27.4 &      50.9 & 140.4 &  64.4 &      12.7 &         nan &                                                NaN \\
       Laurentia-Superior(East) &                                      Minto dikes &     1998\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &         nan &                                  \textbackslash cite\{Evans2010a\} \\
-             Baltica - Karelia &                           Onego 1975 Ma Intrusions &     1976\$\textasciicircum \{+9\}\$\$\_\{-9\}\$ &      A &      34.0 &      62.1 & 101.5 &  44.4 &       6.3 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                                    Konchozero Sill &   1974\$\textasciicircum \{+27\}\$\$\_\{-27\}\$ &      B &      34.0 &      62.1 & 281.9 & -14.2 &      10.4 &         nan &                                                NaN \\
                Laurentia-Slave &                                  Rifle Formation &     1963\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &         nan &                                  \textbackslash cite\{Evans1981a\} \\
-          Baltica-Fennoscandia &             Tsuomasavarri Gabbro-Diorite Intrusion &     1931\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &      28.3 &      69.9 & 247.3 &  40.2 &       6.0 &         nan &                                                NaN \\
                  Laurentia-Rae &                           Clearwater Anorthosite &     1917\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &         nan &                                  \textbackslash cite\{Halls1999a\} \\
@@
                Laurentia-Slave &                                 Ghost Dike Swarm &     1887\$\textasciicircum \{+5\}\$\$\_\{-9\}\$ &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &         nan &                                 \textbackslash cite\{Buchan2016a\} \\
-          Baltica-Fennoscandia &                         Kiuruvesi-Pielavesi gabbro &     1886\$\textasciicircum \{+6\}\$\$\_\{-5\}\$ &      B &      26.8 &      63.2 & 231.3 &  41.1 &       5.4 &         nan &                                                NaN \\
                Laurentia-Slave &                         Mean Seton/Akaitcho/Mara &     1885\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &         nan &                               \textbackslash cite\{Mitchell2010c\} \\
@@
       Laurentia-Superior(West) &                              Molson (B+C2) dikes &     1879\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &        47.6 &                                  \textbackslash cite\{Evans2010a\} \\
+               Laurentia-Slave &        Douglas Peninsula Formation, Pethei Group &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &         nan &                                 \textbackslash cite\{Irving1979a\} \\
                Laurentia-Slave &                               Takiyuak Formation &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &         nan &                                 \textbackslash cite\{Irving1979a\} \\
-               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &   1876\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &         nan &                                 \textbackslash cite\{Irving1979a\} \\
-               Laurentia-Slave &               Pearson A/Peninsular/Kilohigok sills &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &         nan &                               \textbackslash cite\{Mitchell2010c\} \\
             Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &         nan &  Nordic workshop calculation based on data of \textbackslash ...
 \\
-        Baltica - Fennoscandia &                                      Keuruu dikes  &   1869\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &     B  &      24.7 &      62.3 & 230.9 &  45.7 &       5.5 &         nan &                                                NaN \\
+               Laurentia-Slave &             Pearson A/Peninsular/Kilohigok sills &     1870\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &         nan &                               \textbackslash cite\{Mitchell2010c\} \\
  Laurentia-Trans-Hudson orogen &                              Boot-Phantom Pluton &     1838\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &        73.8 &                                 \textbackslash cite\{Symons1999a\} \\
@@
                  Laurentia-Rae &                                 Martin Formation &     1818\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &         nan &                                  \textbackslash cite\{Evans1973a\} \\
-          Baltica-Fennoscandia &                                    Notträsk gabbro &     1806\$\textasciicircum \{+6\}\$\$\_\{-6\}\$ &      B &      21.8 &      65.8 & 216.2 &  43.5 &       6.1 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                                      Kallax gabbro &     1800\$\textasciicircum \{+5\}\$\$\_\{-6\}\$ &      B &      22.0 &      65.5 & 209.0 &  49.0 &       nan &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                 Lake Ladoga, Mean intr.
-\& dikes, A &   1800\$\textasciicircum \{+19\}\$\$\_\{-56\}\$ &      B &      30.0 &      61.3 & 229.1 &  50.9 &       nan &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                                      Hoting gabbro &   1786\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      A &      16.2 &      64.2 & 233.3 &  43.0 &      10.9 &         nan &                                                NaN \\
                      Laurentia &                                    Dubawnt Group &   1785\$\textasciicircum \{+35\}\$\$\_\{-35\}\$ &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &        49.4 &                                   \textbackslash cite\{Park1973a\} \\
-          Baltica-Fennoscandia &                              Småland intrusives -C &     1777\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &      A &      15.7 &      57.1 & 182.7 &  45.7 &       8.0 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                                 Shoksha Sandstones &   1775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &      B &      33.8 &      61.3 & 221.1 &  39.7 &       4.0 &         nan &                                                NaN \\
  Laurentia-Trans-Hudson orogen &                          Deschambault Pegmatites &     1766\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &        68.9 &                                 \textbackslash cite\{Symons2000a\} \\
  Laurentia-Trans-Hudson orogen &                                 Jan Lake Granite &     1758\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &        67.3 &                                   \textbackslash cite\{Gala1995a\} \\
-              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &   1755\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      A &      28.6 &      50.6 & 169.1 &  26.5 &       3.9 &         nan &                                                NaN \\
-          Baltica-Fennoscandia &                                MEAN Ropruchey sill &     1751\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      B &      35.3 &      61.4 & 217.0 &  39.1 &       8.6 &         nan &                                                NaN \\
                      Laurentia &                                    Cleaver dikes &     1741\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &        61.7 &                                 \textbackslash cite\{Irving2004a\} \\
-              Baltica-Sarmatia &                       Volyn-Dniestr-Bug intrusions &   1722\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      B &      29.6 &      50.0 & 163.2 &  10.7 &      10.2 &         nan &                                                NaN \\
-        Baltica - Fennoscandia &                                   Häme dike swarm  &   1647\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &      B &      25.1 &      61.4 & 209.8 &  23.6 &      14.7 &         nan &                                                NaN \\
-                       Baltica &                        Sipoo Quartz Porphyry dikes &   1633\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      25.3 &      60.3 & 180.4 &  26.4 &       9.4 &         nan &                                                NaN \\
            Laurentia-Greenland &                      Melville Bugt diabase dikes &     1633\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &        45.5 &                                  \textbackslash cite\{Halls2011a\} \\
-                       Baltica &                              Quartz porphyry dikes &   1631\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      26.8 &      61.3 & 175.4 &  30.2 &       9.4 &         nan &                                                NaN \\
-                       Baltica &                                 Hoting basic dikes &   1614\$\textasciicircum \{+24\}\$\$\_\{-24\}\$ &      B &      16.2 &      64.2 & 146.7 &  21.9 &      13.7 &         nan &                                                NaN \\
                      Laurentia &                          Western Channel Diabase &     1590\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &        47.5 &                                 \textbackslash cite\{Irving1972a\} \\
-                       Baltica &                                   Åland dike swarm &   1578\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      20.0 &      60.0 & 191.4 &  23.7 &       2.8 &         nan &                                                NaN \\
-                       Baltica &                            Satakunta dike swarm -C &   1578\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      21.5 &      62.0 & 188.1 &  29.3 &       6.6 &         nan &                                                NaN \\
-                       Baltica &                                  Ragunda Formation &   1506\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &      B &      16.1 &      63.3 & 166.6 &  51.6 &       7.1 &         nan &                                                NaN \\
                      Laurentia &               St.Francois Mountains Acidic Rocks &   1476\$\textasciicircum \{+16\}\$\$\_\{-16\}\$ &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &        15.8 &                                  \textbackslash cite\{Meert2002b\} \\
-                       Baltica &                           MEAN Bunkris/Glysjon/Oje &     1469\$\textasciicircum \{+9\}\$\$\_\{-9\}\$ &      B &      13.5 &      61.5 & 179.8 &  28.3 &      13.2 &         nan &                                                NaN \\
                      Laurentia &                             Michikamau Intrusion &     1460\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &        24.7 &                                 \textbackslash cite\{Emslie1976a\} \\
                      Laurentia &                                Spokane Formation &   1458\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &         4.2 &                                 \textbackslash cite\{Elston2002a\} \\
-                       Baltica &                    Lake Ladoga basalt, sill, dikes &   1452\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      31.2 &      61.5 & 177.1 &  15.2 &       5.5 &         nan &                                                NaN \\
                      Laurentia &                               Snowslip Formation &   1450\$\textasciicircum \{+14\}\$\$\_\{-14\}\$ &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &         1.4 &                                 \textbackslash cite\{Elston2002a\} \\
@@
                      Laurentia &       Pilcher, Garnet Range and Libby Formations &   1385\$\textasciicircum \{+23\}\$\$\_\{-23\}\$ &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &         8.8 &                                 \textbackslash cite\{Elston2002a\} \\
+           Laurentia-Greenland &                            Midsommersoe Dolerite &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &        39.0 &                              \textbackslash cite\{Marcussen1983a\} \\
            Laurentia-Greenland &                              Zig-Zag Dal Basalts &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &        43.8 &                              \textbackslash cite\{Marcussen1983a\} \\
            Laurentia-Greenland &                    Victoria Fjord dolerite dikes &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &        36.6 &                             \textbackslash cite\{Abrahamsen1987a\} \\
-           Laurentia-Greenland &                              Midsommersoe Dolerite &     1382\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &        39.0 &                              \textbackslash cite\{Marcussen1983a\} \\
-                       Baltica &                                             Mashak &   1376\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      57.0 &      54.0 & 193.0 &   1.8 &      14.8 &         nan &                                                NaN \\
                      Laurentia &                                 Nain Anorthosite &   1305\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &        28.1 &                                 \textbackslash cite\{Murthy1978a\} \\
+           Laurentia-Greenland &                                Kungnat Ring dike &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &        11.1 &                                  \textbackslash cite\{Piper1977b\} \\
            Laurentia-Greenland &                           North Qoroq intrusives &     1275\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &        21.0 &                                  \textbackslash cite\{Piper1992a\} \\
-           Laurentia-Greenland &                                  Kungnat Ring dike &     1275\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &        11.1 &                                  \textbackslash cite\{Piper1977b\} \\
                      Laurentia &                       Mackenzie dikes grand mean &     1267\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &        11.2 &                                 \textbackslash cite\{Buchan2000a\} \\
-                       Baltica &                       MEAN post-Jotnian intrusions &   1258\$\textasciicircum \{+12\}\$\$\_\{-12\}\$ &      A &      18.1 &      62.4 & 159.1 &  -1.8 &       3.4 &         nan &                                                NaN \\
            Laurentia-Greenland &                       West Gardar Dolerite dikes &     1244\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &        17.1 &                                  \textbackslash cite\{Piper1977b\} \\
@@
                      Laurentia &                                    Abitibi dikes &     1141\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \textbackslash cite\{Ernst1993a\} \\
-                       Baltica &                        Salla dike (width 60-100 m) &     1122\$\textasciicircum \{+5\}\$\$\_\{-3\}\$ &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &         nan &                                                NaN \\
                      Laurentia &                          Nipigon sills and lavas &     1109\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \textbackslash ...
@@
                      Laurentia &                                  Freda Sandstone &   1070\$\textasciicircum \{+14\}\$\$\_\{-10\}\$ &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &         2.4 &                                  \textbackslash cite\{Henry1977a\} \\
-                       Baltica &                             MEAN Bamble Intrusions &   1048\$\textasciicircum \{+73\}\$\$\_\{-73\}\$ &     C+ &       9.5 &      58.9 & 222.5 & -15.1 &      19.3 &         nan &                                                NaN \\
-                       Baltica &                           Laanila-Ristijarvi dikes &   1044\$\textasciicircum \{+52\}\$\$\_\{-52\}\$ &      B &      28.1 &      68.7 & 212.2 &  -2.1 &      13.8 &         nan &                                                NaN \\
                      Laurentia &                            Haliburton Intrusions &   1015\$\textasciicircum \{+15\}\$\$\_\{-15\}\$ &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &       -47.0 &                                \textbackslash cite\{Warnock2000a\} \\
@@
                      Laurentia &                     Adirondack Microcline gneiss &    960\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &       -31.5 &                                  \textbackslash cite\{Brown2012a\} \\
-                       Baltica &                            MEAN Blekinga dolerites &      950\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &      15.0 &      56.1 & 247.0 &  13.0 &      16.0 &         nan &                                                NaN \\
-                       Baltica &               dikes of central and southern Sweden &    945\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      A &      14.2 &      58.4 & 240.7 &  -0.9 &       6.7 &         nan &                                                NaN \\
             Laurentia-Scotland &                                   Torridon Group &  925\$\textasciicircum \{+145\}\$\$\_\{-145\}\$ &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &        -8.6 &                        Nordic workshop calculation \\
-        Baltica - Fennoscandia &                Bjerkreim-Sokndal layered intrusion &    921\$\textasciicircum \{+17\}\$\$\_\{-17\}\$ &      B &       6.1 &      58.5 & 217.9 & -35.9 &       6.3 &         nan &                                                NaN \\
-                       Baltica &                    Bratton and Algon Igneous Rocks &    916\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &      B &      11.7 &      57.9 & 249.0 &   5.0 &       3.9 &         nan &                                                NaN \\
-                       Baltica &                           MEAN Rogaland igneous cx &    903\$\textasciicircum \{+33\}\$\$\_\{-33\}\$ &      B &       6.9 &      58.3 & 207.9 & -43.2 &      10.1 &         nan &                                                NaN \\
-                       Baltica &                                   Hunnedalen dikes &    848\$\textasciicircum \{+27\}\$\$\_\{-27\}\$ &      B &       7.0 &      58.9 & 222.0 & -41.0 &      15.6 &         nan &                                                NaN \\
             Laurentia-Svalbard &                    Lower Grusdievbreen Formation &    831\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &        -5.3 &                                 \textbackslash cite\{Maloof2006a\} \\
             Laurentia-Svalbard &                    Upper Grusdievbreen Formation &    800\$\textasciicircum \{+11\}\$\$\_\{-11\}\$ &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &        11.5 &                                 \textbackslash cite\{Maloof2006a\} \\
-                       Baltica &                                    Katav Formation &  800\$\textasciicircum \{+100\}\$\$\_\{-100\}\$ &      B &      57.4 &      54.8 & 169.9 &  35.7 &      11.4 &         nan &                                                NaN \\
                      Laurentia &                                  Gunbarrel dikes &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &       -18.4 &  Calculation from \textbackslash cite\{Eyster2019a\} based on d...
@@
                      Laurentia &                        Franklin event grand mean &      718\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &        -5.7 &                               \textbackslash cite\{Denyszyn2009b\} \\
-                       Baltica &                                     Egersund dikes &      616\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &      A &       5.9 &      58.8 &  44.1 &  31.4 &      15.6 &         nan &                                                NaN \\
                      Laurentia &                                 Long Range dikes &      615\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &       -15.5 &                                 \textbackslash cite\{Murthy1992a\} \\
@@
                      Laurentia &                                 Catoctin Basalts &      572\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &       -69.0 &                                  \textbackslash cite\{Meert1994a\} \\
-                       Baltica &                                       Volhyn lavas &    571\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      26.0 &      51.2 &   4.4 &  19.8 &      28.3 &         nan &                                                NaN \\
                      Laurentia &                      Sept-Iles layered intrusion &      565\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &        -7.9 &                                \textbackslash cite\{Tanczyk1987a\} \\
-                       Baltica &                               Kurgashlya formation &      565\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &      57.5 &      53.3 & 314.5 &  50.9 &       5.3 &         nan &                                                NaN \\
-                       Baltica &                                  Bakeevo formation &      565\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      A &      58.2 &      54.9 & 299.1 &  42.3 &       5.3 &         nan &                                                NaN \\
-                       Baltica &                        Basu-Kukkarauk formation -C &      562\$\textasciicircum \{+5\}\$\$\_\{-5\}\$ &      B &      57.0 &      54.0 & 187.3 &   1.1 &       5.8 &         nan &                                                NaN \\
-                 Baltica-Urals &                          Basu formation  (precise) &    560\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &      B &      56.9 &      53.9 & 186.1 &   1.7 &       3.8 &         nan &                                                NaN \\
-                       Baltica &                 Chernokamenskay group sediments -C &    557\$\textasciicircum \{+13\}\$\$\_\{-13\}\$ &      B &      58.4 &      58.5 & 306.7 &  17.3 &       6.0 &         nan &                                                NaN \\
-                       Baltica &                             MEAN Vendian sediments &      556\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &      B &      40.0 &      65.5 & 293.0 &  31.3 &       9.9 &         nan &                                                NaN \\
-                 Baltica-Urals &                                Zigan HTC thin beds &   548\$\textasciicircum \{+27\}\$\$\_\{--12\}\$ &      B &      56.7 &      54.4 & 138.3 & -18.0 &      10.7 &         nan &                                                NaN \\
-                       Baltica &                   Zigan formation clastic rocks -C &      548\$\textasciicircum \{+4\}\$\$\_\{-4\}\$ &      B &      56.7 &      53.7 & 138.4 & -16.2 &       4.1 &         nan &                                                NaN \\
 \end{longtable}

### 2020-09-24 Manuscript/Laurentia_Paleogeo_Manuscript.tex
implement edits across all sections

@@
 A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
-This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints.
+This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints.
 Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
@@
 Proterozoic orogens are labeled with \textit{italicized text} (or.
--- orogen; THO  -- Trans-Hudson orogen; prov.
--- province).
+-- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
-Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where is called the Labradorian orogeny that is interpreted to have been active from ca.
-1.71 to 1.60 Ma \citep{Gower1992a, Gower2008b}.
+Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca.
+1.71 to 1.60 Ma (Fig.
+\ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).
+
 Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
@@
 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}.
-Younger magmatic activity ca.
-1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}.
-While an active margin interpretation for the Granite-Rhyolite Province with arc and back-arc magmatism in a back-arc setting has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+Younger magmatic activity (ca.
+1.37 Ga) of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}.
+While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.
-Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}
+Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
 \label{fig:tectonic_history}
@@
 Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision.
-If this interpretation is correct, it is similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (\ref{fig:tectonic_history}).
+If this interpretation is correct, it is similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (Fig.
+\ref{fig:tectonic_history}).
 
@@
 This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
-Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.
+Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Cannon1994a, Swanson-Hysell2019a}.
 
@@
 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}.
-This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
+This constraint is incorporated into paleogeographic models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g.
 Amazonia in \citealp{Elming2009a, Elming2021a}).
 The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca.
-1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
+1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
 This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
@@
 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}.
-However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca.
+However, this extensional basin development predates the rifting that that is interpreted to have been followed by passive margin thermal subsidence closer to the ca.
 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}.
@@
 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard.
-In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
+In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
 
@@
 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}.
-Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age -- older than the Fifteenmile Group and younger than the ca.
+Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca.
 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).
@@
 850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}.
-These data suggest that conjugate continental lithosphere had rifted away from northeast Greenland by ca.
+These data suggest that conjugate continental lithosphere rifted away from northeast Greenland by ca.
 820 Ma.
@@
 
-Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
+Even without considering other continents on Earth, the geological record of Paleoproterozoic collision of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years.
 This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.
@@
 \includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
-\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard).
-Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}
+\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates).
+Bottom panel: Paleomagnetic data for Archean provinces prior to the amalgamation of Laurentia.}
 \label{fig:Laurentia_poles}
@@
 Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
-These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}.
+These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported.
 Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions.
@@
 1015 Ma Haliburton intrusions (Table 2).
-However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below).
+However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below).
 In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2).
@@
 Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca.
-760 Ma)  as presented in \cite{Eyster2019a} are also included.
+760 Ma)  as presented in \cite{Eyster2019a}, incorporating data from \cite{Weil2004a}, are also included.
 
@@
 \caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
-This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of narrower Manikewan with a pivot-like closure.
+This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
 Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
@@
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles from 2.23 and 1.89 Ga (Fig.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles between 2.23 and 1.89 Ga (Fig.
 \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}).
@@
 & Euler pole & Euler pole & rotation & note and \\
-Block & longitude & latitude & angle & citation \\
+Terrane & longitude & latitude & angle & citation \\
 \hline
@@
 Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage.
-These data result in progressive paths such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian (ca.
+These data result in progressive paths such as ascending up to the Logan Loop \citep{Robertson1971a}, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
 775 to 720 Ma) path \citep{Eyster2019a}.
@@
 In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia.
-Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent.
+Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire craton.
 The reliability of this approach gains credence through the good agreement between the ca.
@@
 
-An exception to this overall agreement between poles from Greenland and mainland Laurentia occurs ca.
+An exception to this overall agreement between coeval poles from Greenland and mainland Laurentia occurs ca.
 1382 Ma.
@@
 However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts.
-There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}.
+However the overall coherency of the pole directions from the Belt Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
 At present, it is unclear which poles are a better representation of Laurentia's position ca.
@@
 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization.
-Notably the lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole.
+The lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole.
 Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}.
 Data from the ca.
-585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree\ differences in direction within dikes dated within to 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
+585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
 The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
@@
 \includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
-\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W).
+\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes.
 The paleomagnetic poles are compiled in Table 2.
@@
 TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}.
-Bottom panel: velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
+Bottom panel: the velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
 \label{fig:Laurentia_paleolatitude}
@@
 \includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
-\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
+\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
 Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.
@@
 1000 Ma), which is a striking departure from the paleomagnetic record and standard paleogeographic models.
-Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating.
+Additionally, the implemented plate motion of Laurentia in the \cite{Merdith2017b} model strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating (Fig.
+\ref{fig:Laurentia_paleolatitude}).
 The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence both for the supercontinent Rodinia being equatorial and for ice sheets associated with the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.
@@
 The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record.
-The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of (\citealp{Li2013a}; while the model parameters were not published with that study they have now been made available by the authors).
+The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (while the model parameters were not published with that study they have now been made available by the authors).
+This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again.
 
@@
 The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.
-Right panel: The present-day location of evaporites reconstructed to the left.
-1: Altyn Formation (Belt Supergroup), 2: Wallace/Helena Formations (Belt Supergroup), 3: Iqqittuq Formation (Bylot Supergroup), 4: Ten Stone Formation (Mackenzie Mountains Supergroup), 5: Minto Inlet Formation (Shaler Supergroup), 6: Redstone River Formation (Mackenzie Mountains Supergroup), 7: Kilian Formation (Shaler Supergroup), 8: Carboniferous Canadian Maritime, 9: Carboniferous Sverdrup, 10: Permian Midcontinental USA.}
+The evaporite points are labeled with numbers both in the paleolatitude plot and the map of their present-day location in the right panel.
+The formations associated with these numbers are: 1 -- Altyn Formation (Belt Supergroup); 2: Wallace/Helena Formations -- Belt Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Carboniferous Canadian Maritime; 9 -- Carboniferous Sverdrup; 10 -- Permian Midcontinental USA.}
 \label{fig:Laurentia_evaporites}
@@
 The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}.
-The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation \citep{Rooney2014a}.
+The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os date from the overlying Coppercap Formation \citep{Rooney2014a}.
 \end{itemize}
@@
 
-Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia.
-This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models -- particularly those following the assembly of Laurentia.
+Many different paleogeographic connections between Laurentia and other Proterozoic cratons have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces.
+This section is not comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models --- particularly those following the assembly of Laurentia.
 
@@
 
-Within this volume, \cite{Salminen2021a} describe proposed Neoarchean to Paleoproterozoic supercratons (terminology of \cite{Bleeker2003a}) prior to Laurentia assembly.
+Within this volume, \cite{Salminen2021a} describe proposed Neoarchean to Paleoproterozoic groupings of Archean provinces (``supercratons'' in the terminology of \citealp{Bleeker2003a}) prior to Laurentia assembly.
 In particular, they discuss the hypothesis of Superia (named after the Superior province) wherein the Superior province of Laurentia is central to a group of Archean lithospheric blocks \citep{Bleeker2006a}, including the Kola and Karelia provinces of Baltica and the Wyoming province of Laurentia, that broke up prior to 2.0 Ga.
 This hypothesis has largely been based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeast margin of the Superior Province in present-day coordinates \citep{Davey2020a}.
-While \cite{Salminen2021a}  argue that the originally proposed Superia of fit is not consistent with Baltica paleomagnetic poles, they propose a Superia (II) fit where there is a connection between the blocks between 2.68 and 2.05 Ga, but with some internal rotations.
-Another proposed connection, also evaluated by \cite{Salminen2021a}, based on the correlation of mafic dikes is one between the Slave province of Laurentia and the Dharwar province of India \citep{French2010a} as part of Sclavia (named after the Slave province; \citealp{Bleeker2003a}).
+\cite{Salminen2021a}  argue that the originally proposed Superia fit is not consistent with Baltica paleomagnetic poles.
+Instead, they propose a Superia (II) fit where there is a connection between the blocks between 2.68 and 2.05 Ga with some internal rotations.
+Another proposed connection evaluated by \cite{Salminen2021a}, also based on the correlation of mafic dikes, is one between the Slave province of Laurentia and the Dharwar province of India \citep{French2010a} as part of Sclavia (named after the Slave province; \citealp{Bleeker2003a}).
 This Slave-Dharwar connection is found to be consistent with ca.
 2.23 Ga and ca.
-2.19 Ga pairs of paleomagnetic poles if modified into a Sclavia (II) orientation and these blocks would have had a distinct drift history from Superia for most of the Paleoproterozoic \citep{Salminen2021a}.
+2.19 Ga pairs of paleomagnetic poles if modified into a Sclavia (II) orientation \citep{Salminen2021a}.
+These blocks would have had a distinct drift history from Superia for most of the Paleoproterozoic \citep{Salminen2021a}.
 
@@
 On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed.
-In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers.
+In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers.
 This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Hoffman1991a, Evans2013b, Cawood2017a}.
 While the lack of ca.
-1100 and ca.
-1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca.
+1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca.
 1200 Nova Floresta pole and ca.
@@
 Departure of Amazonia potentially occurred as early as ca.
-700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.
+700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia.
 
@@
 It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia.
-Such correlations has led to a number of different proposed connections that place Australia + East Antarctica that
-
-Goodge2017a
-
-The
-
-\textit{Goodge2008a}
+Such correlations has led to a number of different proposed connections that place Australia + East Antarctica...
+
+%Goodge2017a
+%
+%The 
+%
+%\textit{Goodge2008a}
 
@@
 A clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g.
-\citealp{Evans2009a}).
-
- orogen in Labrador         as advocated by \cite{Gower1990b}
-
-Tonian period (ca.
+\citealp{Evans2009a})...
+
+% orogen in Labrador         as advocated by \cite{Gower1990b} 
+%
+%Tonian period (ca.
 975 to 915 Ma; \citealp{McClelland2019a}) intrusive rocks in Svalbard and East Greenland are interpreted to be related to subduction and accretionary orogenesis that has been termed the Valhalla orogeny \citep{Cawood2010a}.
-
-The NENA connection
-
- -- a constraint that is honored in . This reconstruction is brings paleomagnetic poles
-
-compatible with existing paleomagnetic constraints from ca.
+%
+%The NENA connection 
+%
+% -- a constraint that is honored in . This reconstruction is brings paleomagnetic poles
+%
+%compatible with existing paleomagnetic constraints from ca.
 1750 to 1270 Ma \cite{Evans2008a} and these conjoined cratons feature as a major component of the hypothesized Nuna supercontinent (\citealp{Evans2011a, Zhang2012a}; .
@@
 However, with the subsequent rapid motion of Laurentia to low latitudes (Figs.
-\ref{Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny.
+\ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny.
 This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g.
 \citealp{Li2008a}).
-Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift.
+Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and could have been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift.
 The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} was that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
@@
 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean.
-If they indeed became conjoined in Rodinia, the separation of Kalahari from Laurentia likely initiated ca.
-795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a}.
+If they indeed became conjoined in Rodinia, the separation of Kalahari from Laurentia may have initiated ca.
+795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in eastern Laurentia.
 
@@
 
-The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume).
+The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}.
 While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca.
@@
 1110 Ma and moved rapidly with Laurentia as it transited towards the equator.
-\cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
+A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation).
+Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes.
+As additional support for a North China--Laurentia connection,\cite{Zhao2019a} argue similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
 In particular, sediment transport from Laurentia could provide a source for ca.
@@
 1220 Ma dikes pole of the North China craton is not coincident with the ca.
-1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al.
-(this volume) to suggest that North China arrived on the Laurentian margin between ca.
+1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca.
 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
@@
 
-\subsubsection{Siberia}
+%\subsubsection{Siberia}
 
@@
 \label{sec:plate_tectonics}
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{../Figures/baked_contact_timeline_all.pdf}
+\caption{\textbf{Paleomagnetic poles with positive baked contact tests from Laurentia and other cratons}.
+This timeline shows the age of paleomagnetic poles with positive baked contact tests within the Nordic Paleogeography Workshop compilation \citep{Evans2021a}.
+Positive baked contact tests require the presence of an appreciable geomagnetic field.
+In turn, the presence of geomagnetic field requires heat flow across the core mantle boundary that is maintained by plate tectonics, but that would be stifled by a stagnant lid.}
+\label{fig:baked_contact}
+\end{figure*}
 
@@
 From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
-From a geological perspective, the Trans-Hudson orogenic cycle, the Grenville orogenic cycle, and the Appalachian orogenic cycle are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig.
+From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig.
 \ref{fig:tectonic_history}).
 One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics.
-However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}.
+However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there is also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}.
 These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
@@
 Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}.
-Outside of Laurentia, there are examples of ecologites with geochemical affinity to oceanic crust such as that documented in the ca.
+Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca.
 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.
@@
 In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion.
-In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}.
+In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}.
 The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
 \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
-
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{../Figures/baked_contact_timeline_all.pdf}
-\caption{\textbf{Paleomagnetic poles with positive baked contact tests from Laurentia and other cratons}.
-This timeline shows the age of paleomagnetic poles with positive baked contact tests within the Nordic Paleogeography Workshop compilation \citep{Evans2021a}.
-.}
-\label{fig:baked_contact}
-\end{figure*}
-
-An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig.
-\ref{fig:baked_contact}).
-In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.
-Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction.
-Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
-Therefore, the record of abundant positive baked contact tests (Fig.
-\ref{fig:baked_contact}) and coherent paleomagnetic poles (Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
-This interpretation of a significant persistent geomagnetic field throughout the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g.
-\citealp{Macouin2006a}) and elsewhere.
-
-The record of these poles also show that there was progressive motion of Laurentia through the Proterozoic (Figs.
+The record of paleomagnetic poles also show that there was progressive motion of Laurentia through the Proterozoic (Figs.
 \ref{fig:Laurentia_paleolatitude} and \ref{fig:Laurentia_reconstructions}).
@@
 These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid.
-Rather, the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
+Rather, the orogenic interval of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
+
+An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig.
+\ref{fig:baked_contact}).
+In a prolonged stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.
+One way to get insight into the ancient geomagnetic field is through paleointensity experiments on igneous rocks that enable estimates of ancient field strength to be developed.
+Paleointensity data developed from units that are also included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}.
+However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory or non-ideal rock magnetic behavior.
+As a result, it is significantly more challenging to develop reliable paleointensity data than the reliable paleodirectional data used to calculate paleomagnetic poles and the paleointensity database is sparser the compilation of reliable paleomagnetic poles.
+Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests.
+Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction.
+Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
+Therefore, the record of abundant positive baked contact tests (Fig.
+\ref{fig:baked_contact}) and coherent paleomagnetic poles (Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
 
@@
 Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
-Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic interval with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
 While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic intervals bear similarities.
@@
 The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson.
-Break-up following the Grenvillian and Appalachian orogenic cycles occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic cycle have remained sutured.
+Break-up following the Grenvillian and Appalachian orogenic intervals occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic interval have remained sutured.
 As a result, Laurentia has been a formidable continent for the past 1.8 billion years.
@@
 \subsection*{Acknowledgements}
+This work was supported by NSF CAREER Grant EAR-1847277.
+The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Francis Macdonald, Yuem Park and Sarah Slotznick.
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
-Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists as well as additional efforts of Lauri Pesonen in maintaining the Paleomagia database \citep{Veikkolainen2014a}.
+Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists.
 GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}.
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
-The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Yuem Park and Francis Macdonald.
-This work was supported by NSF CAREER Grant EAR-1847277.
 The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0)  in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
@@
 
-\noindent\textbf{allochthonous } An adjective denoting that a rock or terrane originated in a position at significant distance from the lithospheric block that it is currently a part of.
-
-\noindent\textbf{Archean } A geologic eon spanning from 4,000 to 2,500 million years ago (4 to 2.5 Ga)
+\noindent\textbf{allochthonous } An adjective denoting that a rock or terrane originated in a position at significant distance from the lithospheric block where it currently resides.
+
+\noindent\textbf{Archean } A geologic eon spanning from 4,000 to 2,500 million years ago (4 to 2.5 Ga).
 
@@
 
-\noindent\textbf{Canadian shield } Large area of exposed Precambrian rock of Laurentia, or rock covered by thin soil, that is well-exposed due to Pleistocene glaciation erosion
+\noindent\textbf{Canadian shield } The large area of Canada with exposed Precambrian rock, or rock covered by thin soil, that is well-exposed due to Pleistocene glaciation erosion.
 
@@
 
-\noindent\textbf{evaporite } Minerals that crystallize from water that becomes supersaturated in salts through evaporation.
+\noindent\textbf{evaporite } A chemical sedimentary deposit consisting of minerals that crystallize from water that supersaturated in salts due to evaporation.
 
@@
 
-\noindent\textbf{geodynamo } The mechanism whereby convective flow in Earth's fluid outer core generates Earth's magnetic field
-
-\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from southerneastern Ontario, Canada to west Texas, USA.
+\noindent\textbf{geodynamo } The mechanism whereby convective flow in Earth's fluid outer core generates Earth's magnetic field.
+
+\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from southeastern Ontario, Canada to west Texas, USA.
 
@@
 
-\noindent\textbf{juvenile } Rocks that have formed through melt recently extracted from the mantle.
-
-\noindent\textbf{Laurentia } The Precambrian cratonic core of the North America continent and Greenland that formed through the amalgamation of Archean provinces in the Paleoproterozoic.
-
-\noindent\textbf{large igneous province (LIP) } Voluminous and rapidly emplaced volcanics and intrusions that are typically of mafic composition.
+\noindent\textbf{juvenile } An adjuective referring to rocks that have formed through melt recently extracted from the mantle.
+
+\noindent\textbf{Laurentia } The Precambrian cratonic core of the North America continent and Greenland that formed through the amalgamation of Archean provinces in the Paleoproterozoic and subsequent accretion.
+
+\noindent\textbf{large igneous province (LIP) } A region of voluminous and rapidly emplaced volcanics and intrusions that are typically of mafic composition.
 These provinces are often interpreted to be due to decompression melting of an upwelling mantle plume.
@@
 It is framed by a suture with the Hearne province to the north, the Trans-Hudson orogen to the east, and the Great Falls tectonic zone to the south.
-It is also referred to as the Medicine Hat block
+It is also referred to as the Medicine Hat block.
 
@@
 
-\noindent\textbf{Neoproterozoic } A geologic era spanning from 1,000 to 541 million years ago.
+\noindent\textbf{Neoproterozoic } A geologic era spanning from 1,000 to 539 million years ago.
 
@@
 
-\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and broke apart in the middle of the Mesoproterozoic.
-
-\noindent\textbf{orogen } A region of lithosphere that has undergone deformation during a mountain-building event (an orogeny).
-
-\noindent\textbf{paleolatitude } The past latitude of a given point on Earth's surface at a given time.
+\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and to have broken apart in the middle of the Mesoproterozoic.
+
+\noindent\textbf{orogen } A region of lithosphere that has undergone deformation during a mountain-building event (an \textbf{orogeny}).
+
+\noindent\textbf{paleolatitude } The past latitude of a given point on Earth's surface at a given time typically calculated from paleomagnetic data using the geocentric axial dipole hypothesis.
 
@@
 
-\noindent\textbf{Phanerozoic } A geologic eon spanning from 541 million years ago to the present day.
+\noindent\textbf{Phanerozoic } A geologic eon spanning from 539 million years ago to the present day.
 
@@
 
-\noindent\textbf{Precambrian } A commonly used informal term to refer to geologic time prior to the Cambrian Period that started 541 million years ago.
-
-\noindent\textbf{Proterozoic } A geologic eon spanning from 2,500 to 541 million years ago.
+\noindent\textbf{Precambrian } A commonly used informal term to refer to geologic time prior to the Cambrian Period that started 539 million years ago.
+
+\noindent\textbf{Proterozoic } A geologic eon spanning from 2,500 to 539 million years ago.
 
@@
 The term is used in this chapter to refer to Archean provinces that moved as independent cratonic blocks prior to Laurentia's amalgmation (e.g.
-Superior province).
+the Superior province).
 It is also used to refer to zones of crustal growth associated with orogens and the products of contemporaneous magmatic activity (large igneous provinces).
 
-\noindent\textbf{Rae province } An Archean province of Laurentia extending the region of Lake Athabasca northeast to northern Baffin Island in arctic Canada.
+\noindent\textbf{Rae province } An Archean province of Laurentia extending from the region of Lake Athabasca northeast to northern Baffin Island in arctic Canada.
 It is framed by the Thelon orogen to the west, the Taltson orogen to the southwest and the Hearne province to the east.
@@
 
-\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era and broke apart in the middle of the Neoproterozoic.
+\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era at the time of the Grenvillian orogeny and to have broken apart in the middle of the Neoproterozoic.
 
@@
 
-\noindent\textbf{Shawinigan orogen } The orogen resulting from the Mesoproterozoic Shawinigan orogeny when there was accretion of terranes to eastern Laurentia.
+\noindent\textbf{Shawinigan orogen } An orogen resulting from the Mesoproterozoic Shawinigan orogeny when there was accretion of terranes to eastern Laurentia.
 
@@
 The supercontinent Pangea that existed ca.
-200 Ma is the archtypal supercontinent.
+200 Ma is the archetypal supercontinent.
 A threshold of 75$\%$ of extant continental crust has been proposed for a continent to be considered a supercontinent \citep{Meert2012a}.
-Gondwana (Australia + India + Africa + South America) constituted $\sim$60$\%$ and was a constituent of Pangea such that \cite{Evans2016a} proposed that it and similar landmasses should be called \textbf{semi-supercontinents}.
-
-\noindent\textbf{supercraton } A landmass that subsequently split into constituent crustal provinces. The term is typically applied to Archean provinces (cratons).
+Gondwana (Australia + India + Africa + South America) constituted $\sim$60$\%$ of continental lithosphere and was a constituent of Pangea such that \cite{Evans2016a} proposed that it and similar landmasses should be called \textbf{semi-supercontinents}.
+
+\noindent\textbf{supercraton } A landmass that subsequently split into constituent crustal provinces. The term is typically applied to groupings of Archean provinces (cratons).
 
@@
 
-\noindent\textbf{thermal remanent magnetization } Magnetization acquired by magnetic minerals rocks as they cool following crystallization from magma.
+\noindent\textbf{thermal remanent magnetization } Magnetization acquired by magnetic minerals in rocks as they cool typically following crystallization from magma.
 

### 2020-09-28 Manuscript/Laurentia_Paleogeo_Manuscript.tex
small typo fixes caught by Yiming

@@
 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly (Fig.
-\ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}.
+\ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}).
 Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
@@
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
-The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a}.}
+The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a} while the diamond (not rated -- NR) poles are discussed in the text.}
 \label{fig:Laurentia_map}
@@
 Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca.
-1.71 to 1.60 Ma (Fig.
+1.71 to 1.60 Ga (Fig.
 \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).
@@
 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}.
-However, this extensional basin development predates the rifting that that is interpreted to have been followed by passive margin thermal subsidence closer to the ca.
+However, this extensional basin development predates the rifting that is interpreted to have been followed by passive margin thermal subsidence closer to the ca.
 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}.
@@
 
-The eastern margin of Laurentia then went through the multiple phases of Appalachian orogenesis.
+The eastern margin of Laurentia then went through multiple phases of Appalachian orogenesis.
 As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano orogenies in the Grenville interval; Acadian orogeny in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian orogeny in the Appalachian interval).
@@
 
-Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic era where there are inconsistencies between poles of similar age (Figs.
+Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic Era where there are inconsistencies between poles of similar age (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
@@
 1000 Ma), which is a striking departure from the paleomagnetic record and standard paleogeographic models.
-Additionally, the implemented plate motion of Laurentia in the \cite{Merdith2017b} model strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating (Fig.
+Additionally, the implemented plate motion of Laurentia in the \cite{Merdith2017b} model strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \citep{Denyszyn2009b} despite its `A' Nordic rating (Fig.
 \ref{fig:Laurentia_paleolatitude}).
@@
 In the Tonian period, evaporite deposition may have occurred equatorward of 10\textdegree\ (Fig.
-\ref{fig:Laurentia_evaporites} which may reflect increased aridity in the tropical interior of the Rodinia supercontinent.
+\ref{fig:Laurentia_evaporites}) which may reflect increased aridity in the tropical interior of the Rodinia supercontinent.
 
@@
 Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes.
-As additional support for a North China--Laurentia connection,\cite{Zhao2019a} argue similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
+As additional support for a North China--Laurentia connection, \cite{Zhao2019a} argue similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
 In particular, sediment transport from Laurentia could provide a source for ca.

### 2020-10-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
write Siberia connection section and make progress on Australia

@@
 There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig.
-\ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin.
-The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
+\ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin.
+There are thick successions of Paleoproterozoic sediments such as the ca.
+1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada that comprises $\sim$14 km of  fine-grained siliciclastic and carbonate rocks\citep{Furlanetto2016a}.
+This succession is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}.
+These metasedimentary rocks were metamorphosed during the ca.
+1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent \citep{Furlanetto2013a, Furlanetto2016a}.
+Further south along Laurentia's western margin, the 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
 1.47 to 1.40 rift \citep{Evans2000c}.
@@
 \caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic}.
+The hypothesized connection between Siberia and Laurentia is implemented here following \citep{Evans2016b}.
 This reconstruction is modified from \cite{Swanson-Hysell2019a}.
@@
 It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia.
-Such correlations has led to a number of different proposed connections that place Australia + East Antarctica...
+Such correlations has led to a number of different proposed connections that place Australia + East Antarctica along the western margin of Laurentia.
+
+Metamorphism ca.
+1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to the be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}.
+That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}.
+The model of \cite{Nordsvan2018a} the inlier to have been a continental ribbon that rifted from Laurentia ca.
+1.68 Ga and was then caught up in ca.
+1.60 Ga collision between the North Australia craton and northwest Laurentia.
+The ca.
+1.60 Ga Racklan-Forward orogeny recorded in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Furlanetto2013a}.
+
+Metamorphism a. 1.60 Ga Paleoproterozoic
 
@@
 
-%\subsubsection{Siberia}
+\subsubsection{Siberia}
+
+Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoporterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}.
+U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous province  between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca.
+1.8 Ga all the way up to the time of the ca.
+720 Ma Franklin large igneous province (LIP).
+Mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}.
+Comparisons of paleomagnetic poles between Laurentia and Siberia support such a tight and internally stable fit between southern Siberia and northern Laurentia from ca.
+1.64 to at least ca.
+0.76 Ga  \citep{Evans2016b}.
+Overlap in Laurentia and Siberia pole positions with such a reconstruction are achieved in the Statherian Period of the Paleoproterozoic (the ca.
+1.64 Ga Nersa complex of Siberia compared to the Melville Bugt dikes of Laurentia), the Calymmian Period of the Mesoproterozoic (the ca.
+1.50 Ga Anabar intrusions with the ca.
+1.48 Ga St.
+Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.
+Kitoi pole of Siberia with Tonian Laurentia poles).
+The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}).
+This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca.
+1100 Ma.
+
+Connections may have existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}.
+The emplacement of the Franklin LIP ca.
+720 Ma could have initiated break-up of a joint Laurentia-Siberia continent ending the long Proterozoic connection between the cratons.
 

### 2020-10-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
draft of Australia - Laurentia connection section

@@
 This succession is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}.
-These metasedimentary rocks were metamorphosed during the ca.
-1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent \citep{Furlanetto2013a, Furlanetto2016a}.
+These metasedimentary rocks were deformed metamorphosed during the ca.
+1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent with associated orogenesis (referred to as the Forward orogeny) further into the craton\citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}.
 Further south along Laurentia's western margin, the 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
@@
 \includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
-\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic}.
-The hypothesized connection between Siberia and Laurentia is implemented here following \citep{Evans2016b}.
-This reconstruction is modified from \cite{Swanson-Hysell2019a}.
-The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
+\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}.
+The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interprets this relationship as persistent from 1.7 to 0.7 Ga.
+The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny.
+The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a}.
+This configuration back to ca.
+1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt.
+Isa dikes pole from North Australia and is consistent with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca.
+1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca.
+1070 Ma.
+The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
 \label{fig:Grenville_reconstructions}
@@
 It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia.
-Such correlations has led to a number of different proposed connections that place Australia + East Antarctica along the western margin of Laurentia.
-
+Such correlations have led to a number of different proposed connections that place Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic.
 Metamorphism ca.
@@
 That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}.
-The model of \cite{Nordsvan2018a} the inlier to have been a continental ribbon that rifted from Laurentia ca.
+In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca.
 1.68 Ga and was then caught up in ca.
@@
 The ca.
-1.60 Ga Racklan-Forward orogeny recorded in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Furlanetto2013a}.
-
-Metamorphism a. 1.60 Ga Paleoproterozoic
-
-%Goodge2017a
-%
-%The 
-%
-%\textit{Goodge2008a}
-
-%Could discuss Gibson2019b model related to hypothesized Paleoproterozoic tectonic history
+1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Thorkelson2005a, Furlanetto2013a}.
+This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations.
+Detrital zircons dating from ca.
+1.61 to 1.50 Ga in the ca.
+1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}.
+Additionally, ca.
+1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}.
+The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet).
+The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}.
+Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}.
+A comparison between paleomagnetic data from the ca.
+1.32 Ga Derim Derim sills of the North Australia craton and the ca.
+1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}.
+A paleomagnetic pole from the ca.
+1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}.
+There is not a coeval pole from Laurentia for comparison at ca.
+1.21 Ga and data do indicate poleward motion for Laurentia between the ca.
+1.24 and 1.18 Ga constraints.
+Nevertheless, comparison between latest Mesoproterozoic (ca.
+1070 Ma) and Neoproterozoic (ca.
+750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit.
+Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}.
+There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca.
+1070 Ma and ca.
+770 Ma poles \citep{Swanson-Hysell2012a}.
+This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca.
+650 Ma \citep{Li2011a}.
 
@@
 1.48 Ga St.
-Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.
+Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.
 Kitoi pole of Siberia with Tonian Laurentia poles).
@@
 This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca.
-1100 Ma.
+1100 Ma (Fig.
+\ref{fig:Grenville_reconstructions}).
 

### 2020-10-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits to Australia section and a commit of a compiled PDF

@@
 The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny.
-The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a}.
+The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia.
 This configuration back to ca.
 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt.
-Isa dikes pole from North Australia and is consistent with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca.
+Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca.
 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca.
@@
 
-It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia.
-Such correlations have led to a number of different proposed connections that place Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic.
+It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia \citep{Moores1991a}.
+The extent of Antarctic lithosphere that was conjoined with Antarctica prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009a}.
+Separation between these Antarctic Provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway.
+Correlations with Laurentia have led to a number of different proposed reconstructions of Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic.
+
 Metamorphism ca.
-1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to the be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}.
+1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}.
 That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}.
@@
 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}.
-A paleomagnetic pole from the ca.
+If the North Australian craton was continous with the South Australia his interpretation would have the ca.
+1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca.
+1.35 Ga East Kootenay orogeny.
+Another challenge with interpreting data from North Australia A paleomagnetic pole from the ca.
 1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}.
@@
 
-There is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga.
+Even without considering other continents, there is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga.
 The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well.

### 2020-10-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
Baltica connection edits

@@
 This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
-In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogeny of Baltica is a continuation of the the Mazatzal-Labradorian orogenies of Laurentia\citep{Karlstrom2001a}.
-Accretionary orogenesis within the Gothian orogeny was active ca.
-1.66 to 1.52 Ga \citep{Bergstrom2020a}.
+In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca.
+1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia\citep{Karlstrom2001a}.
 Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}.
@@
 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca.
-1.12 Ga if a virtual geomagnetic pole from
-the Salla dike is considered to be representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
+1.12 Ga if a virtual geomagnetic pole from the Salla dike is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
 These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.
@@
 Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic.
-This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b}.
-The timing of the main phase of the Sveconorwegian orogeny from ca.
+This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}.
+However, as geochronology has improved the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.
+
+In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny was from ca.
 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}.
@@
 1.06 Ga.
+
+Another basis for a continued connection between
 

### 2020-10-15 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits/additions to Baltica connection text

@@
 The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
-Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak which could support a deviation from stable axial dipolar geomagnetic field behavior at the time \citep{Thallner2020a}.
+Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak \citep{Thallner2020a} as are data from coeval volcanics in Ukraine \citep{Shcherbakova2019a} which could support an anomalous deviation from stable axial dipolar geomagnetic field behavior at the time as interpreted by \cite{Halls2015a}.
 Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig.
@@
 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca.
-1.12 Ga if a virtual geomagnetic pole from the Salla dike is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
+1.12 Ga if a virtual geomagnetic pole from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
 These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.
@@
 This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}.
-However, as geochronology has improved the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.
-
-In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny was from ca.
+However, as geochronology has improved, the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.
+
+In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny from ca.
 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}.
-However, it is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}.
+It is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}.
 \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca.
@@
 
-Another basis for a continued connection between
-
-A clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g.
-\citealp{Evans2009a})...
-
-% orogen in Labrador         as advocated by \cite{Gower1990b} 
-%
-%Tonian period (ca.
-975 to 915 Ma; \citealp{McClelland2019a}) intrusive rocks in Svalbard and East Greenland are interpreted to be related to subduction and accretionary orogenesis that has been termed the Valhalla orogeny \citep{Cawood2010a}.
-%
-%The NENA connection 
-%
-% -- a constraint that is honored in . This reconstruction is brings paleomagnetic poles
-%
-%compatible with existing paleomagnetic constraints from ca.
-1750 to 1270 Ma \cite{Evans2008a} and these conjoined cratons feature as a major component of the hypothesized Nuna supercontinent (\citealp{Evans2011a, Zhang2012a}; .
+Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related terranes including East Svalbard \citep{McClelland2019a}.
+These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
+The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation \citep{Cawood2010a}.
+Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g.
+\citealp{Evans2009a} and as shown in Fig.
+\ref{fig:Grenville_reconstructions}).
+A challenge with this model is it is unclear what was the conjugate lithosphere that rifted from East Greenland leading to thick sedimentary successions interpreted to have been deposited on a thermally-subsiding passive margin from ca.
+850 Ma into the Cryogenian (Fig.
+\ref{fig:tectonic_history}; \citealp{Maloof2006a}).
+\cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting although the paucity of tuffs within successions such as the Akademikerbreen Group of East Svalbard is difficult to reconcile with such a setting.
+The separation of Baltica from Laurentia is hypothesized to have occurred ca.
+600 Ma associated with the opening of the Iapetus Ocean (Fig.
+\ref{fig:tectonic_history}; \citep{Cawood2001a}).
 

### 2020-10-21 Manuscript/Laurentia_Paleogeo_Manuscript.tex
develop a brief introduction prior to splitting out \subsubsection{Laurentia's initial formation}

@@
 Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca.
-1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly (Fig.
+1.8 billion years ago \[Ga\]), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig.
 \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}).
-Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}.
+That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents.
+In contrast, South America and Africa are the product of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca.
+0.6 Ga \citep{Goscombe2019a}.
+Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca.
+0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}.
+The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly.
+Rich geologic, paleomagnetic and geochronologic data give deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.
+
+\subsubsection{Laurentia's initial formation}
+Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig.
+\ref{fig:Laurentia_map}; \citealp{Corrigan2009a}).
 Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca.
@@
 This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints.
-Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
+Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca.
@@
 Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
-1.86 to 1.82 \citep{Schulz2007a}.
+1.86 to 1.82 (Fig.
+\ref{fig:Laurentia_map}; \citealp{Schulz2007a}).
 Firm evidence of the end of the Penokean orogeny comes from the ca.

### 2020-10-21 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add subsubheadings and make small edits through tectonic history intro

@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
+\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}.
+Proterozoic orogens are labeled with \textit{italicized text} (or.
+-- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift).
+The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
+The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a} while the diamond (not rated -- NR) poles are discussed in the text.}
+\label{fig:Laurentia_map}
+\end{figure*}
+
 \subsubsection{Laurentia's initial formation}
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
-\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}.
-Proterozoic orogens are labeled with \textit{italicized text} (or.
--- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift).
-The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
-The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a} while the diamond (not rated -- NR) poles are discussed in the text.}
-\label{fig:Laurentia_map}
-\end{figure*}
+\subsubsection{Protracted Proterozoic accretionary growth followed by collisional orogenesis}
 
@@
 
-The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}.
+The collisions of provinces and terranes leading up to the Trans-Hudson orogeny are interpreted to be associated with the initial assembly of the supercontinent Nuna in some paleogeographic models (e.g.
+\citealp{Zhang2012a,Pehrsson2015a}.
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
@@
 \ref{fig:Superior_Slave_recons}).
-The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
+The paleogeographic model of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
 In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography.
@@
 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
-This plutonism is interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region Laurentia's long-live active margin \citep{Bickford2015a}.
+This plutonism is interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-live active margin \citep{Bickford2015a}.
 Metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca.
@@
 The subsequent ca.
-1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of terrane comprised of a previously rifted fragment of Laurentia and led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}.
+1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of a previously rifted fragment of Laurentia that led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}.
 The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
@@
 There are thick successions of Paleoproterozoic sediments such as the ca.
-1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada that comprises $\sim$14 km of  fine-grained siliciclastic and carbonate rocks\citep{Furlanetto2016a}.
+1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada that comprises $\sim$14 km of  fine-grained siliciclastic and carbonate rocks \citep{Furlanetto2016a}.
 This succession is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}.
-These metasedimentary rocks were deformed metamorphosed during the ca.
-1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent with associated orogenesis (referred to as the Forward orogeny) further into the craton\citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}.
+These metasedimentary rocks were deformed and metamorphosed during the ca.
+1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent with associated orogenesis (referred to as the Forward orogeny) further into the craton \citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}.
 Further south along Laurentia's western margin, the 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
@@
 
+\subsubsection{Neoproterozoic rifting}
+
 The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig.
@@
 The emplacement of the ca.
-780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia).
-If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
+780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia; \citealp{Li2008a}).
+If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
 The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
@@
 In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
-Tectonic extension is recorded in units of the lower Fifteenmile Group with maximum depositional ages of ca.
+Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca.
 1050 Ma with ongoing basin development ca.
@@
 This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys).
-Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia margin ca.
+In this mode, subsequent to this rifting of the major continental blocks, smaller terranes rift off the east Laurentia margin ca.
 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.
+
+\subsubsection{Similarities in Laurentia's Proterozoic and Phanerozoic tectonic histories}
 

### 2020-11-03 Manuscript/Laurentia_Paleogeo_Manuscript.tex
manuscript updates

@@
 Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca.
-1.8 billion years ago \[Ga\]), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig.
+1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig.
 \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}).
@@
 Proterozoic orogens are labeled with \textit{italicized text} (or.
--- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift).
+-- orogen; THO  -- Trans-Hudson orogen; GFTZ -- Great Falls Tectonic Zone; MCR -- Midcontinent Rift).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 
-As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca.
+As for the suturing of the Wyoming province to Laurentia (Fig.
+\ref{fig:Laurentia_map}), many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca.
 1.80 Ga \citep{Whitmeyer2007a}.
@@
 1.72 Ga \citep{Kilian2016b}.
-This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
+This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca.
 1.75 to 1.72 Ga \citep{Condit2015a} and ca.
@@
 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.
-Overall, the collision of these Archean microcontinents between ca.
-1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig.
-\ref{fig:Laurentia_map}).
+The Mojave province of southwestern Laurentia (Fig.
+\ref{fig:Laurentia_map}) comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}.
+more enigmatic both in terms of whether it constituents a province of Archean lithosphere or whether it should
+
+%ADD MOJAVE PROVINCE TEXT
+Overall, the collision of Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).
 
@@
 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
-Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca.
+Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeast Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca.
 1.71 to 1.60 Ga (Fig.
@@
 
-Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
-An interval of major plutonism occurred ca.
-1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
-This plutonism is interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-live active margin \citep{Bickford2015a}.
-Metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca.
+Laurentia's growth continued into the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
+Continental arc magmatism is interpreted to have occurred associated with the Pinwarian orogeny in the northeast Grenville province in Labrador from ca.
+1.52 to 1.46 Ga \citep{Gower2002a}.
+Accretionary orogenesis recorded in the Grenville province includes accretion of the Quebecia composite arc terrane to Laurentia ca.
+1.43 to 1.37 Ga \citep{Groulier2020a}.
+Far to the southwest along the margin, metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca.
 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}.
+In the midcontinent region,  deformation and metamorphism of post-Mazatzal orogeny sedimentary rocks is constrained to have occurred ca.
+1.49 to 1.46 Ga associated with the Baraboo orogeny \citep{Medaris2003a, Holm2019a}.
+Picuris to Baraboo to Pinwarian orogenesis indicates convergent tectonism along the entirety of southeast Laurentia.
+It was in this context of an active margin that there was major plutonism ca.
+1.48 to 1.35 Ga that resulted in the A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
+This plutonism has been interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-live active margin \citep{Bickford2015a}.
 Younger magmatic activity (ca.
 1.37 Ga) of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}.
-While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature and is consistent with evidence for accretionary orogenesis in the Picuris, Baraboo and Pinware  orogens, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 This region is interpreted to have been subsequently deformed during a ca.
-1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
-
-This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions.
+1.37 to 1.33 Ga event known as the East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates\citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+
+Taken together, this late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions.
 In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca.
@@
 \ref{fig:tectonic_history}).
-Along the western margin of Laurentia, rifting occurred ca.
-780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}.
-However, this extensional basin development predates the rifting that is interpreted to have been followed by passive margin thermal subsidence closer to the ca.
-539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}.
+Along the western margin of Laurentia, by ca.
+760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwest Laurentia to the Mackenzie Mountains of northwest Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}.
 The emplacement of the ca.
-780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia; \citealp{Li2008a}).
-If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
-The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
-The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
+780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia; \citealp{Li2008a}).
+Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca.
+717 to 656 Ma) and Marinoan (ca.
+645 to 635 Ma) glacial deposits in the Death Valley region of southwest Laurentia\citep{Nelson2020a}.
+Additionally, Cryogenian volcanics along the western Laurentia margin (e.g.
+\citealp{Eyster2018a}) are interpreted to be the result of active rifting.
+A puzzlingly feature of this record of active rifting is that it significant predates interpreted passive margin thermal subsidence closer to the ca.
+539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}.
+If the interpretation of a conjugate continent rifting of the margin prior to the ca.
+717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
+While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
+This record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models).
-Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca.
-780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard.
-In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
+Another possibility, along the lines of that proposed in \citet{Ross1991a} and \citet{Colpron2002a}, is that ca.
+760 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further to the west.
+In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
 
@@
 760 to 680 Ma in the region of modern-day North Carolina and Virginia.
-Rifting ca.
-620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca.
+Rifting initiated in the region from modern-day New York to Newfoundland ca.
+620 to 580 Ma and by ca.
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
@@
 \includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
-\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates).
-Bottom panel: Paleomagnetic data for Archean provinces prior to the amalgamation of Laurentia.}
+\caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland).
+Portions of the apparent polar wander path (APWP) are referred to by the names Logan loop, Keweenawan track, and Grenville loop in the literature and those are labeled next to the associated poles.
+Upper-right panel:  Paleomagnetic poles from 1800 to 580 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates).
+The youngest poles from the Ediacaran Period have unusually variable positions as discussed in the text.
+Lower-left panel: Running mean APWP calculated with a 20 million year moving window.
+Lower-right panel: Poles for the Archean provinces of Laurentia prior to Laurentia's Paleoproterozoic amalgamation.}
 \label{fig:Laurentia_poles}
@@
 The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
-Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
+I refer to additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR').
 These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported.
@@
 In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2).
-An additional not-rated pole included in the present compilation is the new pole for the ca.
+An additional not-rated pole included in the present compilation is a more recently published pole for the ca.
 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca.
@@
 1.4 Ga Belt Supergroup and the ca.
-1.1 Ga Midcontinent Rift (Fig.
-\ref{fig:tectonic_history}).
+1.1 Ga Midcontinent Rift (Figs.
+\ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
 The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}.
@@
 Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton.
-There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation.
+There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation.
 However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton.
@@
 Data from the ca.
-585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
+585 Ma Grenville dyke swarm of Laurentia, that are interpreted as primary, reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}.
 The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
@@
 \includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
-\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes.
+\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (present-day coordinates of lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes.
 The paleomagnetic poles are compiled in Table 2.
@@
 A synthesized pole path for Laurentia is developed here and used to develop a paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles.
-The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
+The paleolatitude implied by this continuous simple Laurentia pole interpolation model is shown in Figure \ref{fig:Laurentia_paleolatitude}.
 This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example).
@@
 This method follows the approach taken in the Phanerozoic (e.g.
-\citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps.
+\citealp{Torsvik2012a}) wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps.
 When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model.
@@
 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
-The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China (albeit one that needs to be revisited given updates to the paleomagnetic and geochronologic record from South China; \citealp{Zhang2021a}).
+The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to partially incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China (albeit one that needs to be revisited given updates to the paleomagnetic and geochronologic record from South China; \citealp{Zhang2021a}).
 
@@
 While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block.
-Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes with specified paleolongitudes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}.
-In Figure \ref{fig:Laurentia_reconstructions}, the map projections are centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig.
+Other approaches to obtain paleolongitude utilize geodynamic hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes with specified paleolongitudes that switch through time in conjunction with hypothesized supercontinent cyclicity \citep{Mitchell2012a}.
+In Figure \ref{fig:Laurentia_reconstructions}, the map projections are centered on the longitudinal position of Duluth with the orientation and paleolatitude of Laurentia being constrained by the paleomagnetic pole compilation as synthesized in the simple pole interpolation model (Fig.
 \ref{fig:Laurentia_paleolatitude}).
@@
 \includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
-\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
+\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
 Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.
@@
 
-Developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types.
+Developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data.
 Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles.
@@
 The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
-As noted above, the development of such models is challenging and the researchers need to balance varying constraints.
+As noted above, the development of such models is challenging and  researchers need to balance varying constraints.
 The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia.
@@
 This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again.
+The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}.
+Over much of the record, Laurentia's pole positions can be satisfied through motion of the content at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig.
+\ref{fig:Laurentia_paleolatitude}).
 
@@
 Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation.
-Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003b}.
+Interpretations of paleolatitude based on glacial deposits during the Proterozoic are complicated by the evidence for multiple global and low-latitude glacial intervals associated with the Snowball Earth climate state \citep{Evans2003b}.
 Evaporite deposits are particularly compelling as paleolatitude constraints given that their deposition is interpreted to be associated with arid regions resulting from large-scale Hadley cell downwelling \citep{Evans2006a}.
@@
 In Figure \ref{fig:Laurentia_evaporites}, the paleolatitude of these evaporite deposits are reconstructed using the simple Laurentia model developed in this work as well as with the \cite{Li2013a} model for the late Mesoproterozoic to Neoproterozoic.
-The position of major Phanerozoic evaporite basins of North America are also shown with their paeleolatitude reconstructed with the paleogeographic mode of \cite{Torsvik2017a}.
+The position of major Phanerozoic evaporite basins of North America are also shown with their paleolatitude reconstructed with the paleogeographic model of \cite{Torsvik2017a}.
 These paleogeographic models reconstruct evaporite deposition to have been within 30\textdegree\ of the equator in both the Phanerozoic and Proterozoic.
@@
 Many different paleogeographic connections between Laurentia and other Proterozoic cratons have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces.
-This section is not comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models --- particularly those following the assembly of Laurentia.
-
-\textit{Note that this section is currently under development and is incomplete}
+This section is not comprehensive in terms of proposed connections, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
+These connections are often discussed in the context of hypothesized supercontinents given the hypothesis that Laurentia was an important constituent of Nuna, following ca.
+1.85 Ga Trans-Hudson orogenesis, and of Rodinia, following ca.
+1.05 Ga Grenvillian orogenesis.
 
@@
 In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers.
-This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Hoffman1991a, Evans2013b, Cawood2017a}.
+This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig.
+\ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}).
 While the lack of ca.
@@
 It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia \citep{Moores1991a}.
-The extent of Antarctic lithosphere that was conjoined with Antarctica prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009a}.
+The extent of Antarctic lithosphere that was conjoined with Antarctica prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009b}.
 Separation between these Antarctic Provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway.

### 2020-11-04 Manuscript/Laurentia_Paleogeo_Manuscript.tex
Update Laurentia_Paleogeo_Manuscript.tex

@@
 
-\subsubsection{Paleogeographic connections prior to Laurentia assembly}
+\subsubsection{Paleogeographic connections prior to initial Laurentia assembly}
 
@@
 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia.
+A significantly later separation ca.
+600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until ca.
+620 Ma north of the New York promontory.
 
@@
 Metamorphism ca.
-1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}.
+1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}.
 That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}.
@@
 1.68 Ga and was then caught up in ca.
-1.60 Ga collision between the North Australia craton and northwest Laurentia.
+1.60 Ga collision between the North Australia craton and northwest Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}.
 The ca.
-1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Thorkelson2005a, Furlanetto2013a}.
+1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}.
 This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations.

### 2020-11-04 Manuscript/Laurentia_Paleogeo_Manuscript.tex
Update Laurentia_Paleogeo_Manuscript.tex

@@
 A significantly later separation ca.
-600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until ca.
+600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until after  ca.
 620 Ma north of the New York promontory.
@@
 Metamorphism ca.
-1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}.
+1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to be the  result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}.
 That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}.
@@
 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}.
-If the North Australian craton was continous with the South Australia his interpretation would have the ca.
+If the North Australian craton was continuous with the South Australia his interpretation would have the ca.
 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca.
 1.35 Ga East Kootenay orogeny.
-Another challenge with interpreting data from North Australia A paleomagnetic pole from the ca.
-1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}.
+A paleomagnetic pole from the ca.
+1.21 Ga Gnowangerup-Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia ca.
+1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}.
 There is not a coeval pole from Laurentia for comparison at ca.
@@
 In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca.
-1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia\citep{Karlstrom2001a}.
+1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia \citep{Karlstrom2001a}.
 Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}.
@@
 Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic.
-This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}.
+This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica \citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}.
 However, as geochronology has improved, the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.
@@
 1.06 Ga.
+In this model, Baltica could have been on the same plate as Laurentia during the rapid late Mesoporoterozoic mortion leading up to collisional Grenvillian orogenesis and the associated assembly of Rodinia (Fig.
+\ref{fig:Grenville_reconstructions}).
 
@@
 These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
-The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation \citep{Cawood2010a}.
+The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}.
 Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g.
@@
 \ref{fig:Grenville_reconstructions}).
-A challenge with this model is it is unclear what was the conjugate lithosphere that rifted from East Greenland leading to thick sedimentary successions interpreted to have been deposited on a thermally-subsiding passive margin from ca.
+This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica's APWP that would have Laurentia journey after the Grenville loop northward across the equator and back south to a similar Grenville loop position prior to the ca.
+780 Ma Laurentia poles \citep{Evans2015a,Fairchild2017a}.
+A challenge with this model in terms of regional tectonics, is that it is unclear what conjugate lithosphere rifted from East Greenland leading to the thick sedimentary succession interpreted to have been deposited on a thermally-subsiding passive margin from ca.
 850 Ma into the Cryogenian (Fig.
@@
 \cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting although the paucity of tuffs within successions such as the Akademikerbreen Group of East Svalbard is difficult to reconcile with such a setting.
-The separation of Baltica from Laurentia is hypothesized to have occurred ca.
+The separation of Baltica from Laurentia, once in the configuration shown in Figure \ref{fig:Grenville_reconstructions}, is hypothesized to have occurred ca.
 600 Ma associated with the opening of the Iapetus Ocean (Fig.
-\ref{fig:tectonic_history}; \citep{Cawood2001a}).
+\ref{fig:tectonic_history}; \citealp{Cawood2001a}).
+The cratons would once again be conjoined ca.
+430 Ma during their collision that led to the Caledonian orogeny resulting in a continent that is referred to as Laurussia \citep{Torsvik2017a}.
 
@@
 Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and could have been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift.
-The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} was that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
+The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} is that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
 \citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia.
-The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them become conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}.
+The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them becoming conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}.
 The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca.
@@
 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean.
-If they indeed became conjoined in Rodinia, the separation of Kalahari from Laurentia may have initiated ca.
+If they indeed became conjoined in Rodinia as in Figure \ref{fig:Grenville_reconstructions}, the separation of Kalahari from Laurentia may have initiated ca.
 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in eastern Laurentia.
+This late Tonian timing for separation could be  consistent with a position along the southeast Laurentia margin given the ca.
+750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Cuyania \citep{Martin2019a}.
 
@@
 A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation).
-Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes.
-As additional support for a North China--Laurentia connection, \cite{Zhao2019a} argue similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction.
+Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation.
+As additional support for a North China--Laurentia connection, \cite{Zhao2019a} point to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction.
 In particular, sediment transport from Laurentia could provide a source for ca.

### 2020-11-05 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits prior to submission of revisions

@@
 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.
-The Mojave province of southwestern Laurentia (Fig.
-\ref{fig:Laurentia_map}) comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}.
-more enigmatic both in terms of whether it constituents a province of Archean lithosphere or whether it should
 
@@
 Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}.
+Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia (Fig.
+\ref{fig:Laurentia_map}).
+The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}.
+It is unclear whether the Mojave province should be considered to be a distinct province of Archean lithosphere akin to other Archean provinces or rather it should dominantly be considered to be Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}.
 Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 A significantly later separation ca.
-600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until after  ca.
+600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of rifting until after ca.
 620 Ma north of the New York promontory.
@@
 
-Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoporterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}.
-U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous province  between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca.
+Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoproterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}.
+U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous provinces between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca.
 1.8 Ga all the way up to the time of the ca.
 720 Ma Franklin large igneous province (LIP).
-Mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}.
-Comparisons of paleomagnetic poles between Laurentia and Siberia support such a tight and internally stable fit between southern Siberia and northern Laurentia from ca.
+For example, mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}.
+Comparisons of paleomagnetic poles between Laurentia and Siberia support this tight and internally stable fit between southern Siberia and northern Laurentia from ca.
 1.64 to at least ca.
@@
 
-Connections may have existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}.
+Connections between Archean provinces may have also existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}.
 The emplacement of the Franklin LIP ca.
@@
 From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
-From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig.
+From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary orogenesis followed by collisional orogenesis (Fig.
 \ref{fig:tectonic_history}).
 One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics.
-However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there is also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}.
+However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonicsas as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}.
 These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
@@
 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.
+Mesoproterozoic ophiolites were also obducted to Laurentia such as the Pyrites ophiolite complex within the Shawinigan orogen and the Coal Creek Domain of the Llano uplift \citep{Chiarenzelli2011a, McLelland2013a}.
 
@@
 These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid.
-Rather, the orogenic interval of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
+Rather, the orogenic interval of the Mesoproterozoic (1.6 to 1.0 Ga) bears similarity with that of the Paleozoic (0.54 to 0.25 Ga) and reveals Laurentia to have been a central player in amalgamation of continents associated with both Rodinia and Pangea.
 
@@
 Paleointensity data developed from units that are also included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}.
-However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory or non-ideal rock magnetic behavior.
-As a result, it is significantly more challenging to develop reliable paleointensity data than the reliable paleodirectional data used to calculate paleomagnetic poles and the paleointensity database is sparser the compilation of reliable paleomagnetic poles.
+However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory hearing or non-ideal rock magnetic behavior.
+As a result, it is significantly more challenging to develop reliable paleointensity data than to develop the reliable paleodirectional data used to calculate paleomagnetic poles.
+Therefore, the paleointensity database is sparser the compilation of reliable paleomagnetic poles.
 Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests.
 Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction.
-Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
+Additionally, since paleomagnetic poles are typically developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
 Therefore, the record of abundant positive baked contact tests (Fig.
@@
 
-The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles.
+The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological data on tectonism and the record of paleomagnetic poles.
 Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny.
@@
 The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles.
-Following, a return to high latitudes as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent.
-Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era.
+Following a return to high latitudes, as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent.
+Rifting and passive margin development then isolated Laurentia in late Ediacaran period and into the early Paleozoic Era.
 Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic interval with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
@@
 In each case, accretionary collision of arc terranes was followed by continent-continent collision.
-The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson.
+The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson orogeny.
 Break-up following the Grenvillian and Appalachian orogenic intervals occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic interval have remained sutured.
@@
 This work was supported by NSF CAREER Grant EAR-1847277.
-The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Francis Macdonald, Yuem Park and Sarah Slotznick.
+The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Francis Macdonald, Yuem Park, Sarah Slotznick and Justin Strauss.
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
@@
 
+\noindent\textbf{Cryogenian Period } The geologic period that lasted from ca.
+717 to 635 million years ago during which time there were two global glaciations.
+The start of the period is provisionally defined as the first evidence of low-latitude glaciation.
+It is the second geologic period of the Neoproterozoic Era being preceded by the Tonian Period and followed by the Ediacaran Period.
+
+\noindent\textbf{Ediacaran Period } The third geologic period of the Neoproterozoic Era from ca.
+635 to 539 Ma million years ago.
+It is the final period of the Proterozoic Eon and is followed by the Cambrian Period.
+
 \noindent\textbf{Elzevirian orogen } The orogen resulting from the Mesoproterozoic Elzevirian orogeny when there was accretion of arc terranes to eastern Laurentia.
@@
 \noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and to have broken apart in the middle of the Mesoproterozoic.
+
+\noindent\textbf{ophiolite } Oceanic lithosphere that as been accreted onto continental lithosphere.
 
@@
 \noindent\textbf{thermal remanent magnetization } Magnetization acquired by magnetic minerals in rocks as they cool typically following crystallization from magma.
+
+\noindent\textbf{Tonian Period } The first geologic period of the Neoproterozoic Era from ca. 1000 to 717 Ma million years ago. It followed by the Cryogenian Period.
 

### 2020-11-07 Manuscript/Laurentia_Paleogeo_Manuscript.tex
update evaporite figure caption

@@
 1.37 Ga) of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}.
-While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature and is consistent with evidence for accretionary orogenesis in the Picuris, Baraboo and Pinware  orogens, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
+While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature and is consistent with evidence for accretionary orogenesis in the Picuris, Baraboo and Pinwarian orogens, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
 
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Laurentia_evaporite_figure.pdf}
-\caption{\textbf{Paleolatitude of Laurentia evaporites.} Left panel: The paleolatitude of evaporite deposits following the amalgamation of Laurentia as reconstructed by the simple Laurentia model shown in Fig.
+\includegraphics[width=7 in]{../Figures/Laurentia_evaporite_figure.pdf}
+\caption{\textbf{Paleolatitude of Laurentia evaporites.} Left panel: The paleolatitude of evaporite deposits as reconstructed by the simple Laurentia model shown in Fig.
 \ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic.
 Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}.
-The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.
-The evaporite points are labeled with numbers both in the paleolatitude plot and the map of their present-day location in the right panel.
-The formations associated with these numbers are: 1 -- Altyn Formation (Belt Supergroup); 2: Wallace/Helena Formations -- Belt Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Carboniferous Canadian Maritime; 9 -- Carboniferous Sverdrup; 10 -- Permian Midcontinental USA.}
+The evaporite lines extend from the maximum to minimum age constraints with points at the preferred depositional age.
+The evaporite paleolatitude points are labeled with numbers that correspond to the numbers on the present-day location map.
+These numbers correspond to: 1 -- Altyn Formation (Belt Supergroup); 2: Wallace/Helena Formations -- Belt Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Silurian Michigan Basin ; 9 -- Devonian Western Canada; 10 -- Carboniferous Canadian Maritime; 11 -- Carboniferous Sverdrup; 12 -- Permian Midcontinental USA; 13 -- Jurassic Gulf of Mexico.}
 \label{fig:Laurentia_evaporites}

### 2020-11-11 Manuscript/Laurentia_Paleogeo_Manuscript.tex
addition of THO text and other small edits

@@
 Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca.
-1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}.
+1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}.
 A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
@@
 1.75 to 1.72 Ga \citep{Condit2015a} and ca.
-1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}.
-However, the evidence for earlier orogenesis ca.
-1.78 to 1.75 in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca.
+1.72 Ga tectonomagmatic activity in the Black Hills region \citep{Redden1990a}.
+However, evidence for earlier orogenesis ca.
+1.78 to 1.75 Ga in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca.
 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca.
@@
 
-%ADD MOJAVE PROVINCE TEXT
 Overall, the collision of Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).
@@
 Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
-1.86 to 1.82 (Fig.
+1.86 to 1.82 Ga (Fig.
 \ref{fig:Laurentia_map}; \citealp{Schulz2007a}).
 Firm evidence of the end of the Penokean orogeny comes from the ca.
-1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
-
-The collisions of provinces and terranes leading up to the Trans-Hudson orogeny are interpreted to be associated with the initial assembly of the supercontinent Nuna in some paleogeographic models (e.g.
-\citealp{Zhang2012a,Pehrsson2015a}.
+1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
+
+The collisions of provinces and terranes leading up to the Trans-Hudson orogeny are interpreted to be associated with the initial assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}.
 The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
@@
 In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}).
-This core then subsequently grows to be a semi-supercontinent with India and Australia, however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
+This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
 
 Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
-This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs.
+This change marks a shift in the predominant style of Laurentia's growth, as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs.
 \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
@@
 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
-Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeast Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca.
+Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeast Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca.
 1.71 to 1.60 Ga (Fig.
@@
 It was in this context of an active margin that there was major plutonism ca.
-1.48 to 1.35 Ga that resulted in the A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
+1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces forming the Granite-Rhyolite Province that extends from the southwest United States to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
 This plutonism has been interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-live active margin \citep{Bickford2015a}.
@@
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}.
-Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, up through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and up to the Labrador Sea \citep{Rivers2008a}.
+1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{McLelland2010a}.
+Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a}.
 The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca.
@@
 This region is interpreted to have been subsequently deformed during a ca.
-1.37 to 1.33 Ga event known as the East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates\citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+1.37 to 1.33 Ga event known as the East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
 
@@
 Amazonia in \citealp{Elming2009a, Elming2021a}).
-The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca.
+The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and another continent ca.
 1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
@@
 717 to 656 Ma) and Marinoan (ca.
-645 to 635 Ma) glacial deposits in the Death Valley region of southwest Laurentia\citep{Nelson2020a}.
+645 to 635 Ma) glacial deposits in the Death Valley region of southwest Laurentia \citep{Nelson2020a}.
 Additionally, Cryogenian volcanics along the western Laurentia margin (e.g.
 \citealp{Eyster2018a}) are interpreted to be the result of active rifting.
-A puzzlingly feature of this record of active rifting is that it significant predates interpreted passive margin thermal subsidence closer to the ca.
+A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca.
 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}.
@@
 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
-The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}.
+The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca.
+550 Ma \citep{Dickerson1998a, Martin2019a}.
 As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar.
-Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca.
-700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia).
-This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys).
-In this mode, subsequent to this rifting of the major continental blocks, smaller terranes rift off the east Laurentia margin ca.
-600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.
+Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca.
+700 Ma extension is associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the east Laurentia margin ca.
+600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}.
+In these models, the separation of a conjugate continent ca.
+700 Ma (interpreted to be Amazonia) leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}.
+The subsequent rifting of terranes off the Laurentia margin ca.
+600 Ma would have resulted in the formation of the (Neo-)Iapetus Ocean and the record of thermal subsidence associated with passive margin development on Laurentia \citep{Escayola2011a, Robert2020a}.
 
@@
 However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age.
-For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
-However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen.
+For example, there are many poles that have been developed from lithologies of the Trans-Hudson orogen (e.g.
+\citealp{Symons2005a}.
+However, these poles typically lack field tests, have poor control on paleohorizontal and are hypothesized to suffer from variable overprints as discussed in \cite{Raub2008a} and \cite{DAgrella-Filho2020a}.
+Given the preponderance of similar directions, three representative poles from the Trans-Hudson orogen are included in the Nordic Workshop compilation (Table 2) as `B' rated poles.
+However, they should be used with caution given that they suffer from the same deficiencies.
+
+There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
+The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen.
 Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions.
@@
 \hline
-Svalbard & 125.0 & -81.0 & 68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
+Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
 \hline
@@
 Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage.
-These data result in progressive paths such as ascending up to the Logan Loop \citep{Robertson1971a}, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
+These data result in progressive paths such as ascending to the Logan Loop \citep{Robertson1971a}, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
 775 to 720 Ma) path \citep{Eyster2019a}.
@@
 As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed.
-However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts.
+However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts.
 There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}.
@@
 In portions of the record with a more dense record of poles, such as ca.
-1450 Ma, a calculated running mean is used to integrate constraints from multiple poles.
+1450 Ma, a calculated running mean is used to integrate constraints from multiple poles (shown in Fig \ref{fig:Laurentia_poles}).
 This method follows the approach taken in the Phanerozoic (e.g.
@@
 As a result, for the portion of the reconstruction during the interval of time ca.
-1110 to 1070 Ma where there is high data density from the Midcontinent Rift, I utilize an Euler pole inversion from \cite{Swanson-Hysell2019a}.
+1110 to 1070 Ma where there is high data density from the Midcontinent Rift, I use the time-calibrated path from \cite{Swanson-Hysell2019a}.
 
@@
 The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}.
-Over much of the record, Laurentia's pole positions can be satisfied through motion of the content at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig.
+Over much of the record, Laurentia's pole positions can be satisfied through motion of the continent at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig.
 \ref{fig:Laurentia_paleolatitude}).
@@
 \item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in an intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
-The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}.
+The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone \citep{Jefferson1989a, Jones2010a}.
 The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os date from the overlying Coppercap Formation \citep{Rooney2014a}.
@@
 \caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}.
-The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interprets this relationship as persistent from 1.7 to 0.7 Ga.
+The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga.
 The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny.
-The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia.
+The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia relative to South and West Australia.
 This configuration back to ca.
@@
 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}.
-If the North Australian craton was continuous with the South Australia his interpretation would have the ca.
+If the North Australian craton was continuous with the South Australia, this interpretation would have the ca.
 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca.
-1.35 Ga East Kootenay orogeny.
+1.35 Ga East Kootenay orogeny on that margin.
 A paleomagnetic pole from the ca.
@@
 1070 Ma) and Neoproterozoic (ca.
-750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit.
+750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tightly positioned against western Laurentia as they are in the proto-SWEAT fit.
 Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}.
@@
 Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration.
+This configuration had originally been proposed by \citet{Patchett1978a} and \citet{Piper1980a} largely on the basis of paleomagnetic pole comparisons .
 This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
@@
 This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica \citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}.
-However, as geochronology has improved, the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.
+However, as geochronology has improved (e.g.
+\citealp{Gong2018b}), the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.
 
@@
 1.06 Ga.
-In this model, Baltica could have been on the same plate as Laurentia during the rapid late Mesoporoterozoic mortion leading up to collisional Grenvillian orogenesis and the associated assembly of Rodinia (Fig.
+In this model, Baltica could have been on the same plate as Laurentia during the rapid late Mesoporoterozoic motion leading up to collisional Grenvillian orogenesis and the associated assembly of Rodinia (Fig.
 \ref{fig:Grenville_reconstructions}).
@@
 The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} is that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g.
-\citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia.
+\citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia (Fig.
+\ref{fig:Grenville_reconstructions}).
 The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them becoming conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}.
@@
 If they indeed became conjoined in Rodinia as in Figure \ref{fig:Grenville_reconstructions}, the separation of Kalahari from Laurentia may have initiated ca.
-795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in eastern Laurentia.
+795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in southeastern Laurentia.
 This late Tonian timing for separation could be  consistent with a position along the southeast Laurentia margin given the ca.
-750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Cuyania \citep{Martin2019a}.
-
-%This interpretation is also consistent with a record of continental arc magmatism ca.
-1160 to 1070 Ma with a a peak of activity ca.
-1110 to 1090 Ma from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica  \citep{Wang2020a}.
-This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca.
-1080 and 1030 Ma consistent with orogenesis at the time of the Grenvillian orogeny \citep{Wang2020a}.
-They argue that this continued as an accretionary orogen with the TOAST coming in, but that Namaqua-Natal was on the lower plate with Maud belt on the upper plate.
+750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.
 
@@
 The ca.
-1220 Ma dikes pole of the North China craton is not coincident with the ca.
-1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca.
+1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca.
 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
-In terms of departing from this position, one possibility is that its departure is associated with early Neoproterozoic extension in northwest Laurentia.
+Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma KEEP EDITTING HERE pole when it needs to have asce In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwest Laurentia.
 

### 2020-11-11 Manuscript/Laurentia_Paleogeo_Manuscript.tex
update figure numbers and add NW Laurentia rifting text

@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
 \caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
@@
 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).
+As in southwest Laurentia, extensional basin development continued into the Cryogenian and Ediacaran which accommodated deposition of the Windermere Supergroup \citep{Moynihan2019a}.
+Ediacaran rifting transitioned to thermal subsidence in the Cambrian and an early Paleozoic passive margin \citep{Moynihan2019a}.
+A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} suggest that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig3_Laurentia_poles_combined.pdf}
 \caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland).
@@
 \centering
-\includegraphics[width=6 in]{../Figures/Superior_Slave_reconstructions.pdf}
+\includegraphics[width=6 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
 \caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
+\includegraphics[width=0.9\textwidth]{../Figures/Fig5_Laurentia_paleolatitude.pdf}
 \caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (present-day coordinates of lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes.
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig6_Laurentia_reconstructions.pdf}
 \caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}.
@@
 \centering
-\includegraphics[width=7 in]{../Figures/Laurentia_evaporite_figure.pdf}
+\includegraphics[width=7 in]{../Figures/Fig7_Laurentia_evaporite_figure.pdf}
 \caption{\textbf{Paleolatitude of Laurentia evaporites.} Left panel: The paleolatitude of evaporite deposits as reconstructed by the simple Laurentia model shown in Fig.
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig8_Grenville_Reconstruction.pdf}
 \caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}.
@@
 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
-Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma KEEP EDITTING HERE pole when it needs to have asce In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwest Laurentia.
-
-%CHECK THE SUDBURY POINT FOR MYSELF
+Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma during which it ascends to the Logan Loop (Figs.
+\ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
+In its reconstructed position, the ca.
+1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwest Laurentia margin prior to 1220 Ma.
+In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwest Laurentia.
 
@@
 \ref{fig:Grenville_reconstructions}).
-
 Connections between Archean provinces may have also existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}.
-The emplacement of the Franklin LIP ca.
-720 Ma could have initiated break-up of a joint Laurentia-Siberia continent ending the long Proterozoic connection between the cratons.
+
+The interpretation that the Franklin LIP is recorded in both Laurentia and Siberia constrains the break-up between the continents to post-date 720 Ma.
+While there could have been some extension in the Tonian Period and at the time of the Franklin emplacement, the record of northern Laurentia subsidence indicates Ediacaran to early Cambrian extensional tectonism (e.g.
+\citealp{Strauss2019a}).
+Therefore, it might not have been until the late Ediacaran Period that the long-term Proterozoic connection between the cratons ended.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/baked_contact_timeline_all.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig9_baked_contact_timeline_all.pdf}
 \caption{\textbf{Paleomagnetic poles with positive baked contact tests from Laurentia and other cratons}.
@@
 One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics.
-However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonicsas as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}.
+However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonics as as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}.
 These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
@@
 As a result, it is significantly more challenging to develop reliable paleointensity data than to develop the reliable paleodirectional data used to calculate paleomagnetic poles.
-Therefore, the paleointensity database is sparser the compilation of reliable paleomagnetic poles.
+Therefore, the paleointensity database is sparser than the compilation of reliable paleomagnetic poles.
 Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests.
@@
 \textbf{Table 2}: Compilation of paleomagnetic poles from Laurentia
-\input{pole_table_edited.tex}
+\begin{ThreePartTable}
+\begin{TableNotes}
+\footnotesize
+site lon -- longitude of paleomagnetic locality;
+site lat -- latitude of paleomagnetic locality;
+plon -- longitude of the paleomagnetic pole position;
+plat -- latitude of the paleomagnetic pole position;
+A$_{95}$ -- angle of 95$\%$ confidence on the pole position;
+Duluth lat -- latitude of Duluth, MN implied by the paleomagnetic pole
+\end{TableNotes}
+
+\begin{longtable}{p{1.4 in}p{1.2 in}rrrrrrrrp{1.2 in}}
+\toprule
+terrane & unit name & age (Ma) & rating & site lon & site lat & plon & plat & A$_{95}$ & Duluth lat & pole reference \\ \hline
+\midrule
+\endhead
+\midrule
+\multicolumn{11}{r}{{Continued on next page}} \\ \hline
+\midrule
+\endfoot
+
+\bottomrule
+\insertTableNotes
+\endlastfoot
+             Laurentia-Wyoming &                            Stillwater Complex - C2 &     2705$^{+4}_{-4}$ &      A &     249.2 &      45.2 & 335.8 & -83.6 &       4.0 &          &                                 \cite{Selkin2008a} \\ \hline
+      Laurentia-Superior(East) &                       Otto Stock dikes and aureole &     2676$^{+5}_{-5}$ &      B &     279.9 &      48.0 & 227.0 &  69.0 &       4.8 &          &                               \cite{Pullaiah1975b} \\ \hline
+               Laurentia-Slave &                                       Defeat Suite &     2625$^{+5}_{-5}$ &      B &     245.5 &      62.5 &  64.0 &  -1.0 &      15.0 &          &                               \cite{Mitchell2014a} \\ \hline
+      Laurentia-Superior(East) &                         Ptarmigan-Mistassini dikes &     2505$^{+2}_{-2}$ &      B &     287.0 &      54.0 & 213.0 & -45.3 &      13.8 &          &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                 Matachewan dikes R &   2466$^{+23}_{-23}$ &      A &     278.0 &      48.0 & 238.3 & -44.1 &       1.6 &          &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                 Matachewan dikes N &     2446$^{+3}_{-3}$ &      A &     278.0 &      48.0 & 239.5 & -52.3 &       2.4 &          &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                       Malley dikes &     2231$^{+2}_{-2}$ &      A &     249.8 &      64.2 & 310.0 & -50.8 &       6.7 &          &                                 \cite{Buchan2012a} \\ \hline
+      Laurentia-Superior(East) &                                   Senneterre dikes &     2218$^{+6}_{-6}$ &      A &     283.0 &      49.0 & 284.3 & -15.3 &       5.5 &          &                                 \cite{Buchan1993a} \\ \hline
+      Laurentia-Superior(East) &                                 Nipissing N1 sills &     2217$^{+4}_{-4}$ &      A &     279.0 &      47.0 & 272.0 & -17.0 &      10.0 &          &                                 \cite{Buchan2000a} \\ \hline
+               Laurentia-Slave &                                       Dogrib dikes &     2193$^{+2}_{-2}$ &      A &     245.5 &      62.5 & 315.0 & -31.0 &       7.0 &          &                               \cite{Mitchell2014a} \\ \hline
+      Laurentia-Superior(East) &                                  Biscotasing dikes &     2170$^{+3}_{-3}$ &      A &     280.0 &      48.0 & 223.9 &  26.0 &       7.0 &          &                                  \cite{Evans2010a} \\ \hline
+             Laurentia-Wyoming &    Rabbit Creek, Powder River and South Path dikes &    2160$^{+11}_{-8}$ &      A &     252.8 &      43.9 & 339.2 &  65.5 &       7.6 &          &                                 \cite{Kilian2015a} \\ \hline
+               Laurentia-Slave &                                        Indin dikes &    2126$^{+3}_{-18}$ &      A &     245.6 &      62.5 & 256.0 & -36.0 &       7.0 &          &                                 \cite{Buchan2016a} \\ \hline
+      Laurentia-Superior(West) &                                   Marathon dikes N &     2124$^{+3}_{-3}$ &      A &     275.0 &      49.0 & 198.2 &  45.4 &       7.7 &        43.3 &                                  \cite{Halls2008a} \\ \hline
+      Laurentia-Superior(West) &                                   Marathon dikes R &     2104$^{+3}_{-3}$ &      A &     275.0 &      49.0 & 182.2 &  55.1 &       7.5 &        38.8 &                                  \cite{Halls2008a} \\ \hline
+      Laurentia-Superior(West) &                                 Cauchon Lake dikes &     2091$^{+2}_{-2}$ &      A &     263.0 &      56.0 & 180.9 &  53.8 &       7.7 &        37.5 &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(West) &                                 Fort Frances dikes &     2077$^{+5}_{-5}$ &      A &     266.0 &      48.0 & 184.6 &  42.8 &       6.1 &        33.6 &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Superior(East) &                                   Lac Esprit dikes &     2069$^{+1}_{-1}$ &      A &     282.0 &      53.0 & 170.5 &  62.0 &       6.4 &          &                                  \cite{Evans2010a} \\ \hline
+      Laurentia-Greenland-Nain &                                    Kangamiut dikes &   2042$^{+12}_{-12}$ &      B &     307.0 &      66.0 & 273.8 &  17.1 &       2.7 &         &                                 \cite{Fahrig1976b} \\ \hline
+               Laurentia-Slave &                                  Lac de Gras dikes &     2026$^{+5}_{-5}$ &      A &     249.6 &      64.4 & 267.9 &  11.8 &       7.1 &          &                                 \cite{Buchan2009a} \\ \hline
+      Laurentia-Superior(East) &                                        Minto dikes &     1998$^{+2}_{-2}$ &      A &     285.0 &      57.0 & 171.5 &  38.7 &      13.1 &          &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                    Rifle Formation &     1963$^{+6}_{-6}$ &      B &     252.9 &      65.9 & 341.0 &  14.0 &       7.7 &          &                                  \cite{Evans1981a} \\ \hline
+                 Laurentia-Rae &                             Clearwater Anorthosite &     1917$^{+7}_{-7}$ &      B &     251.6 &      57.1 & 311.8 &   6.5 &       2.9 &          &                                  \cite{Halls1999a} \\ \hline
+             Laurentia-Wyoming &                         Sourdough mafic dike swarm &     1899$^{+5}_{-5}$ &      A &    -108.3 &      44.7 & 292.0 &  49.2 &       8.1 &          &                                 \cite{Kilian2016b} \\ \hline
+               Laurentia-Slave &                                   Ghost Dike Swarm &     1887$^{+5}_{-9}$ &      A &     244.6 &      62.6 & 286.0 &  -2.0 &       6.0 &          &                                 \cite{Buchan2016a} \\ \hline
+               Laurentia-Slave &                           Mean Seton/Akaitcho/Mara &     1885$^{+5}_{-5}$ &      B &     250.0 &      65.0 & 260.0 &  -6.0 &       4.0 &          &                               \cite{Mitchell2010c} \\ \hline
+               Laurentia-Slave &                     Mean Kahochella, Peacock Hills &     1882$^{+4}_{-4}$ &      B &     250.0 &      65.0 & 285.0 & -12.0 &       7.0 &          &                               \cite{Mitchell2010c} \\ \hline
+      Laurentia-Superior(West) &                                Molson (B+C2) dikes &     1879$^{+6}_{-6}$ &      A &     262.0 &      55.0 & 218.0 &  28.9 &       3.8 &        47.6 &                                  \cite{Evans2010a} \\ \hline
+               Laurentia-Slave &                                 Takiyuak Formation &   1876$^{+10}_{-10}$ &      B &     246.9 &      66.1 & 249.0 & -13.0 &       8.0 &          &                                 \cite{Irving1979a} \\ \hline
+               Laurentia-Slave &          Douglas Peninsula Formation, Pethei Group &   1876$^{+10}_{-10}$ &      B &     249.7 &      62.8 & 258.0 & -18.0 &      14.2 &          &                                 \cite{Irving1979a} \\ \hline
+               Laurentia-Slave &               Pearson A/Peninsular/Kilohigok sills &     1870$^{+4}_{-4}$ &      A &     250.0 &      65.0 & 269.0 & -22.0 &       6.0 &          &                               \cite{Mitchell2010c} \\ \hline
+            Laurentia-Superior &                         Haig/Flaherty/Sutton Mean  &     1870$^{+1}_{-1}$ &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &          &  Nordic workshop calculation based on data of \cite{Schmidt1980a, Schwarz1982a} \\ \hline
+ Laurentia-Trans-Hudson orogen &                                Boot-Phantom Pluton &     1838$^{+1}_{-1}$ &      B &     258.1 &      54.7 & 275.4 &  62.4 &       7.9 &        73.8 &                                 \cite{Symons1999a} \\ \hline
+                 Laurentia-Rae &                                      Sparrow dikes &     1827$^{+4}_{-4}$ &      B &     250.2 &      61.6 & 291.0 &  12.0 &       7.9 &          &                                \cite{McGlynn1974a} \\ \hline
+                 Laurentia-Rae &                                   Martin Formation &     1818$^{+4}_{-4}$ &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &          &                                  \cite{Evans1973a} \\ \hline
+                     Laurentia &                                      Dubawnt Group &   1785$^{+35}_{-35}$ &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &        49.4 &                                   \cite{Park1973a} \\ \hline
+ Laurentia-Trans-Hudson orogen &                            Deschambault Pegmatites &     1766$^{+5}_{-5}$ &      B &     256.7 &      54.9 & 276.0 &  67.5 &       7.7 &        68.9 &                                 \cite{Symons2000a} \\ \hline
+ Laurentia-Trans-Hudson orogen &                                   Jan Lake Granite &     1758$^{+1}_{-1}$ &      B &     257.2 &      54.9 & 264.3 &  24.3 &      16.9 &        67.3 &                                   \cite{Gala1995a} \\ \hline
+                     Laurentia &                                      Cleaver dikes &     1741$^{+5}_{-5}$ &      A &     242.0 &      67.5 & 276.7 &  19.4 &       6.1 &        61.7 &                                 \cite{Irving2004a} \\ \hline
+           Laurentia-Greenland &                        Melville Bugt diabase dikes &     1633$^{+5}_{-5}$ &      B &     303.0 &      74.6 & 273.8 &   5.0 &       8.7 &        45.5 &                                  \cite{Halls2011a} \\ \hline
+                     Laurentia &                            Western Channel Diabase &     1590$^{+3}_{-3}$ &      A &     242.2 &      66.4 & 245.0 &   9.0 &       6.6 &        47.5 &                                 \cite{Irving1972a} \\ \hline
+                     Laurentia &                 St.Francois Mountains Acidic Rocks &   1476$^{+16}_{-16}$ &      A &     269.5 &      37.5 & 219.0 & -13.2 &       6.1 &        15.8 &                                  \cite{Meert2002b} \\ \hline
+                     Laurentia &                               Michikamau Intrusion &     1460$^{+5}_{-5}$ &      A &     296.0 &      54.5 & 217.5 &  -1.5 &       4.7 &        24.7 &                                 \cite{Emslie1976a} \\ \hline
+                     Laurentia &                                  Spokane Formation &   1458$^{+13}_{-13}$ &      A &     246.8 &      48.2 & 215.5 & -24.8 &       4.7 &         4.2 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &                                 Snowslip Formation &   1450$^{+14}_{-14}$ &      A &     245.9 &      47.9 & 210.2 & -24.9 &       3.5 &         1.4 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &                                 Tobacco Root dikes &   1448$^{+49}_{-49}$ &      B &     247.6 &      47.4 & 216.1 &   8.7 &      10.5 &        31.9 &                                 \cite{Harlan2008a} \\ \hline
+                     Laurentia &                                       Purcell Lava &     1443$^{+7}_{-7}$ &      A &     245.1 &      49.4 & 215.6 & -23.6 &       4.8 &         5.3 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &                          Rocky Mountain intrusions &   1430$^{+15}_{-15}$ &      B &     253.8 &      40.3 & 217.4 & -11.9 &       9.7 &        16.0 &  Nordic workshop calculation based on data of \cite{Harlan1994a,Harlan1998a} \\ \hline
+                     Laurentia &                                   Mistastin Pluton &   1425$^{+25}_{-25}$ &      B &     296.3 &      55.6 & 201.5 &  -1.0 &       7.6 &        15.1 &                                 \cite{Fahrig1976a} \\ \hline
+                     Laurentia &                                 McNamara Formation &     1401$^{+6}_{-6}$ &      A &     246.4 &      46.9 & 208.3 & -13.5 &       6.7 &         9.6 &                                 \cite{Elston2002a} \\ \hline
+                     Laurentia &         Pilcher, Garnet Range and Libby Formations &   1385$^{+23}_{-23}$ &      A &     246.4 &      46.7 & 215.3 & -19.2 &       7.7 &         8.8 &                                 \cite{Elston2002a} \\ \hline
+           Laurentia-Greenland &                                Zig-Zag Dal Basalts &     1382$^{+2}_{-2}$ &      B &     334.8 &      81.2 & 242.8 &  12.0 &       3.8 &        43.8 &                              \cite{Marcussen1983a} \\ \hline
+           Laurentia-Greenland &                      Victoria Fjord dolerite dikes &     1382$^{+2}_{-2}$ &      B &     315.3 &      81.5 & 231.7 &  10.3 &       4.3 &        36.6 &                             \cite{Abrahamsen1987a} \\ \hline
+           Laurentia-Greenland &                              Midsommersoe Dolerite &     1382$^{+2}_{-2}$ &      B &     333.4 &      81.6 & 242.0 &   6.9 &       5.1 &        39.0 &                              \cite{Marcussen1983a} \\ \hline
+                     Laurentia &                                   Nain Anorthosite &   1305$^{+15}_{-15}$ &      B &     298.2 &      56.5 & 206.7 &  11.7 &       2.2 &        28.1 &                                 \cite{Murthy1978a} \\ \hline
+           Laurentia-Greenland &                             North Qoroq intrusives &     1275$^{+1}_{-1}$ &      B &     314.6 &      61.1 & 202.6 &  13.2 &       8.3 &        21.0 &                                  \cite{Piper1992a} \\ \hline
+           Laurentia-Greenland &                                  Kungnat Ring dike &     1275$^{+2}_{-2}$ &      B &     311.7 &      61.2 & 198.7 &   3.4 &       3.2 &        11.1 &                                  \cite{Piper1977b} \\ \hline
+                     Laurentia &                         Mackenzie dikes grand mean &     1267$^{+2}_{-2}$ &      A &     250.0 &      65.0 & 190.0 &   4.0 &       5.0 &        11.2 &                                 \cite{Buchan2000a} \\ \hline
+           Laurentia-Greenland &                         West Gardar Dolerite dikes &     1244$^{+8}_{-8}$ &      B &     311.7 &      61.2 & 201.7 &   8.7 &       6.6 &        17.1 &                                  \cite{Piper1977b} \\ \hline
+           Laurentia-Greenland &                      West Gardar Lamprophyre dikes &   1238$^{+11}_{-11}$ &      B &     311.7 &      61.2 & 206.4 &   3.2 &       7.2 &        15.9 &                                  \cite{Piper1977b} \\ \hline
+                     Laurentia &                             Sudbury dikes Combined &     1237$^{+5}_{-5}$ &      A &     278.6 &      46.3 & 192.8 &  -2.5 &       2.5 &         8.3 &                                 \cite{Palmer1977a} \\ \hline
+            Laurentia-Scotland &                                        Stoer Group &   1199$^{+70}_{-70}$ &      B &     354.5 &      58.0 & 238.4 &  37.2 &       7.7 &        43.9 &                        Nordic workshop calculation \\ \hline
+           Laurentia-Greenland &                                 Hviddal Giant dike &     1184$^{+5}_{-5}$ &      B &     313.7 &      60.9 & 215.3 &  33.2 &       9.6 &        43.3 &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                                     Narssaq Gabbro &     1184$^{+5}_{-5}$ &      B &     313.8 &      60.9 & 225.4 &  31.6 &       9.7 &        48.8 &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                                  South Qoroq Intr.
+&     1163$^{+2}_{-2}$ &      A &     314.6 &      61.1 & 215.9 &  41.8 &      13.1 &        48.7 &                                  \cite{Piper1992a} \\ \hline
+           Laurentia-Greenland &                                 Giant Gabbro dikes &     1163$^{+2}_{-2}$ &      B &     313.7 &      60.9 & 226.1 &  42.3 &       9.4 &        55.5 &                                  \cite{Piper1977a} \\ \hline
+           Laurentia-Greenland &                               NE-SW Trending dikes &     1160$^{+5}_{-5}$ &      B &     314.6 &      61.1 & 230.8 &  33.4 &       5.7 &        53.5 &                                  \cite{Piper1992a} \\ \hline
+                     Laurentia &                          Ontario lamprophyre dikes &   1143$^{+12}_{-12}$ &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &        61.2 &                                 \cite{Piispa2018a} \\ \hline
+                     Laurentia &                                      Abitibi dikes &     1141$^{+2}_{-2}$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \cite{Ernst1993a} \\ \hline                       Baltica &                        Salla dike (width 60-100 m) &     1122$^{+5}_{-3}$ &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &          &                                                 \\ \hline
+                     Laurentia &                            Nipigon sills and lavas &     1109$^{+2}_{-2}$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
+                     Laurentia &             Lowermost Mamainse Point volcanics -R1 &     1109$^{+2}_{-3}$ &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &        62.9 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &                           Lower Osler volcanics -R &     1108$^{+3}_{-3}$ &      A &     272.3 &      48.8 & 218.6 &  40.9 &       4.8 &        54.6 &                         \cite{Swanson-Hysell2014b} \\ \hline
+                     Laurentia &                          Middle Osler volcanics -R &     1107$^{+4}_{-4}$ &      A &     272.4 &      48.8 & 211.3 &  42.7 &       8.2 &        50.5 &                         \cite{Swanson-Hysell2014b} \\ \hline
+                     Laurentia &                           Upper Osler volcanics -R &     1105$^{+1}_{-1}$ &      A &     272.4 &      48.7 & 203.4 &  42.3 &       3.7 &        45.1 &  \cite{Halls1974a, Swanson-Hysell2014b, Swanson-Hysell2019a} \\ \hline
+                     Laurentia &                 Lower Mamainse Point volcanics -R2 &     1105$^{+3}_{-4}$ &      A &     275.3 &      47.1 & 205.2 &  37.5 &       4.5 &        43.9 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &     Mamainse Point volcanics -C (lower N, upper R) &     1101$^{+1}_{-1}$ &      A &     275.3 &      47.1 & 189.7 &  36.1 &       4.9 &        32.9 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &                               North Shore lavas -N &     1097$^{+3}_{-3}$ &      A &     268.7 &      46.3 & 181.7 &  31.1 &       2.1 &        24.5 &              \cite{Tauxe2009a,Swanson-Hysell2019a} \\ \hline
+                     Laurentia &                              Chengwatana Volcanics &     1095$^{+2}_{-2}$ &      B &     267.3 &      45.4 & 186.1 &  30.9 &       8.2 &        27.3 &                                   \cite{Kean1997a} \\ \hline
+                     Laurentia &                             Portage Lake Volcanics &     1095$^{+3}_{-3}$ &      A &     271.2 &      47.0 & 182.5 &  27.5 &       2.3 &        22.7 &  \cite{Books1972a, Hnat2006a} as calculated in \cite{Swanson-Hysell2019a} \\ \hline
+                     Laurentia &              Uppermost Mamainse Point volcanics -N &     1094$^{+6}_{-4}$ &      A &     275.3 &      47.1 & 183.2 &  31.2 &       2.5 &        25.6 &                         \cite{Swanson-Hysell2014a} \\ \hline
+                     Laurentia &                    Cardenas Basalts and Intrusions &     1091$^{+5}_{-5}$ &      B &     248.1 &      36.1 & 185.0 &  32.0 &       8.0 &        27.3 &                                   \cite{Weil2003a} \\ \hline
+                     Laurentia &                           Schroeder Lutsen Basalts &     1090$^{+2}_{-7}$ &      A &     269.1 &      47.5 & 187.8 &  27.1 &       3.0 &        25.9 &                              \cite{Fairchild2017a} \\ \hline
+                     Laurentia &                        Central Arizona diabases -N &   1088$^{+11}_{-11}$ &      A &     249.2 &      33.7 & 175.3 &  15.7 &       7.0 &         9.6 &                               \cite{Donadini2011b} \\ \hline
+                     Laurentia &                                   Lake Shore Traps &     1086$^{+1}_{-1}$ &      A &     271.9 &      47.6 & 186.4 &  23.1 &       4.0 &        22.3 &                                \cite{Kulakov2013a} \\ \hline
+                     Laurentia &                      Michipicoten Island Formation &     1084$^{+1}_{-1}$ &      A &     274.3 &      47.7 & 174.7 &  17.0 &       4.4 &        10.2 &                              \cite{Fairchild2017a} \\ \hline
+                     Laurentia &                                     Nonesuch Shale &    1080$^{+4}_{-10}$ &      B &     271.5 &      47.0 & 178.1 &   7.6 &       5.5 &         5.7 &                                  \cite{Henry1977a} \\ \hline
+                     Laurentia &                                    Freda Sandstone &   1070$^{+14}_{-10}$ &      B &     271.5 &      47.0 & 179.0 &   2.2 &       4.2 &         2.4 &                                  \cite{Henry1977a} \\ \hline
+                     Laurentia &                              Haliburton Intrusions &   1015$^{+15}_{-15}$ &      B &     281.4 &      45.0 & 141.9 & -32.6 &       6.3 &       -47.0 &                                \cite{Warnock2000a} \\ \hline
+                     Laurentia &                        Adirondack fayalite granite &    990$^{+20}_{-20}$ &     NR &     285.5 &      44.0 & 132.7 & -28.4 &       6.9 &       -50.7 &                                  \cite{Brown2012a} \\ \hline
+                     Laurentia &               Adirondack metamorphic\ orthosites &    970$^{+20}_{-20}$ &     NR &     286.0 &      44.0 & 149.0 & -25.1 &      11.6 &       -37.5 &                                  \cite{Brown2012a} \\ \hline
+                     Laurentia &                       Adirondack Microcline gneiss &    960$^{+20}_{-20}$ &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &       -31.5 &                                  \cite{Brown2012a} \\ \hline
+            Laurentia-Scotland &                                     Torridon Group &  925$^{+145}_{-145}$ &      B &     354.3 &      57.9 & 220.9 & -17.7 &       7.1 &        -8.6 &                        Nordic workshop calculation \\ \hline
+            Laurentia-Svalbard &                      Lower Grusdievbreen Formation &    831$^{+20}_{-20}$ &      B &      18.0 &      79.0 & 204.9 &  19.6 &      10.9 &        -5.3 &                                 \cite{Maloof2006a} \\ \hline
+            Laurentia-Svalbard &                      Upper Grusdievbreen Formation &    800$^{+11}_{-11}$ &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &        11.5 &                                 \cite{Maloof2006a} \\ \hline
+                     Laurentia &                                    Gunbarrel dikes &      778$^{+2}_{-2}$ &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &       -18.4 &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a} \\ \hline
+            Laurentia-Svalbard &                          Svanbergfjellet Formation &    770$^{+19}_{-40}$ &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &        12.8 &                                 \cite{Maloof2006a} \\ \hline
+                     Laurentia &                               Uinta Mountain Group &     760$^{+6}_{-10}$ &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &       -10.7 &                                   \cite{Weil2006b} \\ \hline
+                     Laurentia &                                      Carbon Canyon &      757$^{+7}_{-7}$ &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &        -8.5 &  \cite{Weil2004a} as calculated in \cite{Eyster2019a} \\ \hline
+                     Laurentia &                               Carbon Butte/Awatubi &      751$^{+8}_{-8}$ &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &         1.0 &                                 \cite{Eyster2019a} \\ \hline
+                     Laurentia &                          Franklin event grand mean &      718$^{+2}_{-2}$ &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &        -5.7 &                               \cite{Denyszyn2009b} \\ \hline
+                     Laurentia &                                   Long Range dikes &      615$^{+2}_{-2}$ &      B &     303.3 &      53.7 & 175.3 & -19.0 &      17.4 &       -15.5 &                                 \cite{Murthy1992a} \\ \hline
+                     Laurentia &                           Baie des Moutons complex &      583$^{+2}_{-2}$ &      B &     301.0 &      50.8 & 152.7 & -42.6 &      12.0 &       -45.1 &                             \cite{McCausland2011a} \\ \hline
+                     Laurentia &                           Baie des Moutons complex &      583$^{+2}_{-2}$ &      B &     301.0 &      50.8 & 141.5 &  34.2 &      15.4 &         4.2 &                             \cite{McCausland2011a} \\ \hline
+                     Laurentia &                         Callander Alkaline Complex &      575$^{+5}_{-5}$ &      B &     280.6 &      46.2 & 121.4 & -46.3 &       6.0 &       -67.1 &                                 \cite{Symons1991a} \\ \hline
+                     Laurentia &                                   Catoctin Basalts &      572$^{+5}_{-5}$ &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &       -69.0 &                                  \cite{Meert1994a} \\ \hline
+                     Laurentia &                        Sept-Iles layered intrusion &      565$^{+4}_{-4}$ &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &        -7.9 &                                \cite{Tanczyk1987a} \\ \hline
+\end{longtable}
+\end{ThreePartTable}
 \end{landscape}

### 2020-11-13 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add what the 1184 Ma Laurentia pole constraint is

@@
 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time.
-Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma during which it ascends to the Logan Loop (Figs.
+Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma Narssaq Gabbro and Hvidaal dike poles during which it ascends to the Logan Loop (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).

### 2020-11-18 Manuscript/Laurentia_Paleogeo_Manuscript.tex
implementing some edits from JS suggestions

@@
 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{McLelland2010a}.
+Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage.
+Properly referring to the Grenvillian orogeny as distinct from the Elzevirian and Shawinigan orogenies enables the available constraints to be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
+\ref{fig:tectonic_history}).
 Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a}.
@@
 \ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin.
-There are thick successions of Paleoproterozoic sediments such as the ca.
-1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada that comprises $\sim$14 km of  fine-grained siliciclastic and carbonate rocks \citep{Furlanetto2016a}.
-This succession is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}.
+There are thick successions of Paleoproterozoic siliciclastic and carbonate sedimentary rocks such as the ca.
+1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada \citep{Delaney1986a, Furlanetto2016a}.
+The Wernecke Supergroup is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}, although the tectonic setting is poorly understood and could be intracratonic.
 These metasedimentary rocks were deformed and metamorphosed during the ca.
-1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent with associated orogenesis (referred to as the Forward orogeny) further into the craton \citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}.
-Further south along Laurentia's western margin, the 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca.
+1.60 Ga Racklan-Forward orogeny that is interpreted to be associated with collision of an arc terrane and potentially a conjugate continent \citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}.
+Further south along Laurentia's western margin, sedimentary rocks of the 15 to 20 km thick Belt-Purcell Supergroup are associated with a ca.
 1.47 to 1.40 rift \citep{Evans2000c}.
@@
 1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
-This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
-Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage.
-These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
-\ref{fig:tectonic_history}).
+This extensive and major collisional orogenesis on Laurentia's margin, and that of other Proterozoic continents, remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.
 
@@
 Along the western margin of Laurentia, by ca.
-760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwest Laurentia to the Mackenzie Mountains of northwest Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}.
+760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwest Laurentia to the Ogilvie Mountains of northwest Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}.
 The emplacement of the ca.
-780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia; \citealp{Li2008a}).
+780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g.
+\citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development.
 Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca.
@@
 Ediacaran rifting transitioned to thermal subsidence in the Cambrian and an early Paleozoic passive margin \citep{Moynihan2019a}.
-A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} suggest that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.
-
-Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig.
+A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northeastern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} interpret that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.
+
+Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin, including the displaced terrane of northeastern Svalbard, Norway (Fig.
 \ref{fig:tectonic_history}).

### 2021-01-05 Manuscript/Laurentia_Paleogeo_Manuscript.tex
additional small edits resulting from JS feedback

@@
 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
-Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeast Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca.
+Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca.
 1.71 to 1.60 Ga (Fig.
@@
 1.49 to 1.46 Ga associated with the Baraboo orogeny \citep{Medaris2003a, Holm2019a}.
-Picuris to Baraboo to Pinwarian orogenesis indicates convergent tectonism along the entirety of southeast Laurentia.
+Picuris to Baraboo to Pinwarian orogenesis indicates convergent tectonism along the entirety of southeastern Laurentia.
 It was in this context of an active margin that there was major plutonism ca.
-1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces forming the Granite-Rhyolite Province that extends from the southwest United States to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
+1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces forming the Granite-Rhyolite Province that extends from the southwestern United States to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
 This plutonism has been interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-live active margin \citep{Bickford2015a}.
@@
 Along the western margin of Laurentia, by ca.
-760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwest Laurentia to the Ogilvie Mountains of northwest Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}.
+760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}.
 The emplacement of the ca.
@@
 717 to 656 Ma) and Marinoan (ca.
-645 to 635 Ma) glacial deposits in the Death Valley region of southwest Laurentia \citep{Nelson2020a}.
+645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}.
 Additionally, Cryogenian volcanics along the western Laurentia margin (e.g.
@@
 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
-While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
-This record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
+While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia, it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
+One explanation is that Tonian extensional basin development is associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}.
+The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models).
-Another possibility, along the lines of that proposed in \citet{Ross1991a} and \citet{Colpron2002a}, is that ca.
-760 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further to the west.
+Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca.
+760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west.
 In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
 
-In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
+In northwestern Laurentia from the Ogilvie Mountains (Yukon, Canada) to Victoria Island (Nunavut, Canada), the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
 Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca.
 1050 Ma with ongoing basin development ca.
-812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}.
-Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca.
+812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) that may have been accommodated through thermal subsidence \citep{Macdonald2012a}.
+Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on the tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca.
 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).
-As in southwest Laurentia, extensional basin development continued into the Cryogenian and Ediacaran which accommodated deposition of the Windermere Supergroup \citep{Moynihan2019a}.
+As in southwestern Laurentia, extensional basin development continued into the Cryogenian and Ediacaran which accommodated deposition of the Windermere Supergroup \citep{Moynihan2019a}.
 Ediacaran rifting transitioned to thermal subsidence in the Cambrian and an early Paleozoic passive margin \citep{Moynihan2019a}.
-A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northeastern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} interpret that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.
-
-Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin, including the displaced terrane of northeastern Svalbard, Norway (Fig.
+The record of Neoproterozoic basin development on the northern Franklinian margin of Laurentia is more difficult to resolve given the truncation of the margin by the Paleozoic Ellesmerian orogeny \citep{Cocks2011a}.
+The North Slope terrane of Arctic Alaska is considered to have been part of northeastern Laurentia in the Neoproterozoic and shows a similar history of Tonian to Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation \citep{Strauss2019a}.
+
+Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeastern Greenland margin, including the displaced terrane of northeastern Svalbard, Norway (Fig.
 \ref{fig:tectonic_history}).
 Available geochronological constraints and thermal subsidence modeling indicate ca.
-850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}.
-These data suggest that conjugate continental lithosphere rifted away from northeast Greenland by ca.
-820 Ma.
-
-Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
+850 to 820 Ma rifting followed by thermal subsidence of a stable carbonate platform \citep{Maloof2006a, Sonderholm2008a, Halverson2018a}.
+These data suggest that conjugate continental lithosphere rifted away from northeastern Greenland by ca.
+820 Ma although some models place northeastern Greenland in a distal backarc setting \citep{Malone2014a}.
+
+Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
 A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
@@
 Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca.
-700 Ma extension is associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the east Laurentia margin ca.
+700 Ma extension is associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca.
 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}.
@@
 This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
-Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca.
-760 Ma)  as presented in \cite{Eyster2019a}, incorporating data from \cite{Weil2004a}, are also included.
+Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca.
+760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.
 
@@
 Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland.
-These terranes were once part of contiguous Laurentia, but have subsequently rifted away.
+These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting.
 These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}.
@@
 The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction.
-This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
+This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.
 
@@
 780 Ma) and Franklin LIP (ca.
-720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}.
+720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}.
 Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage.
 These data result in progressive paths such as ascending to the Logan Loop \citep{Robertson1971a}, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
-775 to 720 Ma) path \citep{Eyster2019a}.
+775 to 720 Ma) path \citep{Eyster2020a}.
 
@@
 
-Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic Era where there are inconsistencies between poles of similar age (Figs.
+The interval of the record when there are the most significant inconsistencies between poles of similar age is the Ediacaran Period at the end of the Neoproterozoic Era (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
-As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 583 and 565 Ma (Fig.
+Between 583 and 565 Ma, paleomagnetic poles imply both low-latitude and high-latitude positions of Laurentia  (Fig.
 \ref{fig:Laurentia_paleolatitude}).
@@
 While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain.
-These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca.
+These poles are not utilized in the simple Laurentia model, which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca.
 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
@@
 
-Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude.
+Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude and therefore can be a tool for the evaluation of paleogeographic models.
 Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation.
@@
 
-There is high evaporation in the subtropics and tropics.
-Within the tropical rain belt (0 to $\sim$10\textdegree\ latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree\ of the equator (within $\sim$8\textdegree$\;$ of the equator over land) with evaporation exceeding precipitation from those latitudes towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree\ \citep{Park2020b}.
-However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to the precipitation of evaporites.
-For example, Lake Magadi in Kenya at a latitude of 1.9\textdegree S is a saline lake where thick bedded evaporites have accumulated \citep{Eugster1980a}.
-Caution is therefore needed when interpreting paleolatitude from evaporites in terrestrial and intracratonic settings given that they could occur both in tropical and subtropical latitudes.
-
-Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
+Proterozoic evaporite deposits are documented within the following units that were deposited following the amalgamation of Laurentia (Fig. \ref{fig:Laurentia_evaporites}): 
 \begin{itemize}
@@
 Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
-There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.
+There are also halite and gypsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.
 These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 Ma within the Helena Formation \citep{Evans2000c}.
@@
 These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
-\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} .
-These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca.
+\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a}.
+These thick bedded sulfate deposits are older than cross-cutting 775.1 $\pm$ 0.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Milton2017a}) and younger than ca.
 1005 Ma detrital zircons \citep{Turner2016a}.
@@
 \item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in an intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}.
-The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone \citep{Jefferson1989a, Jones2010a}.
+The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the Mackenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone \citep{Jefferson1989a, Jones2010a}.
 The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os date from the overlying Coppercap Formation \citep{Rooney2014a}.
+This units are also interpreted to be correlative with the Callison Lake Formation which contains pseudomorphs after gypsum and is constrained to have been deposited between Re-Os dates of 752.7 $\pm$ 5.5 Ma and 739.9 $\pm$ 6.5 Ma \citep{Strauss2015a}.
 \end{itemize}
@@
 In the Tonian period, evaporite deposition may have occurred equatorward of 10\textdegree\ (Fig.
-\ref{fig:Laurentia_evaporites}) which may reflect increased aridity in the tropical interior of the Rodinia supercontinent.
+\ref{fig:Laurentia_evaporites}).
+In interpreting these data, it is important to note that there is high evaporation in the subtropics and the tropics.
+Within the tropical rain belt (0 to $\sim$10\textdegree\ latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree\ of the equator (within $\sim$8\textdegree$\;$ of the equator over land).
+Evaporation typically exceeding precipitation from those latitudes ($\sim$10 to 15\textdegree) towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree\ \citep{Park2020b}.
+However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to aridity and the formation of evaporites.
+For example, Lake Magadi in Kenya at a latitude of 1.9\textdegree S is a saline lake where thick bedded evaporites have accumulated \citep{Eugster1980a}.
+Caution is therefore needed when interpreting paleolatitude from evaporites in terrestrial and intracratonic settings given that they could occur both in tropical and subtropical latitudes.
+Therefore low-latitude evaporites ca.
+750 Ma could reflect increased aridity in the tropical interior of the Rodinia supercontinent.
+Overall, the paleomagnetic data as synthesized in the paleogeographic models are in agreement with the evaporite record from the Mesoproterozoic to the Cenozoic (Fig.
+\ref{fig:Laurentia_evaporites}).
 
@@
 In particular, they discuss the hypothesis of Superia (named after the Superior province) wherein the Superior province of Laurentia is central to a group of Archean lithospheric blocks \citep{Bleeker2006a}, including the Kola and Karelia provinces of Baltica and the Wyoming province of Laurentia, that broke up prior to 2.0 Ga.
-This hypothesis has largely been based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeast margin of the Superior Province in present-day coordinates \citep{Davey2020a}.
+This hypothesis has largely been based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeastern margin of the Superior Province \citep{Davey2020a}.
 \cite{Salminen2021a}  argue that the originally proposed Superia fit is not consistent with Baltica paleomagnetic poles.
@@
 In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}.
-On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed.
-In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers.
+On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the Grenvillian orogeny and left behind when the Iapetus Ocean formed.
+In particular, \cite{Fisher2010a} suggested that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers.
 This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig.
@@
 It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia \citep{Moores1991a}.
-The extent of Antarctic lithosphere that was conjoined with Antarctica prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009b}.
+The extent of Antarctic lithosphere that was conjoined with Australia prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009b}.
 Separation between these Antarctic Provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway.
@@
 1.68 Ga and was then caught up in ca.
-1.60 Ga collision between the North Australia craton and northwest Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}.
+1.60 Ga collision between the North Australia craton and northwestern Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}.
 The ca.
-1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}.
+1.60 Ga Racklan-Forward orogeny in northwestern Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}.
 This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations.
@@
 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}.
-The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet).
-The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}.
+The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwestern Laurentia A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet).
+The correlation of the eastern North Australia craton with northwestern Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}.
 Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}.
@@
 1.21 Ga Gnowangerup-Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia ca.
-1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}.
+1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path \citep{Pisarevsky2014a}.
 There is not a coeval pole from Laurentia for comparison at ca.
@@
 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tightly positioned against western Laurentia as they are in the proto-SWEAT fit.
-Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}.
+Rather, while these pole comparisons could be consistent with East Antarctica against southwestern Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}.
 There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca.
@@
 Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration.
-This configuration had originally been proposed by \citet{Patchett1978a} and \citet{Piper1980a} largely on the basis of paleomagnetic pole comparisons .
+This configuration had originally been proposed by \citet{Patchett1978a} and \citet{Piper1980a} largely on the basis of paleomagnetic pole comparisons.
 This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
@@
 
-Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related terranes including East Svalbard \citep{McClelland2019a}.
+Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related Arctic terranes including northeast Svalbard \citep{McClelland2019a}.
 These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
 The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}.
-Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador, Canada is implemented in many paleogeographic models (e.g.
+Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador (northeast Canada) is implemented in many paleogeographic models (e.g.
 \citealp{Evans2009a} and as shown in Fig.
 \ref{fig:Grenville_reconstructions}).
-This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica's APWP that would have Laurentia journey after the Grenville loop northward across the equator and back south to a similar Grenville loop position prior to the ca.
+This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica's APWP.
+These oscillations would imply that following the Grenville loop, Laurentia moved northward across the equator and then back south to a similar Grenville loop position prior to the ca.
 780 Ma Laurentia poles \citep{Evans2015a,Fairchild2017a}.
@@
 \ref{fig:tectonic_history}; \citealp{Maloof2006a}).
-\cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting although the paucity of tuffs within successions such as the Akademikerbreen Group of East Svalbard is difficult to reconcile with such a setting.
+\cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting.
+
 The separation of Baltica from Laurentia, once in the configuration shown in Figure \ref{fig:Grenville_reconstructions}, is hypothesized to have occurred ca.
@@
 High-quality paleomagnetic poles constrain the coeval ca.
-1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced with the craton margins separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}.
+1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced.
+These poles reconstruct the craton margins as having been separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}.
 These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone.
 However, with the subsequent rapid motion of Laurentia to low latitudes (Figs.
-\ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)east margin of Laurentia during the time of the Grenvillian orogeny.
+\ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)eastern margin of Laurentia during the time of the Grenvillian orogeny.
 This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g.
@@
 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in southeastern Laurentia.
-This late Tonian timing for separation could be  consistent with a position along the southeast Laurentia margin given the ca.
+This late Tonian timing for separation could be  consistent with a position along the southeastern Laurentia margin given the ca.
 750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.
@@
 The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}.
-While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca.
+While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} proposed that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwestern margin of Laurentia from prior to ca.
 1110 Ma into the early Neoproterozoic.
@@
 In its reconstructed position, the ca.
-1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwest Laurentia margin prior to 1220 Ma.
-In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwest Laurentia.
+1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwestern Laurentia margin prior to 1220 Ma.
+In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwestern Laurentia.
 
@@
 
-Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoproterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}.
+Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoproterozoic \citep{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}.
 U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous provinces between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca.
@@
 One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics.
-However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonics as as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}.
+However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonics as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}.
 These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca.
@@
 
-Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}.
+Another perspective on Proterozoic tectonics is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}.
 In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion.
@@
 
-An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig.
-\ref{fig:baked_contact}).
+An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from paleomagnetic evidence of a sustained geomagnetic field.
 In a prolonged stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}.
@@
 Paleointensity data developed from units that are also included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}.
-However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory hearing or non-ideal rock magnetic behavior.
+However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory heating or non-ideal rock magnetic behavior.
 As a result, it is significantly more challenging to develop reliable paleointensity data than to develop the reliable paleodirectional data used to calculate paleomagnetic poles.
 Therefore, the paleointensity database is sparser than the compilation of reliable paleomagnetic poles.
-Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests.
+Paleodirectional data themselves can given insight into the presence of a significant geomagnetic field --- particularly those with positive baked contact tests (Fig.
+\ref{fig:baked_contact}).
 Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction.
@@
 The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological data on tectonism and the record of paleomagnetic poles.
-Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny.
+Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these microcontinents during the Trans-Hudson orogeny.
 The collisions of these and other Archean provinces led to the formation of the core of Laurentia.
@@
 This work was supported by NSF CAREER Grant EAR-1847277.
-The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions with Francis Macdonald, Yuem Park, Sarah Slotznick and Justin Strauss.
+The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions and manuscript feedback from Francis Macdonald, Yuem Park, Toby Rivers, Sarah Slotznick and Justin Strauss.
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
@@
 
-\noindent\textbf{Wopmay orogen } An orogen resulting from Paleoproterozoic collision between the Hottah terrane, a continental magmatic arc, and the west margin of the Slave province.
+\noindent\textbf{Wopmay orogen } An orogen resulting from Paleoproterozoic collision between the Hottah terrane, a continental magmatic arc, and the western margin of the Slave province.
 
@@
             Laurentia-Svalbard &                      Upper Grusdievbreen Formation &    800$^{+11}_{-11}$ &      B &      18.2 &      78.9 & 252.6 &  -1.1 &       6.2 &        11.5 &                                 \cite{Maloof2006a} \\ \hline
-                     Laurentia &                                    Gunbarrel dikes &      778$^{+2}_{-2}$ &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &       -18.4 &  Calculation from \cite{Eyster2019a} based on data of \cite{Harlan1993a, Harlan1997a} \\ \hline
+                     Laurentia &                                    Gunbarrel dikes &      778$^{+2}_{-2}$ &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &       -18.4 &  Calculation from \cite{Eyster2020a} based on data of \cite{Harlan1993a, Harlan1997a} \\ \hline
             Laurentia-Svalbard &                          Svanbergfjellet Formation &    770$^{+19}_{-40}$ &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &        12.8 &                                 \cite{Maloof2006a} \\ \hline
                      Laurentia &                               Uinta Mountain Group &     760$^{+6}_{-10}$ &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &       -10.7 &                                   \cite{Weil2006b} \\ \hline
-                     Laurentia &                                      Carbon Canyon &      757$^{+7}_{-7}$ &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &        -8.5 &  \cite{Weil2004a} as calculated in \cite{Eyster2019a} \\ \hline
-                     Laurentia &                               Carbon Butte/Awatubi &      751$^{+8}_{-8}$ &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &         1.0 &                                 \cite{Eyster2019a} \\ \hline
+                     Laurentia &                                      Carbon Canyon &      757$^{+7}_{-7}$ &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &        -8.5 &  \cite{Weil2004a} as calculated in \cite{Eyster2020a} \\ \hline
+                     Laurentia &                               Carbon Butte/Awatubi &      751$^{+8}_{-8}$ &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &         1.0 &                                 \cite{Eyster2020a} \\ \hline
                      Laurentia &                          Franklin event grand mean &      718$^{+2}_{-2}$ &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &        -5.7 &                               \cite{Denyszyn2009b} \\ \hline

### 2021-03-06 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add Bruce Buffett acknowledgement
small manuscript edits

@@
 That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents.
-In contrast, South America and Africa are the product of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca.
+In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca.
 0.6 Ga \citep{Goscombe2019a}.
@@
 The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly.
-Rich geologic, paleomagnetic and geochronologic data give deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.
+Rich geologic, paleomagnetic and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.
 
@@
 \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}).
-Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca.
+Terminal collision recorded in the Trans-Hudson orogen is estimated to have occured ca.
 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}.
-A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
+It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
 This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints.
@@
 
-Crustal growth also progressed in the Paleoproterozoic through accretionary orogenesis.
+Growth of Laurentia in the Paleoproterozoic also occured through accretionary orogenesis.
 This accretion occurred within the Wopmay orogen through ca.
@@
 The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}.
-It is unclear whether the Mojave province should be considered to be a distinct province of Archean lithosphere akin to other Archean provinces or rather it should dominantly be considered to be Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}.
+However, it remains unclear whether the Mojave province should be considered a Archean province or a Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}.
 Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}.
@@
 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca.
-1.12 Ga if a virtual geomagnetic pole from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
+1.12 Ga if a virtual geomagnetic pole (that is a pole position calculated from paleomagnetic data of a single cooling unit) from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
 These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.
@@
 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}.
-It is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}.
+There is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}.
 \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca.
@@
 
-Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related Arctic terranes including northeast Svalbard \citep{McClelland2019a}.
-These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
-The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}.
-Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador (northeast Canada) is implemented in many paleogeographic models (e.g.
+Early Tonian intrusive granitoids and extrusive calc-alkaline volcanics in East Greenland and related Arctic terranes including northeast Svalbard, with ages between 975 to 915 Ma \citep{McClelland2019a}, are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a}.
+This magmatic activity is hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
+This geological evidence for an active margin in East Greenland in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}.
+Such a clockwise rotation of Baltica relative to Laurentia from the earlier Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador (northeast Canada) is implemented in many paleogeographic models (e.g.
 \citealp{Evans2009a} and as shown in Fig.
 \ref{fig:Grenville_reconstructions}).
-This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica's APWP.
+This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica poles.
 These oscillations would imply that following the Grenville loop, Laurentia moved northward across the equator and then back south to a similar Grenville loop position prior to the ca.
@@
 If they indeed became conjoined in Rodinia as in Figure \ref{fig:Grenville_reconstructions}, the separation of Kalahari from Laurentia may have initiated ca.
-795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in southeastern Laurentia.
-This late Tonian timing for separation could be  consistent with a position along the southeastern Laurentia margin given the ca.
-750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.
+795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and volcanic rocks in southeastern Laurentia (e.g., the ca.
+750 Ma Mount Rogers volcanics and associated rift-related sediments \cite{Aleinikoff1995a, MacLennan2020a}).
+This late Tonian timing for separation could be consistent with a position along the southeastern Laurentia margin, although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.
 
@@
 Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation.
-As additional support for a North China--Laurentia connection, \cite{Zhao2019a} point to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction.
+As additional support for a North China--Laurentia connection, \cite{Zhao2019a} pointed to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction.
 In particular, sediment transport from Laurentia could provide a source for ca.
@@
 0.8 Ga \citep{Hamilton2011a, Stern2018a}.
-These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}.
+These arguments rest on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}.
 An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}.
@@
 1.8 Ga.
-Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}.
-Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca.
+Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep subduction and exhumation of continental crust during convergent tectonics \citep{Weller2017a}.
+Outside of Laurentia, there are examples of Paleoproterozoic eclogites with geochemical affinity to oceanic crust such as that documented in the ca.
 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.
@@
 In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion.
-In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}.
+In particular, it has been argued that the Mesoproterozoic Era is an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}.
 The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
@@
 Using data from Laurentia alone, however, it is difficult to ascertain whether this motion is due to plate tectonic motion or rotation of the entire solid Earth through true polar wander.
-True polar wander can lead to changing position relative to the spin axis even with a stagnant lid.
+True polar wander can lead to a changing position relative to the spin axis even with a stagnant lid.
 One interval when the Laurentian paleomagnetic record demands that some of the motion is through differential plate tectonics is in the latest Mesoproterozoic.
@@
 Rifting and passive margin development then isolated Laurentia in late Ediacaran period and into the early Paleozoic Era.
-Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic interval with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea.
-While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic intervals bear similarities.
+Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic interval with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form the supercontinent Pangea.
+While the details of the conjugate continents to Laurentia are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic intervals bear similarities.
 In each case, accretionary collision of arc terranes was followed by continent-continent collision.
@@
 
-\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from southeastern Ontario, Canada to west Texas, USA.
-
-\noindent\textbf{Grenvillian orogen } An orogen resulting from collision orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.
+\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from Labrador, Canada to west Texas, USA.
+
+\noindent\textbf{Grenville orogen } An orogen resulting from the collision Grenvillian orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.
 
@@
 
-\noindent\textbf{juvenile } An adjuective referring to rocks that have formed through melt recently extracted from the mantle.
+\noindent\textbf{juvenile } An adjective referring to rocks formed from melt recently extracted from the mantle.
 
@@
 
-\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and to have broken apart in the middle of the Mesoproterozoic.
+\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and to have broken apart during the Mesoproterozoic.
 
@@
 
-\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era at the time of the Grenvillian orogeny and to have broken apart in the middle of the Neoproterozoic.
+\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era at the time of the Grenvillian orogeny and to have broken apart during the Neoproterozoic.
 

### 2021-03-06 Manuscript/Laurentia_Paleogeo_Manuscript.tex
updates prior to final revisions

@@
 \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}).
-Terminal collision recorded in the Trans-Hudson orogen is estimated to have occured ca.
+Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca.
 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}.
@@
 
-Growth of Laurentia in the Paleoproterozoic also occured through accretionary orogenesis.
+Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis.
 This accretion occurred within the Wopmay orogen through ca.
@@
 The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions and manuscript feedback from Francis Macdonald, Yuem Park, Toby Rivers, Sarah Slotznick and Justin Strauss.
+Bruce Buffett provided insights on the implications of a stagnant lid regime on the generation of the geomagnetic field.
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.

### 2021-03-06 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits associated with comments from FM

@@
 The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly.
-Rich geologic, paleomagnetic and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.
+Rich geologic, paleomagnetic, and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.
 
@@
 It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
-This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints.
+Throughout this chapter, I will utilize accretionary orogenesis in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia \citep{Stall2020a}.
+Collisional orogenesis will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure \citep{Stall2020a}.
+As discussed in \cite{Staal2020a}, there is ambiguity in this distinction as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale.
+Nevertheless, the categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.
+
+The overall story of the rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints.
 Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca.
@@
 Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}.
-Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia (Fig.
+Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia which experienced metamorphic events between ca.
+1.76 and 1.67 Ga (\citealp{Strickland2013a}; Fig.
 \ref{fig:Laurentia_map}).
@@
 If the interpretation of a conjugate continent rifting of the margin prior to the ca.
-717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}).
+717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}).
 While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia, it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
@@
 The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland.
-The reconstruction of Svalbard is more challenging given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting.
+The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting.
 The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction.
@@
 A significantly later separation ca.
-600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of rifting until after ca.
-620 Ma north of the New York promontory.
+560 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of earlier rifting until after ca.
+620 Ma north of the New York promontory \citep{Allen2010a}.
 
@@
 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}.
+The U-Pb-Hf signatures of detrital zircons from ca.
+1.45 to 1.30 Ga sedimentary rocks of Tasmania have also been interpreted to indicate a Laurentia source consistent with such a paleogeographic connection \citep{Mulder2015a}.
 Additionally, ca.
@@
 
-The separation of Baltica from Laurentia, once in the configuration shown in Figure \ref{fig:Grenville_reconstructions}, is hypothesized to have occurred ca.
-600 Ma associated with the opening of the Iapetus Ocean (Fig.
-\ref{fig:tectonic_history}; \citealp{Cawood2001a}).
+Magmatism ca.
+615 Ma in both Laurentia and Baltica (in the relative configuration shown in Figure \ref{fig:Grenville_reconstructions}) is interpreted to be associated with the plume-related Central Iapetus Magmatic Province (CIMP; \citealp{Tegner2019a}).
+This ca.
+615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Fig.
+\ref{fig:tectonic_history}; \citealp{Cawood2001a, Tegner2019a}).
 The cratons would once again be conjoined ca.
@@
 750 Ma Mount Rogers volcanics and associated rift-related sediments \cite{Aleinikoff1995a, MacLennan2020a}).
+A ca.
+750 Ma volcano-sedimentary sequence in northwest Kalahari is interpreted to be rift-related \citep{Borg2003a}  and could be correlative with the Mount Rogers Formation.
 This late Tonian timing for separation could be consistent with a position along the southeastern Laurentia margin, although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.
@@
 These arguments rest on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}.
-An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}.
-While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
+Alternative interpretations for the lack of blueschists in the Proterozoic call upon both their low preservation potential or secular changes in mantle chemistry and/or temperature \citep{Brown2019a}.
+\cite{Palin2015a} proposed that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry and changing MgO composition of oceanic crust rather than a harbinger of the onset of plate tectonics.
+An alternative hypothesis is that earlier in the Proterozoic slab breakoff occurred at shallower depths limiting the formation and preservation of low-temperature/high-pressure metamorphic rocks \citep{Brown2019a}.
+In this model, secular mantle cooling led to increasingly strong lithosphere that enabled deeper continental subduction and slab breakoff depths.
+While the secular change in the abundance of low-temperature/high-pressure metamorphic rocks is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
 From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.

### 2021-03-07 Manuscript/Laurentia_Paleogeo_Manuscript.tex
address comments and suggested edits from TR

@@
 \includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
-\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}.
+\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}.
 Proterozoic orogens are labeled with \textit{italicized text} (or.
@@
 It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
-Throughout this chapter, I will utilize accretionary orogenesis in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia \citep{Stall2020a}.
-Collisional orogenesis will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure \citep{Stall2020a}.
+Throughout this chapter, I will utilize accretionary orogenesis in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia \citep{Staal2020a}.
+Collisional orogenesis will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure \citep{Staal2020a}.
 As discussed in \cite{Staal2020a}, there is ambiguity in this distinction as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale.
@@
 
-Overall, the collision of Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).
-
-\subsubsection{Protracted Proterozoic accretionary growth followed by collisional orogenesis}
-
-Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis.
-This accretion occurred within the Wopmay orogen through ca.
-1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
-1.88 to 1.84 Ga \citep{Hildebrand2009a}.
-Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca.
-1.86 to 1.82 Ga (Fig.
-\ref{fig:Laurentia_map}; \citealp{Schulz2007a}).
-Firm evidence of the end of the Penokean orogeny comes from the ca.
-1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
-
-The collisions of provinces and terranes leading up to the Trans-Hudson orogeny are interpreted to be associated with the initial assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}.
-The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g.
+Overall, the collision of Archean microcontinents between ca.
+1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig.
+\ref{fig:Laurentia_map}).
+These collisional orogenies, including the Trans-Hudson orogeny, are interpreted to be associated with the initial assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}.
+
+The Trans-Hudson orogeny is taken to be the terminal collision associated with closure of the Manikewan Ocean, a large oceanic tract that had previously separated the Superior and the North Atlantic provinces from the composite Slave+Rae+Hearne provinces (often referred to as the Churchill domain or plate; e.g.
 \citealp{Skipton2016a, Weller2017a}; Fig.
@@
 
-Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia.
-This change marks a shift in the predominant style of Laurentia's growth, as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs.
+\subsubsection{Protracted Proterozoic accretionary growth followed by collisional orogenesis}
+
+Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis.
+In the northwest, this accretion occurred within the Wopmay orogen through ca.
+1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca.
+1.88 to 1.84 Ga \citep{Hildebrand2009a}.
+Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margins of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the North Atlantic orogeny (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}).
+The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southeastern margin of the west Superior province ca.
+1.86 to 1.82 Ga (Fig.
+\ref{fig:Laurentia_map}; \citealp{Schulz2007a}).
+Firm evidence of the end of the Penokean orogeny comes from the ca.
+1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
+
+Following the Trans-Hudson orogeny, Laurentia's growth was dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs.
 \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).
-Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
-Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
+Determining the extent of these belts can be complicated by poor exposure in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
+Nevertheless, it is well-constrained that major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
 1.71 to 1.68 Ga Yavapai orogeny (Fig.
@@
 1.43 to 1.37 Ga \citep{Groulier2020a}.
-Far to the southwest along the margin, metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca.
+Far to the southwest along the margin in northern New Mexico, metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths have been interpreted to indicate an interval of ca.
 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}.
@@
 1.49 to 1.46 Ga associated with the Baraboo orogeny \citep{Medaris2003a, Holm2019a}.
-Picuris to Baraboo to Pinwarian orogenesis indicates convergent tectonism along the entirety of southeastern Laurentia.
-It was in this context of an active margin that there was major plutonism ca.
-1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces forming the Granite-Rhyolite Province that extends from the southwestern United States to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}.
-This plutonism has been interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-live active margin \citep{Bickford2015a}.
-Younger magmatic activity (ca.
-1.37 Ga) of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}.
+Coveval Picuris-Baraboo-Pinwarian orogenesis is indicative of convergent tectonism along the entire length of the southeastern margin of Laurentia.
+It was in this active-margin, upper-plate setting that was of site of voluminous widespread plutonism from ca.
+1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout the previously accreted Mesoproterozoic and Paleoproterozoic provinces.
+Known as the Granite-Rhyolite Province, these magmatic products extend from the southwestern United States to the Central Gneiss Belt of Ontario northeast of Georgian Bay \citep{Slagstad2009a}.
+This magmatism has been interpreted to be a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-lived active margin \citep{Bickford2015a}.
+Magmatic activity at the younger end of this range (ca.
+1.37 Ga) is abundant in the Southern Granite-Rhyolite Province, suggesting a similar active margin setting \citep{Bickford2015a}.
 While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature and is consistent with evidence for accretionary orogenesis in the Picuris, Baraboo and Pinwarian orogens, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).
@@
 
-Accretionary orogenesis continued along the eastern margin of Laurentia with the amalgamation and accretion of arcs associated with the ca.
-1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}.
+Accretionary orogenesis continued locally along the along the southeastern margin of Laurentia with the amalgamation and accretion of arcs and back-arcs associated with the ca.
+1.25 to 1.22 Ga Elzevirian orogeny \citep{Carr2000a,McLelland2013a}.
 The subsequent ca.
-1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of a previously rifted fragment of Laurentia that led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}.
+1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision and accretion of a previously rifted fragment of Laurentia that led to obduction of the Pyrites Complex ophiolite \citep{Carr2000a,McLelland2010a, Chiarenzelli2011a}.
 The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}.
@@
 This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}.
-Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision.
-If this interpretation is correct, it is similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (Fig.
+Taken together, this context is suggestive of a microcontinent collision leading to Llano orogenesis prior to terminal Grenvillian continental collision.
+If this interpretation is correct, it would be similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (Fig.
 \ref{fig:tectonic_history}).
@@
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{McLelland2010a}.
-Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage.
+1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen throughout the orogen \citep{Carr2000a, Rivers2008a, Rivers2012a, Indares2020a}.
+Note that while the terms Grenvillian orogeny and Grenville belt have been used rather loosely in the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage.
 Properly referring to the Grenvillian orogeny as distinct from the Elzevirian and Shawinigan orogenies enables the available constraints to be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig.
 \ref{fig:tectonic_history}).
-Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a}.
+Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a, Rivers2012a}.
 The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca.
@@
 In the latest Mesoproterozoic (ca.
-1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}.
-This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
+1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior, leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}.
+This Midcontinent Rift is associated with the emplacement of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
 Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Cannon1994a, Swanson-Hysell2019a}.
 
-There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig.
-\ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin.
+There is significantly less preserved Mesoproterozoic crust on the western margin of Laurentia (Fig.
+\ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well constrained as on the southern to eastern margin.
 There are thick successions of Paleoproterozoic siliciclastic and carbonate sedimentary rocks such as the ca.
@@
 While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated.
-\citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}.
-This region is interpreted to have been subsequently deformed during a ca.
-1.37 to 1.33 Ga event known as the East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
+\citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others have envisioned it being associated with continental rifting associated with separation of a conjugate continent \citep{Jones2015a}.
+This region was subsequently deformed during the ca.
+1.37 to 1.33 Ga East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.
 
@@
 Amazonia in \citealp{Elming2009a, Elming2021a}).
-The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and another continent ca.
+The great extent of the high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and another continent ca.
 1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}.
@@
 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}).
-While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia, it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up.
-One explanation is that Tonian extensional basin development is associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}.
+While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia
+break-up.
+One explanation is that Tonian extensional basin development was associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}.
 The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
@@
 Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca.
-700 Ma extension is associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca.
+700 Ma extension was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca.
 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}.
@@
 The eastern margin of Laurentia then went through multiple phases of Appalachian orogenesis.
-As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano orogenies in the Grenville interval; Acadian orogeny in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian orogeny in the Appalachian interval).
+As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville and Appalachian orogenic intervals in that in both cases there was a period of arc-continent collision (Elzevirian orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Shawinigan/Llano orogenies in the Grenville interval; Acadian orogeny in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian orogeny in the Appalachian interval).
 These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until continent-continent collision.
@@
 The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.
-I refer to additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR').
+In addition, in this chapter I refer to select additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR').
 These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported.
-Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions.
+Many additional poles in the Paleomagia database are valuable and should not be dismissed from being considered in paleogeographic reconstructions.
 However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age.
@@
 The evaporite paleolatitude points are labeled with numbers that correspond to the numbers on the present-day location map.
-These numbers correspond to: 1 -- Altyn Formation (Belt Supergroup); 2: Wallace/Helena Formations -- Belt Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Silurian Michigan Basin ; 9 -- Devonian Western Canada; 10 -- Carboniferous Canadian Maritime; 11 -- Carboniferous Sverdrup; 12 -- Permian Midcontinental USA; 13 -- Jurassic Gulf of Mexico.}
+These numbers correspond to: 1 -- Altyn Formation (Belt-Purcell Supergroup); 2: Wallace/Helena Formations -- Belt-Purcell Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Silurian Michigan Basin ; 9 -- Devonian Western Canada; 10 -- Carboniferous Canadian Maritime; 11 -- Carboniferous Sverdrup; 12 -- Permian Midcontinental USA; 13 -- Jurassic Gulf of Mexico.}
 \label{fig:Laurentia_evaporites}
@@
 The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited
-\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preserved within secondary silica \citep{White1984a}.
+\item The Altyn Formation of the Belt-Purcell Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preserved within secondary silica \citep{White1984a}.
 The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}.
@@
 Outside of Laurentia, there are examples of Paleoproterozoic eclogites with geochemical affinity to oceanic crust such as that documented in the ca.
-1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.
+1.9 Ga Ubendian belt of the Congo craton \citep{Boniface2012a}.
 Mesoproterozoic ophiolites were also obducted to Laurentia such as the Pyrites ophiolite complex within the Shawinigan orogen and the Coal Creek Domain of the Llano uplift \citep{Chiarenzelli2011a, McLelland2013a}.

### 2021-03-07 Manuscript/Laurentia_Paleogeo_Manuscript.tex
more TR comment-related edits

@@
 \ref{fig:Laurentia_poles} and Table 2).
-The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021a}.
+The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades to convey the confidence that the community has in these results \citep{Evans2021a}.
 While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment.
@@
 There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary.
-The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen.
+The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenville orogen.
 Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions.
@@
 However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below).
-In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2).
+In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2).
 An additional not-rated pole included in the present compilation is a more recently published pole for the ca.
@@
 
-Following the amalgamation of the Archean provinces in Laurentia ca.
+Following the amalgamation of the Archean provinces in Laurentia by ca.
 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
@@
 \hline
-Scotland & 161.9 & 78.6 & -31.0 & Reconstructing Atlantic opening following \cite{Torsvik2017a} \\
+Scotland & 161.9 & 78.6 & -31.0 & Reconstruction of Atlantic Ocean opening following \cite{Torsvik2017a} \\
 \hline
@@
 Intracontinental rifts have led to the highest density of poles both in the case of the ca.
-1.4 Ga Belt Supergroup and the ca.
+1.4 Ga Belt-Purcell Supergroup and the ca.
 1.1 Ga Midcontinent Rift (Figs.
@@
 Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage.
-These data result in progressive paths such as ascending to the Logan Loop \citep{Robertson1971a}, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
+These data result in progressive paths (Fig.
+\ref{fig:Laurentia_poles}) such as ascending to the ca.
+1140 to 1110 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca.
+1110 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca.
+980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
 775 to 720 Ma) path \citep{Eyster2020a}.
@@
 There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}.
-However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g.
+However, these poles are in a distinct location from poles of similar age associated with the Belt-Purcell Supergroup (e.g.
 the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}).
-Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton.
-There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation.
+Additionally, the older Belt-Purcell Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton.
+There are potential complications associated with the Belt-Purcell Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation.
 However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton.
-Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite.
+Another potential complication is that the remanence used for the development of the Belt-Purcell Supergroup poles resides in hematite.
 As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed.
@@
 There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}.
-However the overall coherency of the pole directions from the Belt Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+However the overall coherency of the pole directions from the Belt-Purcell Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
 At present, it is unclear which poles are a better representation of Laurentia's position ca.
@@
 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.
-\cite{Bono2015a} interpret steeper directions also recovered from the intrusives as the result of remagnetization.
-They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca.
+\cite{Bono2015a} interpreted steeper directions also recovered from the intrusives as the result of remagnetization.
+They suggested that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca.
 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization.
@@
 Paleomagnetic poles within 25 million years of each reconstruction time are plotted.
-All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}
+All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.
+The colors follow Figure \ref{fig:Laurentia_map} where the light grey represents Archean provinces, dark grey represents Paleoproterozoic collisional orogens, and pink/orange represents Paleoproterozoic/Mesoproterozoic accretionary orogens and granitoid intrusions.}
 \label{fig:Laurentia_reconstructions}
@@
 There is an overall lack of models in the literature for the Proterozoic with published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein.
-The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters, although some studies have published the Euler rotations associated with specified times.
+The approach in the community for many years has been to publish models as snapshots at time intervals presented in figures without publishing continuous rotation parameters, although some studies have published the Euler rotations associated with specified times.
 With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g.
@@
 The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}.
-As noted above, the development of such models is challenging and  researchers need to balance varying constraints.
-The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia.
+As noted above, the development of such models is challenging and researchers need to simultaneously balance many constraints.
+The focus here is on the extent to which this model satisfies the available paleomagnetic poles for Laurentia.
 The model does not honor the Grenville loop (e.g.
@@
 
-There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model.
+There are more published models that show snapshots and publish rotation parameters associated with particular time intervals, such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model.
 The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record.
-The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (while the model parameters were not published with that study they have now been made available by the authors).
+The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (although the model parameters were not published with that study they have now been made available by the authors).
 This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again.
@@
 The evaporite lines extend from the maximum to minimum age constraints with points at the preferred depositional age.
-The evaporite paleolatitude points are labeled with numbers that correspond to the numbers on the present-day location map.
+The evaporite paleolatitude points are labeled with numbers that correspond to the numbers on the present-day location map in the right panel.
 These numbers correspond to: 1 -- Altyn Formation (Belt-Purcell Supergroup); 2: Wallace/Helena Formations -- Belt-Purcell Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Silurian Michigan Basin ; 9 -- Devonian Western Canada; 10 -- Carboniferous Canadian Maritime; 11 -- Carboniferous Sverdrup; 12 -- Permian Midcontinental USA; 13 -- Jurassic Gulf of Mexico.}
@@
 Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}.
-Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
+Higher in the Belt-Purcell Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}.
 There are also halite and gypsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.
@@
 In particular, they discuss the hypothesis of Superia (named after the Superior province) wherein the Superior province of Laurentia is central to a group of Archean lithospheric blocks \citep{Bleeker2006a}, including the Kola and Karelia provinces of Baltica and the Wyoming province of Laurentia, that broke up prior to 2.0 Ga.
-This hypothesis has largely been based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeastern margin of the Superior Province \citep{Davey2020a}.
+This hypothesis was based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeastern margin of the Superior Province \citep{Davey2020a}.
 \cite{Salminen2021a}  argue that the originally proposed Superia fit is not consistent with Baltica paleomagnetic poles.
@@
 \ref{fig:tectonic_history}).
-Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart.
+Subsequent separation of Amazonia in the Neoproterozoic would have led to the formation of the Iapetus Ocean as Rodinia rifted apart.
 Departure of Amazonia potentially occurred as early as ca.
@@
 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}.
-If the North Australian craton was continuous with the South Australia, this interpretation would have the ca.
+If the North Australian craton was continuous with South Australia, this interpretation would have the ca.
 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca.
@@
 This configuration had originally been proposed by \citet{Patchett1978a} and \citet{Piper1980a} largely on the basis of paleomagnetic pole comparisons.
-This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
+This connection proposes a tight fit between present-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
 In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca.
@@
 This magmatic activity is hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
-This geological evidence for an active margin in East Greenland in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}.
+This geological evidence for an active margin in East Greenland in the earliest Neoproterozoic has motivated a tectonic model wherein Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}.
 Such a clockwise rotation of Baltica relative to Laurentia from the earlier Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador (northeast Canada) is implemented in many paleogeographic models (e.g.
@@
 \ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)eastern margin of Laurentia during the time of the Grenvillian orogeny.
-This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g.
+This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was proposed in \cite{Hoffman1991a} and is implemented in most reconstructions of Rodinia (e.g.
 \citealp{Li2008a}).
@@
 1110 Ma into the early Neoproterozoic.
-North China would have been at polar latitudes ca.
+In this model, North China would have been at polar latitudes ca.
 1110 Ma and moved rapidly with Laurentia as it transited towards the equator.
@@
 The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well.
-While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca.
+While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be proponents of a stagnant lid throughout the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca.
 0.8 Ga \citep{Hamilton2011a, Stern2018a}.
@@
 In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion.
-In particular, it has been argued that the Mesoproterozoic Era is an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}.
+In particular, it has been argued that the Mesoproterozoic Era was an interval when Earth was in a stagnant regime without mobile plate tectonics \citep{Silver2008a, ONeill2013a}.
 The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs.
@@
 These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid.
-Rather, the orogenic interval of the Mesoproterozoic (1.6 to 1.0 Ga) bears similarity with that of the Paleozoic (0.54 to 0.25 Ga) and reveals Laurentia to have been a central player in amalgamation of continents associated with both Rodinia and Pangea.
+Rather, the orogenic interval within the Mesoproterozoic bears similarity with that of the Paleozoic (0.54 to 0.25 Ga) and reveals Laurentia to have been a central player in amalgamation of continents associated with both Rodinia and Pangea.
 

### 2021-03-10 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add Slave Province endnote and some minor edits

@@
 \usepackage{preprint}
+
+\usepackage{enotez}
+\let\footnote=\endnote
 
@@
 \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}).
-That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents.
+That the vast majority of the present-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents.
 In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca.
@@
 Proterozoic orogens are labeled with \textit{italicized text} (or.
--- orogen; THO  -- Trans-Hudson orogen; GFTZ -- Great Falls Tectonic Zone; MCR -- Midcontinent Rift).
+-- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift).
 The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.
@@
 \subsubsection{Laurentia's initial formation}
-Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig.
+Collision between the Superior province and the composite Slave\endnote{The term Slave Province is rather jarring when read for the first time given that it brings to mind evil human oppression.
+The geologic Slave Province as well as the eponymous Great Slave Lake get their name from a name given to the First Nation indigenous peoples of the Dene Group who are indigenous to the region.
+The origin of the name Slave is commonly explained as a French translation of the name given to the Dene people by the Cree people \citep{Britannica-Slave2017a}.
+Given its negative connotations, the name is not preferred now and there are efforts to remove it from Canadian place names \citep{Mandeville2016a}.
+Given that this name for the Archean geologic province is deeply entrenched in the geologic literature it will be used here.}+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig.
 \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}).
@@
 It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny.
-Throughout this chapter, I will utilize accretionary orogenesis in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia \citep{Staal2020a}.
-Collisional orogenesis will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure \citep{Staal2020a}.
-As discussed in \cite{Staal2020a}, there is ambiguity in this distinction as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale.
-Nevertheless, the categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.
+Throughout this chapter, I will utilize the term ``accretionary orogenesis'' in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia.
+``Collisional orogenesis'' will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure.
+This usage follows \cite{Staal2020a} who discuss the ambiguity in such a division as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale.
+Nevertheless, this categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.
 
@@
 
-Overall, the collision of Archean microcontinents between ca.
-1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig.
-\ref{fig:Laurentia_map}).
-These collisional orogenies, including the Trans-Hudson orogeny, are interpreted to be associated with the initial assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}.
-
+These collisional orogenies, particularly the Trans-Hudson orogeny, are interpreted to be associated with assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}.
 The Trans-Hudson orogeny is taken to be the terminal collision associated with closure of the Manikewan Ocean, a large oceanic tract that had previously separated the Superior and the North Atlantic provinces from the composite Slave+Rae+Hearne provinces (often referred to as the Churchill domain or plate; e.g.
@@
 \ref{fig:Superior_Slave_recons}).
-The paleogeographic model of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
+The paleogeographic model of \cite{Pehrsson2015a} posits the closure of the Manikewan Ocean not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
 In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography.
@@
 This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
+
+Overall, the collision of Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the core of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).
 
@@
 1.88 to 1.84 Ga \citep{Hildebrand2009a}.
-Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margins of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the North Atlantic orogeny (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}).
-The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southeastern margin of the west Superior province ca.
+Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margin of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the southern margin of the North Atlantic province (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}).
+The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southern margin of the west Superior province ca.
 1.86 to 1.82 Ga (Fig.
@@
 Determining the extent of these belts can be complicated by poor exposure in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield.
-Nevertheless, it is well-constrained that major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca.
+Nevertheless, it is well-constrained that major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with arc-continent collision during the ca.
 1.71 to 1.68 Ga Yavapai orogeny (Fig.
@@
 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}.
-Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca.
+Yavapai to Mazatzal-age accretionary orogenesis extended across the southeastern margin of Laurentia from the southwestern USA to eastern Canada where it is called the Labradorian orogeny.
+In eastern Canada, where the orogen was overprinted in the Grenvillian orogeny, it is interpreted to have been active from ca.
 1.71 to 1.60 Ga (Fig.
@@
 
-Laurentia's growth continued into the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion.
+Laurentia's growth continued into the Mesoproterozoic along the southeastern margin through further juvenile terrane and arc accretion.
 Continental arc magmatism is interpreted to have occurred associated with the Pinwarian orogeny in the northeast Grenville province in Labrador from ca.
@@
 Coveval Picuris-Baraboo-Pinwarian orogenesis is indicative of convergent tectonism along the entire length of the southeastern margin of Laurentia.
-It was in this active-margin, upper-plate setting that was of site of voluminous widespread plutonism from ca.
-1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout the previously accreted Mesoproterozoic and Paleoproterozoic provinces.
-Known as the Granite-Rhyolite Province, these magmatic products extend from the southwestern United States to the Central Gneiss Belt of Ontario northeast of Georgian Bay \citep{Slagstad2009a}.
+This active-margin, upper-plate setting was the site of voluminous widespread plutonism from ca.
+1.48 to 1.35 Ga that resulted in the emplacement of A-type granitoids throughout the previously accreted Paleoproterozoic and Mesoproterozoic provinces.
+Known as the Granite-Rhyolite Province, these magmatic products extend from the southwestern United States to the Central Gneiss Belt of Ontario northeast of Georgian Bay (Fig.
+\ref{fig:Laurentia_map}; \citealp{Slagstad2009a}).
 This magmatism has been interpreted to be a combination of continental arc magmatism and melt generation within a back-arc region of Laurentia's long-lived active margin \citep{Bickford2015a}.
@@
 \includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
-\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
-A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown.
-The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.
+\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past 1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
+A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) magmatism is also shown.
+The interpreted ages of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic (Table 2) and in \cite{Torsvik2012a} for the Phanerozoic is shown.
 Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
@@
 
-Accretionary orogenesis continued locally along the along the southeastern margin of Laurentia with the amalgamation and accretion of arcs and back-arcs associated with the ca.
+Accretionary orogenesis continued locally along the southeastern margin of Laurentia with the amalgamation and accretion of arcs and back-arcs associated with the ca.
 1.25 to 1.22 Ga Elzevirian orogeny \citep{Carr2000a,McLelland2013a}.
@@
 The Grenvillian orogeny was a protracted interval of continent-continent collision (ca.
-1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen throughout the orogen \citep{Carr2000a, Rivers2008a, Rivers2012a, Indares2020a}.
+1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{Carr2000a, Rivers2008a, Rivers2012a, Indares2020a}.
 Note that while the terms Grenvillian orogeny and Grenville belt have been used rather loosely in the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage.
@@
 In the latest Mesoproterozoic (ca.
-1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior, leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}.
+1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior, leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south (MCR in Fig.
+\ref{fig:Laurentia_map}; \citealp{Cannon1992b}).
 This Midcontinent Rift is associated with the emplacement of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.
@@
 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}.
-If the interpretation of a conjugate continent rifting of the margin prior to the ca.
+If the interpretation of a conjugate continent rifting off the margin prior to the ca.
 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}).
-While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia
-break-up.
+While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up.
 One explanation is that Tonian extensional basin development was associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}.
 The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca.
-780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models).
+780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence, there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models).
 Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca.
@@
 
-In northwestern Laurentia from the Ogilvie Mountains (Yukon, Canada) to Victoria Island (Nunavut, Canada), the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
+In northwestern Laurentia from the Ogilvie Mountains (Yukon, Canada) to Victoria Island (Nunavut, Canada), the sedimentary rock record is distinct from that further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}.
 Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca.
@@
 Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
-A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
+
+On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
 \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}.
-The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}.
+The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}.
 In this model, spatially-restricted rifting occurs ca.
-760 to 680 Ma in the region of modern-day North Carolina and Virginia.
-Rifting initiated in the region from modern-day New York to Newfoundland ca.
+760 to 680 Ma in the region of present-day North Carolina and Virginia.
+Rifting initiated in the region from present-day New York to Newfoundland ca.
 620 to 580 Ma and by ca.
-580 to 550 Ma rifting extends along the length of Laurentia's eastern margin.
+580 to 550 Ma rifting extended along the length of Laurentia's eastern margin.
 The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca.
@@
 The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}.
-High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles between 2.23 and 1.89 Ga (Fig.
+High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their present-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles between 2.23 and 1.89 Ga (Fig.
 \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}).
@@
 Within the tropical rain belt (0 to $\sim$10\textdegree\ latitude) these high evaporation rates are typically overwhelmed by precipitation such that global zonal mean precipitation exceeds evaporation within $\sim$10\textdegree\ of the equator (within $\sim$8\textdegree$\;$ of the equator over land).
-Evaporation typically exceeding precipitation from those latitudes ($\sim$10 to 15\textdegree) towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree\ \citep{Park2020b}.
+Evaporation typically exceeding precipitation from those latitudes ($\sim$10 to 15\textdegree) towards higher ones with evaporation minus precipitation being at a maximum at $\sim$20-25\textdegree\ \citep{Park2021a}.
 However, continental interiors near the equator can also be arid due to regional precipitation patterns leading to aridity and the formation of evaporites.
@@
 Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca.
-1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca.
+1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018b}) with further motion by ca.
 1070 Ma.
@@
 1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path \citep{Pisarevsky2014a}.
-There is not a coeval pole from Laurentia for comparison at ca.
+There is not a coeval pole from Laurentia for comparison ca.
 1.21 Ga and data do indicate poleward motion for Laurentia between the ca.
@@
 The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
+
+\printendnotes
 

### 2021-06-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
small update to Laurentia rifted margin text

@@
 550 Ma \citep{Dickerson1998a, Martin2019a}.
-As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar.
+Rifting of the Laurentia margin north of the New York promontory was occuring by ca.
+560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{}Hodych2007.
+
+As with other rifted margins, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar.
 Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca.

### 2021-06-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add ECMB pole to table and a reference to it in the text; add new paleogeo reconstruction figure

@@
 Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.
-
 On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig.
@@
 The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca.
-550 Ma \citep{Dickerson1998a, Martin2019a}.
+540 Ma \citep{Dickerson1998a, Martin2019a}.
 Rifting of the Laurentia margin north of the New York promontory was occuring by ca.
-560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{}Hodych2007.
+560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{Hodych2007a}.
 
@@
 Following the amalgamation of the Archean provinces in Laurentia by ca.
-1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
+1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton \citep{Swanson-Hysell2021b}.
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Fig8_Grenville_Reconstruction.pdf}
-\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}.
+\includegraphics[width=\textwidth]{../Figures/Fig8_Rodinia_Reconstruction.pdf}
+\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select conjugate Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic and to its initial break-up in the Neoproterozoic}.
 The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga.
-The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny.
+The reconstruction of North China to Laurentia following \cite{Ding2021a}.
+The reconstruction implements Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny \citep{Hoffman1991a}.
 The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia relative to South and West Australia.
@@
 1070 Ma.
-The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
+The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.
+The assembled Rodinia persisted until initial rifting ca.
+775 Ma with episodic rifting continuing until ca.
+530 Ma.}
 \label{fig:Grenville_reconstructions}
@@
 This ca.
-615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Fig.
-\ref{fig:tectonic_history}; \citealp{Cawood2001a, Tegner2019a}).
+615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Figs.
+\ref{fig:tectonic_history} and \ref{fig:Grenville_reconstructions}; \citealp{Cawood2001a, Tegner2019a}).
 The cratons would once again be conjoined ca.
@@
 
-The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}.
+The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Ding2021a, Zhang2021a}.
 While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} proposed that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwestern margin of Laurentia from prior to ca.
@@
                  Laurentia-Rae &                                   Martin Formation &     1818$^{+4}_{-4}$ &      A &     251.4 &      59.6 & 288.0 &  -9.0 &       8.5 &          &                                  \cite{Evans1973a} \\ \hline
+                 Laurentia &                                   East Central Minnesota Batholith &     1779$^{+2}_{-2}$ &      NR &     265.8 &      45.5 & 265.8 &  20.4 &      4.5 &        63.5 &                                   \cite{Swanson-Hysell2021b} \\ \hline
                      Laurentia &                                      Dubawnt Group &   1785$^{+35}_{-35}$ &      B &     265.6 &      64.1 & 277.0 &   7.0 &       8.0 &        49.4 &                                   \cite{Park1973a} \\ \hline

### 2021-06-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add ECMB pole to tables and Fig 1 map
further integrate ECMB results and make other small text edits

@@
 Firm evidence of the end of the Penokean orogeny comes from the ca.
-1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.
+1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a, Swanson-Hysell2021a}.
 
@@
 Rifting initiated in the region from present-day New York to Newfoundland ca.
-620 to 580 Ma and by ca.
+620 to 580 Ma associated with the Central Iapetus Magmatic Province (CIMP; Fig.
+\ref{fig:Grenville_reconstructions}) and by ca.
 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin.
@@
 540 Ma \citep{Dickerson1998a, Martin2019a}.
-Rifting of the Laurentia margin north of the New York promontory was occuring by ca.
-560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{Hodych2007a}.
 
@@
 Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca.
-700 Ma extension was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca.
+700 Ma extension (which occured in the southern part of the eastern margin) was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca.
 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}.
 In these models, the separation of a conjugate continent ca.
-700 Ma (interpreted to be Amazonia) leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}.
+700 Ma leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}.
 The subsequent rifting of terranes off the Laurentia margin ca.
@@
 These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported.
-Many additional poles in the Paleomagia database are valuable and should not be dismissed from being considered in paleogeographic reconstructions.
+Many additional poles in the database to those that are rated are valuable and should not be dismissed from being considered in paleogeographic reconstructions.
 However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age.
 For example, there are many poles that have been developed from lithologies of the Trans-Hudson orogen (e.g.
-\citealp{Symons2005a}.
+\citealp{Symons2005a}).
 However, these poles typically lack field tests, have poor control on paleohorizontal and are hypothesized to suffer from variable overprints as discussed in \cite{Raub2008a} and \cite{DAgrella-Filho2020a}.
@@
 In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2).
-An additional not-rated pole included in the present compilation is a more recently published pole for the ca.
-1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca.
+Additional not-rated poles, published after the ratings were conferred, that are included in the present compilation are: 1) a pole from the ca.
+1780 Ma East Central Minnesota Batholith that confirms that the Slave, Hearne, Rae, and Superior provinces were part of coherent amalgamated Laurentia following the Trans-Hudson orogeny \citep{Swanson-Hysell2021a}; 2) a pole for the ca.
+1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at that time and coincides with the position of the poles from the ca.
 1140 Ma Abitibi dikes \citep{Ernst1993a}.
-This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
+These poles will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop.
 Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca.
@@
 I follow this interpretation and group the poles into Superior (West) and Superior (East).
-Uncertainty remains with respect to whether the ca.
-1.88 Ga Molson dikes pole pre-dates or post-dates this rotation \citep{Evans2010a} and thus for the time being should be considered solely in the western Superior province reference frame.
-
-%The poles on this figure should be labeled
+
 \begin{figure*}
 \centering
-\includegraphics[width=6 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
+\includegraphics[width=5.25 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
 \caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
-This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
+This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
 Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
-Poles with ages that are within 25 million years of the given time slice are shown.
+Poles with ages that are within 20 million years of the given time slice are shown.
 The relatively well-resolved pole paths from the Superior and Slave provinces (Fig.
@@
 Following the amalgamation of the Archean provinces in Laurentia by ca.
-1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton \citep{Swanson-Hysell2021b}.
+1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton.
+This interpretation of a coherent Laurentia is confirmed paleomagnetically by the consistent positions of ca.
+1780 to 1740 Ma poles from the Slave, Hearne/Rae, and Superior Provinces (Fig.
+\ref{fig:Superior_Slave_recons}; \citealp{Swanson-Hysell2021b}).
 It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton.
@@
 \ref{fig:Laurentia_poles}) such as ascending to the ca.
-1140 to 1110 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca.
-1110 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca.
+1140 to 1108 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca.
+1108 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca.
 980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca.
@@
 1237 Ma Sudbury dikes and ca.
-1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs.
+1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with older and younger constraints from mainland Laurentia while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs.
 \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).
@@
 As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed.
-However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts.
+However, applying a moderate inclination correction factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts.
 There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}.
-However the overall coherency of the pole directions from the Belt-Purcell Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
+However the overall coherency of the pole directions from the Belt-Purcell Supergroup and the presence of geomagnetic reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}.
 At present, it is unclear which poles are a better representation of Laurentia's position ca.
@@
 Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.
-TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}.
+TC2017 refers to \cite{Torsvik2017a} and SHM2017 refers to \cite{Swanson-Hysell2017a}.
 Bottom panel: the velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
@@
 The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os date from the overlying Coppercap Formation \citep{Rooney2014a}.
-This units are also interpreted to be correlative with the Callison Lake Formation which contains pseudomorphs after gypsum and is constrained to have been deposited between Re-Os dates of 752.7 $\pm$ 5.5 Ma and 739.9 $\pm$ 6.5 Ma \citep{Strauss2015a}.
+These units are also interpreted to be correlative with the Callison Lake Formation which contains pseudomorphs after gypsum and is constrained to have been deposited between Re-Os dates of 752.7 $\pm$ 5.5 Ma and 739.9 $\pm$ 6.5 Ma \citep{Strauss2015a}.
 \end{itemize}
@@
 750 Ma could reflect increased aridity in the tropical interior of the Rodinia supercontinent.
-Overall, the paleomagnetic data as synthesized in the paleogeographic models are in agreement with the evaporite record from the Mesoproterozoic to the Cenozoic (Fig.
+Overall, the paleomagnetic data as synthesized in the paleogeographic models are in agreement with the Laurentia evaporite record from the Mesoproterozoic to the Cenozoic (Fig.
 \ref{fig:Laurentia_evaporites}).
@@
 The extent of Antarctic lithosphere that was conjoined with Australia prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009b}.
-Separation between these Antarctic Provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway.
-Correlations with Laurentia have led to a number of different proposed reconstructions of Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic.
+Separation between these Antarctic provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway.
+Correlations with Laurentia have led to a number of distinct proposed reconstructions of Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic.
 
@@
 This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations.
-Detrital zircons dating from ca.
+Detrital zircons dates of ca.
 1.61 to 1.50 Ga in the ca.
@@
 This connection proposes a tight fit between present-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}.
-In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca.
-1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia \citep{Karlstrom2001a}.
+In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica (where accretionary orogenesis was active ca.
+1.66 to 1.52 Ga; \citealp{Bergstrom2020a}) is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia \citep{Karlstrom2001a}.
 Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}.
 Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca.
-1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca.
+1.78 to 1.26 Ga \citep{Buchan2000a, Evans2008a, Swanson-Hysell2021a} that could be extended to ca.
 1.12 Ga if a virtual geomagnetic pole (that is a pole position calculated from paleomagnetic data of a single cooling unit) from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
@@
 This ca.
-615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Figs.
+615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the Iapetus Ocean (Figs.
 \ref{fig:tectonic_history} and \ref{fig:Grenville_reconstructions}; \citealp{Cawood2001a, Tegner2019a}).
@@
 Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.
-Kitoi pole of Siberia with Tonian Laurentia poles).
+758 Kitoi pole of Siberia with Tonian Laurentia poles).
 The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}).
@@
 While the secular change in the abundance of low-temperature/high-pressure metamorphic rocks is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
-From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
+From a paleomagnetic perspective, there is strong support for independent and differential motion between the Slave and Superior provinces from 2.2 to 1.8 Ga from as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
 From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary orogenesis followed by collisional orogenesis (Fig.
@@
 These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid.
-Rather, the orogenic interval within the Mesoproterozoic bears similarity with that of the Paleozoic (0.54 to 0.25 Ga) and reveals Laurentia to have been a central player in amalgamation of continents associated with both Rodinia and Pangea.
+Rather, the orogenic interval within the Mesoproterozoic bears similarity with that of the Paleozoic (0.54 to 0.25 Ga) and reveals Laurentia to have been a central player in amalgamation of continents associated with the supercontinents Rodinia and Pangea.
 
@@
 Additionally, since paleomagnetic poles are typically developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
-Therefore, the record of abundant positive baked contact tests (Fig.
-\ref{fig:baked_contact}) and coherent paleomagnetic poles (Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
+Therefore, the record of abundant positive baked contact tests and coherent paleomagnetic poles (Fig.
+\ref{fig:baked_contact}; Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
 
@@
 In each case, accretionary collision of arc terranes was followed by continent-continent collision.
-The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor crustal growth compared to the Trans-Hudson orogeny.
-Break-up following the Grenvillian and Appalachian orogenic intervals occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic interval have remained sutured.
+The major difference is that the collisions of the Grenvillian and Appalachian orogenic intervals resulted in relatively minor growth of Laurentia compared to the Trans-Hudson orogeny.
+This difference is the result of break-up following the Grenvillian and Appalachian orogenic intervals having occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic interval have remained sutured.
 As a result, Laurentia has been a formidable continent for the past 1.8 billion years.
@@
                      Laurentia &                          Ontario lamprophyre dikes &   1143$^{+12}_{-12}$ &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &        61.2 &                                 \cite{Piispa2018a} \\ \hline
-                     Laurentia &                                      Abitibi dikes &     1141$^{+2}_{-2}$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \cite{Ernst1993a} \\ \hline                       Baltica &                        Salla dike (width 60-100 m) &     1122$^{+5}_{-3}$ &      B &      28.8 &      66.8 & 111.6 &  70.6 &       8.1 &          &                                                 \\ \hline
-                     Laurentia &                            Nipigon sills and lavas &     1109$^{+2}_{-2}$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
+                     Laurentia &                                      Abitibi dikes &     1141$^{+2}_{-2}$ &      A &     279.0 &      48.0 & 215.5 &  48.8 &      14.1 &        55.4 &                                  \cite{Ernst1993a} \\ \hline                                            Laurentia &                            Nipigon sills and lavas &     1109$^{+2}_{-2}$ &      A &     270.9 &      49.1 & 217.8 &  47.2 &       4.0 &        56.4 &  Nordic workshop calculation based on data of \cite{Palmer1970a, Robertson1971a, Pesonen1979a, Pesonen1979b, Middleton2004a, Borradaile2006a} \\ \hline
                      Laurentia &             Lowermost Mamainse Point volcanics -R1 &     1109$^{+2}_{-3}$ &      A &     275.3 &      47.1 & 227.0 &  49.5 &       5.3 &        62.9 &                         \cite{Swanson-Hysell2014a} \\ \hline

### 2021-06-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
updates prior to proofs

@@
 Given its negative connotations, the name is not preferred now and there are efforts to remove it from Canadian place names \citep{Mandeville2016a}.
-Given that this name for the Archean geologic province is deeply entrenched in the geologic literature it will be used here.}+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig.
+Given that this name for the Archean geologic province is deeply entrenched in the geologic literature, it will be used here.}+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig.
 \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}).
@@
 This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.
+
+\begin{figure*}
+\centering
+\includegraphics[width=5.25 in]{../Figures/Fig2_Superior_Slave_reconstructions.pdf}
+\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
+This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
+Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
+Poles with ages that are within 20 million years of the given time slice are shown.
+The relatively well-resolved pole paths from the Superior and Slave provinces (Fig.
+\ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
+\label{fig:Superior_Slave_recons}
+\end{figure*}
 
@@
 \centering
-\includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig3_Tectonic_history.pdf}
 \caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past 1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text.
@@
 
-\begin{figure*}
+\begin{figure*}[!h]
 \centering
-\includegraphics[width=\textwidth]{../Figures/Fig3_Laurentia_poles_combined.pdf}
+\includegraphics[width=\textwidth]{../Figures/Fig4_Laurentia_poles_combined.pdf}
 \caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland).
@@
 I follow this interpretation and group the poles into Superior (West) and Superior (East).
-
-\begin{figure*}
-\centering
-\includegraphics[width=5.25 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
-\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
-This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
-Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
-Poles with ages that are within 20 million years of the given time slice are shown.
-The relatively well-resolved pole paths from the Superior and Slave provinces (Fig.
-\ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
-\label{fig:Superior_Slave_recons}
-\end{figure*}
 
@@
 
-\begin{figure*}
+\begin{figure*}[!h]
 \centering

### 2021-07-23 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits associated with corrections of chapter proofs

@@
 In: Pesonen, L.J., Salminen, J., Evans, D.A.D., Elming, S.-Å., Veikkolainen, T.
-(eds.) Ancient Supercontinents and the Paleogeography of the Earth.}
+(eds.) Ancient Supercontinents and the Paleogeography of the Earth, doi:10.1016/B978-0-12-818533-9.00009-6.}
 \vspace{0.3 cm}
@@
 1.80 Ga \citep{Whitmeyer2007a}.
-A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca.
+A contrasting view has been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca.
 1.72 Ga \citep{Kilian2016b}.
@@
 \ref{fig:Superior_Slave_recons}).
-The paleogeographic model of \cite{Pehrsson2015a} posits the closure of the Manikewan Ocean not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
+The paleogeographic model of \cite{Pehrsson2015a} posits that the closure of the Manikewan Ocean not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia.
 In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography.
@@
 \caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length.
-This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
+This model follows \cite{Swanson-Hysell2021b} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure.
 Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel.
@@
 Firm evidence of the end of the Penokean orogeny comes from the ca.
-1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a, Swanson-Hysell2021a}.
+1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a, Swanson-Hysell2021b}.
 
@@
 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west.
-In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.
+In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence (Fig.
+\ref{fig:.
 
@@
 Additional not-rated poles, published after the ratings were conferred, that are included in the present compilation are: 1) a pole from the ca.
-1780 Ma East Central Minnesota Batholith that confirms that the Slave, Hearne, Rae, and Superior provinces were part of coherent amalgamated Laurentia following the Trans-Hudson orogeny \citep{Swanson-Hysell2021a}; 2) a pole for the ca.
+1780 Ma East Central Minnesota Batholith that confirms that the Slave, Hearne, Rae, and Superior provinces were part of coherent amalgamated Laurentia following the Trans-Hudson orogeny \citep{Swanson-Hysell2021b}; 2) a pole for the ca.
 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at that time and coincides with the position of the poles from the ca.
@@
 The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of the Manikewan Ocean between the Superior province and the Hearne+Rae+Slave provinces.
-Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.
+Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the differences in implied orientation and paleolatitude that result from these well-constrained poles.
 
@@
 \hline
-Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
+Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin correlation following to \cite{Maloof2006a}\\
 \hline
@@
 
-Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig.
+Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig.
 \ref{fig:Laurentia_poles}).
@@
 In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia.
-Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire craton.
+Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once they have been rotated these poles can be used for reconstruction of the entire craton.
 The reliability of this approach gains credence through the good agreement between the ca.
@@
 They suggested that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca.
-583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization.
-The lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole.
+583 Ma Baie des Moutons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization.
+The lower inclination Baie des Moutons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole.
 Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}.
@@
 The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions.
-Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak \citep{Thallner2020a} as are data from coeval volcanics in Ukraine \citep{Shcherbakova2019a} which could support an anomalous deviation from stable axial dipolar geomagnetic field behavior at the time as interpreted by \cite{Halls2015a}.
+Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak \citep{Thallner2021a} as are data from coeval volcanics in Ukraine \citep{Shcherbakova2019a} which could support an anomalous deviation from stable axial dipolar geomagnetic field behavior at the time as interpreted by \cite{Halls2015a}.
 Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig.
@@
 This method follows the approach taken in the Phanerozoic (e.g.
-\citealp{Torsvik2012a}) wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps.
+\citealp{Torsvik2012a}) wherein all poles within a 20-Myr interval are averaged with the interval than progressively moved forward in 10-Myr steps.
 When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model.
-Where there are no constraints a simple interpolation between constraints is made.
+Where there are no constraints, a simple interpolation between constraints is made.
 While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain.
-These poles are not utilized in the simple Laurentia model, which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca.
+These poles are not utilized in the simple Laurentia model, which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred between ca.
 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}.
@@
 The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (although the model parameters were not published with that study they have now been made available by the authors).
-This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again.
+This paleogeographic model implements large oscillations, including ones in the Ediacaran, that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again.
 The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}.
@@
 These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
-\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a}.
+\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500-meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a}.
 These thick bedded sulfate deposits are older than cross-cutting 775.1 $\pm$ 0.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Milton2017a}) and younger than ca.
@@
 The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga.
-The reconstruction of North China to Laurentia following \cite{Ding2021a}.
+The reconstruction of North China to Laurentia follows \cite{Ding2021a}.
 The reconstruction implements Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny \citep{Hoffman1991a}.
@@
 It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia \citep{Moores1991a}.
-The extent of Antarctic lithosphere that was conjoined with Australia prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009b}.
+The extent of Antarctic lithosphere that was conjoined with Australia prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extend to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009b}.
 Separation between these Antarctic provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway.
@@
 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}.
-The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwestern Laurentia A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet).
+The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwestern Laurentia magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet).
 The correlation of the eastern North Australia craton with northwestern Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}.
@@
 Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca.
-1.78 to 1.26 Ga \citep{Buchan2000a, Evans2008a, Swanson-Hysell2021a} that could be extended to ca.
+1.78 to 1.26 Ga \citep{Buchan2000a, Evans2008a, Swanson-Hysell2021b} that could be extended to ca.
 1.12 Ga if a virtual geomagnetic pole (that is a pole position calculated from paleomagnetic data of a single cooling unit) from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}.
@@
 
-Early Tonian intrusive granitoids and extrusive calc-alkaline volcanics in East Greenland and related Arctic terranes including northeast Svalbard, with ages between 975 to 915 Ma \citep{McClelland2019a}, are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a}.
+Early Tonian intrusive granitoids and extrusive calc-alkaline volcanics in East Greenland and related Arctic terranes including northeast Svalbard, with ages between 975 and 915 Ma \citep{McClelland2019a}, are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a}.
 This magmatic activity is hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}.
@@
 The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca.
-1090 followed by peak granulite metamorphism ca.
+1090 Ma followed by peak granulite metamorphism ca.
 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean.
@@
 1.08 Ga zircons (from the Grenville orogen).
-In the Laurentia basins, ca.
-1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g.
-\citealp{Wang2020a}).
 If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear.
@@
 Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.
-758 Kitoi pole of Siberia with Tonian Laurentia poles).
+758 Ma Kitoi pole of Siberia with Tonian Laurentia poles).
 The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}).
@@
 The interpretation that the Franklin LIP is recorded in both Laurentia and Siberia constrains the break-up between the continents to post-date 720 Ma.
-While there could have been some extension in the Tonian Period and at the time of the Franklin emplacement, the record of northern Laurentia subsidence indicates Ediacaran to early Cambrian extensional tectonism (e.g.
-\citealp{Strauss2019a}).
-Therefore, it might not have been until the late Ediacaran Period that the long-term Proterozoic connection between the cratons ended.
+Separation between the continents likely initiated associated with the Franklin LIP given that the southern margin of Siberia became an active margin during the Cryogenian Period \citep{Powerman2015a}.
 
@@
 While the secular change in the abundance of low-temperature/high-pressure metamorphic rocks is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data.
-From a paleomagnetic perspective, there is strong support for independent and differential motion between the Slave and Superior provinces from 2.2 to 1.8 Ga from as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
+From a paleomagnetic perspective, there is strong support for independent and differential motion between the Slave and Superior provinces from 2.2 to 1.8 Ga as is illustrated in Figure \ref{fig:Superior_Slave_recons}.
 From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary orogenesis followed by collisional orogenesis (Fig.
@@
 One interval when the Laurentian paleomagnetic record demands that some of the motion is through differential plate tectonics is in the latest Mesoproterozoic.
-At that time, the pole path is very well-resolved with many high-quality paleomagnetic poles between 1110 and 1070 Ma (Table 2; \ref{fig:Laurentia_poles}).
+At that time, the pole path is very well-resolved with many high-quality paleomagnetic poles between 1110 and 1070 Ma (Table 2; Fig.
+\ref{fig:Laurentia_poles}).
 The progression of the poles requires rotation about an Euler pole that is distinct from a great circle path which would result if the motion were solely due to true polar wander \citep{Swanson-Hysell2019a}.
@@
 One way to get insight into the ancient geomagnetic field is through paleointensity experiments on igneous rocks that enable estimates of ancient field strength to be developed.
-Paleointensity data developed from units that are also included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}.
+Paleointensity data developed from units that are included within the Laurentia paleomagnetic poles database indicate a significant geomagnetic field in the Neoarchean \citep{Selkin2000a} through to the Mesoproterozoic \citep{Macouin2006a, Sprain2018a}.
 However, paleointensity experiments are challenging and prone to failure due to alteration during laboratory heating or non-ideal rock magnetic behavior.
@@
 Additionally, since paleomagnetic poles are typically developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.
-Therefore, the record of abundant positive baked contact tests and coherent paleomagnetic poles (Fig.
-\ref{fig:baked_contact}; Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
+The record of abundant positive baked contact tests and coherent paleomagnetic poles (Fig.
+\ref{fig:baked_contact}; Table 2) supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic.
+This record implies ongoing active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.
 
@@
 This work was supported by NSF CAREER Grant EAR-1847277.
-The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions and manuscript feedback from Francis Macdonald, Yuem Park, Toby Rivers, Sarah Slotznick and Justin Strauss.
-Bruce Buffett provided insights on the implications of a stagnant lid regime on the generation of the geomagnetic field.
+The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions and manuscript feedback from Francis Macdonald, Yuem Park, Toby Rivers, Sarah Slotznick, Justin Strauss, and Yiming Zhang.
+Bruce Buffett provided insights on the implications of a stagnant lid regime for the generation of the geomagnetic field.
 Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein.
@@
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
-The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
+The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography} which is also archived on Zenodo.
 
@@
 
-\noindent\textbf{Canadian shield } The large area of Canada with exposed Precambrian rock, or rock covered by thin soil, that is well-exposed due to Pleistocene glaciation erosion.
+\noindent\textbf{Canadian shield } The large area of Canada with exposed Precambrian rock, or rock covered by thin soil, that is well-exposed due to Pleistocene glacial erosion.
 
@@
 
-\noindent\textbf{evaporite } A chemical sedimentary deposit consisting of minerals that crystallize from water that supersaturated in salts due to evaporation.
+\noindent\textbf{evaporite } A chemical sedimentary deposit consisting of minerals that crystallize from water that were supersaturated in salts due to evaporation.
 
@@
 
-\noindent\textbf{geocentric axial dipole hypothesis } The hypothesis that when it is time-averaged, Earth's magnetic field is dominantly a dipole aligned with the spin axis.
+\noindent\textbf{geocentric axial dipole hypothesis } The hypothesis that, when it is time-averaged, Earth's magnetic field is dominantly a dipole aligned with the spin axis.
 
@@
 \noindent\textbf{large igneous province (LIP) } A region of voluminous and rapidly emplaced volcanics and intrusions that are typically of mafic composition.
-These provinces are often interpreted to be due to decompression melting of an upwelling mantle plume.
+These provinces are often interpreted to result from decompression melting of an upwelling mantle plume.
 
@@
 
-\noindent\textbf{Rae province } An Archean province of Laurentia extending from the region of Lake Athabasca northeast to northern Baffin Island in arctic Canada.
+\noindent\textbf{Rae province } An Archean province of Laurentia extending from the region of Lake Athabasca northeast to northern Baffin Island in Arctic Canada.
 It is framed by the Thelon orogen to the west, the Taltson orogen to the southwest and the Hearne province to the east.
@@
                      Laurentia &                                   Catoctin Basalts &      572$^{+5}_{-5}$ &      B &     281.8 &      38.5 & 116.7 & -42.0 &      17.5 &       -69.0 &                                  \cite{Meert1994a} \\ \hline
-                     Laurentia &                        Sept-Iles layered intrusion &      565$^{+4}_{-4}$ &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &        -7.9 &                                \cite{Tanczyk1987a} \\ \hline
+                     Laurentia &                        Sept-\^Iles layered intrusion &      565$^{+4}_{-4}$ &      B &     293.5 &      50.2 & 141.0 &  20.0 &       6.7 &        -7.9 &                                \cite{Tanczyk1987a} \\ \hline
 \end{longtable}

### 2021-07-23 Manuscript/Laurentia_Paleogeo_Manuscript.tex
manuscript edits associated with proof corrections

@@
 The reconstruction of North China to Laurentia follows \cite{Ding2021a}.
+The omission of South China from Rodinia follows ​\cite{Park2021b}​.
 The reconstruction implements Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny \citep{Hoffman1991a}.
@@
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
-The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography} which is also archived on Zenodo.
+The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography} which is also archived on Zenodo (https://doi.org/10.5281/zenodo.5129624).
 

### 2021-07-23 Manuscript/Laurentia_Paleogeo_Manuscript.tex
update Zenodo link

@@
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
-The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography} which is also archived on Zenodo (https://doi.org/10.5281/zenodo.5129624).
+The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography} which is also archived on Zenodo (https://doi.org/10.5281/zenodo.5129927).
 

### 2021-12-11 Manuscript/Laurentia_Paleogeo_Manuscript.tex
update .tex/PDF to indicate author's version rather than preprint

@@
 %\keywords{First keyword \and Second keyword \and More} % (optional)
-\textit{This manuscript is a preprint of the chapter: \vspace{0.1 cm} \\
+\textit{This manuscript is the author's version of the following chapter and is licensed for reuse and distribution under a Creative Commons Attribution (CC BY 4.0) license: \vspace{0.1 cm} \\
 Swanson-Hysell, N.
@@
 Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}.
-The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography} which is also archived on Zenodo (https://doi.org/10.5281/zenodo.5129927).
+The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
+It is  also archived on Zenodo: \url{https://doi.org/10.5281/zenodo.5129927}.
 
